### 2008-02-23 papers/reasoning-graphically/paper.tex
added paper skeleton, working on adding name-ordering.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1345 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[runningheads]{llncs}
%\usepackage[dvips]{color}
%\usepackage{graphics}
%\usepackage{multicol}
%\usepackage{tabls}
%\usepackage{ttbox}
\usepackage{allmtt}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{pstricks,pst-node,pst-tree} % PS Tricks for diagrams
%\usepackage{proof2}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\urldef{\mailsa}\path|{alfred.hofmann,ursula.barth,ingrid.beyer,natalie.brecht,|
\urldef{\mailsb}\path|christine.guenther,frank.holzwarth,piamaria.karbach,|
\urldef{\mailsc}\path|anna.kramer,erika.siebert-cole,lncs}@springer.com|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

% -=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
%   Random maybe useful things...
% -=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
\newcommand{\cmnt}[1]{\textcolor[rgb]{ 0.8,      0,    0  }{#1}}
%\newcommand{\intabp}[2]{\parbox[t]{#1}{\raggedright{#2}\vspace{0.2cm}}}
%\newcommand{\vsmall}[1]{\footnotesize{#1}}
\newcommand{\ors}{\oplus}
\newcommand{\tensor}{\otimes}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Reasoning Graphically about Quantum Computation}

% a short form should be given in case it is too long for the running head
%\titlerunning{Reasoning Graphically about Quantum Computation}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Lucas Dixon\inst{1} \and Ross Duncan\inst{2}~\thanks{EPSRC Grants...}%
}
%
%\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{University of Edinburgh \and
University of Oxford
%\mailsa\\
%\mailsb\\
%\mailsc
}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%
%\toctitle{Lecture Notes in Computer Science}
%\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
  Systems of quantum computation are typically represented by large
  matrix compositions. However, recent graph-based formalisms offer an
  approach which exposes the structure of these systems in a clearer
  and more accessible way. Although these provide significantly
  simpler way of reading and presenting quantum computations, manual
  manipulation of such graphs is slow, arduous, and error prone. We
  present a formalism that supports a mechanised reasoning about such
  graphs. This involves a compositional account of graph rewriting
  that meets the underlying categorical semantics for the model of
  quantum computation. In particular, we present system, which has
  been implemented, with a fixed logical kernel and which gives a
  declarative account and supports derived rules. Soundness of the
  system is proved with respect to the underlying model of quantum
  computation and completeness is an open problem.

  \keywords{graph rewriting, quantum computing, categorical
    logic, interactive theorem proving, graphical calculi}
\end{abstract}

\section{Introduction}

blah blah

\bibliographystyle{plain}
\bibliography{bibfile}

\end{document}

### 2008-02-23 papers/reasoning-graphically/paper.tex
added secions and edited the abstract
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1349 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
   approach which exposes the structure of these systems in a clearer
-  and more accessible way. Although these provide significantly
+  and more accessible way. Although these provide a significantly
   simpler way of reading and presenting quantum computations, manual
-  manipulation of such graphs is slow, arduous, and error prone. We
+  manipulation of such graphs is slow and error prone. We
   present a formalism that supports a mechanised reasoning about such
   graphs. This involves a compositional account of graph rewriting
-  that meets the underlying categorical semantics for the model of
-  quantum computation. In particular, we present system, which has
-  been implemented, with a fixed logical kernel and which gives a
-  declarative account and supports derived rules. Soundness of the
-  system is proved with respect to the underlying model of quantum
-  computation and completeness is an open problem.
+  that preserves the underlying categorical semantics. We present a
+  system, which has been implemented, with a fixed logical kernel and
+  support for derived rules, which allows the model of quantum
+  computation  to be specified declarativley.
+  Soundness of the 
+  system is proved with respect to the underlying model and
+  completeness is an open problem.
 
@@
 \section{Introduction}
+\label{sec:introduction}
 
-blah blah
+\section{Compact Categories and Graphs}
+\label{sec:mono-categ-graphs}
+
+\section{Representation and Compositionality}
+\label{sec:repr-comp}
+
+\section{Graph Patterns and Matching}
+\label{sec:graph-patt-match}
+
+\section{Conclusions}
+\label{sec:conclusions}
+
+Ideas for further work
+\begin{itemize}
+\item simplification ordering
+\item confluence arguments for graphs
+\item formalise algorithms in a thm prover
+\item richer graph structures - quantification over edges
+\item richer expression language in vertices
+\item completeness of rewrite rules with respect to wpHilb
+\end{itemize}
 
 \bibliographystyle{plain}
+\bibliography{all}
 \bibliography{bibfile}

### 2008-02-24 papers/reasoning-graphically/paper.tex
trivial test change
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1350 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
um this is an accident... you probably want to ignore
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1351 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
some restructuring.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1352 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\section{Compact Categories and Graphs}
+\section{Compact Closed Categories and Graphs}
 \label{sec:mono-categ-graphs}
 
-\section{Representation and Compositionality}
-\label{sec:repr-comp}
+\section{Graph Patterns}
+\label{sec:patterns}
 
-\section{Graph Patterns and Matching}
-\label{sec:graph-patt-match}
+The representation of Compact Closed Categories as Graphs, discuseed
+in the previous section, is too restrictive for automated reasoning
+about quantum computation. In particular, pictures such as the
+following are frequently needed:
+
+\begin{center}
+  \scalebox{1.0}{\includegraphics{images/spider_lhs.eps}}
+\end{center}
+
+\noindent Such pictures represent an infinite family of concrete
+graphs. In this section we describe an extension of the representation
+for Compact Closed Categories that allows a graph-based representation
+that captures an infinite family of graphs. Notably, this allows us to
+represent laws about quantum computation, such as
+Spider-Law~\ref{spider-law-fig}. Furthermore, this representation is
+still compositional in the sense that we can compose graphs. The
+extension, which we call \ref{graph patterns}, adds two extra kinds of
+nodes to the graph: {\em banged nodes} which informally represent an
+arbitrary number of copies of that node, and {\em variable nodes}
+which informally represent some as yet unkown node in a graph.
+
+\subsection{Banged Nodes}
+
+\subsection{Variable Nodes}
+
+\subsection{Compositonality}
+
+\section{Graph Rewriting}
+\label{sec:rewriting}
+
+\subsection{Rules}
+\subsection{Matching}
+\subsection{Substitution}
+
+\section{Related Work}
+\label{sec:relatedwork}
 

### 2008-02-24 papers/reasoning-graphically/paper.tex
some stuff, added picture
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1353 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \usepackage{url}
-\urldef{\mailsa}\path|{alfred.hofmann,ursula.barth,ingrid.beyer,natalie.brecht,|
-\urldef{\mailsb}\path|christine.guenther,frank.holzwarth,piamaria.karbach,|
-\urldef{\mailsc}\path|anna.kramer,erika.siebert-cole,lncs}@springer.com|
+\urldef{\mailsa}\path|l.dixon@ed.ac.uk|
+\urldef{\mailsb}\path|ross.duncan@comlab.ox.ac.uk|
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
@@
 % unless you accept that it will be published
-\institute{University of Edinburgh \and
-University of Oxford
+\institute{\mailsa, University of Edinburgh
+\and \mailsb, University of Oxford
 %\mailsa\\
@@
 extension, which we call \ref{graph patterns}, adds two extra kinds of
-nodes to the graph: {\em banged nodes} which informally represent an
-arbitrary number of copies of that node, and {\em variable nodes}
-which informally represent some as yet unkown node in a graph.
+nodes to the graph. Namely, {\em banged boxes} which informally
+represent an arbitrary number of copies of a node (where the node is
+defined as some graph), and {\em variable nodes} which informally
+represent some, as yet unkown, node in a graph.
 
-\subsection{Banged Nodes}
+\subsection{Bang-Boxes}
+
+Bang-Boxes
 
@@
 
+In graph patterns, variable nodes generalise the concept of a
+half-edge. A graph with variable nodes is given the following
+interpretation in the underyling semantics of our concrete graphs:
+
+Definition: super-type: given a node with type T
+
+\begin{itemize}
+\item let the node kinds be denoted by the set $K$
+\item given a graph $G$ containing a set of variable nodes $V$ 
+  a given node $n$, it's type is written $T_n$,
+\end{itemize}
+
 \subsection{Compositonality}
 
-\section{Graph Rewriting}
+\subsection{Instantiation}
+
+\section{Graph Pattern Rewriting}
 \label{sec:rewriting}
@@
 \subsection{Rules}
-\subsection{Matching}
+\subsection{Unification}
 \subsection{Substitution}
+
+\section{Node Expressions}
+\label{sec:node-expressions}
 
@@
 \label{sec:relatedwork}
+
+Initially one might be tempted to think that the usual notion of
+subgraph might surfice for expressing matching, however it allows
+additional edges on nodes where, because of the tensorial semantics of
+our graphs, we do not. Our work provides a restriction of the usual
+subgraph definition that is sound for Compact Closed Categories.
+
+Link Graphs and their extention to BiGraphs also encapsualte a very
+different kind of semantics where they allow edges to go to multiple
+nodes. An interesting area of further work would be to consider our
+graph patterns for these, and other, graph based formalisms.
+
+Graph Grammars...
+
+Other graph rewriting...
+
+Matrix Based Graph Tansformation...
+
+Other Graph Transformation...
 

### 2008-02-24 papers/reasoning-graphically/paper.tex
more stuff
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1354 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\subsection{Bang-Boxes}
+\subsection{!-Boxes}
 
-Bang-Boxes
+!-Boxes
 
@@
 
+instantiation of a variable node:
+
 \begin{itemize}
-\item let the node kinds be denoted by the set $K$
-\item given a graph $G$ containing a set of variable nodes $V$ 
-  a given node $n$, it's type is written $T_n$,
+\item let the set of node kinds be denoted $K$
+\item given a graph $G$, containing a variable node $x$, where for a
+  given non-vaiable node $n$ it's type is written $T_n$, the concrete
+  graph is:
+
+$$
+\{G[x/]  | x \in V \}
+$$
+
+\item
+
 \end{itemize}
@@
 \subsection{Compositonality}
+
+By enriching graphs with variable-nodes and banged-boxes, to form
+graph-patterns, we have lost the compositionality principle from the
+underlying concrete graphs. However, this can be restored and shown to
+correspond to the underyling compositionality principle.
+
+Type Expressions: We introduce a new form of type for a pattern
+graph. In particular,
 

### 2008-02-24 papers/reasoning-graphically/paper.tex
working in the comonoid equations
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1355 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 %\usepackage[dvips]{color}
-%\usepackage{graphics}
 %\usepackage{multicol}
@@
 \usepackage{allmtt}
-%\usepackage{amsmath}
-%\usepackage{amssymb}
 %\usepackage{pstricks,pst-node,pst-tree} % PS Tricks for diagrams
 %\usepackage{proof2}
-
-\usepackage{amssymb}
 \setcounter{tocdepth}{3}
 \usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{latexsym}
+\usepackage{stmaryrd}
+%\usepackage{mbboard}
+\usepackage{xspace}
+%\usepackage{diagrams}
 
@@
 \newcommand{\tensor}{\otimes}
+
+%%-----------------------------------------------------------------
+%%%% Ross's handy macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%-----------------------------------------------------------------
+
+\newcommand{\figureline}{\rule{\textwidth}{0.5pt}}
+\newcommand{\figureend}{\rule{\textwidth}{0.5pt}}
+
+\newcommand{\dimm}{\mathrm{dim}}
+
+% aliases
+\newcommand{\infinity}{\infty}
+\newcommand{\iso}{\cong}
+\newcommand{\isomorphism}{\cong}
+
+% ``semantic'' brakets
+\newcommand{\denote}[1]{% ---------
+\llbracket #1 \rrbracket} 
+% name / coname 
+\newcommand{\name}[1]{%--------
+\ulcorner #1 \urcorner}
+\newcommand{\coname}[1]{%
+\llcorner #1 \lrcorner}
+
+\newcommand{\sizeof}[1]{% \sizeof{x} == |x|
+  \left|#1\right|}
+
+%-------------------------------------------------------
+%  Useful macros for all things categorical
+%-------------------------------------------------------
+
+\newcommand{\dom}{\operatorname{dom}}
+\newcommand{\cod}{\operatorname{cod}}
+\newcommand{\Tr}{\operatorname{Tr}}
+
+% Objects of ???
+\newcommand{\OBJ}[1]{\ensuremath{\mathrm{Obj}_{#1}}}
+%
+% Objects of {\cal ??}
+\newcommand{\OBJC}[1]{\ensuremath{\mathrm{Obj}_{{\cal #1}}}}
+
+%
+% Arrows of ???
+\newcommand{\ARR}[1]{\ensuremath{\mathrm{Arr}_{#1}}}
+%
+% Arrows of {\cal ??}
+\newcommand{\ARRC}[1]{\ensuremath{\mathrm{Arr}_{{\cal #1}}}}
+
+% Caligraphic category names
+\newcommand{\catA}{\ensuremath{{\cal A}}\xspace}
+\newcommand{\catB}{\ensuremath{{\cal B}}\xspace}
+\newcommand{\catC}{\ensuremath{{\cal C}}\xspace}
+\newcommand{\catD}{\ensuremath{{\cal D}}\xspace}
+\newcommand{\catE}{\ensuremath{{\cal E}}\xspace}
+\newcommand{\catF}{\ensuremath{{\cal F}}\xspace}
+\newcommand{\catG}{\ensuremath{{\cal G}}\xspace}
+\newcommand{\catH}{\ensuremath{{\cal H}}\xspace}
+\newcommand{\catP}{\ensuremath{{\cal P}}\xspace}
+\newcommand{\catQ}{\ensuremath{{\cal Q}}\xspace}
+
+%sub scripted identity morphisms
+\newcommand{\id}[1]{\ensuremath{\mathrm{id}_{#1}}}
+
+% Boldified names of useful categories
+%
+\newcommand{\vectfd}[1]{% category of finite dimensional vector spaces over some field
+\ensuremath{\textbf{Vect}^{\mathrm{fd}}_{#1}}\xspace}
+\newcommand{\vectfdc}{% category of finite dimensional vector spaces over complexes
+\vectfd{\mathbb{C}}}
+\newcommand{\catRel}{% the category of sets and relations
+\ensuremath{\textbf{Rel}}\xspace}
+\newcommand{\catSet}{% the category of sets and functions
+\ensuremath{\textbf{Set}}\xspace}
+\newcommand{\catCat}{% the category of (small) categories
+\ensuremath{\textbf{Cat}}\xspace}
+\newcommand{\catInvCat}{% the category of involutive categories
+\ensuremath{\textbf{InvCat}}\xspace}
+\newcommand{\catComCl}{% the category of compact closed  categories
+\ensuremath{\textbf{ComCl}}\xspace}
+\newcommand{\catSComCl}{% the category of strongly compact closed  categories
+\ensuremath{\textbf{SComCl}}\xspace}
+\newcommand{\catGrph}{% the category of graphs
+\ensuremath{\textbf{Grph}}\xspace}
+\newcommand{\qubit}{% the category of qubits
+\ensuremath{\textbf{Qubit}}\xspace}
+\newcommand{\fdhilb}{% the category of finite dimensional hilbert space
+\ensuremath{\textbf{FDHilb}}\xspace}
+\newcommand{\catInvCom}{% the category of involutive categories
+\ensuremath{\textbf{InvCom}}\xspace}
+\newcommand{\catCom}{% the category of compact closed  categories
+\ensuremath{\textbf{Com}}\xspace}
+%%%% End of Ross's lovely categorcal macros--------
+%%--------------------------------------------
 
@@
 \label{sec:mono-categ-graphs}
+
+representation of freely generated compact closed  categories  as
+graphs;
+
+theorem: two arrows in a a freely generated compact closed  category
+are equal iff and only if their graphs are equal.
+
+Note that the notion of vertex and graph have the same external
+structure so it is possible to view a subgraph as a vertex and vice versa.
+
+\section{Quotients and Rewriting}
+\label{sec:quotients-rewriting}
+
+Unfortunately. or fortunately, few examples of interesting
+compact closed categories are freely generated.  In practice we work
+with cateories presented as generators and \emph{equations}.  For
+representing quantum computations, Coecke and Duncan
+\cite{Coecke:2008jo} propose the following collection of generators:
+\begin{description}
+\item[Objects] the self dual object $Q$;
+\item[Arrows] $\epsilon_Z : Q \to I$, $\delta_Z : Q \to Q \otimes Q$,
+  $\alpha_Z : Q \to Q$, $\epsilon_X : Q \to I$, $\delta_X : Q \to Q \otimes Q$,
+  $\alpha_X : Q \to Q$ and $H:Q\to Q$.
+\end{description}
+subject to  the following equations:
+\begin{gather}
+  (\delta \otimes \id{Q}) \circ \delta 
+   =   ( \id{Q} \otimes \delta) \circ \delta  
+  \tag{A}\label{comonoid-assoc}
+  \\
+  (\epsilon \otimes \id{Q}) \circ \delta 
+  = (\id{Q} \otimes \epsilon)\circ \delta 
+  = \id{Q} \tag{U} \label{comonoid-unit}
+  \\
+  \delta = \sigma \circ \delta \tag{C}\label{comonoid-comm}
+  \\
+  \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
+  \\
+  (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
+  = \delta \circ \delta^\dag \tag{F}\label{frobenius}
+\end{gather}
+
+its denotation is the
+  2-dimensional Hilbert space $\mathbb{C}^2$
 

### 2008-02-24 papers/reasoning-graphically/paper.tex
added make file generator to convert EPS to PDF files
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1356 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 %%%% End of Ross's lovely categorcal macros--------
+
+%%%%  use this to include graphics in the right place.
+\newcommand{\inlinegraphic}[2]{
+  %% todo -- make this thing calculate the height 
+  %% itself based on a global scaling factor
+  \dimendef\grafheight=255\dimendef\grafvshift=254
+  \grafheight=#1
+  \grafvshift=-0.5\grafheight
+  \advance\grafvshift by 0.5ex
+  \raisebox{\grafvshift}{\includegraphics[height=\grafheight]{#2}\xspace}
+}
+
 %%--------------------------------------------
@@
 
-Unfortunately. or fortunately, few examples of interesting
-compact closed categories are freely generated.  In practice we work
-with cateories presented as generators and \emph{equations}.  For
+Few examples of interesting compact closed categories are freely
+generated.  In practice we work with categories presented as
+generators and \emph{equations}.  For 
 representing quantum computations, Coecke and Duncan
@@
 \item[Objects] the self dual object $Q$;
-\item[Arrows] $\epsilon_Z : Q \to I$, $\delta_Z : Q \to Q \otimes Q$,
-  $\alpha_Z : Q \to Q$, $\epsilon_X : Q \to I$, $\delta_X : Q \to Q \otimes Q$,
-  $\alpha_X : Q \to Q$ and $H:Q\to Q$.
+\item[Arrows] Two families of arrows:
+  \begin{align}
+  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
+  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
+  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
+  \end{align}
+  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
 \end{description}
-subject to  the following equations:
+We represent these two families, and their adjoints, by colours in our
+graphical notation:
+\begin{gather*}
+  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
+  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
+  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
+  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
+  \alpha_Z = \inlinegraphic{1.5em}{alpha} \qquad
+\\  
+  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
+  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
+%  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
+%  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
+  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
+\end{gather*}
+The $H$ arrow is denoted by \inline
+Each family $(\delta, \epsilon, \alpha)$ is governed by a set of
+equations defining a forms a \emph{classical structure}
+\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima} with a
+\emph{phase group}.  The requisite equations state that $(\delta,
+\epsilon )$ form a cocommutative comonoid:
 \begin{gather}
@@
   \\
-  \delta = \sigma \circ \delta \tag{C}\label{comonoid-comm}
+  \delta = \sigma \circ \delta. \tag{C}\label{comonoid-comm}
+\end{gather}
+Their adjoints $(\delta^\dag,\epsilon^\dag)$ form a commutative
+monoid':
+\begin{gather}
+  \delta^\dag \circ (\delta^\dag \otimes \id{Q})
+  =    \delta^\dag   \circ ( \id{Q} \otimes \delta^\dag) 
+  \tag{A}\label{monoid-assoc}
   \\
+  \delta^\dag  \circ (\epsilon^\dag \otimes \id{Q}) 
+  = \delta^\dag \circ (\id{Q} \otimes \epsilon^\dag)
+  = \id{Q} \tag{U} \label{monoid-unit}
+  \\
+  \delta^\dag = \delta^\dag \circ
+  \sigma. \tag{C}\label{monoid-comm}.
+\end{gather}
+Futher, these jointly satisfy the \emph{isometry} and \emph{frobenius} laws:
+\begin{gather}
   \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
@@
   (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
-  = \delta \circ \delta^\dag \tag{F}\label{frobenius}
+  = \delta \circ \delta^\dag. \tag{F}\label{frobenius}
 \end{gather}

### 2008-02-24 papers/reasoning-graphically/paper.tex
more stuff abut quantum computayion
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1358 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
   \advance\grafvshift by 0.5ex
-  \raisebox{\grafvshift}{\includegraphics[height=\grafheight]{#2}\xspace}
+  \raisebox{\grafvshift}{\includegraphics[height=\grafheight]{images/#2}\xspace}
 }
@@
   \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
-  \alpha_Z = \inlinegraphic{1.5em}{alpha} \qquad
+  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
 \\  
@@
   \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
-%  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
-%  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
+  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
+  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
   \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
 \end{gather*}
-The $H$ arrow is denoted by \inline
+The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  The free
+compact closed category  is then given by all graphs formed by
+composing and tensoring these basic graphs.
+
+To model the behaviour of quantum systems these generators must
+satisfy certain equations.
+
 Each family $(\delta, \epsilon, \alpha)$ is governed by a set of
@@
 
-its denotation is the
-  2-dimensional Hilbert space $\mathbb{C}^2$
+The need to formalise these equations leads naturally to a rewriting
+theory
 

### 2008-02-24 papers/reasoning-graphically/paper.tex
more pictures
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1359 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-To model the behaviour of quantum systems these generators must
-satisfy certain equations.
+To model the behaviour of quantum systems certain additional equations
+must be satisfied.  Each family $(\delta, \epsilon)$ must form a
+\emph{classical structure}
+\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
+$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
+$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
+together they must satisfy the isometry and frobenius equations. In
+addition the family $\alpha$ must form an abelian group with $\alpha^\dag
+= -\alpha$.  These equations are presented graphically below for the
+green family;  the red vertices obey all the same equations
+\begin{description}
+\item[Comonoid Laws] 
+\[
+\begin{array}{ccccccccccccccc}
+  \inlinegraphic{3.5em}{comonoid-assoc1}  &=\;&
+  \inlinegraphic{3.5em}{comonoid-assoc2}
+  &\qquad\qquad&
+  \inlinegraphic{3.7em}{comonoid-unit1}  &=\;&
+  \inlinegraphic{3.7em}{comonoid-unit2}  &=\;&
+  \inlinegraphic{3.7em}{comonoid-unit3} 
+  &\qquad\qquad&
+  \inlinegraphic{3.5em}{comonoid-comm1}  &=\;&
+  \inlinegraphic{3.5em}{comonoid-comm2}
+\end{array}
+\]
+\item[Monoid Laws] 
+\[
+\begin{array}{ccccccccccccccc}
+  \inlinegraphic{3.5em}{monoid-assoc1}  &=\;&
+  \inlinegraphic{3.5em}{monoid-assoc2}
+  &\qquad\qquad&
+  \inlinegraphic{3.7em}{monoid-unit1}  &=\;&
+  \inlinegraphic{3.7em}{monoid-unit2}  &=\;&
+  \inlinegraphic{3.7em}{monoid-unit3} 
+  &\qquad\qquad&
+  \inlinegraphic{3.5em}{monoid-comm1}  &=\;&
+  \inlinegraphic{3.5em}{monoid-comm2}
+\end{array}
+\]
+\item[Isometry] 
+\item[Frobenius]
+  \[
+  \begin{array}{ccccc}
+    \inlinegraphic{3.5em}{frobenius1}  &=\;&
+    \inlinegraphic{3.5em}{frobenius2} &=\;&
+    \inlinegraphic{3.5em}{frobenius3}
+  \end{array}
+  \]
+  \item[Abelian Unitary Group]
+\end{description}
 
@@
 equations defining a forms a \emph{classical structure}
-\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima} with a
+ with a
 \emph{phase group}.  The requisite equations state that $(\delta,

### 2008-02-24 papers/reasoning-graphically/paper.tex
some changes
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1361 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \begin{center}
-  \scalebox{1.0}{\includegraphics{images/spider_lhs.eps}}
+  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
 \end{center}
@@
 In graph patterns, variable nodes generalise the concept of a
-half-edge. A graph with variable nodes is given the following
-interpretation in the underyling semantics of our concrete graphs:
+half-edge. Variable nodes, unlike boundary-nodes in the concrete
+graph, are not classified into domain and co-domain. The intuition of
+a variable node is that it can be replaced by any concrete node. Thus,
+unlike nodes in conrete a graph, variable node's do not define an
+ordering on the edges that are incident. Instead, the edges that touch
+a variable node are simply a set.
+
+The type of a variable nodes is thus defined as a multiset rather than
+a tensor. Any tensor type can be lifted into a multiset type in the
+obvious manner, which we denote by the operation {\tt
+  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
+  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
+is an instance of a multiset type, $M$, when it's lifting into a
+multiset is a super-set of $M$: $ M \subseteq
+\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
+here is that an instance of variable node is allowed to have as many
+extra edges as it likes as long as it at least has the ones mentioned
+in the original graph. A variable node, $a$ is be said to an instance
+of another variable node $b$ when the multiset represnting the type of
+$a$ is a superset of the multiset for the type of $b$. Thus, with
+regard to the underlying semantics, $a$ being a subtype of $b$ means
+that the set of instances of $a$ are a subset of the instances of $b$.
+
+\begin{figure}[t]
+%  \scalebox{1.0}{\includegraphics{images/node-var-instance.eps}}
+*** Add figure *** 
+\label{node-variable-instances-fig}\caption{The concrete graphs $a$,
+  $b$, and $c$ are instances of $x$, a graph with variables nodes.}
+\end{figure}
+
+A graph with variable nodes is given a formal semantics by being
+interpretated as a set of concrete graphs. We refer members of
+interpretation set as instances. Formally, an instance of a graph with
+variable-nodes is the graph where every variable node has been
+replaced by some instance of it. For example, in
+Figure~\ref{node-variable-instances-fig}, the graphs $a$, $b$ and $c$
+are instances of the graph $x$. Observe that when a variable-node has
+degree one, then it can be instantiated as a boundary edge of either
+the domain or co-domain.
+
+\subsection{Compositonality}
+
+In this section, we define a compositonality principle for graphs
+patterns, that is graphs with variable-nodes and !-boxes. This is
+related to compositionality of the underlying concrete graphs by the
+following instantiation homomorphism:
+
+By enriching graphs with to form
+graph-patterns, we need to define compositionality principle from the
+underlying concrete graphs. In this section, we restore
+compositionality and show its correspondence to the underyling
+principle in concrete graphs.
 
@@
 
-instantiation of a variable node:
-
-\begin{itemize}
-\item let the set of node kinds be denoted $K$
-\item given a graph $G$, containing a variable node $x$, where for a
-  given non-vaiable node $n$ it's type is written $T_n$, the concrete
-  graph is:
-
-$$
-\{G[x/]  | x \in V \}
-$$
-
-\item
-
-\end{itemize}
-
-\subsection{Compositonality}
-
-By enriching graphs with variable-nodes and banged-boxes, to form
-graph-patterns, we have lost the compositionality principle from the
-underlying concrete graphs. However, this can be restored and shown to
-correspond to the underyling compositionality principle.
-
-Type Expressions: We introduce a new form of type for a pattern
-graph. In particular,
+Type Expressions: we introduce a new form of type for a pattern
+graph. In particular, a variable node with
 

### 2008-02-25 papers/reasoning-graphically/paper.tex
couple more diags
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1362 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
minor changes
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1366 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \begin{array}{ccccccccccccccc}
-  \inlinegraphic{3.5em}{comonoid-assoc1}  &=\;&
-  \inlinegraphic{3.5em}{comonoid-assoc2}
+  \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
+  \inlinegraphic{3em}{comonoid-assoc2}
   &\qquad\qquad&
-  \inlinegraphic{3.7em}{comonoid-unit1}  &=\;&
-  \inlinegraphic{3.7em}{comonoid-unit2}  &=\;&
-  \inlinegraphic{3.7em}{comonoid-unit3} 
+  \inlinegraphic{3em}{comonoid-unit1}  &=\;&
+  \inlinegraphic{3em}{comonoid-unit2}  &=\;&
+  \inlinegraphic{3em}{comonoid-unit3} 
   &\qquad\qquad&
-  \inlinegraphic{3.5em}{comonoid-comm1}  &=\;&
-  \inlinegraphic{3.5em}{comonoid-comm2}
+  \inlinegraphic{3em}{comonoid-comm1}  &=\;&
+  \inlinegraphic{3em}{comonoid-comm2}
 \end{array}
@@
 \begin{array}{ccccccccccccccc}
-  \inlinegraphic{3.5em}{monoid-assoc1}  &=\;&
-  \inlinegraphic{3.5em}{monoid-assoc2}
+  \inlinegraphic{3em}{monoid-assoc1}  &=\;&
+  \inlinegraphic{3em}{monoid-assoc2}
   &\qquad\qquad&
-  \inlinegraphic{3.7em}{monoid-unit1}  &=\;&
-  \inlinegraphic{3.7em}{monoid-unit2}  &=\;&
-  \inlinegraphic{3.7em}{monoid-unit3} 
+  \inlinegraphic{3em}{monoid-unit1}  &=\;&
+  \inlinegraphic{3em}{monoid-unit2}  &=\;&
+  \inlinegraphic{3em}{monoid-unit3} 
   &\qquad\qquad&
-  \inlinegraphic{3.5em}{monoid-comm1}  &=\;&
-  \inlinegraphic{3.5em}{monoid-comm2}
+  \inlinegraphic{3em}{monoid-comm1}  &=\;&
+  \inlinegraphic{3em}{monoid-comm2}
 \end{array}
@@
 \item[Isometry]
+\[
+\begin{array}{ccc}
+  \inlinegraphic{3.5em}{isometry1}  &=\;&
+  \inlinegraphic{3.5em}{isometry2}  
+\end{array}
+\]
 \item[Frobenius]
@@
   \begin{array}{ccccc}
-    \inlinegraphic{3.5em}{frobenius1}  &=\;&
-    \inlinegraphic{3.5em}{frobenius2} &=\;&
-    \inlinegraphic{3.5em}{frobenius3}
+    \inlinegraphic{3em}{frobenius1}  &=\;&
+    \inlinegraphic{3em}{frobenius2} &=\;&
+    \inlinegraphic{3em}{frobenius3}
   \end{array}
@@
 a variable node is that it can be replaced by any concrete node. Thus,
-unlike nodes in conrete a graph, variable node's do not define an
+unlike nodes in conrete a graph, variable nodes do not define an
 ordering on the edges that are incident. Instead, the edges that touch
@@
 \mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
-here is that an instance of variable node is allowed to have as many
+here is that an instance of a variable node is allowed to have as many
 extra edges as it likes as long as it at least has the ones mentioned
 in the original graph. A variable node, $a$ is be said to an instance
-of another variable node $b$ when the multiset represnting the type of
+of another variable node $b$ when the multiset representing the type of
 $a$ is a superset of the multiset for the type of $b$. Thus, with
@@
 In this section, we define a compositonality principle for graphs
-patterns, that is graphs with variable-nodes and !-boxes. This is
+patterns, that is for graphs with variable-nodes and !-boxes. This is
 related to compositionality of the underlying concrete graphs by the
 following instantiation homomorphism:
+
+$\llbracket G \cric H \rrbracket = \llbracket G \rrbracket \cric \llbracket H \rrbracket$
 

### 2008-02-25 papers/reasoning-graphically/paper.tex
a couple more graphical equations
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1367 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
+need to define embedding vs strict embedding.
+
 theorem: two arrows in a a freely generated compact closed  category
@@
   \item[Abelian Unitary Group]
+    \[
+    \begin{array}{ccccccccccccccc}
+      \inlinegraphic{2.2em}{group1}  &:=\;&
+      \inlinegraphic{2.2em}{group2}  &=\;&
+      \inlinegraphic{2.2em}{group3} 
+      &\qquad\qquad&
+      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
+      \inlinegraphic{2.2em}{group5} 
+      &\qquad\qquad&
+      \inlinegraphic{3.5em}{group6}  &=\;&
+      \inlinegraphic{3.5em}{group7}  &=\;&
+      \inlinegraphic{3.5em}{group8} 
+    \end{array}
+    \]
 \end{description}
-
-Each family $(\delta, \epsilon, \alpha)$ is governed by a set of
-equations defining a forms a \emph{classical structure}
- with a
-\emph{phase group}.  The requisite equations state that $(\delta,
-\epsilon )$ form a cocommutative comonoid:
-\begin{gather}
-  (\delta \otimes \id{Q}) \circ \delta 
-   =   ( \id{Q} \otimes \delta) \circ \delta  
-  \tag{A}\label{comonoid-assoc}
-  \\
-  (\epsilon \otimes \id{Q}) \circ \delta 
-  = (\id{Q} \otimes \epsilon)\circ \delta 
-  = \id{Q} \tag{U} \label{comonoid-unit}
-  \\
-  \delta = \sigma \circ \delta. \tag{C}\label{comonoid-comm}
-\end{gather}
-Their adjoints $(\delta^\dag,\epsilon^\dag)$ form a commutative
-monoid':
-\begin{gather}
-  \delta^\dag \circ (\delta^\dag \otimes \id{Q})
-  =    \delta^\dag   \circ ( \id{Q} \otimes \delta^\dag) 
-  \tag{A}\label{monoid-assoc}
-  \\
-  \delta^\dag  \circ (\epsilon^\dag \otimes \id{Q}) 
-  = \delta^\dag \circ (\id{Q} \otimes \epsilon^\dag)
-  = \id{Q} \tag{U} \label{monoid-unit}
-  \\
-  \delta^\dag = \delta^\dag \circ
-  \sigma. \tag{C}\label{monoid-comm}.
-\end{gather}
-Futher, these jointly satisfy the \emph{isometry} and \emph{frobenius} laws:
-\begin{gather}
-  \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
-  \\
-  (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
-  = \delta \circ \delta^\dag. \tag{F}\label{frobenius}
-\end{gather}
 

### 2008-02-25 papers/reasoning-graphically/paper.tex
working...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1370 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \newcommand{\tensor}{\otimes}
+
+\newcommand{\vinterp}[1]{\lfloor\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rfloor}
+\newcommand{\binterp}[1]{\lceil\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rceil}
+\newcommand{\minterp}[1]{\llbracket #1 \rrbracket}
+\newcommand{\sinterp}[1]{| #1 |}
 
@@
 
-\subsection{!-Boxes}
-
-!-Boxes
-
 \subsection{Variable Nodes}
@@
 In graph patterns, variable nodes generalise the concept of a
-half-edge. Variable nodes, unlike boundary-nodes in the concrete
+boundary node. Variable nodes, unlike boundary-nodes in the concrete
 graph, are not classified into domain and co-domain. The intuition of
@@
 The type of a variable nodes is thus defined as a multiset rather than
-a tensor. Any tensor type can be lifted into a multiset type in the
+a tensor. Any tensor type can be lifted to a multiset type in the
 obvious manner, which we denote by the operation {\tt
@@
 \label{node-variable-instances-fig}\caption{The concrete graphs $a$,
-  $b$, and $c$ are instances of $x$, a graph with variables nodes.}
+  $b$, and $c$ are instances of $x$, a graph with variable nodes.}
 \end{figure}
 
-A graph with variable nodes is given a formal semantics by being
-interpretated as a set of concrete graphs. We refer members of
-interpretation set as instances. Formally, an instance of a graph with
-variable-nodes is the graph where every variable node has been
-replaced by some instance of it. For example, in
-Figure~\ref{node-variable-instances-fig}, the graphs $a$, $b$ and $c$
-are instances of the graph $x$. Observe that when a variable-node has
-degree one, then it can be instantiated as a boundary edge of either
-the domain or co-domain.
+A graph with variable nodes, $G$, can be given a formal semantics by
+being interpretated as a set of concrete graphs which is denoted by
+$\vinterp{G}$. We refer to members of this set as
+instances. Formally, an instance of a graph with variable nodes is the
+graph where every variable node has been replaced by some instance of
+it. For example, in Figure~\ref{node-variable-instances-fig}, the
+graphs $a$, $b$ and $c$ are instances of the graph $x$. Observe that
+when a variable node has degree one, then it can be instantiated as a
+boundary node in either the domain or co-domain.
 
-\subsection{Compositonality}
+\subsection{Plugging}
+
+\subsection{!-Boxes}
+
+The set of concrete graphs for a graph $G$ with !-boxes, is denoted by
+$\binterp{G}$.
+
+\subsection{Matching}
+
+The purpose of our graph patterns is to express and formalise, in a
+declarative manner, the intuitive idea of graph matching for graphs
+that represent Compact Closed Categories. In the previous sections we
+defined ways to extend the concrete graph representation with variable
+nodes and !-boxes. In this section we combine these to first define
+the matching relation between graph patterns and concrete graphs and
+then introduce a matching relation between graph patterns.
+
+\subsubsection{Concrete Matches}
+
+We give a semantics to these in terms of a set of concrete graphs,
+denoted by $\minterp{G}$, and say that a member of this set is
+\emph{matched} by the graph pattern $G$. Formally, the set of matches
+are defined as follows:
+
+$\minterp{G} = \sinterp{\vinterp{\binterp{G}}}$
+
+where $\sinterp{G}$ represents the set of graphs that strictly contain
+the graph $G$.
+
+\subsubsection{Pattern Matches}
 
 In this section, we define a compositonality principle for graphs
-patterns, that is for graphs with variable-nodes and !-boxes. This is
+patterns, that is for graphs with variable nodes and !-boxes. This is
 related to compositionality of the underlying concrete graphs by the
@@
 
-$\llbracket G \cric H \rrbracket = \llbracket G \rrbracket \cric \llbracket H \rrbracket$
+$\llbracket G \circ H \rrbracket = \llbracket G \rrbracket \circ \llbracket H \rrbracket$
 

### 2008-02-25 papers/reasoning-graphically/paper.tex
work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1371 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-In graph patterns, variable nodes generalise the concept of a
-boundary node. Variable nodes, unlike boundary-nodes in the concrete
-graph, are not classified into domain and co-domain. The intuition of
-a variable node is that it can be replaced by any concrete node. Thus,
-unlike nodes in conrete a graph, variable nodes do not define an
-ordering on the edges that are incident. Instead, the edges that touch
-a variable node are simply a set.
-
-The type of a variable nodes is thus defined as a multiset rather than
-a tensor. Any tensor type can be lifted to a multiset type in the
-obvious manner, which we denote by the operation {\tt
-  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
-  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
+In graph patterns, variable nodes generalise the concept of a boundary
+node. Variable nodes, unlike boundary-nodes in the concrete graph, are
+not classified into domain and co-domain. The intuition of a variable
+node is that it can be replaced by any concrete node. Thus, unlike
+nodes in conrete a graph, variable nodes do not define an ordering on
+the edges that are incident. Instead, the edges that touch a variable
+node are simply a set. The type of a variable nodes is thus defined as
+a multiset rather than a tensor. Any tensor type can be lifted to a
+multiset type in the obvious manner, which we denote by the operation
+{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
+  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
 is an instance of a multiset type, $M$, when it's lifting into a
@@
 extra edges as it likes as long as it at least has the ones mentioned
-in the original graph. A variable node, $a$ is be said to an instance
-of another variable node $b$ when the multiset representing the type of
-$a$ is a superset of the multiset for the type of $b$. Thus, with
-regard to the underlying semantics, $a$ being a subtype of $b$ means
-that the set of instances of $a$ are a subset of the instances of $b$.
+in the original graph.
+
+A variable node, $a$ is be said to an instance of another variable
+node $b$ when the multiset representing the type of $a$ is a superset
+of the multiset for the type of $b$. Thus, with regard to the
+underlying semantics, $a$ being a subtype of $b$ means that the set of
+instances of $a$ are a subset of the instances of $b$.
 
@@
 
+A graph without boundary nodes but which may have variable nodes is
+called a \emph{variable-node graph}. Such graphs are complete in the
+sense that for every concrete graph there is a variable node graph
+that can be instantiated to it: $\forall g.\,\exists G.\,g \in
+\vinterp{G}$. The witness for this proof is made by simply replacing
+all bounary nodes with variable nodes.
+
+\subsection{Open Embedding}
+
 \subsection{Plugging}
+
+While concrete graphs can be composed sequentially as well as tensored
+in an adjacent manner, variable-node graphs enjoy a single but
+somewhat wilder joining operation which we call \emph{plugging}.
 
@@
 
+\subsubsection{Making !-Box Graphs}
+
 \subsection{Matching}
@@
 declarative manner, the intuitive idea of graph matching for graphs
-that represent Compact Closed Categories. In the previous sections we
-defined ways to extend the concrete graph representation with variable
-nodes and !-boxes. In this section we combine these to first define
-the matching relation between graph patterns and concrete graphs and
-then introduce a matching relation between graph patterns.
+that represent Compact Closed Categories. In this section we combine
+the two extensions to the concrete graph representation, namely
+variable nodes and !-boxes, to provide a matching relation between
+graph patterns and concrete graphs. We then introduce a matching
+relation directly between graph patterns. The former relation gives a
+semantics for our system and ensures soundness, while the latter
+supports deirved equations and thus provides the theoretical basis for
+our formulation of rewriting.
 
-\subsubsection{Concrete Matches}
+\subsubsection{Concrete Matches.}
 
-We give a semantics to these in terms of a set of concrete graphs,
-denoted by $\minterp{G}$, and say that a member of this set is
-\emph{matched} by the graph pattern $G$. Formally, the set of matches
-are defined as follows:
+We give a semantics for a pattern graph $G$ in terms of a set of
+concrete graphs, denoted by $\minterp{G}$, and say that members of
+this set are \emph{matched} by $G$. Formally, the set of matches are
+defined as follows:
 
@@
 
-where $\sinterp{G}$ represents the set of graphs that strictly contain
-the graph $G$.
+\noindent where $\sinterp{G}$ represents the set of graphs that
+strictly contain the graph $G$. Thus, a concrete graph $g$ is matched
+by a graph pattern $G$, written $g \in G$, when an instance of the
+pattern is a strict subgraph of the matched graph.
 
-\subsubsection{Pattern Matches}
+Graph patterns are complete in the sense that for every concrete graph
+there is a pattern that matches it. This is a trivial consequence of
+the completeness of variable-node extension to graphs, which is
+unaffected by allowing !-boxes.
 
-In this section, we define a compositonality principle for graphs
-patterns, that is for graphs with variable nodes and !-boxes. This is
-related to compositionality of the underlying concrete graphs by the
-following instantiation homomorphism:
+\subsubsection{Graph Pattern Matches.}
 
-$\llbracket G \circ H \rrbracket = \llbracket G \rrbracket \circ \llbracket H \rrbracket$
+The definition of one pattern matching another one is simply that it
+is more general. Formally, a graph pattern, $G$, matches another graph
+pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This provides
+the specification for a matching algorithm which we provide an
+implementation of in~\S\ref{sec:rewriting}.
 
-By enriching graphs with to form
-graph-patterns, we need to define compositionality principle from the
-underlying concrete graphs. In this section, we restore
-compositionality and show its correspondence to the underyling
-principle in concrete graphs.
+Our representation has been carefully designed so that this concept
+corresponds to what we call a \emph{respectful embedding} of $G$ into
+$H$.
 
-Definition: super-type: given a node with type T
+***
 
-Type Expressions: we introduce a new form of type for a pattern
-graph. In particular, a variable node with
-
-\subsection{Instantiation}
+This is the basis for our implementation and is thus also the basis
+for its correctness argument.
 

### 2008-02-25 papers/reasoning-graphically/paper.tex
hjdfkshfj
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1372 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
+\begin{definition}
+  A \emph{directed graph} consists of a 4-tuple $(N,E,s,t)$ where $V$
+  and $E$ are sets, respectively of \emph{nodes} and \emph{edges}, 
+\end{definition}
+
 representation of freely generated compact closed  categories  as
@@
 
-need to define embedding vs strict embedding.
+need to define embedding and subgraph
 
@@
   \]
+\item[Compact Structure] 
+  \[
+  \begin{array}{ccccccc}
+    \inlinegraphic{2.5em}{compact1}  &=\;&
+    \inlinegraphic{1.3em}{compact2}  
+    &\qquad\qquad&
+    \inlinegraphic{2.5em}{compact3}  &=\;&
+    \inlinegraphic{1.3em}{compact4}  
+  \end{array}
+  \]
   \item[Abelian Unitary Group]
@@
       \inlinegraphic{3.5em}{group8} 
+    \end{array}
+    \]
+  \item[Bilinearity] 
+    \[
+    \begin{array}{cccccccccccc}
+      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
+      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
+      \inlinegraphic{3.5em}{alpha-commute3} 
+      &\qquad\qquad&
+      \inlinegraphic{3.5em}{alpha-commute4}  &=\;&
+      \inlinegraphic{3.5em}{alpha-commute5}  &=\;&
+      \inlinegraphic{3.5em}{alpha-commute6} 
+    \end{array}
+    \]
+\end{description}
+Additionally, the two families satisfy equations which make them a
+\emph{scaled bialgebra}.
+\begin{description}
+\item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
+      \inlinegraphic{1.5em}{bialgebra2}$;  then:
+  \[
+  \begin{array}{ccccccccccccccc}
+      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
+      \inlinegraphic{3em}{bialgebra4}
+      \qquad\qquad
+      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
+      \inlinegraphic{1.5em}{bialgebra6}
+      \qquad\qquad
+      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
+      \inlinegraphic{1.5em}{bialgebra8} \\ \\
+      \inlinegraphic{3.5em}{bialgebra9}  &=\;&
+      \inlinegraphic{3em}{bialgebra10}
+      \qquad\qquad
+      \inlinegraphic{2.5em}{bialgebra11}  &=\;&
+      \inlinegraphic{1.5em}{bialgebra12}
+      \qquad\qquad
+      \inlinegraphic{2.5em}{bialgebra13}  &=\;&
+      \inlinegraphic{1.5em}{bialgebra14} 
   \end{array}
@@
 
+Interaction nets and interaction combinators (Lafont and others)
+
 \section{Conclusions}

### 2008-02-26 papers/reasoning-graphically/paper.tex
updated paper and .gvlock for svn:ignore
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1373 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 in the previous section, is too restrictive for automated reasoning
-about quantum computation. In particular, pictures such as the
-following are frequently needed:
+about quantum computation. In particular, pictures such the Spider
+Law, in Figure~\ref{spider-law-fig} are frequently needed. Such
+pictures represent an infinite family of concrete graphs. In this
+section we describe an extension of the representation for Compact
+Closed Categories that allows a graph-based representation that
+captures an infinite family of graphs. Notably, this allows us to
+represent laws about quantum computation, such as the Spider
+Law. Furthermore, this representation is still compositional in the
+sense that we can plug graphs together. The extension, which we call
+\ref{graph patterns}, adds two extra kinds of nodes to the
+graph. Namely, {\em banged boxes} which informally represent an
+arbitrary number of copies of a node (where the node itself can
+contain a graph of the appropriate type), and {\em variable nodes}
+which informally represent some, as yet unkown, node in a graph.
 
-\begin{center}
-  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
-\end{center}
-
-\noindent Such pictures represent an infinite family of concrete
-graphs. In this section we describe an extension of the representation
-for Compact Closed Categories that allows a graph-based representation
-that captures an infinite family of graphs. Notably, this allows us to
-represent laws about quantum computation, such as
-Spider-Law~\ref{spider-law-fig}. Furthermore, this representation is
-still compositional in the sense that we can compose graphs. The
-extension, which we call \ref{graph patterns}, adds two extra kinds of
-nodes to the graph. Namely, {\em banged boxes} which informally
-represent an arbitrary number of copies of a node (where the node is
-defined as some graph), and {\em variable nodes} which informally
-represent some, as yet unkown, node in a graph.
+%\begin{center}
+%  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
+%\end{center}
 
@@
 
-The set of concrete graphs for a graph $G$ with !-boxes, is denoted by
-$\binterp{G}$.
+The key feature of rules like the Spider Law is that an arbirary
+number of nodes can be matched. To support this we introduce an
+operation, !-boxing, on graph representations. We draw !-boxes around
+a set of chosen nodes and intuitively think of the resulting !-boxed
+graph as representing a set of graphs with an arbirary number of
+copies of the !-boxed nodes, where any connections to nodes outside
+the !-box are also copied. The !-boxes can also be instantiated with
+zero copies errasing all edges from the !-box. See
+Figure~\ref{bang-box-ex1-fig} for a simple example.
 
-\subsubsection{Making !-Box Graphs}
+A graph with !-boxes is simply a graph paired with a set of !-box node
+sets, which define the !-boxes. The !-box node sets must either be
+disjoint or strict subsets of each other. This is needed to avoid
+confusion with overlaping boxes.
+
+\subsubsection{Semantics.}
+
+We give a formal semantics to !-boxing a representation of graphs by
+defining the a !-box graph in terms of a set of graphs in the
+underlying representation. We denote this interpretation of a graph
+$G$ by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are
+instances of $G$. The definition of $\binterp{G}$ involves the
+following three operations\footnote{These correspond closely to the
+  linear logic rules for weakening, contraction and deriliction.} on a
+!-box and a graph that contains it:
+
+\begin{definition}
+  \emph{copy} copies a box in a graph to produce a new graph with two
+  copies of the boxed subgraph, with a second box around the newly
+  copied nodes. Any edges incident on both a node in the !-box and
+  outside it get copied so that there the old edge remains and the new
+  edge goes from the outer node to the new copy of the internal node.
+\end{definition}
+
+\begin{definition}
+  \emph{drop} simply removes the box, but leaves its contents in the
+  graph.
+\end{definition}
+
+\begin{definition}
+  \emph{kill} removes from the graph all nodes in the !-box set as
+  well as any incident edges to nodes in the box.
+\end{definition}
+
+\noindent The semantics of a !-boxed graph is then defined as the
+members without any !-boxes of the set closed by these operation.  For
+the purposes of this paper, the underlying graph representation of
+!-boxed graphs is variable-node graphs. The set of variable-node
+graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.
+
+\subsubsection{}
 
@@
 
-The purpose of our graph patterns is to express and formalise, in a
-declarative manner, the intuitive idea of graph matching for graphs
-that represent Compact Closed Categories. In this section we combine
-the two extensions to the concrete graph representation, namely
-variable nodes and !-boxes, to provide a matching relation between
-graph patterns and concrete graphs. We then introduce a matching
-relation directly between graph patterns. The former relation gives a
-semantics for our system and ensures soundness, while the latter
-supports deirved equations and thus provides the theoretical basis for
-our formulation of rewriting.
+The purpose of graph patterns is to formally express, in a declarative
+manner, the intuitive idea of graph matching commonly used in for
+Quantum computation. To do this, we combine the two extensions to the
+concrete graph representation, namely variable nodes and !-boxes. This
+provides a matching relation between graph patterns and concrete
+graphs as well as directly between graph patterns. The former relation
+gives a semantics for our system and ensures soundness, while the
+latter supports deirved equations and thus provides the theoretical
+basis for our formulation of rewriting.
 
@@
 
-We give a semantics for a pattern graph $G$ in terms of a set of
-concrete graphs, denoted by $\minterp{G}$, and say that members of
-this set are \emph{matched} by $G$. Formally, the set of matches are
-defined as follows:
+The semantics for a graph pattern $G$ in terms of a set of concrete
+graphs, is denoted by $\minterp{G}$. We say that members of this set
+are \emph{matched} by $G$. Formally, the set of matches are defined as
+follows:
 
-$\minterp{G} = \sinterp{\vinterp{\binterp{G}}}$
-
-\noindent where $\sinterp{G}$ represents the set of graphs that
-strictly contain the graph $G$. Thus, a concrete graph $g$ is matched
-by a graph pattern $G$, written $g \in G$, when an instance of the
-pattern is a strict subgraph of the matched graph.
+$\minterp{G} = (\vinterp{G'} | G' \in \binterp{G})$
 
@@
 pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This provides
-the specification for a matching algorithm which we provide an
-implementation of in~\S\ref{sec:rewriting}.
+the specification for a matching algorithm which we implement of
+in~\S\ref{sec:rewriting}. To make such a proof easier, we extend
+embedding and plugging of variable-node graphs to graph patterns.
 
 Our representation has been carefully designed so that this concept
-corresponds to what we call a \emph{respectful embedding} of $G$ into
+corresponds to what we call a \emph{open embedding} of $G$ into
 $H$.

### 2008-02-26 papers/reasoning-graphically/paper.tex
working on !box graphs
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1374 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
adding th start of the graphs stuff
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1375 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \usepackage{xspace}
-%\usepackage{diagrams}
+\usepackage{diagrams}
 
@@
 
+%%%% More handy things for me - rwd
+\newcommand{\TODO}[1]{%
+\typeout{WARNING!!! there is still a TODO left}
+\marginpar{\textbf{!TODO: }\emph{#1}}
+}
 %%--------------------------------------------
@@
 
+Want to do reasoning about quantum computation by manipulating graphs
+
+CAn formalise quantum mechanics in an abstract setting using compact
+or $\dag$-compact categories.
+
+We will give a description of graphs in general;  a specialisation of
+this gives the structure required for compact closed  categories.  The
+graph structure suffices to give a tight representation.
+
+We then introduce a concrete set of generators and equations for use
+in quantum computation.  We need rewrites to capture the non-logical
+axioms of this structure.   It turns out that we can most easily express
+these equations using an infinite family of ``spider rules'';  this
+motivates the  more general notion of pattern graphs.
+
+next section introduces the  notion of pattern graph, which is
+essentially a graph structured
+
 \section{Compact Closed Categories and Graphs}
@@
 
-\begin{definition}
-  A \emph{directed graph} consists of a 4-tuple $(N,E,s,t)$ where $V$
-  and $E$ are sets, respectively of \emph{nodes} and \emph{edges}, 
-\end{definition}
+\subsection{Graphs}
+\label{sec:graphs}
+
+A \emph{directed graph}\footnote{
+Equivalently:  a directed graph is a functor $G$ from  $\bullet
+\pile{\rTo\\\rTo} \bullet$ to \catSet;  a graph morphism is then a natural
+transformation  $f: G \Rightarrow H$.
+} consists of a 4-tuple $(V,E,s,t)$ where $V$
+and $E$ are sets, respectively of \emph{vertices}\footnote{We will use
+  the words ``vertex'' and ``node'' interchangeably.} and \emph{edges},
+and $s$ and $t$ are maps 
+\begin{diagram}
+  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
+\end{diagram}
+which we call \emph{source} and \emph{target}.  The \emph{in-degree}
+of a vertex $v$ is the cardinality of $t^{-1}(v)$; its out-degree is
+the cardinality of $s^{-1}(v)$.  The \emph{degree} of a vertex is
+the sum of its in- and out-degrees.
+
+Given graphs $G$ and $H$, a \emph{graph morphism} $f : G\to H$ consists of
+functions $f_E : E_G \to E_H$ and $f_V:V_G\to V_H$ such that:
+\begin{gather*}
+  s_H\circ f_E = f_V \circ s_G,\\
+  t_H\circ f_E = f_V \circ t_G.
+\end{gather*}
+\TODO{different kinds of morphism?}
+
+Augmented by some additional structure, graphs form a compact closed
+category.  We present an overview of this structure:  the details
+omitted here can be found in \cite{Duncan:thesis:2006}.
+
+A \emph{concrete graph} $\Gamma$ is 5-tuple $(G, \dom\Gamma, \cod\Gamma,
+<_{\text{in}(\cdot)}, <_{\text{out}(\cdot)})$ where:
+\begin{itemize}
+\item $G = (V,E,C,s,t)$ is a graph;
+\item $\dom\Gamma$ and $\cod\Gamma$ are totally ordered disjoint of
+  degree one vertices of $G$.
+\item $<_{\text{in}(\cdot)}$ is a family of maps, indexed by $V$ such
+  that $<_{\text{in}(v)} : \text{in}(v) \rTo^\isomorphism
+  \mathbb{N}_k$ where $k = \sizeof{\text{in}(v)}$.
+\item $<_{\text{out}(\cdot)}$ is a family of maps, indexed by $V$ such
+  that $<_{\text{out}(v)} : \text{out}(v) \rTo^\isomorphism
+  \mathbb{N}_{k'}$ where $k' = \sizeof{\text{out}(v)}$.
+\end{itemize}
+
+Since the sets $\dom\Gamma$ and $\cod\Gamma$ consist of vertices of degree
+one, we can assign a polarity to each one:  $v \mapsto +$ if the edge
+incident at $v$ is an incoming edge; $v \mapsto -$ otherwise.  Hence
+$\cod \Gamma$ and $\dom \Gamma$ are \emph{signed sets}.  Concrete
+graphs form a category whose objects are
+
+can be seen as giving a signing to each element of the ordinal
+one, we can view these sets as maps from the ordinal
+
+can be viewed as a maps from an
 
@@
 
-\subsubsection{}
+\subsubsection{Matching}
+
+\subsubsection{Plugging}
 

### 2008-02-27 papers/reasoning-graphically/paper.tex
plugging of variable node graphs
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1376 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 pictures represent an infinite family of concrete graphs. In this
-section we describe an extension of the representation for Compact
-Closed Categories that allows a graph-based representation that
-captures an infinite family of graphs. Notably, this allows us to
-represent laws about quantum computation, such as the Spider
-Law. Furthermore, this representation is still compositional in the
-sense that we can plug graphs together. The extension, which we call
-\ref{graph patterns}, adds two extra kinds of nodes to the
-graph. Namely, {\em banged boxes} which informally represent an
+section we describe an extension to the representation that captures
+an infinite family of graphs as a single finite graph. The extension,
+which we call \ref{graph patterns}, adds two extra kinds of nodes to
+the graph. Namely, {\em banged boxes} which informally represent an
 arbitrary number of copies of a node (where the node itself can
 contain a graph of the appropriate type), and {\em variable nodes}
-which informally represent some, as yet unkown, node in a graph.
+which informally represent some, as yet unkown, node in a graph. We
+also introduce a notion to replace graph composition, called graph
+plugging that allows graph patterns to put together to form larger
+graph patterns.
+
+%Notably, this
+%allows us to represent laws about quantum computation, such as the
+%Spider Law. Furthermore, this representation is still compositional in
+%the sense that we can plug graphs together.
 
@@
 not classified into domain and co-domain. The intuition of a variable
-node is that it can be replaced by any concrete node. Thus, unlike
-nodes in conrete a graph, variable nodes do not define an ordering on
-the edges that are incident. Instead, the edges that touch a variable
-node are simply a set. The type of a variable nodes is thus defined as
-a multiset rather than a tensor. Any tensor type can be lifted to a
-multiset type in the obvious manner, which we denote by the operation
-{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
-  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
-is an instance of a multiset type, $M$, when it's lifting into a
-multiset is a super-set of $M$: $ M \subseteq
+node is that it can be replaced by any subgraph of the right
+type. Thus, unlike nodes in conrete a graph, variable nodes do not
+define an ordering on the edges that are incident. Instead, the edges
+that touch a variable node are simply a set. The type of a variable
+nodes is thus defined as a multiset rather than a tensor. Any tensor
+type can be lifted to a multiset type in the obvious manner, which we
+denote by the operation {\tt mset\_of\_tensor}. For example, {\tt
+  mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
+say that a tensor-type, $T$, is an instance of a multiset type, $M$,
+when it's lifting into a multiset is a super-set of $M$: $ M \subseteq
 \mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
@@
 being interpretated as a set of concrete graphs which is denoted by
-$\vinterp{G}$. We refer to members of this set as
-instances. Formally, an instance of a graph with variable nodes is the
-graph where every variable node has been replaced by some instance of
-it. For example, in Figure~\ref{node-variable-instances-fig}, the
-graphs $a$, $b$ and $c$ are instances of the graph $x$. Observe that
-when a variable node has degree one, then it can be instantiated as a
-boundary node in either the domain or co-domain.
+$\vinterp{G}$. We refer to members of this set as instances. An
+instance of a graph with variable nodes is the graph with every
+variable node replaced by a subgraph of the right type. When a graph
+has several varibale-nodes, the replacement subgraphs may be
+intersecting. For example, in
+Figure~\ref{node-variable-instances-fig}, the graphs $a$, $b$ and $c$
+are instances of the graph $x$. A graph with variable nodes, and
+without boundary nodes, is called a \emph{variable-node graph}. Such
+graphs are complete in the sense that for every concrete graph there
+is a variable node graph that can be instantiated to it: $\forall
+g.\,\exists G.\,g \in \vinterp{G}$. The witness for this proof is made
+by simply replacing all bounary nodes with variable nodes.
 
-A graph without boundary nodes but which may have variable nodes is
-called a \emph{variable-node graph}. Such graphs are complete in the
-sense that for every concrete graph there is a variable node graph
-that can be instantiated to it: $\forall g.\,\exists G.\,g \in
-\vinterp{G}$. The witness for this proof is made by simply replacing
-all bounary nodes with variable nodes.
+Below, we give a more formal and constuctive account the semnatics by
+introducing a notion of an \emph{open embedding} and note that this
+corresponds to the intuitive idea of graph-matching. We then go on and
+introduce a way to join graphs with variable-nodes and note it's
+relationship to composition on the underlying concrete graphs.
 
-\subsection{Open Embedding}
+\subsection{Matching and Open Embedding}
 
-\subsection{Plugging}
+\begin{definition}
+  An \emph{open embedding} from $G$ to $H$ a strict embedding from the
+  non-variable subgraph of $G$ into the non-variable subgraph of $H$
+  and a mapping from the variable nodes in $G$ ...
+\end{definition}
+
+\subsection{Variable-node Plugging}
 
@@
 
+\begin{definition}
+  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of
+  strict partial homomorpihc mappings, $f_G : G \Longrightarrow H$ and
+  $f_H : H \Longrightarrow G$ such that $f_G$ and $f_H$ do not map
+  concrete nodes and for the nodes that are mapped \emph{agree}
+  following the definition given below.
+\end{definition}
+
+\begin{definition}
+  Two mappings $f_G$ and $f_H$ \emph{agree} when $\forall v \in
+  f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
+  f_G(v) \in f^{-1}_H(v)$. Intuitively this captures the idea a node
+  is one function is always mapped to one the nodes that maps to it in
+  the other function.
+\end{definition}
+
+We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
+overload the notation $\pi_p(G,H)$ to let it denote the graph that
+results from the plugging. This is defined to be the union of the
+nodes and edges from $G$ and $H$ where the nodes and edges mapped by
+the pair of plugging function are identified. Identification of nodes
+instantiates variable nodes: when a variable node is mapped to a
+concrete one, the result is a concrete node.
+
+While this is simply the intuitive way to combine graphs with variable
+nodes, the definition takes some care. In particular, it is important
+that plugging preserves the types of concrete nodes and results in
+subtypes in variable nodes. Figure~\ref{plugging-examples-fig}
+presents some examples to help the reader get the intuition as well as
+understand some of the more subtlte cases.
+
+\begin{theorem}
+  \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
+  = \pi_q(H,G)$. Proof follows simply: given $p = (f_G,f_H)$, $q =
+  (f_H,f_G)$. As a corrolary, we get $\pi(G,H) \Leftrightarrow
+  \pi(H,G)$.
+\end{theorem}
+
+\begin{theorem}
+  \emph{Match after plugging}: $G \leq \pi_f(G,H)$. And as a corrolary
+  we get $H \leq \pi_f(G,H)$ by symmetry of plugging. 
+\end{theorem}
+
+\begin{theorem}
+  \emph{Plugging preserves matching}: $V \leq G \Rightarrow V \leq
+  \pi_f(G,H)$. As corrolaries we get $V \leq H \Rightarrow V \leq
+  \pi_f(G,H)$ from symmetry of plugging and $\vinterp{G} \cap
+  \vinterp{H} \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the
+  definition of interpretation. 
+\end{theorem}
+
 \subsection{!-Boxes}
 
-The key feature of rules like the Spider Law is that an arbirary
-number of nodes can be matched. To support this we introduce an
-operation, !-boxing, on graph representations. We draw !-boxes around
-a set of chosen nodes and intuitively think of the resulting !-boxed
-graph as representing a set of graphs with an arbirary number of
-copies of the !-boxed nodes, where any connections to nodes outside
-the !-box are also copied. The !-boxes can also be instantiated with
-zero copies errasing all edges from the !-box. See
-Figure~\ref{bang-box-ex1-fig} for a simple example.
+A crucial feature of rules like the Spider Law is that an arbirary
+number of repetitions of a subgraph to be matched. To support this we
+introduce an operation, !-boxing, on graph representations. Given a
+graph representation, this introduces a new notation by outlining
+(called !-boxing) a set of chosen nodes. Intuitively, the resulting
+!-boxed graph can be thought of as representing a set of graphs with
+an arbirary number of copies of the !-boxed nodes, where every copy
+has the connects to the same nodes outside the !-box. The !-boxes can
+also be instantiated with zero copies which is erases all edges from
+the !-box. See Figure~\ref{bang-box-ex1-fig} for a simple example.
 
-A graph with !-boxes is simply a graph paired with a set of !-box node
-sets, which define the !-boxes. The !-box node sets must either be
-disjoint or strict subsets of each other. This is needed to avoid
-confusion with overlaping boxes.
+More formaly, a graph with !-boxes is simply a graph paired with a set
+!-boxes. Each !-box is just the set of nodes inside it and each !-box
+must be disjoint from the others.\footnote{The disjointness condition
+  avoids a slightly more complex, although also more expressive,
+  semantics for overlaping node sets in the !-boxes. While such
+  expressivity is interesting from a representational point of view,
+  it is not needed to express the currently known rules of quantum
+  computation.}
 
-\subsubsection{Semantics.}
+For the purposes of this paper, the underlying graph representation of
+!-boxed graphs is variable-node graphs. The set of variable-node
+graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
+now define the semantics for !-boxed graphs, then show how to extend
+an the embedding function which corresponds to matching as well as the
+plugging operation for !-boxes.
+
+\subsubsection{!-Box Semantics.}
 
@@
 
-\noindent The semantics of a !-boxed graph is then defined as the
-members without any !-boxes of the set closed by these operation.  For
-the purposes of this paper, the underlying graph representation of
-!-boxed graphs is variable-node graphs. The set of variable-node
-graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.
+\noindent The semantics of a !-boxed graph is then defined as those
+members of the set closed by these operation that do have any
+!-boxes. Observe that an instance of a !-box graph is defined by
+pairing each !-box with the natural number that defines how many
+copies are made of it. Thus the $\binterp{G}$ is isomorphic to the set
+of $k$-tuples of natural numbers, where $k$ is the number of !-boxes.
 
-\subsubsection{Matching}
+\subsubsection{!-Box Matching.}
 
-\subsubsection{Plugging}
+Matching for !-box graphs is defined in terms of an underlying
+matching operation. A !-box graph $G$ matches another one, $H$, iff
+$\binterp{G} \supseteq \binterp{H}$. Matching can be given a more
+constructive flavour by extending the open-embeddings of variable-node
+graphs to !-boxes.
+
+There is a !-open embedding of $G$ into $H$ if
+
+\subsubsection{!-Box Plugging.}
 

### 2008-02-27 papers/reasoning-graphically/paper.tex
section on grph reps done
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1377 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
+History of diagrammatic representation going back to ???
+
 CAn formalise quantum mechanics in an abstract setting using compact
@@
 
-\section{Compact Closed Categories and Graphs}
+\section{Graphs and Compact Closed  Categories  }
 \label{sec:mono-categ-graphs}
@@
 \end{diagram}
-which we call \emph{source} and \emph{target}.  The \emph{in-degree}
-of a vertex $v$ is the cardinality of $t^{-1}(v)$; its out-degree is
-the cardinality of $s^{-1}(v)$.  The \emph{degree} of a vertex is
-the sum of its in- and out-degrees.
+which we call \emph{source} and \emph{target}.  Define sets
+$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ of \emph{incoming}
+and \emph{outgoing} edges at a vertex $v$.  
+The \emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
+\sizeof{\text{out}(v)}$.
 
@@
 Augmented by some additional structure, graphs form a compact closed
-category.  We present an overview of this structure:  the details
-omitted here can be found in \cite{Duncan:thesis:2006}.
+category.  In section \ref{sec:graph-repr-comp} we will describe this
+structure, but first we review the concept of compact closed  category.
+
+\subsection{Compact Closed Categories}
+\label{sec:comp-clos-categ}
+
+\subsection{Graph Representations for Compact Closed Categories}
+\label{sec:graph-repr-comp}
+
+Graphs with certain additional structure give a
+representation for compact closed  categories; we now give an overview
+of this construction.  The details omitted here can be found in
+\cite{Duncan:thesis:2006}.
 
@@
 \begin{itemize}
-\item $G = (V,E,C,s,t)$ is a graph;
-\item $\dom\Gamma$ and $\cod\Gamma$ are totally ordered disjoint of
+\item $G = (V,E,s,t)$ is a graph;
+\item $\dom\Gamma$ and $\cod\Gamma$ are totally ordered disjoint sets of
   degree one vertices of $G$.
@@
 incident at $v$ is an incoming edge; $v \mapsto -$ otherwise.  Hence
-$\cod \Gamma$ and $\dom \Gamma$ are \emph{signed sets}.  Concrete
-graphs form a category whose objects are
+$\cod \Gamma$ and $\dom \Gamma$ are \emph{ordered signed sets}.  Given any
+ordered signed set $S$ we write $S^*$ for the same ordered set, but
+the opposite signing.   Given two such sets we can define their disjoint
+union $R+S$ as the disjoint union of the underlying sets, inheriting
+the signing and the order from $R$ and $S$, with the convention that
+$r < s$ for all $r\in R, s\in S$.
 
-can be seen as giving a signing to each element of the ordinal
-one, we can view these sets as maps from the ordinal
+\begin{proposition}
+Concrete graphs form a compact closed category whose objects are
+ordered signed sets and whose arrows $f:A\to B$ are concrete graphs with
+$\cod f = B$  and $\dom f = A^*$.
+\end{proposition}
+For each ordered signed set $A$, the identity map for $A$ has
+$\dom \id{A} = A^*$ and $\cod \id{A} = A$; its underlying
+graph has $E = A$ and $V = A^* + A$ with $t(a) = a$ and $s(a) =
+a^*$.  Given a pair of concrete graphs $f:A\to B$ and $g:B\to C$ their
+composition $g\circ f:A\to C$ is constructed by merging the two
+graphs, erasing the vertices of $\cod f$ and $\dom g$, and identifying
+the edges previously incident at the deleted vertices.  (Due to the
+opposite polarity of the domain and codomain the edges have compatible
+direction.)  The tensor product on objects $A,B$ is simply $A+B$;
+given $f:A\to B$, $g: C\to D$, the graph of $f \otimes g$ is the
+disjoint union of the graphs of $f$ and $g$.  The unit for the tensor
+is the empty set.  The morphisms $d : I \to A \otimes A^*$,
+$e: A^* \otimes A \to I$ have the same underlying graph, but $\dom d =
+\emptyset$, $\cod d = A+A^*$, $\dom e = A^*+A$ and $\cod e = \emptyset$.
 
-can be viewed as a maps from an
+This category captures exactly the axioms for compact closed
+structure, in the sense that any freely generated compact closed
+category can be represented by concrete graphs.  We will consider
+a collection of basic terms\footnote{See \cite{Duncan:thesis:2006} for
+  a more thorough description of the nature of the terms here} $F$
+whose types are vectors of some set of basic types $T$.  Then:
 
-representation of freely generated compact closed  categories  as
-graphs;
+\begin{definition}
+  A \emph{$T,F$-labelling} $\theta$ for a concrete graph $\Gamma$ is a pair of
+  maps  $\theta_T : E \to T$ and $\theta_F : (V - \cod\Gamma -
+  \dom\Gamma) \to F$  such that for each vertex  $v$, if
+  $\text{in}(v) = \langle a_1, \ldots, a_n\rangle$ and $\text{out}(v)
+  = \langle b_1, \ldots, b_m\rangle$ then 
+  \[
+  \theta v : \langle \theta a_1, \ldots, \theta a_n \rangle
+  \to 
+  \langle \theta b_1, \ldots, \theta b_m \rangle
+  \]
+  Call a concrete graph $\Gamma$ \emph{$T,F$-labellable} if there exists an 
+  $T,F$-labelling for it; if $\theta$ is a labelling for $\Gamma$, then
+ the pair $(\Gamma,\theta)$ is an \emph{$T,F$-labelled graph}.
+\end{definition}
 
-need to define embedding and subgraph
+The $T,F$-labelled graphs form a compact closed category in the same
+way as the the concrete graphs, subject to the further restriction
+that arrows are composable only when their labellings agree.
 
-theorem: two arrows in a a freely generated compact closed  category
-are equal iff and only if their graphs are equal.
+\begin{theorem}[\cite{Duncan:thesis:2006}]
+  Let \catC be a compact closed category  freely generated by some set
+  of arrows $F$ and ground types $T$;  then \catC is equivalent to the
+  category  of $T,F$-labelled graphs.
+\end{theorem}
 
-Note that the notion of vertex and graph have the same external
-structure so it is possible to view a subgraph as a vertex and vice versa.
+Given a compact closed  category  \catC generated by some basic set of
+operations,  the arrows of \catC have a canonical representation as
+labelled graphs.  A consequence of the theorem is then that two arrows
+are equal by the equations of the compact closed  structure if and
+only if their graph representations are equal.
+
+As a final remark before moving on, note that the external structre of vertex
+in a concrete graph is esentially the same as that of a complete
+graph;  hence one can consistently view subgraphs as vertices, and
+abstract over the their internal structure.
 
@@
 \end{description}
+Last group of equations
+
+Statement and Discussion of spider law.
 

### 2008-02-27 papers/reasoning-graphically/paper.tex
re-organised sections, !-box graphs section written... more stuff...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1378 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\section{Graph Patterns}
-\label{sec:patterns}
+\section{Graphs with Variable Nodes}
 
-The representation of Compact Closed Categories as Graphs, discuseed
-in the previous section, is too restrictive for automated reasoning
-about quantum computation. In particular, pictures such the Spider
-Law, in Figure~\ref{spider-law-fig} are frequently needed. Such
-pictures represent an infinite family of concrete graphs. In this
-section we describe an extension to the representation that captures
-an infinite family of graphs as a single finite graph. The extension,
-which we call \ref{graph patterns}, adds two extra kinds of nodes to
-the graph. Namely, {\em banged boxes} which informally represent an
-arbitrary number of copies of a node (where the node itself can
-contain a graph of the appropriate type), and {\em variable nodes}
-which informally represent some, as yet unkown, node in a graph. We
-also introduce a notion to replace graph composition, called graph
-plugging that allows graph patterns to put together to form larger
-graph patterns.
-
-%Notably, this
-%allows us to represent laws about quantum computation, such as the
-%Spider Law. Furthermore, this representation is still compositional in
-%the sense that we can plug graphs together.
-
-%\begin{center}
-%  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
-%\end{center}
-
-\subsection{Variable Nodes}
-
-In graph patterns, variable nodes generalise the concept of a boundary
-node. Variable nodes, unlike boundary-nodes in the concrete graph, are
-not classified into domain and co-domain. The intuition of a variable
-node is that it can be replaced by any subgraph of the right
-type. Thus, unlike nodes in conrete a graph, variable nodes do not
-define an ordering on the edges that are incident. Instead, the edges
-that touch a variable node are simply a set. The type of a variable
-nodes is thus defined as a multiset rather than a tensor. Any tensor
-type can be lifted to a multiset type in the obvious manner, which we
-denote by the operation {\tt mset\_of\_tensor}. For example, {\tt
-  mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
-say that a tensor-type, $T$, is an instance of a multiset type, $M$,
-when it's lifting into a multiset is a super-set of $M$: $ M \subseteq
+Variable nodes generalise the concept of a boundary nodes. Unlike
+boundary-nodes in the concrete graph, they are not classified into
+domain and co-domain. The intuition of a variable node is that it can
+be replaced by any subgraph of the right type. Thus, unlike nodes in
+conrete a graph, variable nodes do not define an ordering on the edges
+that are incident. Instead, the edges that touch a variable node are
+simply a set. The type of a variable nodes is thus defined as a
+multiset rather than a tensor. Any tensor type can be lifted to a
+multiset type in the obvious manner, which we denote by the operation
+{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
+  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
+is an instance of a multiset type, $M$, when it's lifting into a
+multiset is a super-set of $M$: $ M \subseteq
 \mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
@@
 
-\subsection{!-Boxes}
+\section{!-Boxes}
 
@@
 graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
-now define the semantics for !-boxed graphs, then show how to extend
-an the embedding function which corresponds to matching as well as the
-plugging operation for !-boxes.
+now define the concept of !-box matching which we use to provide a
+semantics for the !-boxed graphs.
 
-\subsubsection{!-Box Semantics.}
+\subsection{!-Matching.}
 
-We give a formal semantics to !-boxing a representation of graphs by
-defining the a !-box graph in terms of a set of graphs in the
-underlying representation. We denote this interpretation of a graph
-$G$ by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are
-instances of $G$. The definition of $\binterp{G}$ involves the
-following three operations\footnote{These correspond closely to the
-  linear logic rules for weakening, contraction and deriliction.} on a
-!-box and a graph that contains it:
+The set of !-box graphs matched by a !-box graph is defined as the set
+closed by the following operations:
 
 \begin{definition}
-  \emph{copy} copies a box in a graph to produce a new graph with two
-  copies of the boxed subgraph, with a second box around the newly
-  copied nodes. Any edges incident on both a node in the !-box and
-  outside it get copied so that there the old edge remains and the new
-  edge goes from the outer node to the new copy of the internal node.
+  \emph{copy} copies a !-box in a graph to produce a new graph with
+  two copies of the boxed subgraph, with a second box around the newly
+  copied nodes. Any edges between a node inside the !-box, $n$ and one
+  outside it $m$ get copied so that there is a new edge from $m$ to
+  the new copy of $n$.
 \end{definition}
@@
   \emph{kill} removes from the graph all nodes in the !-box set as
-  well as any incident edges to nodes in the box.
+  well as any incident edges to those nodes.
 \end{definition}
 
-\noindent The semantics of a !-boxed graph is then defined as those
-members of the set closed by these operation that do have any
-!-boxes. Observe that an instance of a !-box graph is defined by
-pairing each !-box with the natural number that defines how many
-copies are made of it. Thus the $\binterp{G}$ is isomorphic to the set
-of $k$-tuples of natural numbers, where $k$ is the number of !-boxes.
+\begin{definition}
+  \emph{merge} combines two !-boxes, $B_1$ and $B_2$ into a single
+  larger !-box $B_1 \union B_2$.
+\end{definition}
 
-\subsubsection{!-Box Matching.}
+Like the notation for variable-node graphs, we use $G \leq H$ to
+denote that $G$ matches $H$.
 
-Matching for !-box graphs is defined in terms of an underlying
-matching operation. A !-box graph $G$ matches another one, $H$, iff
-$\binterp{G} \supseteq \binterp{H}$. Matching can be given a more
-constructive flavour by extending the open-embeddings of variable-node
-graphs to !-boxes.
+\subsection{!-Box Semantics.}
 
-There is a !-open embedding of $G$ into $H$ if
+We give a formal semantics to !-boxing a representation of graphs by
+defining the !-box graph in terms of a set of graphs in the
+underlying representation. We denote the interpretation of a graph $G$
+by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are instances
+of $G$.
 
-\subsubsection{!-Box Plugging.}
+\begin{definition}
+  \emph{Interpretation of !-box Graphs}: $\binterp{G}$ is the subset
+  of graphs matched by the !-box graph that have no !-boxes.
+\end{definition}
+
+Observe that an instance of a !-box graph is defined by pairing each
+!-box with the natural number that defines how many copies are made of
+it. Thus the $\binterp{G}$ is isomorphic to the set of $k$-tuples of
+natural numbers, where $k$ is the number of !-boxes.
+
+\begin{theorem}
+  \emph{!-Matching respects !-box semantics}: $G \leq H
+  \Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
+  simple consequence from the defintion of $\binterp{G}$ being a
+  subset of the graphs that match $G$.
+\end{theorem}
+
+\section{Graph Patterns}
+\label{sec:patterns}
+
+The representation of Compact Closed Categories as Graphs, discuseed
+in the previous section~\S\ref{graph-repr-comp}, is too restrictive
+for reasoning about quantum computation. In particular, pictures such
+the Spider Law, in Figure~\ref{spider-law-fig} are frequently
+needed. Such pictures apply to an infinite family of concrete
+graphs. In this section we combine {\em !-boxes} with {\em variable
+  nodes} to form a representation of graphs that allows us to finitely
+express certain infinite families of concrete graphs. We call this
+extension \ref{graph patterns}. Using this representation Spider Law
+can now be represented by the equation on graph paterns shown in
+Figure~\ref{graph-pattern-spider-law-fig}. Graph matching for pattens
+graphs is defined as is a suitable extension to plugging. This
+provides the foundational machinery for a rewriting system, based on
+pattern graphs, for reasoning about quantum computation.
 
@@
 $H$.
+
+We then extend the plugging
+operation on variable-node graphs to graphs with !-boxes.
+
+\subsection{Plugging}
 

### 2008-02-27 papers/reasoning-graphically/paper.tex
fixed latex bug
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1379 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
more changes
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1380 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 {\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
-  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
-is an instance of a multiset type, $M$, when it's lifting into a
-multiset is a super-set of $M$: $ M \subseteq
-\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
-here is that an instance of a variable node is allowed to have as many
-extra edges as it likes as long as it at least has the ones mentioned
-in the original graph.
-
-A variable node, $a$ is be said to an instance of another variable
-node $b$ when the multiset representing the type of $a$ is a superset
-of the multiset for the type of $b$. Thus, with regard to the
-underlying semantics, $a$ being a subtype of $b$ means that the set of
-instances of $a$ are a subset of the instances of $b$.
+  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$
+is an instance of a multiset type $M$, written $T \in M$, when $M
+\subseteq \mathtt{mset\_of\_tensor}(T)$. The intuition here is that
+when a variable node is instantiated to some other node, it's type
+will have at least the currently attached edges.
 
@@
 by simply replacing all bounary nodes with variable nodes.
+
+%% not needed for this story...
+%A variable node, $a$ is be said to an instance of another variable
+%node $b$ when the multiset representing the type of $a$ is a superset
+%of the multiset for the type of $b$. Thus, with regard to the
+%underlying semantics, $a$ being a subtype of $b$ means that the set of
+%instances of $a$ are a subset of the instances of $b$.
 
@@
   \emph{merge} combines two !-boxes, $B_1$ and $B_2$ into a single
-  larger !-box $B_1 \union B_2$.
+  larger !-box $B_1 \cup B_2$.
 \end{definition}
@@
 
-The representation of Compact Closed Categories as Graphs, discuseed
-in the previous section~\S\ref{graph-repr-comp}, is too restrictive
-for reasoning about quantum computation. In particular, pictures such
-the Spider Law, in Figure~\ref{spider-law-fig} are frequently
-needed. Such pictures apply to an infinite family of concrete
-graphs. In this section we combine {\em !-boxes} with {\em variable
-  nodes} to form a representation of graphs that allows us to finitely
-express certain infinite families of concrete graphs. We call this
-extension \ref{graph patterns}. Using this representation Spider Law
-can now be represented by the equation on graph paterns shown in
-Figure~\ref{graph-pattern-spider-law-fig}. Graph matching for pattens
-graphs is defined as is a suitable extension to plugging. This
-provides the foundational machinery for a rewriting system, based on
-pattern graphs, for reasoning about quantum computation.
+The representation of Compact Closed Categories as Graphs, discussed
+in~\S\ref{graph-repr-comp}, is too restrictive for reasoning about
+quantum computation. In particular, pictures such the Spider Law, in
+Figure~\ref{spider-law-fig} are frequently needed, but not
+expressible. Such pictures are used to denote an infinite family of
+concrete graphs that can be rewritten. In this section we combine the
+{\em !-boxes} with {\em variable nodes} graph extensions to form a
+representation that allows us to finitely express these infinite
+families of concrete graphs. We call this extension \ref{graph
+  patterns}. The Spider Law can now be represented by the equation in
+Figure~\ref{graph-pattern-spider-law-fig}. We also extend the notion
+of matching for graph pattens as well as present a corresponding
+notion of plugging. This provides the foundations for the rewriting
+machinery in~\S\ref{rewriting} which can then be used to reasoning
+about quantum computation.
 

### 2008-02-28 papers/reasoning-graphically/paper.tex
slight restructuring and some comcl stuff
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1381 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
inprogress on reasoning section
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1382 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
+A strict symmetric monoidal  category
+\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact
+closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
+object $A^*$, and morphisms
+\begin{gather*}
+  d_A : I \to A^* \otimes A\\
+  e_A : A \otimes A^* \to A
+\end{gather*}
+such that 
+\begin{align*}
+  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A 
+  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A
+\end{align*}
 \subsection{Graph Representations for Compact Closed Categories}
@@
   f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
-  f_G(v) \in f^{-1}_H(v)$. Intuitively this captures the idea a node
-  is one function is always mapped to one the nodes that maps to it in
-  the other function.
+  f_G(v) \in f^{-1}_H(v)$. This states when a set of nodes in one
+  graph all map to a common node $n$ in the other graph, $n$ maps back
+  to one of the nodes that maps to it. 
 \end{definition}
@@
 manner, the intuitive idea of graph matching commonly used in for
-Quantum computation. To do this, we combine the two extensions to the
-concrete graph representation, namely variable nodes and !-boxes. This
+Quantum Computation. To do this, we combine the two extensions to the
+concrete graph representation, namely variable nodes and !-boxes, to
 provides a matching relation between graph patterns and concrete
@@
 
-The semantics for a graph pattern $G$ in terms of a set of concrete
-graphs, is denoted by $\minterp{G}$. We say that members of this set
-are \emph{matched} by $G$. Formally, the set of matches are defined as
-follows:
+The semantics for a graph pattern $G$ is a set of concrete graphs
+denoted by $\minterp{G}$. We say that members of this set are
+\emph{matched} by $G$. Formally, the set of matches are defined as:
 
-$\minterp{G} = (\vinterp{G'} | G' \in \binterp{G})$
+$$\minterp{G} = (\vinterp{G'}\;|\; G' \in \binterp{G})$$
 
@@
 
-The definition of one pattern matching another one is simply that it
-is more general. Formally, a graph pattern, $G$, matches another graph
-pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This provides
+The specification for one pattern matching another one is simply that
+it is more general. Formally, a graph pattern, $G$, matches another
+graph pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This is
 the specification for a matching algorithm which we implement of
-in~\S\ref{sec:rewriting}. To make such a proof easier, we extend
-embedding and plugging of variable-node graphs to graph patterns.
+in~\S\ref{sec:rewriting}. We can use the definition of open embedding
+from variable-node graphs as an intermediate mechanism and show that a
+suitable definition of graph pattern matching is simply the !-Box
+matching followed by variable-node matching:
 
-Our representation has been carefully designed so that this concept
-corresponds to what we call a \emph{open embedding} of $G$ into
-$H$.
+\begin{definition}
+  $G \leq H \Leftrightarrow \forall H' \in \binterp{H}.\,\exists G' \in \binterp{G}.\, G' \leq_v H'$
+\end{definition}
+
+Because viable-node matching preserves the semantics, it is trivial to
+prove that this definition meets the specification. This gives us
+matching between pattern-nodes.
+
+\subsubsection{Graph Pattern Equality.}
+
+Equality of graph patterns can simply be defined as mutual matching:
+
+\begin{definition}
+  $G = H \Leftrightarrow G \leq H \land H \leq G$
+\end{definition}
+
+\subsection{Plugging}
 
@@
 
-\subsection{Plugging}
+\subsection{Adequacy}
 
-***
-
-This is the basis for our implementation and is thus also the basis
-for its correctness argument.
-
-\section{Graph Pattern Rewriting}
+\section{Reasoning with Graph Patterns}
 \label{sec:rewriting}
 
+The formalism we have presented for graph patterns provides a
+\emph{meta-level} framework for reasoning in some model of concrete
+graphs. The chosen model provides the \emph{object-level} graph
+formalism. This includes a particular representation for data at the
+nodes and edges as well as node-matching and edge-matching
+algorithms. The object-level definition also defines a set of
+axiomatic equational rules. These are derived
+
+\subsubsection{Soundness.} An object-level graph system is sound if
+their is some equation which can not be derived.
+
+ The soundness
+of a set of axioms in our framework
+
+ framework
+requres the model is an instance of a compact closed category, and
+that the
+
+For the model we use of qantum computation, this is finite
+dimensional hilbert spaces. Because the basic equational rules come
+from a model and have been proved for that model, the resulting
+preserve the soundness of our graphical system.
+
+ and the families of them represneted by graph
+patterns.
+
+An object graph formalsism uses the
+
+By defining graph patterns as a conservative extension of concrete
+graphs we get the relative soundness for them.
+
+Such a theorem prover starts with a system of equations between
+graphs. These are not true of the graphs abstractly, but are proved of
+the particular model of concrete graphs being considered.
+
 \subsection{Rules}
+
 \subsection{Unification}
@@
 
-\section{Node Expressions}
-\label{sec:node-expressions}
+\subsection{Node in Quantum Computations}
+\label{ssec:node-expressions}
+
+ Expressions
 

### 2008-02-28 papers/reasoning-graphically/paper.tex
working on reaosning section
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1383 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-Like the notation for variable-node graphs, we use $G \leq H$ to
-denote that $G$ matches $H$.
+\noindent We use $G \leq_! H$ to denote that $G$ matches $H$.
 
@@
 \begin{theorem}
-  \emph{!-Matching respects !-box semantics}: $G \leq H
+  \emph{!-Matching respects !-box semantics}: $G \leq_! H
   \Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
@@
 
+Rules that have been proved for concrete graphs, such as the
+Frombenius Law, are in fact a family of proofs that correspond to
+replacing the boundary edges with variable nodes. This is beacuse of
+the symmetry of rules in a compact closed categories:
+
+By defining graph patterns as a conservative extension of concrete
+graphs we get the relative soundness for them.
+
 \section{Reasoning with Graph Patterns}
@@
 \emph{meta-level} framework for reasoning in some model of concrete
-graphs. The chosen model provides the \emph{object-level} graph
-formalism. This includes a particular representation for data at the
-nodes and edges as well as node-matching and edge-matching
-algorithms. The object-level definition also defines a set of
-axiomatic equational rules. These are derived
+graphs.
 
-\subsubsection{Soundness.} An object-level graph system is sound if
-their is some equation which can not be derived.
+An \emph{object-level} graph formalism is defined by some
+model of a compact closed category.
 
- The soundness
-of a set of axioms in our framework
-
- framework
-requres the model is an instance of a compact closed category, and
-that the
-
-For the model we use of qantum computation, this is finite
-dimensional hilbert spaces. Because the basic equational rules come
-from a model and have been proved for that model, the resulting
-preserve the soundness of our graphical system.
-
- and the families of them represneted by graph
-patterns.
-
-An object graph formalsism uses the
-
-By defining graph patterns as a conservative extension of concrete
-graphs we get the relative soundness for them.
+This includes a particular
+representation for data at the nodes and edges as well as
+node-matching and edge-matching algorithms. The object-level
+definition also defines a set of axiomatic equational rules.
 
@@
 
+\subsubsection{Soundness.} An object-level graph system is sound if
+their is some pair of graphs which can not be derived as an equation.
+
+To ensure soundness, it is sufficient to show that the object-level
+model is a compact-closed category and that the basic equational rules
+of the object-level can be proved in the underlying model.
+
+\subsubsection{Completeness.} A formalism for graphs can be shown to
+be complete with respect to the underlying semantics if the equational
+rules have been shown to be complete.
+
 \subsection{Rules}
 
-\subsection{Unification}
-\subsection{Substitution}
+\section{A Case Study in Quantum Computation}
 
-\subsection{Node in Quantum Computations}
+The model we are interested in is finite dimensional hilbert spaces as
+an account of qantum computation.
+
+\subsection{Node Expressions}
 \label{ssec:node-expressions}
 
- Expressions
+Completeness is a challenging and still open problem for graph based
+models of quantum computation. However, even without completeness
+interesting quantum computations can be expressed.
 

### 2008-02-28 papers/reasoning-graphically/paper.tex
added comcl
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1384 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 functions $f_E : E_G \to E_H$ and $f_V:V_G\to V_H$ such that:
-\begin{gather*}
-  s_H\circ f_E = f_V \circ s_G,\\
-  t_H\circ f_E = f_V \circ t_G.
-\end{gather*}
-\TODO{different kinds of morphism?}
+\begin{gather}
+  s_H\circ f_E = f_V \circ s_G,\label{eq:graph-hom1}\\
+  t_H\circ f_E = f_V \circ t_G\label{eq:graph-hom2}.
+\end{gather}
 
-Augmented by some additional structure, graphs form a compact closed
-category.  In section \ref{sec:graph-repr-comp} we will describe this
-structure, but first we review the concept of compact closed  category.
+Let $f: G \to H$ be  graph morphism and  let $V' \subseteq V_G$. We say
+that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
+\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
+that $f_E(e') = e$.  We call $f$ an \emph{open embedding} for $V'$ if:
+\begin{enumerate}
+\item $f_E$ is injective;
+\item $f_V$ restricted to $V'$ is injective; and,
+\item $f$ is strict for $V'$.
+\end{enumerate}
+The intuition behind this definition is that the subgraph of $G$
+determined by $V'$ should be preserved exactly by $f$, whereas the
+other vertices may be identified.
+
+Given two graphs $G,H$, a pair of partial maps $r_V: V_G
+\rightharpoonup V_H$, $r_E: E_G \rightharpoonup E_H$ is called a
+partial graph morphism if whenever $r(e)$ is defined then $r(s_G(e))$
+and $r(t_G(e))$ are defined, and the restriction of $r_E$ to its
+preimage satifies equations \eqref{eq:graph-hom1} and
+\eqref{eq:graph-hom2} above.
+
+Augmented by some additional structure, graphs form a \emph{compact closed
+category}.  In section \ref{sec:graph-repr-comp} we will describe this
+structure, but first we review the basic properties of compact closed
+categories.
 
@@
 
+\begin{definition}
 A strict symmetric monoidal  category
@@
 such that
-\begin{align*}
+\begin{align}
   A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
-  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A
-\end{align*}
+  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
+  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
+  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
+  \id{A^*} \label{eq:comcl2}
+\end{align}
+\end{definition}
+
+For any arrow $f:A\to B$ in a compact closed category \catC 
+its \emph{name} and \emph{coname}:
+\[
+\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
+\]
+and constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
+$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
+isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
+\catC(A\otimes B^*,I)$ making the \catC monoidally closed\footnote{In
+  general compact closed categories  are models of multiplicative
+  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
+Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
+\[
+f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
+\id{B^*}) \circ (d_A \otimes \id{B^*})
+\]
+By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
+f$.  Thus $(\cdot)^*$ lifts to an involutive functor
+$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
+opposite.
+
 \subsection{Graph Representations for Compact Closed Categories}
@@
 of this construction.  The details omitted here can be found in
-\cite{Duncan:thesis:2006}.
+\cite{Duncan:thesis:2006}.   Pictorial representations are in
+Fig.~\ref{fig:comcl-graphs}.
 
@@
 \item $\dom\Gamma$ and $\cod\Gamma$ are totally ordered disjoint sets of
-  degree one vertices of $G$.
+  degree one vertices of $G$.  The union of these sets is the
+  \emph{boundary} of $\Gamma$.
 \item $<_{\text{in}(\cdot)}$ is a family of maps, indexed by $V$ such
@@
 disjoint union of the graphs of $f$ and $g$.  The unit for the tensor
-is the empty set.  The morphisms $d : I \to A \otimes A^*$,
-$e: A^* \otimes A \to I$ have the same underlying graph, but $\dom d =
-\emptyset$, $\cod d = A+A^*$, $\dom e = A^*+A$ and $\cod e = \emptyset$.
+is the empty set.  The morphisms $d_A : I \to A^* \otimes A$,
+$e_: A \otimes A^* \to I$ have the same underlying graph, but $\dom d =
+\emptyset$, $\cod d = A^*+A$, $\dom e = A+A^*$ and $\cod e = \emptyset$.
+
+\begin{figure}{t}
+  We make the convention that domain of an arrow is at the top of the
+  picture, and its codomain is at the bottom.
+  \centering
+  \[
+  \begin{array}{lllll}
+      \id{A} \otimes \id{B^*} = \id{A \otimes B^*} = PIC 
+      &\qquad&
+      d_A = PIC  
+      &\qquad& 
+      f = PIC \\\\
+      \name{f} = PIC 
+      & \qquad &
+      \coname{f} = PIC 
+      &\qquad &
+      f^* = PIC
+  \end{array}
+  \]
+  \caption{Compact Closed Structure as Graphs}
+  \label{fig:comcl-graphs}
+\end{figure}
 
@@
 a collection of basic terms\footnote{See \cite{Duncan:thesis:2006} for
-  a more thorough description of the nature of the terms here} $F$
+  a more thorough description of the nature of the terms.} $F$
 whose types are vectors of some set of basic types $T$.  Then:
@@
 
-\begin{theorem}[\cite{Duncan:thesis:2006}]
+\begin{theorem}
   Let \catC be a compact closed category  freely generated by some set
@@
 \bibliographystyle{plain}
-\bibliography{all}
-\bibliography{bibfile}
+\bibliography{all,bibfile}
 

### 2008-02-28 papers/reasoning-graphically/paper.tex
fixed typo
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1385 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
some images, intro for var-nodes and reasoning section work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1386 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
   simpler way of reading and presenting quantum computations, manual
-  manipulation of such graphs is slow and error prone. We
-  present a formalism that supports a mechanised reasoning about such
+  manipulation of such graphs is slow and error prone. We present a
+  formalism that supports a mechanised reasoning about such
   graphs. This involves a compositional account of graph rewriting
@@
   support for derived rules, which allows the model of quantum
-  computation  to be specified declarativley.
-  Soundness of the 
-  system is proved with respect to the underlying model and
-  completeness is an open problem.
+  computation to be specified declarativley.  Soundness of the system
+  is proved with respect to the underlying model and completeness is
+  an open problem.
 
@@
 
-\begin{figure}{t}
+\begin{figure}[t]
   We make the convention that domain of an arrow is at the top of the
@@
 
-Variable nodes generalise the concept of a boundary nodes. Unlike
-boundary-nodes in the concrete graph, they are not classified into
-domain and co-domain. The intuition of a variable node is that it can
-be replaced by any subgraph of the right type. Thus, unlike nodes in
-conrete a graph, variable nodes do not define an ordering on the edges
-that are incident. Instead, the edges that touch a variable node are
-simply a set. The type of a variable nodes is thus defined as a
-multiset rather than a tensor. Any tensor type can be lifted to a
-multiset type in the obvious manner, which we denote by the operation
-{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
-  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$
-is an instance of a multiset type $M$, written $T \in M$, when $M
+In the concrete representation, the following graphs represent
+different computations:
+
+\begin{center}
+\inlinegraphic{2.5em}{compact1} \quad\quad
+\inlinegraphic{2.5em}{compact3} \quad\quad
+\inlinegraphic{2.5em}{compact5}
+\end{center}
+
+\noindent However, composition with semicircle line(s) on the domain
+or co-domain allows a proof involving one of the above to easily
+derive a corresponding one for the other two. Proving any of the
+following allows a trivial derivation of the others:
+
+\begin{center}
+\inlinegraphic{2.5em}{compact1} $=$
+\inlinegraphic{1.3em}{compact2} $\Leftrightarrow$
+\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
+$\Leftrightarrow$ \inlinegraphic{2.5em}{compact5} $=$
+\inlinegraphic{2.5em}{comonoid-unit4}
+\end{center}
+
+\noindent This motivates a representation that has only one
+representative for such equations. Such a representation abstracts
+over the boundary nodes membership in the domain and co-domain. In
+this section we formalise such a notion and refer to the generalised
+boundary nodes as variable-nodes.
+
+The intuition of a variable node is that it can be replaced by any
+subgraph of the right type. Thus, unlike nodes in conrete a graph,
+variable nodes do not define an ordering on the edges that are
+incident. Instead, the edges that touch a variable node are simply a
+set. The type of a variable nodes is thus defined as a multiset rather
+than a tensor. Any tensor type can be lifted to a multiset type in the
+obvious manner, which we denote by the operation {\tt
+  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
+  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$ is
+an instance of a multiset type $M$, written $T \in M$, when $M
 \subseteq \mathtt{mset\_of\_tensor}(T)$. The intuition here is that
@@
 
-\subsection{Matching and Open Embedding}
-
-\begin{definition}
-  An \emph{open embedding} from $G$ to $H$ a strict embedding from the
-  non-variable subgraph of $G$ into the non-variable subgraph of $H$
-  and a mapping from the variable nodes in $G$ ...
+\subsection{Matching as an Open Embedding}
+
+To formalise the semantics of variable-node graphs, we define
+matching, which captures the intuitive idea with
+
+\begin{definition}
+  Matching is charcterised by an \emph{open embedding} which is strict
+  for the non-variable nodes.
 \end{definition}
@@
 
-The formalism we have presented for graph patterns provides a
-\emph{meta-level} framework for reasoning in some model of concrete
-graphs.
-
-An \emph{object-level} graph formalism is defined by some
-model of a compact closed category.
-
-This includes a particular
-representation for data at the nodes and edges as well as
-node-matching and edge-matching algorithms. The object-level
-definition also defines a set of axiomatic equational rules.
-
-Such a theorem prover starts with a system of equations between
-graphs. These are not true of the graphs abstractly, but are proved of
-the particular model of concrete graphs being considered.
+In this section we show how the formalism we have presented for graph
+patterns can be used to provide a \emph{meta-level} framework for
+reasoning about a given model of a compact closed
+categories. Following the terminology of logical frameworks such as
+Isabelle~\cite{isabelle}, we call the specification provided by the
+underlying model an \emph{object-level} graph formalism. Each
+object-level formalism defines a set of rules which are treated as the
+axioms for the system. The object-level implementation also defines
+the data at the nodes and edges as well as corresponding data-matching
+behaviour.
+
+These are not true of the graphs abstractly, but are proved of the
+particular model of concrete graphs being considered.
+
+\subsection{Substitution with Equational Rules}
+
+A rule in our framework is a pair of pattern graph that represent the
+left and right hand sides of an equation. Substituon with a rule then
+involves replacing a subgraph that matches the left hand side with the
+right hand side. To do this there has to be a isomorphism between
+variable nodes in the left and right hand subgraphs. More concretely,
+there must be a unique naming of the variable nodes in the left hand
+side such that each name occurs in a unique variable node on the right
+hand side. Given a matching on the left hand side, the target subgraph
+can be replaced with right hand side connected by the corresponding
+variable nodes. This provides rewriting for pattern graphs.
 
@@
 
-\subsection{Rules}
+\subsection{Implications}
 

### 2008-02-28 papers/reasoning-graphically/paper.tex
working...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1387 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \usepackage{diagrams}
+\usepackage{proof2}
 
@@
 
-$$\minterp{G} = (\vinterp{G'}\;|\; G' \in \binterp{G})$$
+$$\minterp{G} = \{\vinterp{G'}\;.\; G' \in \binterp{G}\}$$
 
@@
 
-\subsection{Adequacy}
-
-Rules that have been proved for concrete graphs, such as the
-Frombenius Law, are in fact a family of proofs that correspond to
-replacing the boundary edges with variable nodes. This is beacuse of
-the symmetry of rules in a compact closed categories:
-
 By defining graph patterns as a conservative extension of concrete
@@
 
-In this section we show how the formalism we have presented for graph
-patterns can be used to provide a \emph{meta-level} framework for
-reasoning about a given model of a compact closed
+In this section we outline how the formalism we have presented for
+graph patterns can be used to provide a \emph{meta-level} framework
+for reasoning about a given model of a compact closed
 categories. Following the terminology of logical frameworks such as
@@
 
-These are not true of the graphs abstractly, but are proved of the
-particular model of concrete graphs being considered.
-
 \subsection{Substitution with Equational Rules}
 
-A rule in our framework is a pair of pattern graph that represent the
-left and right hand sides of an equation. Substituon with a rule then
-involves replacing a subgraph that matches the left hand side with the
-right hand side. To do this there has to be a isomorphism between
-variable nodes in the left and right hand subgraphs. More concretely,
-there must be a unique naming of the variable nodes in the left hand
-side such that each name occurs in a unique variable node on the right
-hand side. Given a matching on the left hand side, the target subgraph
-can be replaced with right hand side connected by the corresponding
-variable nodes. This provides rewriting for pattern graphs.
-
-\subsubsection{Soundness.} An object-level graph system is sound if
-their is some pair of graphs which can not be derived as an equation.
-
-To ensure soundness, it is sufficient to show that the object-level
-model is a compact-closed category and that the basic equational rules
-of the object-level can be proved in the underlying model.
-
-\subsubsection{Completeness.} A formalism for graphs can be shown to
-be complete with respect to the underlying semantics if the equational
-rules have been shown to be complete.
-
-\subsection{Implications}
+A rule in our framework declarative in the sense that it is a pair of
+pattern graph that represent the left and right hand sides of an
+equation. Substituon with a rule then involves replacing a subgraph
+that matches the left hand side with the right hand side. To do this
+there has to be a isomorphism between variable nodes in the left and
+right hand subgraphs. More concretely, there must be a unique naming
+of the variable nodes in the left hand side such that each name occurs
+in a unique variable node on the right hand side. Given a matching on
+the left hand side, the target subgraph can be replaced with right
+hand side connected by the corresponding variable nodes. This provides
+rewriting for pattern graphs.
+
+\subsection{Meta-Level Logic and Derived Rules}
+
+The meta-logical rules of the framework are quite simple as they only
+involve dealing with object-level equations:
+
+\begin{center}
+\prooftree
+A = B \in \Gamma
+\justifies
+\Gamma \vdash A = B
+\using\mbox{trivial}
+\endprooftree
+\quad\quad
+\prooftree
+\justifies
+\Gamma \vdash A = A
+\using\mbox{reflexivity}
+\endprooftree
+\end{center}
+
+\begin{center}
+\prooftree
+\Gamma \vdash A = B
+\justifies
+\Gamma \vdash B = A
+\using\mbox{symmetry}
+\endprooftree
+\quad\quad
+\prooftree
+\Gamma \vdash A = B
+\Gamma \vdash C = D
+\justifies
+\Gamma \vdash (C = D[A/B])
+\using\mbox{substitution}
+\endprooftree
+\end{center}
+
+\noindent where $\Gamma$ is the set of object-level axioms. For the
+reflexivity rule, $A$ must be a well-formed pattern graph.
+
+This system allows derived rules which allow a combination of steps to
+be abbreviated by representing them as a single new rule.
+
+\subsection{Soundness and Completeness}
+
+The soundness of a system that uses our meta-level rests on two
+arguments:
+
+\begin{itemize}
+\item Representation of the underying model as a compact closed
+  category with a defined set of equations.
+
+\item The soundness and adequacy of the axioms when expressed within
+  the compact closed category. Adquency ensures that the encoding as a
+  graph pattern represents only the intended rules.  The soundness of
+  the rules is obviously needed and requires proving the rule in terms
+  of the underlying semantics.
+\end{itemize}
+
+The completeness of the resulting system, with respect to the intended
+semantics, depends on the completeness of the set of given rules. The
+completeness of
+
+\subsection{Derived Rules}
+
+\subsection{Simplification}
+
+\subsection{Adequacy}
+
+% An object-level graph system is sound if their is some pair of graphs
+% which can not be derived as an equation.
+
+% To ensure soundness, it is sufficient to show that the object-level
+% model is a compact-closed category and that the basic equational rules
+% of the object-level can be proved in the underlying model.
+
+% The axioms should be proved of the particular model of concrete graphs
+% being considered.
+
+% \subsubsection{Completeness.} A formalism for graphs can be shown to
+% be complete with respect to the underlying semantics if the equational
+% rules have been shown to be complete.
+
+% \subsection{Implications}
 
@@
 
-The model we are interested in is finite dimensional hilbert spaces as
-an account of qantum computation.
+The model of qantum computation in which we are interested is finite
+dimensional hilbert spaces.
 

### 2008-02-28 papers/reasoning-graphically/paper.tex
more diagrams and some restructuring
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1388 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\section{Quotients and Rewriting}
+\section{Graphs as Quantum Computations}
 \label{sec:quotients-rewriting}
 
-Few examples of interesting compact closed categories are freely
-generated.  In practice we work with categories presented as
-generators and \emph{equations}.  For 
-representing quantum computations, Coecke and Duncan
-\cite{Coecke:2008jo} propose the following collection of generators:
+While compact closed categories provide a suitable setting for
+reasoning about quantum computation \cite{AbrCoe:CatSemQuant:2004},
+freely generated structure will not suffice:  we need additional
+equations.  These equations will be expressed as rewrites rules for
+graphs.  In this section we will describe the generators and equations
+used to reason about quantum computation, and show how some of its formal
+properties lead to particular issues in for the rewriting machinery.
+
+Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
+system for quantum computation built from the following collection of
+generators:  
 \begin{description}
@@
 \item[Arrows] Two families of arrows:
-  \begin{align}
+  \begin{align*}
   &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
@@
   &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
-  \end{align}
+  \end{align*}
   where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
 \end{description}
-We represent these two families, and their adjoints, by colours in our
-graphical notation:
+To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : b
+\to A$.  Each arrow is represented as a small graph; its adjoint is
+the same graph written upside down.  We use colours to denote two families:
 \begin{gather*}
@@
 \end{gather*}
-The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  The free
-compact closed category  is then given by all graphs formed by
+The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
+for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
+free compact closed category  is then given by all graphs formed by
 composing and tensoring these basic graphs.
@@
 To model the behaviour of quantum systems certain additional equations
-must be satisfied.  Each family $(\delta, \epsilon)$ must form a
-\emph{classical structure}
-\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
-$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
-$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
-together they must satisfy the isometry and frobenius equations. In
-addition the family $\alpha$ must form an abelian group with $\alpha^\dag
-= -\alpha$.  These equations are presented graphically below for the
-green family;  the red vertices obey all the same equations
+must be satisfied.  These are discussed in detail in
+\cite{Coecke:2008jo};  we present them in graphical form in
+Figure~\ref{fig:graph-quant-eqns}.
+
+\begin{figure}
+  \figureline
+
+  The $(\cdot)^\dag$ functor gives a vertical symmetry to the
+  category, hence for every equation we have a second equation
+  obtained by flipping the diagram upside down.  In additon, we have a
+  ``colour symmetry'' so for each equation shown here gives rise to
+  second obtained by exchanging the two colours.
   \begin{description}
@@
     \]
-\item[Monoid Laws] 
+  \item[Isometry, Frobenius, and  Compact Structure]
     \[
-\begin{array}{ccccccccccccccc}
-  \inlinegraphic{3em}{monoid-assoc1}  &=\;&
-  \inlinegraphic{3em}{monoid-assoc2}
-  &\qquad\qquad&
-  \inlinegraphic{3em}{monoid-unit1}  &=\;&
-  \inlinegraphic{3em}{monoid-unit2}  &=\;&
-  \inlinegraphic{3em}{monoid-unit3} 
-  &\qquad\qquad&
-  \inlinegraphic{3em}{monoid-comm1}  &=\;&
-  \inlinegraphic{3em}{monoid-comm2}
-\end{array}
-\]
-\item[Isometry]
-\[
-\begin{array}{ccc}
+    \begin{array}{cccccccccccccccc}
       \inlinegraphic{3.5em}{isometry1}  &=\;&
       \inlinegraphic{3.5em}{isometry2}  
-\end{array}
-\]
-\item[Frobenius]
-  \[
-  \begin{array}{ccccc}
+      &\qquad\qquad &
       \inlinegraphic{3em}{frobenius1}  &=\;&
-    \inlinegraphic{3em}{frobenius2} &=\;&
-    \inlinegraphic{3em}{frobenius3}
-  \end{array}
-  \]
-\item[Compact Structure] 
-  \[
-  \begin{array}{ccccccc}
+      \inlinegraphic{3em}{frobenius2}  
+      &\qquad\qquad&
       \inlinegraphic{2.5em}{compact1}  &=\;&
       \inlinegraphic{1.3em}{compact2}  
-    &\qquad\qquad&
-    \inlinegraphic{2.5em}{compact3}  &=\;&
-    \inlinegraphic{1.3em}{compact4}  
     \end{array}
@@
       \inlinegraphic{3.5em}{alpha-commute3} 
-      &\qquad\qquad&
-      \inlinegraphic{3.5em}{alpha-commute4}  &=\;&
-      \inlinegraphic{3.5em}{alpha-commute5}  &=\;&
-      \inlinegraphic{3.5em}{alpha-commute6} 
     \end{array}
     \]
-\end{description}
-Additionally, the two families satisfy equations which make them a
-\emph{scaled bialgebra}.
-\begin{description}
   \item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
@@
       \inlinegraphic{2.5em}{bialgebra7}  &=\;&
-      \inlinegraphic{1.5em}{bialgebra8} \\ \\
-      \inlinegraphic{3.5em}{bialgebra9}  &=\;&
-      \inlinegraphic{3em}{bialgebra10}
-      \qquad\qquad
-      \inlinegraphic{2.5em}{bialgebra11}  &=\;&
-      \inlinegraphic{1.5em}{bialgebra12}
-      \qquad\qquad
-      \inlinegraphic{2.5em}{bialgebra13}  &=\;&
-      \inlinegraphic{1.5em}{bialgebra14} 
+      \inlinegraphic{1.5em}{bialgebra8} 
   \end{array}
   \]
+\item[Group Actions]
+  \[
+  \begin{array}{cccccccccccccccccccccc}
+    \inlinegraphic{2.5em}{group-int1} &=\;&
+    \inlinegraphic{2.5em}{group-int2} 
+    &\qquad\qquad&
+    \inlinegraphic{2.5em}{group-int3} &=\;&
+    \inlinegraphic{2em}{group-int4} 
+    &\qquad\qquad&
+    \inlinegraphic{2.5em}{group-int5} &=\;&
+    \inlinegraphic{1.5em}{group-int6} 
+    &\qquad\qquad&
+    \inlinegraphic{2.5em}{group-int7} &=\;&
+    \inlinegraphic{1.5em}{group-int8} 
+  \end{array}
+  \]
+    \item[$H$ Properties]
+    \[
+    \begin{array}{cccccccccccccccccccc}
+      \left(\inlinegraphic{1.5em}{H}\right)^\dag &=\;&
+      \inlinegraphic{1.5em}{H} 
+      &\qquad\qquad&
+      \inlinegraphic{2.5em}{Heq1} &=\;&
+      \inlinegraphic{2.5em}{Heq2}
+    \end{array}
+    \]
+  \item[Colour Duality]
+        \[
+    \begin{array}{cccccccccccccccccccc}
+      \inlinegraphic{2.5em}{Heq-delta} &=\;&
+      \inlinegraphic{1.5em}{reddelta}
+      &\qquad\qquad&
+      \inlinegraphic{2em}{Heq-epsilon} &=\;&
+      \inlinegraphic{1.5em}{redepsilon}
+      &\qquad\qquad&
+      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
+      \inlinegraphic{1.5em}{redalpha}
+    \end{array}
+    \]
+
   \end{description}
-Last group of equations
-
-Statement and Discussion of spider law.
-
-The need to formalise these equations leads naturally to a rewriting
-theory
+
+  \caption{Graphical Equations for Quantum Systems}
+  \label{fig:graph-quant-eqns}
+  \figureend
+\end{figure}
 

### 2008-02-29 papers/reasoning-graphically/paper.tex
background finished, intro started
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1389 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-Want to do reasoning about quantum computation by manipulating graphs
+Recent work in quantum computation has emphasised the use of graphical
+languages motivated by the underlying logical structure of quantum
+mechanics itself.  These techniques have a number of advantages over
+the convential matrix-based approach to quantum mechanics, but in
+particular the graph can directly represent the structure of quantum
+states explicitly, including the
 
@@
 
-For any arrow $f:A\to B$ in a compact closed category \catC 
-its \emph{name} and \emph{coname}:
+Every arrow $f:A\to B$ in a compact closed category \catC
+has a \emph{name} and \emph{coname}:
 \[
@@
 \]
-and constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
+are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
 $\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
 isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
-\catC(A\otimes B^*,I)$ making the \catC monoidally closed\footnote{In
+\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
   general compact closed categories  are models of multiplicative
@@
 Fig.~\ref{fig:comcl-graphs}.
+
+\begin{figure}
+  We make the convention that domain of an arrow is at the top of the
+  picture, and its codomain is at the bottom.
+  \centering
+  \[
+  \begin{array}{lllll}
+      \id{A \otimes B^*} = \inlinegraphic{3em}{comcl-id}
+      &\qquad&
+      d_A = \inlinegraphic{2em}{comcl-eta}
+      &\qquad& 
+      e_B = \inlinegraphic{2em}{comcl-epsilon} 
+      \\\\
+      f = \inlinegraphic{3em}{comcl-f} 
+      & \qquad &
+      \name{f} = \inlinegraphic{3em}{comcl-name-f}
+      &\qquad &
+      f^* =  \inlinegraphic{3em}{comcl-dual-f}
+  \end{array}
+  \]
+  \caption{Compact Closed Structure as Graphs}
+  \label{fig:comcl-graphs}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-  We make the convention that domain of an arrow is at the top of the
-  picture, and its codomain is at the bottom.
-  \centering
-  \[
-  \begin{array}{lllll}
-      \id{A} \otimes \id{B^*} = \id{A \otimes B^*} = PIC 
-      &\qquad&
-      d_A = PIC  
-      &\qquad& 
-      f = PIC \\\\
-      \name{f} = PIC 
-      & \qquad &
-      \coname{f} = PIC 
-      &\qquad &
-      f^* = PIC
-  \end{array}
-  \]
-  \caption{Compact Closed Structure as Graphs}
-  \label{fig:comcl-graphs}
-\end{figure}
+\begin{remark}
+  Given concrete graphs $f : A\to B$ and $g:B\to C$ there exists
+  exactly one graph morphism  $\tau:  f \to (g\circ f)$ such that
+  $\tau$ is an open embedding for the non-boundary nodes
+  of $f$.   Indeed the intuition behind an open embedding, is that any
+  such map picks out a subgraph which forms a well defined arrow in
+  its own right.
+\end{remark}
 
@@
 \begin{theorem}
-  Let \catC be a compact closed category  freely generated by some set
+  Let \catC be a compact closed category, freely generated by some set
   of arrows $F$ and ground types $T$;  then \catC is equivalent to the
@@
 equations.  These equations will be expressed as rewrites rules for
-graphs.  In this section we will describe the generators and equations
+graphs.  In this section we will describe a set of generators and equations
 used to reason about quantum computation, and show how some of its formal
-properties lead to particular issues in for the rewriting machinery.
+properties lead to particular issues for the rewriting machinery.
 
@@
 \end{description}
-To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : b
+To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
 \to A$.  Each arrow is represented as a small graph; its adjoint is
@@
 free compact closed category is then given by all graphs formed by
-composing and tensoring these basic graphs.
+composing and tensoring these basic graphs.  Note  that since $Q$ is
+self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.
 
@@
   obtained by flipping the diagram upside down.  In additon, we have a
-  ``colour symmetry'' so for each equation shown here gives rise to
-  second obtained by exchanging the two colours.
+  ``colour duality'': each equation shown here gives rise to
+  second, which is obtained by exchanging the two colours.  (In fact
+  this duality is derivable from the equations for involving $H$.)
   \begin{description}
@@
 \end{figure}
+
+Consider the equations shown in Figure \ref{fig:graph-quant-eqns} which
+involve only one colour:  these allow the remarkable \emph{spider theorem},
+first noted in \cite{Coecke2006Quantum-Measure}, to be proved:
+
+\begin{theorem}
+  Let $G$ be a connected graph generated from $\delta_Z$, $\epsilon_Z$,
+  $\alpha_Z$ and their adjoints;  then $G$ is totally determined by
+  the number of inputs, the number of outputs, and the sum modulo
+  $2\pi$ of the $\alpha$s which occur in it.
+\end{theorem}
+
+Hence any connected subgraph
+involving nodes of only one colour, wherever it occurs in a graph, may
+be collapsed to a single vertex labelled by a single value $\alpha$,
+giving a ``spider'' as depicted below:
+\[
+\inlinegraphic{3em}{gen-spider}
+\]
+Conversely, a spider may be arbitrarily divided into subspiders,
+provided the total in- and out-degree is preserved, along with and the
+sum of the $\alpha $s.  Further,
+one can derive from the  spider theorem $n$-fold versions of many of
+the other rewrites in the system.
+
+Spiders offer a very intuitive way to manipulate graphs,
+and are far more compact and convenient in calculations than the
+graphs built up naively 
+from the generators.  However, formalising spiders requires moving from
+finite graphs, where each vertex has bounded degree, and which are
+subject to a finite number of rewrite rules, to a system where nodes
+may have arbitrarily many edges, and there are infinitely many rewrite
+rules.  The desire to retain these intuitive reasoning methods
+motivates the extension from concrete graphs to \emph{graph patterns}.
+These will comprise the main subject of this paper.
 

### 2008-03-02 papers/reasoning-graphically/paper.tex
working...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1391 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\noindent This motivates a representation that has only one
-representative for such equations. Such a representation abstracts
-over the boundary nodes membership in the domain and co-domain. In
-this section we formalise such a notion and refer to the generalised
-boundary nodes as variable-nodes.
-
-The intuition of a variable node is that it can be replaced by any
-subgraph of the right type. Thus, unlike nodes in conrete a graph,
-variable nodes do not define an ordering on the edges that are
-incident. Instead, the edges that touch a variable node are simply a
-set. The type of a variable nodes is thus defined as a multiset rather
-than a tensor. Any tensor type can be lifted to a multiset type in the
-obvious manner, which we denote by the operation {\tt
-  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
-  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$ is
-an instance of a multiset type $M$, written $T \in M$, when $M
-\subseteq \mathtt{mset\_of\_tensor}(T)$. The intuition here is that
-when a variable node is instantiated to some other node, it's type
-will have at least the currently attached edges.
+\noindent This motivates a characterisation that has only one
+representative for such a set of equations. More presicely, we need a
+representation that abstracts over the boundary nodes membership in
+the domain or co-domain. In this section we formalise such a notion
+and refer to the generalised boundary nodes as \emph{variable
+  nodes}. This gives rise to a \emph{variable-node graphs}, which are
+graphs with variable nodes instead of boundary edges.
+
+Below, we give a more formal and constuctive account the semnatics by
+introducing a notion of an \emph{open embedding} and note that this
+corresponds to the intuitive idea of graph-matching. We then go on and
+introduce a way to join graphs with variable-nodes and note it's
+relationship to composition on the underlying concrete graphs.
+
+\subsection{Semantics}
+
+\subsection{Matching as an Open Embedding}
+
+To formalise the semantics of variable-node graphs, we define
+matching, which captures the intuitive idea with
+
+\begin{definition}
+  Matching is charcterised by an \emph{open embedding} which is strict
+  for the non-variable nodes.
+\end{definition}
+
+The intuition of variable nodes is that they can replaced by any be
+nodes of the right type that are themseleves within any graph.  A
+graph with variable nodes, $G$, can be given a formal semantics by
+being interpretated as a set of concrete graphs, denoted by
+$\vinterp{G}$.
+
+% Thus, unlike nodes in conrete a graph, variable nodes do not define an
+% ordering on the edges that are incident. Instead, the edges that touch
+% a variable node are simply a set. The type of a variable nodes is thus
+% defined as a multiset rather than a tensor. Any tensor type can be
+% lifted to a multiset type in the obvious manner, which we denote by
+% the operation {\tt mset\_of\_tensor}. For example, {\tt
+%   mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
+% say that a tensor-type $T$ is an instance of a multiset type $M$,
+% written $T \in M$, when $M \subseteq
+% \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
+% variable node is instantiated to some other node, it's type will have
+% at least the currently attached edges.
+
+% An instance of a graph with variable nodes is the graph with every
+% variable node replaced by a subgraph of the right type. When a graph
+% has several varibale-nodes, the replacement subgraphs may be
+% intersecting.
 
@@
 
-A graph with variable nodes, $G$, can be given a formal semantics by
-being interpretated as a set of concrete graphs which is denoted by
-$\vinterp{G}$. We refer to members of this set as instances. An
-instance of a graph with variable nodes is the graph with every
-variable node replaced by a subgraph of the right type. When a graph
-has several varibale-nodes, the replacement subgraphs may be
-intersecting. For example, in
-Figure~\ref{node-variable-instances-fig}, the graphs $a$, $b$ and $c$
-are instances of the graph $x$. A graph with variable nodes, and
-without boundary nodes, is called a \emph{variable-node graph}. Such
-graphs are complete in the sense that for every concrete graph there
-is a variable node graph that can be instantiated to it: $\forall
-g.\,\exists G.\,g \in \vinterp{G}$. The witness for this proof is made
-by simply replacing all bounary nodes with variable nodes.
+We refer to members of this set as instances.
+
+For example, in Figure~\ref{node-variable-instances-fig}, the graphs
+$a$, $b$ and $c$ are instances of the graph $x$. A graph with variable
+nodes, and without boundary nodes, is called a \emph{variable-node
+  graph}.
+
+\subsubsection{Completness and Adequacy.}
+
+Variable-node graphs are complete in the sense that for every concrete
+graph there is a variable node graph that can be instantiated to it:
+$\forall g.\,\exists G.\,g \in \vinterp{G}$. The witness for this
+proof is made by simply replacing all bounary nodes with variable
+nodes.
 
@@
 %instances of $a$ are a subset of the instances of $b$.
-
-Below, we give a more formal and constuctive account the semnatics by
-introducing a notion of an \emph{open embedding} and note that this
-corresponds to the intuitive idea of graph-matching. We then go on and
-introduce a way to join graphs with variable-nodes and note it's
-relationship to composition on the underlying concrete graphs.
-
-\subsection{Matching as an Open Embedding}
-
-To formalise the semantics of variable-node graphs, we define
-matching, which captures the intuitive idea with
-
-\begin{definition}
-  Matching is charcterised by an \emph{open embedding} which is strict
-  for the non-variable nodes.
-\end{definition}
 
@@
 A crucial feature of rules like the Spider Law is that an arbirary
-number of repetitions of a subgraph to be matched. To support this we
-introduce an operation, !-boxing, on graph representations. Given a
-graph representation, this introduces a new notation by outlining
-(called !-boxing) a set of chosen nodes. Intuitively, the resulting
-!-boxed graph can be thought of as representing a set of graphs with
-an arbirary number of copies of the !-boxed nodes, where every copy
-has the connects to the same nodes outside the !-box. The !-boxes can
-also be instantiated with zero copies which is erases all edges from
-the !-box. See Figure~\ref{bang-box-ex1-fig} for a simple example.
-
-More formaly, a graph with !-boxes is simply a graph paired with a set
-!-boxes. Each !-box is just the set of nodes inside it and each !-box
-must be disjoint from the others.\footnote{The disjointness condition
-  avoids a slightly more complex, although also more expressive,
-  semantics for overlaping node sets in the !-boxes. While such
-  expressivity is interesting from a representational point of view,
-  it is not needed to express the currently known rules of quantum
-  computation.}
+number of repetitions of a subgraph need to be matched. To support
+this we introduce an operation, !-boxing, on graph
+representations. Given a graph representation, this introduces a new
+notation by outlining (called !-boxing) a set of chosen
+nodes. Intuitively, the resulting !-boxed graph can be thought of as
+representing a set of graphs with an arbirary number of copies of the
+!-boxed nodes, where every copy connects, in the same way, to the
+nodes outside the !-box. The !-boxes can also be instantiated with
+zero copies, which erases all edges to and from the !-box. See
+Figure~\ref{spider-bang-rule} .
+
+More formaly, a graph with !-boxes is a graph paired with a set
+!-boxes. Each of these !-boxes is the set of nodes inside it and each
+!-box must be disjoint from the others.\footnote{The disjointness
+  condition avoids a slightly more complex, although also more
+  expressive, semantics for overlaping node sets in the !-boxes. While
+  such expressivity is interesting from a representational point of
+  view, it is not needed to express the currently known rules of
+  quantum computation.}
 
@@
 
-The set of !-box graphs matched by a !-box graph is defined as the set
-closed by the following operations:
+To formalise the intuitive notion that a !-box represents arbitrary
+number of copies of the subgraph made from it's nodes, we introduce
+\emph{!-box matching}. This is a binary relation on graphs, written
+infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
+matched by a !-box graph is defined as the set closed by the following
+operations on the !-boxes in a graph:
 
 \begin{definition}
-  \emph{copy} copies a !-box in a graph to produce a new graph with
-  two copies of the boxed subgraph, with a second box around the newly
-  copied nodes. Any edges between a node inside the !-box, $n$ and one
-  outside it $m$ get copied so that there is a new edge from $m$ to
-  the new copy of $n$.
+  \emph{copy}: copies a !-box, $b$ in a graph to produce a new graph
+  with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
+  is the new one. The set of !-boxes in the copied graph now also
+  contains the new !-box $b'$. Any edges between a node, $n$, inside
+  the !-box $b$, and a node, $m$, outside it, get copied so that there
+  is a new edge from $m$ to the new copy of $n$ in $b'$.
 \end{definition}
@@
 \begin{definition}
-  \emph{drop} simply removes the box, but leaves its contents in the
+  \emph{drop} simply removes the !-box, but leaves its contents in the
   graph.
@@
 \begin{definition}
-  \emph{kill} removes from the graph all nodes in the !-box set as
-  well as any incident edges to those nodes.
+  \emph{kill} removes from the graph all nodes in the !-box as well as
+  any incident edges.
 \end{definition}
@@
 
-\noindent We use $G \leq_! H$ to denote that $G$ matches $H$.
-
 \subsection{!-Box Semantics.}
 
-We give a formal semantics to !-boxing a representation of graphs by
-defining the !-box graph in terms of a set of graphs in the
-underlying representation. We denote the interpretation of a graph $G$
-by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are instances
-of $G$.
+We give a formal semantics to a !-box graphs by defining the !-box
+graph in terms of a set of graphs in the underlying representation. We
+denote the interpretation of a !-box graph $G$ by $\binterp{G}$ and
+say that members of $\binterp{G}$ are instances of $G$.
 
@@
 The purpose of graph patterns is to formally express, in a declarative
-manner, the intuitive idea of graph matching commonly used in for
-Quantum Computation. To do this, we combine the two extensions to the
+manner, the intuitive idea of graph matching which commonly used in
+quantum computation. To do this, we combine the two extensions to the
 concrete graph representation, namely variable nodes and !-boxes, to
@@
 
-\subsubsection{Concrete Matches.}
-
-The semantics for a graph pattern $G$ is a set of concrete graphs
-denoted by $\minterp{G}$. We say that members of this set are
-\emph{matched} by $G$. Formally, the set of matches are defined as:
-
-$$\minterp{G} = \{\vinterp{G'}\;.\; G' \in \binterp{G}\}$$
+\subsubsection{Completeness.}
 
@@
 unaffected by allowing !-boxes.
+
+\subsubsection{Concrete Matches.}
+
+The semantics for a graph pattern $G$ is a set of concrete graphs
+denoted by $\minterp{G}$. We say that members of this set are
+\emph{matched} by $G$. Formally, the concrete set of matches are
+defined as:
+
+$$\minterp{G} = \{\vinterp{G'}\;.\; G' \in \binterp{G}\}$$
 
@@
 graph patterns can be used to provide a \emph{meta-level} framework
-for reasoning about a given model of a compact closed
-categories. Following the terminology of logical frameworks such as
-Isabelle~\cite{isabelle}, we call the specification provided by the
-underlying model an \emph{object-level} graph formalism. Each
-object-level formalism defines a set of rules which are treated as the
-axioms for the system. The object-level implementation also defines
-the data at the nodes and edges as well as corresponding data-matching
-behaviour.
-
-\subsection{Substitution with Equational Rules}
-
-A rule in our framework declarative in the sense that it is a pair of
-pattern graph that represent the left and right hand sides of an
-equation. Substituon with a rule then involves replacing a subgraph
-that matches the left hand side with the right hand side. To do this
-there has to be a isomorphism between variable nodes in the left and
-right hand subgraphs. More concretely, there must be a unique naming
-of the variable nodes in the left hand side such that each name occurs
-in a unique variable node on the right hand side. Given a matching on
-the left hand side, the target subgraph can be replaced with right
-hand side connected by the corresponding variable nodes. This provides
+for reasoning about models of compact closed categories. Following the
+terminology of logical frameworks such as Isabelle~\cite{isabelle}, we
+call the specification provided by the underlying model an
+\emph{object-level} graph formalism.
+
+An object-level formalisation defines a set of rules which are treated
+as the axioms for the system. The object-level implementation also
+defines the data at the nodes and edges as well as corresponding
+data-matching behaviour. The meta-level provides machinery to
+manipulate graphs and derive new equations. In this section we
+describe the meta-level framework and prove its adequacy for
+rewriting.
+
+\subsection{Equational Rules}
+
+Axioms defined by an object-level model, as well as derived rules in
+our framework, are declarative in the sense that they are pairs of
+pattern graph. This pair represents the left and right hand sides of
+an equation.
+
+Substituon with a rule then involves replacing a subgraph that matches
+the left hand side with the right hand side. To do this there has to
+be a isomorphism between variable nodes in the left and right hand
+subgraphs. More concretely, there must be a unique naming of the
+variable nodes in the left hand side such that each name occurs in a
+unique variable node on the right hand side. Given a matching on the
+left hand side, the target subgraph can be replaced with right hand
+side connected by the corresponding variable nodes. This provides
 rewriting for pattern graphs.
+
+\subsubsection{Adequacy}
+
+A an equation that involves a graph pattern on one side
+
+\emph{Adequacy} says that if one by doing so within an
+equation, we are not
 
@@
   the compact closed category. Adquency ensures that the encoding as a
-  graph pattern represents only the intended rules.  The soundness of
-  the rules is obviously needed and requires proving the rule in terms
-  of the underlying semantics.
+  graph pattern represents only the intended concrete rules. The
+  soundness of the rules is obviously needed and requires proving the
+  rule in terms of the underlying semantics.
 \end{itemize}
 
+The proof of adequancy ...
+
 The completeness of the resulting system, with respect to the intended
-semantics, depends on the completeness of the set of given rules. The
-completeness of
+underlying semantics, depends on the completeness of the set of given
+rules.
 

### 2008-03-02 papers/reasoning-graphically/paper.tex
working...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1392 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 %\usepackage{pstricks,pst-node,pst-tree} % PS Tricks for diagrams
-%\usepackage{proof2}
 \setcounter{tocdepth}{3}
@@
 \usepackage{proof2}
-
 \usepackage{url}
+
 \urldef{\mailsa}\path|l.dixon@ed.ac.uk|
@@
 \begin{abstract}
-  Systems of quantum computation are typically represented by large
-  matrix compositions. However, recent graph-based formalisms offer an
-  approach which exposes the structure of these systems in a clearer
-  and more accessible way. Although these provide a significantly
-  simpler way of reading and presenting quantum computations, manual
-  manipulation of such graphs is slow and error prone. We present a
-  formalism that supports a mechanised reasoning about such
-  graphs. This involves a compositional account of graph rewriting
-  that preserves the underlying categorical semantics. We present a
-  system, which has been implemented, with a fixed logical kernel and
-  support for derived rules, which allows the model of quantum
-  computation to be specified declarativley.  Soundness of the system
-  is proved with respect to the underlying model and completeness is
-  an open problem.
+  Recent graph-based formalisms of quantum computation provide an
+  abstract and symbolic way to represent and simulate the
+  computations. However, manual manipulation of such graphs is slow
+  and error prone. We present a formalism that supports mechanised
+  reasoning about such graphs. This involves a compositional account
+  of graph rewriting that preserves the underlying categorical
+  semantics. We present a generic system with a fixed logical kernel
+  for representing compact closed categories. This provides a
+  declarative account which supports derived rules. We illustrate the
+  framework by instantiating it for the graphical language of quantum
+  computation.
 
@@
 
-History of diagrammatic representation going back to ???
-
-CAn formalise quantum mechanics in an abstract setting using compact
+%History of diagrammatic representation going back to ???
+
+Can formalise quantum mechanics in an abstract setting using compact
 or $\dag$-compact categories.
@@
 these equations using an infinite family of ``spider rules'';  this
-motivates the  more general notion of pattern graphs.
-
-next section introduces the  notion of pattern graph, which is
-essentially a graph structured
+motivates the  more general notion of graph patterns.
+
+%next section introduces the  notion of pattern graph, which is
+%essentially a graph structured
 
@@
 
-A \emph{directed graph}\footnote{
-Equivalently:  a directed graph is a functor $G$ from  $\bullet
-\pile{\rTo\\\rTo} \bullet$ to \catSet;  a graph morphism is then a natural
-transformation  $f: G \Rightarrow H$.
-} consists of a 4-tuple $(V,E,s,t)$ where $V$
-and $E$ are sets, respectively of \emph{vertices}\footnote{We will use
-  the words ``vertex'' and ``node'' interchangeably.} and \emph{edges},
-and $s$ and $t$ are maps 
+A \emph{directed graph}\footnote{ Equivalently: a directed
+  graph is a functor $G$ from $\bullet \pile{\rTo\\\rTo} \bullet$ to
+  \catSet; a graph morphism is then a natural transformation $f: G
+  \Rightarrow H$.  } consists of a 4-tuple $(V,E,s,t)$ where $V$ and
+$E$ are sets, respectively of \emph{vertices}\footnote{We will
+  use the words ``vertex'' and ``node'' interchangeably.} and
+\emph{edges}, and $s$ and $t$ are maps
 \begin{diagram}
@@
 \end{diagram}
-which we call \emph{source} and \emph{target}.  Define sets
-$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ of \emph{incoming}
-and \emph{outgoing} edges at a vertex $v$.  
-The \emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
+which we call \emph{source} and \emph{target}.  The definitions
+$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ express the
+\emph{incoming} and \emph{outgoing} edges at a vertex $v$.  The
+\emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
 \sizeof{\text{out}(v)}$.
 
-Given graphs $G$ and $H$, a \emph{graph morphism} $f : G\to H$ consists of
-functions $f_E : E_G \to E_H$ and $f_V:V_G\to V_H$ such that:
+To distinguish between elements of different graphs, we will use the
+subscript notation $G = (V_G,E_G,s_G,t_G)$. Given graphs $G$ and $H$,
+a \emph{graph morphism} $f : G\to H$ consists of functions $f_E : E_G
+\to E_H$ and $f_V:V_G\to V_H$ such that:
+
 \begin{gather}
@@
 \in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
-that $f_E(e') = e$.  We call $f$ an \emph{open embedding} for $V'$ if:
+that $f_E(e') = e$.
+
+\begin{definition}
+\label{open-embedding-def}
+We call $f$ an \emph{open embedding} for $V'$ if:
 \begin{enumerate}
@@
 \end{enumerate}
+\end{definition}
+
 The intuition behind this definition is that the subgraph of $G$
-determined by $V'$ should be preserved exactly by $f$, whereas the
-other vertices may be identified.
-
-Given two graphs $G,H$, a pair of partial maps $r_V: V_G
-\rightharpoonup V_H$, $r_E: E_G \rightharpoonup E_H$ is called a
-partial graph morphism if whenever $r(e)$ is defined then $r(s_G(e))$
-and $r(t_G(e))$ are defined, and the restriction of $r_E$ to its
-preimage satifies equations \eqref{eq:graph-hom1} and
+determined by $V'$ should be preserved exactly by $f$; it should
+contain $G$ without any extra incident egdes into the verticies $V'$
+whereas other vertices may be identified and may contain additional
+verticies.
+
+Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
+\rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
+partial graph morphism if whenever $r_E(e)$ is defined then
+$r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
+$r_E$ to its preimage satifies equations \eqref{eq:graph-hom1} and
 \eqref{eq:graph-hom2} above.
@@
 \begin{definition}
+\label{compactcat-def}
 A strict symmetric monoidal  category
@@
 
-\begin{figure}
+\begin{figure}[p]
   \figureline
-
-  The $(\cdot)^\dag$ functor gives a vertical symmetry to the
-  category, hence for every equation we have a second equation
-  obtained by flipping the diagram upside down.  In additon, we have a
-  ``colour duality'': each equation shown here gives rise to
-  second, which is obtained by exchanging the two colours.  (In fact
-  this duality is derivable from the equations for involving $H$.)
   \begin{description}
@@
 
-  \caption{Graphical Equations for Quantum Systems}
+  \caption{Graphical Equations for Quantum Systems. The $(\cdot)^\dag$
+    functor gives a vertical symmetry to the category, hence for every
+    equation we have a second equation obtained by flipping the
+    diagram upside down.  In additon, we have a ``colour duality'':
+    each equation shown here gives rise to second, which is obtained
+    by exchanging the two colours. The colour duality is derivable
+    from the equations for involving $H$.}
   \label{fig:graph-quant-eqns}
@@
 
-\noindent However, composition with semicircle line(s) on the domain
-or co-domain allows a proof involving one of the above to easily
-derive a corresponding one for the other two. Proving any of the
-following allows a trivial derivation of the others:
+\noindent However, composition with semi-circles ($d$ and $e$ from
+Definition~\ref{compactcat-def}) on the co-domain or domain allows an
+equation involving one of the above to easily lead to a derivation
+corresponding either of the other two. For instance, proving any of
+the following allows a trivial derivation of the others:
 
@@
 \noindent This motivates a characterisation that has only one
-representative for such a set of equations. More presicely, we need a
-representation that abstracts over the boundary nodes membership in
-the domain or co-domain. In this section we formalise such a notion
-and refer to the generalised boundary nodes as \emph{variable
-  nodes}. This gives rise to a \emph{variable-node graphs}, which are
-graphs with variable nodes instead of boundary edges.
-
-Below, we give a more formal and constuctive account the semnatics by
-introducing a notion of an \emph{open embedding} and note that this
-corresponds to the intuitive idea of graph-matching. We then go on and
-introduce a way to join graphs with variable-nodes and note it's
-relationship to composition on the underlying concrete graphs.
-
-\subsection{Semantics}
-
-\subsection{Matching as an Open Embedding}
+representative for such a set of equations. More presicely, we
+formalise a representation that abstracts over the boundary nodes
+membership in the domain or co-domain. This gives rise to a
+\emph{variable-node graphs}, in which boundary nodes have been
+generalised to \emph{variable nodes}. The intuition of variable nodes
+is that they can replaced by concrete nodes in some graph. We now
+define the semantics for variable-node graphs and then go on and
+introduce a way to join them.
+
+% relationship to composition on the underlying concrete graphs.
+
+\subsection{Semantics by Matching as an Open Embedding}
 
 To formalise the semantics of variable-node graphs, we define
-matching, which captures the intuitive idea with
+matching, which captures the intuitive idea of graph with variable
+nodes occuring within another graph.
 
 \begin{definition}
-  Matching is charcterised by an \emph{open embedding} which is strict
-  for the non-variable nodes.
+  \emph{Matching} is an open embedding
+  (Definition~\ref{open-embedding-def}) which is strict for the
+  non-variable nodes.
 \end{definition}
 
-The intuition of variable nodes is that they can replaced by any be
-nodes of the right type that are themseleves within any graph.  A
-graph with variable nodes, $G$, can be given a formal semantics by
+A graph with variable nodes, $G$, can be given a formal semantics by
 being interpretated as a set of concrete graphs, denoted by
-$\vinterp{G}$.
-
+$\vinterp{G}$. The interpretation is simply the ways a graph with
+variable nodes can be matched to a concrete graph. We refer to members
+of this set as instances. Matching is also well-defined between
+variable-node graphs, in which case we denote that $G$ macthes $H$ by
+$G \leq_v H$.
+
+\begin{theorem}
+  $G \leq_v H \Leftrightarrow \vinterp{G} \superseteq \vinterp{H}$. 
+\end{theorem}
+
+% \begin{figure}[t]
+% %  \scalebox{1.0}{\includegraphics{images/node-var-instance.eps}}
+% *** Add figure *** 
+% \label{node-variable-instances-fig}\caption{The concrete graphs $a$,
+%   $b$, and $c$ are instances of $x$, a graph with variable nodes.}
+% \end{figure}
+
+% For example, in Figure~\ref{node-variable-instances-fig}, the graphs
+% $a$, $b$ and $c$ are instances of the graph $x$. A graph with variable
+% nodes, and without boundary nodes, is called a \emph{variable-node
+%   graph}.
+
+%% TYPES 
 % Thus, unlike nodes in conrete a graph, variable nodes do not define an
@@
 
-\begin{figure}[t]
-%  \scalebox{1.0}{\includegraphics{images/node-var-instance.eps}}
-*** Add figure *** 
-\label{node-variable-instances-fig}\caption{The concrete graphs $a$,
-  $b$, and $c$ are instances of $x$, a graph with variable nodes.}
-\end{figure}
-
-We refer to members of this set as instances.
-
-For example, in Figure~\ref{node-variable-instances-fig}, the graphs
-$a$, $b$ and $c$ are instances of the graph $x$. A graph with variable
-nodes, and without boundary nodes, is called a \emph{variable-node
-  graph}.
-
-\subsubsection{Completness and Adequacy.}
-
-Variable-node graphs are complete in the sense that for every concrete
-graph there is a variable node graph that can be instantiated to it:
-$\forall g.\,\exists G.\,g \in \vinterp{G}$. The witness for this
-proof is made by simply replacing all bounary nodes with variable
-nodes.
+% \subsubsection{Completness and Adequacy.}
+
+% Variable-node graphs are complete in the sense that for every concrete
+% graph there is a variable node graph that can be instantiated to it:
+% $\forall g.\,\exists G.\,g \in \vinterp{G}$. The witness for this
+% proof is made by simply replacing all bounary nodes with variable
+% nodes.
 
@@
 
-In this section we outline how the formalism we have presented for
-graph patterns can be used to provide a \emph{meta-level} framework
-for reasoning about models of compact closed categories. Following the
-terminology of logical frameworks such as Isabelle~\cite{isabelle}, we
-call the specification provided by the underlying model an
-\emph{object-level} graph formalism.
-
-An object-level formalisation defines a set of rules which are treated
-as the axioms for the system. The object-level implementation also
+In this section we describe how the garph pattern formalism can
+provide a \emph{meta-level} framework for reasoning about models of
+compact closed categories. Following the terminology of logical
+frameworks such as Isabelle~\cite{isabelle}, we call the specification
+provided by the underlying model an \emph{object-level} graph
+formalism. In particular, an object-level formalisation defines a set
+of rules which are treated as the axioms for the system. It also
 defines the data at the nodes and edges as well as corresponding
-data-matching behaviour. The meta-level provides machinery to
-manipulate graphs and derive new equations. In this section we
-describe the meta-level framework and prove its adequacy for
+data-matching behaviour. For it's part, the meta-level provides
+machinery to manipulate graphs and derive new rules. In this section
+we describe the meta-level framework and prove its adequacy for
 rewriting.
@@
 Axioms defined by an object-level model, as well as derived rules in
-our framework, are declarative in the sense that they are pairs of
-pattern graph. This pair represents the left and right hand sides of
-an equation.
-
-Substituon with a rule then involves replacing a subgraph that matches
-the left hand side with the right hand side. To do this there has to
-be a isomorphism between variable nodes in the left and right hand
+our framework, are pairs of pattern graph. The pair represents the
+left and right hand sides of an equation. Rules are declarative in
+that they denote set of concrete equational rules.
+
+Substituon with a rule involves replacing a subgraph that matches the
+left hand side with the right hand side. To do this there has to be a
+isomorphism between variable nodes in the left and right hand
 subgraphs. More concretely, there must be a unique naming of the
@@
 unique variable node on the right hand side. Given a matching on the
-left hand side, the target subgraph can be replaced with right hand
-side connected by the corresponding variable nodes. This provides
-rewriting for pattern graphs.
-
-\subsubsection{Adequacy}
-
-A an equation that involves a graph pattern on one side
+left hand side, the target subgraph is replaced with the right hand
+side while keeping the same instantiations for the isomorphic variable
+nodes.
+
+Rules also contain a partial 1-1 mapping between !-boxes on the left
+and right hand sides. The intuition for this mapping is that the
+unfolding used when matching, for a !-box on the left, is applied to
+the mapped !-box on the right before replacement.
+
+Notationally, and implementationally, we annotate !-boxes and variable
+nodes in a graphs with unique names. For example,
+Figure~\ref{spider-law2-fig} illustrates the Spider Law as a rule in
+the general framework. The partial 1-1 mapping for !-boxes is
+represented by !-boxes having the same name on the left and right of a
+rule. Similarly, the isomorphism between variable nodes is captured by
+the set of variable node names on the left and right hand side being
+equal. There is an additional restriction that comes from the
+interplay between !-boxes and variable nodes. When a variable node
+appears within a !-box on one side of a rule it must appear under
+mapped !-box on the other side. This enforces that rules cannot change
+the type of a graph pattern.
+
+\subsection{Adequacy}
+
+This
+provides rewriting for pattern graphs.
+
+A an equation on a graph pattern is
 

### 2008-03-03 papers/reasoning-graphically/paper.tex
intro rewritten
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1396 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 languages motivated by the underlying logical structure of quantum
-mechanics itself.  These techniques have a number of advantages over
-the convential matrix-based approach to quantum mechanics, but in
-particular the graph can directly represent the structure of quantum
-states explicitly, including the
+mechanics
+itself~\cite{AbrCoe:CatSemQuant:2004,Selinger:dagger:2005,Coecke2005Kindergarten-Qu,Coecke2006POVMs-and-Naima,Coecke2006Quantum-Measure}.
+These techniques have a number of advantages over the convential
+matrix-based approach to quantum mechanics:
+
+\begin{itemize}
+\item The visual representation abstracts over the values in the
+  matricies. This removes detail that requires a lot of work for a
+  human to interpret.
+
+\item Many properties have a natural graphical representation. For
+  example, non entanglement is can be infered from disjoint subgraphs.
+
+\item The graphical calculus generalises to domains other than just
+  vector spaces. In particular, it provides a representation for
+  compact closed cartegories.
+
+\end{itemize}
+
+The major problem with these graphical representations is the lack of
+machinery for automating the manipulation of graphs.  Existing
+approaches to graph transformation have a different underlying
+semantics corresponding to graphs in general rather than those which
+describe compact closed categories. The result is that they provide an
+unsound concept of rewriting for the kind of graphs we are interested
+in.
+
+This paper reviews a formalism for graphical representations of
+compact closed categories. We also revisit a model of quantum
+computation based on this calculus. We then extend the graphical
+calculus in two significant ways driven by the need to express rules
+that could not be accounted for in the initial formalism. We provide a
+semantics for these in terms of the initial representation. By
+suitably combining these extensions to graphs, we provide a
+representation that captures an interesting and useful set of graph
+patterns. We use this as the foundation for a simple logical framework
+for manipulating compact closed categories as graphs. This has a
+suitable rewriting mechanism where the axioms of the underlying
+object-formalism are expressed as equations between graphs. We then
+present a short case study that illustrates the framework by
+instantiating it for the introduced model of quantum computation.
+This shows how the framework can be used to symbolically perform
+simplifications to quantum programs as well as simulate quantum
+computations.
 
@@
 
-Can formalise quantum mechanics in an abstract setting using compact
-or $\dag$-compact categories.
-
-We will give a description of graphs in general;  a specialisation of
-this gives the structure required for compact closed  categories.  The
-graph structure suffices to give a tight representation.
-
-We then introduce a concrete set of generators and equations for use
-in quantum computation.  We need rewrites to capture the non-logical
-axioms of this structure.   It turns out that we can most easily express
-these equations using an infinite family of ``spider rules'';  this
-motivates the  more general notion of graph patterns.
+%Can formalise quantum mechanics in an abstract setting using compact
+%or $\dag$-compact categories.
+
+%We will give a description of graphs in general;  a specialisation of
+%this gives the structure required for compact closed  categories.  The
+%graph structure suffices to give a tight representation.
+
+%We then introduce a concrete set of generators and equations for use
+%in quantum computation.  We need rewrites to capture the non-logical
+%axioms of this structure.   It turns out that we can most easily express
+%these equations using an infinite family of ``spider rules'';  this
+%motivates the  more general notion of graph patterns.
 
@@
 
-The intuition behind this definition is that the subgraph of $G$
-determined by $V'$ should be preserved exactly by $f$; it should
+\noindent The intuition behind this definition is that the subgraph of
+$G$ determined by $V'$ should be preserved exactly by $f$; it should
 contain $G$ without any extra incident egdes into the verticies $V'$
@@
 \begin{theorem}
-  $G \leq_v H \Leftrightarrow \vinterp{G} \superseteq \vinterp{H}$. 
+  $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. 
 \end{theorem}

### 2008-03-03 papers/reasoning-graphically/paper.tex
more small changes
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1397 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \begin{theorem}
-  $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. 
+\label{match-interp-thm}
+  $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. Proof... 
+\end{theorem}
+
+\begin{theorem}
+\label{reflexive-match-thm}
+  \emph{Matching is reflexive}: $G \leq_v G$ is a trivial consequence from open-embedding. 
 \end{theorem}
@@
   f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
-  f_G(v) \in f^{-1}_H(v)$. This states when a set of nodes in one
-  graph all map to a common node $n$ in the other graph, $n$ maps back
-  to one of the nodes that maps to it. 
+  f_G(v) \in f^{-1}_H(v)$. This states that if a set of nodes in one
+  graph all map to a common node, $n$, in the other graph, then $n$
+  maps back to one of the nodes that maps to it.
 \end{definition}
@@
 instantiates variable nodes: when a variable node is mapped to a
-concrete one, the result is a concrete node.
+concrete one, identification results in the concrete node.
 
@@
 \begin{theorem}
-  \emph{Match after plugging}: $G \leq \pi_f(G,H)$. And as a corrolary
-  we get $H \leq \pi_f(G,H)$ by symmetry of plugging. 
-\end{theorem}
-
-\begin{theorem}
-  \emph{Plugging preserves matching}: $V \leq G \Rightarrow V \leq
-  \pi_f(G,H)$. As corrolaries we get $V \leq H \Rightarrow V \leq
-  \pi_f(G,H)$ from symmetry of plugging and $\vinterp{G} \cap
-  \vinterp{H} \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the
-  definition of interpretation. 
+  \emph{Plugging preserves matching}: $V \leq_v G \Rightarrow V \leq_v
+  \pi_f(G,H)$. As corrolaries we get $V \leq_v H \Rightarrow V \leq_v
+  \pi_f(G,H)$ from symmetry of plugging, $\vinterp{G} \cap \vinterp{H}
+  \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the definition of
+  interpretation, $G \leq_v \pi_f(G,H)$ from reflexivity of matching
+  and consequently $H \leq_v \pi_f(G,H)$ by symmetry of plugging.
 \end{theorem}

### 2008-03-04 papers/reasoning-graphically/paper.tex
pollishing..
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1398 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-  \keywords{graph rewriting, quantum computing, categorical
-    logic, interactive theorem proving, graphical calculi}
+  \keywords{graph rewriting, quantum computing, categorical logic,
+    interactive theorem proving, graphical calculi; concerning:
+    symbolic computation, automated reasoning, formal mathematics,
+    logic}
 \end{abstract}
@@
 \emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
-\sizeof{\text{out}(v)}$.
-
-To distinguish between elements of different graphs, we will use the
-subscript notation $G = (V_G,E_G,s_G,t_G)$. Given graphs $G$ and $H$,
-a \emph{graph morphism} $f : G\to H$ consists of functions $f_E : E_G
-\to E_H$ and $f_V:V_G\to V_H$ such that:
+\sizeof{\text{out}(v)}$. To distinguish between elements of different
+graphs, we will use the subscript notation $G = (V_G,E_G,s_G,t_G)$.
+
+Given graphs $G$ and $H$, a \emph{graph morphism} $f : G\to H$
+consists of functions $f_E : E_G \to E_H$ and $f_V:V_G\to V_H$ such
+that:
 
@@
 
+\noindent These ensure that the structure of the graph is preserved by
+the morphism: an edge connected to a node gets mapped to a new edge
+that must connected in the same way to the mapped node.
+
 Let $f: G \to H$ be graph morphism and let $V' \subseteq V_G$. We say
@@
 \in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
-that $f_E(e') = e$.
+that $f_E(e') = e$. Strictness ensures that there are no additional
+edges in $H$.
 
@@
 \label{open-embedding-def}
-We call $f$ an \emph{open embedding} for $V'$ if:
+We call a graph morphism $f$ an \emph{open embedding} for $V'$ if:
 \begin{enumerate}
@@
 \]
-are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
+which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
 $\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
@@
 
-\section{Graphs as Quantum Computations}
+\section{Quantum Computations as Graphs}
 \label{sec:quotients-rewriting}
@@
 \begin{theorem}
-  Let $G$ be a connected graph generated from $\delta_Z$, $\epsilon_Z$,
-  $\alpha_Z$ and their adjoints;  then $G$ is totally determined by
-  the number of inputs, the number of outputs, and the sum modulo
-  $2\pi$ of the $\alpha$s which occur in it.
+  The \emph{Spider Theorem}: let $G$ be a connected graph generated
+  from $\delta_Z$, $\epsilon_Z$, $\alpha_Z$ and their adjoints; then
+  $G$ is totally determined by the number of inputs, the number of
+  outputs, and the sum modulo $2\pi$ of the $\alpha$s which occur in
+  it.
 \end{theorem}
@@
 provided the total in- and out-degree is preserved, along with and the
-sum of the $\alpha $s.  Further,
-one can derive from the  spider theorem $n$-fold versions of many of
-the other rewrites in the system.
+sum of the $\alpha $s.  Further, one can derive from the spider
+theorem $n$-fold versions of many of the other rewrites in the system.
+Informally, this can be expressed as the graph equation in
+Figure~\ref{informal-spider-law-fig}.
 
@@
 \noindent However, composition with semi-circles ($d$ and $e$ from
-Definition~\ref{compactcat-def}) on the co-domain or domain allows an
-equation involving one of the above to easily lead to a derivation
-corresponding either of the other two. For instance, proving any of
+Definition~\ref{compactcat-def}), on the co-domain or domain, allows
+an equation involving one of the above to easily lead to a derivation
+corresponding to either of the other two. For instance, proving any of
 the following allows a trivial derivation of the others:
@@
 generalised to \emph{variable nodes}. The intuition of variable nodes
-is that they can replaced by concrete nodes in some graph. We now
-define the semantics for variable-node graphs and then go on and
-introduce a way to join them.
+is that they can replaced by concrete nodes in some graph. This is a
+process analgous to composition. We now define the semantics for
+variable-node graphs and then go on and introduce a way to join them.
 
@@
 To formalise the semantics of variable-node graphs, we define
-matching, which captures the intuitive idea of graph with variable
+matching, which captures the intuitive idea of a graph with variable
 nodes occuring within another graph.
@@
 being interpretated as a set of concrete graphs, denoted by
-$\vinterp{G}$. The interpretation is simply the ways a graph with
-variable nodes can be matched to a concrete graph. We refer to members
-of this set as instances. Matching is also well-defined between
-variable-node graphs, in which case we denote that $G$ macthes $H$ by
-$G \leq_v H$.
+$\vinterp{G}$. The interpretation is simply the set of concrete graphs
+which the variable node graph matches. We refer to members of this set
+as instances. Matching is also well-defined between two variable-node
+graphs, in which case we denote that $G$ macthes $H$ by $G \leq_v H$.
 
@@
 \label{match-interp-thm}
-  $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. Proof... 
+$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. The
+proof from left to right is a consequence of the trivial lemma that a
+compositon of open embeddings is an open embedding. This makes the
+embedding of $G$ into $H$ composed with $H$ into $\vinterp{H}$ an open
+embedding of $G$ into $\vinterp{H}$. From right to left, we compose
+the open embedding $\vinterp{G}$, restricted to it's subset
+$\vinterp{H}$, with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
 \end{theorem}
@@
 \label{reflexive-match-thm}
-  \emph{Matching is reflexive}: $G \leq_v G$ is a trivial consequence from open-embedding. 
+\emph{Matching is reflexive}: $G \leq_v G$ is a trivial instance of
+the same theorem for open-embeddings, where the embedding is the
+identity map.
 \end{theorem}
@@
 \begin{definition}
-  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of
-  strict partial homomorpihc mappings, $f_G : G \Longrightarrow H$ and
-  $f_H : H \Longrightarrow G$ such that $f_G$ and $f_H$ do not map
-  concrete nodes and for the nodes that are mapped \emph{agree}
-  following the definition given below.
+  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of strict
+  partial graph morphisms, $f_G : G \rightarrow H$ and $f_H : H
+  \rightarrow G$ such that $f_G$ and $f_H$ \emph{agree}, following
+  the definition given below, on node mappings and do not map concrete
+  nodes.
 \end{definition}
@@
 We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
-overload the notation $\pi_p(G,H)$ to let it denote the graph that
-results from the plugging. This is defined to be the union of the
-nodes and edges from $G$ and $H$ where the nodes and edges mapped by
-the pair of plugging function are identified. Identification of nodes
-instantiates variable nodes: when a variable node is mapped to a
-concrete one, identification results in the concrete node.
+overload the notation $\pi_p(G,H)$ to denote the graph that results
+from the plugging. This is defined to be the union of the nodes and
+edges from $G$ and $H$ where the nodes and edges mapped by the pair of
+plugging function are identified. Identification of nodes instantiates
+variable nodes: when a variable node is mapped to a concrete one,
+identification results in the concrete node.
 
@@
 nodes, the definition takes some care. In particular, it is important
-that plugging preserves the types of concrete nodes and results in
-subtypes in variable nodes. Figure~\ref{plugging-examples-fig}
-presents some examples to help the reader get the intuition as well as
-understand some of the more subtlte cases.
+that plugging preserves incident edges of concrete nodes.
+Figure~\ref{plugging-examples-fig} presents some examples to help the
+reader get the intuition as well as understand some of the more
+subtlte cases.
 
@@
   \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
-  = \pi_q(H,G)$. Proof follows simply: given $p = (f_G,f_H)$, $q =
-  (f_H,f_G)$. As a corrolary, we get $\pi(G,H) \Leftrightarrow
-  \pi(H,G)$.
+  = \pi_q(H,G)$. Proof, given $p = (f_G,f_H)$ then $q = (f_H,f_G)$. As
+  a corollary, we get $\pi(G,H) \Leftrightarrow \pi(H,G)$.
 \end{theorem}
@@
 
-A crucial feature of rules like the Spider Law is that an arbirary
+A crucial feature of rules like the Spider Theorem is that an arbirary
 number of repetitions of a subgraph need to be matched. To support
@@
 in~\S\ref{graph-repr-comp}, is too restrictive for reasoning about
-quantum computation. In particular, pictures such the Spider Law, in
+quantum computation. In particular, pictures such the Spider Theorem, in
 Figure~\ref{spider-law-fig} are frequently needed, but not

### 2008-03-04 papers/reasoning-graphically/paper.tex
...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1399 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-Graphs with certain additional structure give a
-representation for compact closed  categories; we now give an overview
-of this construction.  The details omitted here can be found in
+Graphs with certain additional structure give a representation for
+compact closed categories; we now give an overview of this
+construction.  The details omitted here can be found in
 \cite{Duncan:thesis:2006}.  Pictorial representations are in
-Fig.~\ref{fig:comcl-graphs}.
-
-\begin{figure}
-  We make the convention that domain of an arrow is at the top of the
-  picture, and its codomain is at the bottom.
+Fig.~\ref{fig:comcl-graphs}. We make the convention that domain of an
+arrow is at the top of the picture, and its codomain is at the bottom.
+
+\begin{figure}[t]
   \centering
@@
   \]
-  \caption{Compact Closed Structure as Graphs}
+  \caption{Compact Closed Structure as Graphs. }
   \label{fig:comcl-graphs}
@@
 
-Hence any connected subgraph
-involving nodes of only one colour, wherever it occurs in a graph, may
-be collapsed to a single vertex labelled by a single value $\alpha$,
-giving a ``spider'' as depicted below:
-\[
-\inlinegraphic{3em}{gen-spider}
-\]
-Conversely, a spider may be arbitrarily divided into subspiders,
-provided the total in- and out-degree is preserved, along with and the
-sum of the $\alpha $s.  Further, one can derive from the spider
-theorem $n$-fold versions of many of the other rewrites in the system.
-Informally, this can be expressed as the graph equation in
-Figure~\ref{informal-spider-law-fig}.
-
-Spiders offer a very intuitive way to manipulate graphs,
-and are far more compact and convenient in calculations than the
-graphs built up naively 
-from the generators.  However, formalising spiders requires moving from
-finite graphs, where each vertex has bounded degree, and which are
-subject to a finite number of rewrite rules, to a system where nodes
-may have arbitrarily many edges, and there are infinitely many rewrite
-rules.  The desire to retain these intuitive reasoning methods
-motivates the extension from concrete graphs to \emph{graph patterns}.
-These will comprise the main subject of this paper.
+\begin{figure}[t]
+\includegraphics{images/gen-spider} = \includegraphics{images/gen-spider}
+\label{fig:informal-spider-law}\caption{An informal equation on graphs
+  that expresses the Spider Theorem. }
+\end{figure}
+
+Hence any connected subgraph involving nodes of only one colour,
+wherever it occurs in a graph, may be collapsed to a single vertex
+labelled by a single value $\alpha$, giving a ``spider''. Informally,
+this can be depicted graphically as the equation in
+Figure~\ref{fig:informal-spider-law}. Conversely, a spider may be
+arbitrarily divided into subspiders, provided the total in- and
+out-degree is preserved, along with and the sum of the $\alpha $s.
+Further, one can derive, from the spider theorem, $n$-fold versions of
+many of the other equations.
+
+Spiders offer a very intuitive way to manipulate graphs, and are far
+more compact and convenient in calculations than the graphs built up
+naively from the generators.  However, formalising spiders requires
+moving from finite graphs, where each vertex has bounded degree, and
+which are subject to a finite number of rewrite rules, to a system
+where nodes may have arbitrarily many edges, and there are infinitely
+many concrete rewrite rules. The desire to retain these intuitive
+reasoning methods motivates the extension from concrete graphs to
+\emph{graph patterns}. These will comprise the main subject of this
+paper.
 
@@
 about quantum computation.
+
+graph-pattern-spider-law-fig
 

### 2008-03-05 papers/reasoning-graphically/paper.tex
more spider pictures
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1400 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
work...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1401 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \begin{figure}[t]
-\includegraphics{images/gen-spider} = \includegraphics{images/gen-spider}
+$$\begin{array}{ccc}
+\inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
+\end{array}$$
 \label{fig:informal-spider-law}\caption{An informal equation on graphs
@@
 generalised to \emph{variable nodes}. The intuition of variable nodes
-is that they can replaced by concrete nodes in some graph. This is a
-process analgous to composition. We now define the semantics for
-variable-node graphs and then go on and introduce a way to join them.
-
+is that they can replaced by concrete nodes in some graph in a process
+analgous to composition.
+
+% We now define the semantics for variable-node
+% graphs and then go on and introduce a way to join them.
 % relationship to composition on the underlying concrete graphs.
 
-\subsection{Semantics by Matching as an Open Embedding}
+% \subsection{Semantics by Matching as an Open Embedding}
 
@@
 
-\subsection{Variable-node Plugging}
-
-While concrete graphs can be composed sequentially as well as tensored
-in an adjacent manner, variable-node graphs enjoy a single but
-somewhat wilder joining operation which we call \emph{plugging}.
-
-\begin{definition}
-  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of strict
-  partial graph morphisms, $f_G : G \rightarrow H$ and $f_H : H
-  \rightarrow G$ such that $f_G$ and $f_H$ \emph{agree}, following
-  the definition given below, on node mappings and do not map concrete
-  nodes.
-\end{definition}
-
-\begin{definition}
-  Two mappings $f_G$ and $f_H$ \emph{agree} when $\forall v \in
-  f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
-  f_G(v) \in f^{-1}_H(v)$. This states that if a set of nodes in one
-  graph all map to a common node, $n$, in the other graph, then $n$
-  maps back to one of the nodes that maps to it.
-\end{definition}
-
-We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
-overload the notation $\pi_p(G,H)$ to denote the graph that results
-from the plugging. This is defined to be the union of the nodes and
-edges from $G$ and $H$ where the nodes and edges mapped by the pair of
-plugging function are identified. Identification of nodes instantiates
-variable nodes: when a variable node is mapped to a concrete one,
-identification results in the concrete node.
-
-While this is simply the intuitive way to combine graphs with variable
-nodes, the definition takes some care. In particular, it is important
-that plugging preserves incident edges of concrete nodes.
-Figure~\ref{plugging-examples-fig} presents some examples to help the
-reader get the intuition as well as understand some of the more
-subtlte cases.
-
-\begin{theorem}
-  \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
-  = \pi_q(H,G)$. Proof, given $p = (f_G,f_H)$ then $q = (f_H,f_G)$. As
-  a corollary, we get $\pi(G,H) \Leftrightarrow \pi(H,G)$.
-\end{theorem}
-
-\begin{theorem}
-  \emph{Plugging preserves matching}: $V \leq_v G \Rightarrow V \leq_v
-  \pi_f(G,H)$. As corrolaries we get $V \leq_v H \Rightarrow V \leq_v
-  \pi_f(G,H)$ from symmetry of plugging, $\vinterp{G} \cap \vinterp{H}
-  \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the definition of
-  interpretation, $G \leq_v \pi_f(G,H)$ from reflexivity of matching
-  and consequently $H \leq_v \pi_f(G,H)$ by symmetry of plugging.
-\end{theorem}
+% \subsection{Variable-node Plugging}
+
+% While concrete graphs can be composed sequentially as well as tensored
+% in an adjacent manner, variable-node graphs enjoy a single but
+% somewhat wilder joining operation which we call \emph{plugging}.
+
+% \begin{definition}
+%   A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of strict
+%   partial graph morphisms, $f_G : G \rightarrow H$ and $f_H : H
+%   \rightarrow G$ such that $f_G$ and $f_H$ \emph{agree}, following
+%   the definition given below, on node mappings and do not map concrete
+%   nodes.
+% \end{definition}
+
+% \begin{definition}
+%   Two mappings $f_G$ and $f_H$ \emph{agree} when $\forall v \in
+%   f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
+%   f_G(v) \in f^{-1}_H(v)$. This states that if a set of nodes in one
+%   graph all map to a common node, $n$, in the other graph, then $n$
+%   maps back to one of the nodes that maps to it.
+% \end{definition}
+
+% We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
+% overload the notation $\pi_p(G,H)$ to denote the graph that results
+% from the plugging. This is defined to be the union of the nodes and
+% edges from $G$ and $H$ where the nodes and edges mapped by the pair of
+% plugging function are identified. Identification of nodes instantiates
+% variable nodes: when a variable node is mapped to a concrete one,
+% identification results in the concrete node.
+
+% While this is simply the intuitive way to combine graphs with variable
+% nodes, the definition takes some care. In particular, it is important
+% that plugging preserves incident edges of concrete nodes.
+% Figure~\ref{plugging-examples-fig} presents some examples to help the
+% reader get the intuition as well as understand some of the more
+% subtlte cases.
+
+% \begin{theorem}
+%   \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
+%   = \pi_q(H,G)$. Proof, given $p = (f_G,f_H)$ then $q = (f_H,f_G)$. As
+%   a corollary, we get $\pi(G,H) \Leftrightarrow \pi(H,G)$.
+% \end{theorem}
+
+% \begin{theorem}
+%   \emph{Plugging preserves matching}: $V \leq_v G \Rightarrow V \leq_v
+%   \pi_f(G,H)$. As corrolaries we get $V \leq_v H \Rightarrow V \leq_v
+%   \pi_f(G,H)$ from symmetry of plugging, $\vinterp{G} \cap \vinterp{H}
+%   \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the definition of
+%   interpretation, $G \leq_v \pi_f(G,H)$ from reflexivity of matching
+%   and consequently $H \leq_v \pi_f(G,H)$ by symmetry of plugging.
+% \end{theorem}
 
@@
 number of repetitions of a subgraph need to be matched. To support
-this we introduce an operation, !-boxing, on graph
-representations. Given a graph representation, this introduces a new
-notation by outlining (called !-boxing) a set of chosen
-nodes. Intuitively, the resulting !-boxed graph can be thought of as
-representing a set of graphs with an arbirary number of copies of the
-!-boxed nodes, where every copy connects, in the same way, to the
+this we introduce an operation, !-boxing (pronounced band-boxing), on
+graph representations. Given a graph representation, this introduces a
+new notation by outlining a set of nodes. These nodes are said to be
+in a !-box. Intuitively, the resulting !-boxed graph can be thought of
+as representing a set of graphs with an arbirary number of copies of
+the !-boxed nodes, where every copy connects, in the same way, to the
 nodes outside the !-box. The !-boxes can also be instantiated with
-zero copies, which erases all edges to and from the !-box. See
-Figure~\ref{spider-bang-rule} .
-
-More formaly, a graph with !-boxes is a graph paired with a set
-!-boxes. Each of these !-boxes is the set of nodes inside it and each
-!-box must be disjoint from the others.\footnote{The disjointness
-  condition avoids a slightly more complex, although also more
-  expressive, semantics for overlaping node sets in the !-boxes. While
-  such expressivity is interesting from a representational point of
-  view, it is not needed to express the currently known rules of
-  quantum computation.}
-
-For the purposes of this paper, the underlying graph representation of
-!-boxed graphs is variable-node graphs. The set of variable-node
-graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
-now define the concept of !-box matching which we use to provide a
-semantics for the !-boxed graphs.
+zero copies. This erases all edges to and from the !-box.
+
+%Figure~\ref{fig:spider-bang-lhs-example} shows an how the left hand side of
+%the spider theorem can be represented with bang boxes and illustrates
+%some instances of it.
+
+%\begin{figure}[t]
+%$$\begin{array}{ccc}
+%\inlinegraphic{6em}{spider_lhs_patt} & = & \inlinegraphic{6em}{spider_rhs_ex1} & , & \inlinegraphic{6em}{spider_rhs_ex2} & , & \inlinegraphic{6em}{spider_rhs_ex3} & \dots
+%\end{array}$$
+%\label{fig:spider-bang-lhs-example}\caption{An informal equation on graphs
+%  that expresses the Spider Theorem. }
+%\end{figure}
+
+More formaly, a !-box graph is a graph paired with a set !-boxes. Each
+of these !-boxes is the set of nodes inside it and each !-box must be
+disjoint from the others.\footnote{The disjointness condition avoids a
+  slightly more complex, although also more expressive, semantics for
+  overlaping node sets in the !-boxes. While such expressivity is
+  interesting from a representational point of view, it is not needed
+  to express the currently known rules for quantum computation.}  We
+now define the concept of !-box matching which we then use to provide
+a formal semantics for the !-boxed graphs.
 
@@
 \emph{!-box matching}. This is a binary relation on graphs, written
-infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
-matched by a !-box graph is defined as the set closed by the following
-operations on the !-boxes in a graph:
+infix as $G \leq_! H$ when $G$ matches $H$. The set matching !-box
+graphs is defined as the set closed by the following operations on the
+!-boxes in a graph:
 
@@
 \begin{theorem}
+\label{thm:bang-box-respect}
 \emph{!-Matching respects !-box semantics}: $G \leq_! H
 \Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
-  simple consequence from the defintion of $\binterp{G}$ being a
-  subset of the graphs that match $G$.
+simple consequence from the defintion of $\binterp{G}$ being a subset
+of the graphs that match $G$.
 \end{theorem}
+
+For the purposes of this paper, the underlying graph representation of
+!-boxed graphs is variable-node graphs. This gives rise to graph
+patterns which we now discuss in more detail.
+% The set of variable-node
+% graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.
 
@@
 The representation of Compact Closed Categories as Graphs, discussed
-in~\S\ref{graph-repr-comp}, is too restrictive for reasoning about
-quantum computation. In particular, pictures such the Spider Theorem, in
-Figure~\ref{spider-law-fig} are frequently needed, but not
-expressible. Such pictures are used to denote an infinite family of
-concrete graphs that can be rewritten. In this section we combine the
-{\em !-boxes} with {\em variable nodes} graph extensions to form a
-representation that allows us to finitely express these infinite
-families of concrete graphs. We call this extension \ref{graph
-  patterns}. The Spider Law can now be represented by the equation in
-Figure~\ref{graph-pattern-spider-law-fig}. We also extend the notion
-of matching for graph pattens as well as present a corresponding
-notion of plugging. This provides the foundations for the rewriting
-machinery in~\S\ref{rewriting} which can then be used to reasoning
-about quantum computation.
-
-graph-pattern-spider-law-fig
+in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
+quantum computation. In particular, graphical rules such the Spider
+Theorem, from Figure~\ref{fig:informal-spider-law}, are frequently
+needed, but not expressible. In this section we combine the {\em
+  !-boxes} and {\em variable-node} extensions to graphs. We call this
+representation \emph{graph patterns}. This forms a representation that
+allows us to express, in a finite way, the infinite family of concrete
+graphs equations. In particular, the Spider Theroem can now be
+represented by the equation shown in Figure~\ref{fig:spider-thm-patt}.
+We also extend the notion of matching for graph pattens.
+%as well as present a corresponding notion of plugging. 
+This provides the foundations for the rewriting machinery
+in~\S\ref{sec:rewriting} which can then be used to reasoning about
+quantum computation.
+
+\begin{figure}[t]
+$$\begin{array}{ccc}
+\inlinegraphic{8em}{spider_lhs_patt} & \;=\; & \inlinegraphic{6em}{spider_rhs_patt}
+\end{array}$$
+\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using graph patterns.}
+\end{figure}
+
+% \subsubsection{Completeness.}
+
+% Graph patterns are complete in the sense that for every concrete graph
+% there is a pattern that matches it. This is a trivial consequence of
+% the completeness of variable-node extension to graphs, which is
+% unaffected by allowing !-boxes.
 
@@
 
-\subsubsection{Completeness.}
-
-Graph patterns are complete in the sense that for every concrete graph
-there is a pattern that matches it. This is a trivial consequence of
-the completeness of variable-node extension to graphs, which is
-unaffected by allowing !-boxes.
-
 \subsubsection{Concrete Matches.}
@@
 
-Because viable-node matching preserves the semantics, it is trivial to
-prove that this definition meets the specification. This gives us
-matching between pattern-nodes.
-
-\subsubsection{Graph Pattern Equality.}
-
-Equality of graph patterns can simply be defined as mutual matching:
-
-\begin{definition}
-  $G = H \Leftrightarrow G \leq H \land H \leq G$
-\end{definition}
-
-\subsection{Plugging}
-
-We then extend the plugging
-operation on variable-node graphs to graphs with !-boxes.
-
-By defining graph patterns as a conservative extension of concrete
-graphs we get the relative soundness for them.
+Because variable-node matching preserves the semantics, it is trivial to
+prove that this definition meets the specification.
+
+% \subsubsection{Graph Pattern Equality.}
+
+% Equality of graph patterns can simply be defined as mutual matching:
+
+% \begin{definition}
+%   $G = H \Leftrightarrow G \leq H \land H \leq G$
+% \end{definition}
+
+% \subsection{Plugging}
+
+% We then extend the plugging
+% operation on variable-node graphs to graphs with !-boxes.
+
+% By defining graph patterns as a conservative extension of concrete
+% graphs we get the relative soundness for them.
+
+% We note that using concrete graphs as the underlying representation
+% results in composition of !-box graphs to disrespect the !-box
+% semantics: Theorem \ref{thm:bang-box-respects-plugging} fails to hold
+% for composision.
 
@@
 left and right hand sides of an equation. Rules are declarative in
-that they denote set of concrete equational rules.
-
-Substituon with a rule involves replacing a subgraph that matches the
-left hand side with the right hand side. To do this there has to be a
-isomorphism between variable nodes in the left and right hand
-subgraphs. More concretely, there must be a unique naming of the
-variable nodes in the left hand side such that each name occurs in a
-unique variable node on the right hand side. Given a matching on the
-left hand side, the target subgraph is replaced with the right hand
-side while keeping the same instantiations for the isomorphic variable
-nodes.
-
-Rules also contain a partial 1-1 mapping between !-boxes on the left
-and right hand sides. The intuition for this mapping is that the
-unfolding used when matching, for a !-box on the left, is applied to
-the mapped !-box on the right before replacement.
+that they denote a set of concrete equational rules.
+
+The intuitive idea of substituon with a rule is to replace a subgraph
+that matches the left hand side with the right hand side. However, not
+all pairs of rules are make sense with respect to the underlying
+semantics. For an equation to be well defined with respect to the
+compact closed structure it must not be possible to changes the type
+of a concrete graph by rewriting. Mapping this restriction back to
+pairs of graph patterns results in the following conditions on rules:
+
+\begin{itemize}
+
+\item There has to be a isomorphism between variable nodes in the left
+  and right hand subgraphs. More concretely, there must be a unique
+  naming of the variable nodes in the left hand side such that each
+  name occurs in a unique variable node on the right hand side. Given
+  a matching on the left hand side, the target subgraph is replaced
+  with the right hand side while keeping the same instantiations for
+  the isomorphic variable nodes.
+
+\item Rules must also define a partial 1-1 mapping between !-boxes on
+  the left and right hand sides. The intuition for this mapping is
+  that the unfolding used when matching a !-box on the left, is
+  applied to the mapped !-box on the right before replacement.
+
+\item The interplay between !-boxes and variable nodes means that when
+  a variable node appears within a !-box on one side of a rule it must
+  also appear under a mapped !-box on the other side.
+
+\end{itemize}
 
 Notationally, and implementationally, we annotate !-boxes and variable
-nodes in a graphs with unique names. For example,
-Figure~\ref{spider-law2-fig} illustrates the Spider Law as a rule in
-the general framework. The partial 1-1 mapping for !-boxes is
-represented by !-boxes having the same name on the left and right of a
-rule. Similarly, the isomorphism between variable nodes is captured by
-the set of variable node names on the left and right hand side being
-equal. There is an additional restriction that comes from the
-interplay between !-boxes and variable nodes. When a variable node
-appears within a !-box on one side of a rule it must appear under
-mapped !-box on the other side. This enforces that rules cannot change
-the type of a graph pattern.
-
-\subsection{Adequacy}
-
-This
-provides rewriting for pattern graphs.
-
-A an equation on a graph pattern is
-
-\emph{Adequacy} says that if one by doing so within an
-equation, we are not
+nodes in a graphs with unique names. For example, see
+Figure~\ref{fig:spider-thm-patt} which illustrates the Spider Theorem
+as a rule between graph patterns. In this figure, the partial 1-1
+mapping for !-boxes is represented by !-boxes having the same name on
+the left and right of a rule.  Similarly, the isomorphism between
+variable nodes is captured by the set of variable node names on the
+left and right hand side being equal.
+
+\subsubsection{Adequacy.} An equation on graph patterns corresponds to
+a possibly infinite set of equations betweens concrete graphs. We
+might expect that such equations express too much: they may allow
+rewrites which are not intended. We prove \emph{adequacy}, which
+states that this is not the case. In particular, given an equation
+between concrete graphs, $G = H$, the equations between the lifting of
+these to variable node graphs by replacing boundary nodes with
+variable nodes is a conservative extension. The proof follows directly
+from the obervation that every instance of the lifted variable node
+equation corresponds to compsition of $G$ and $H$ with some other
+graph (the non-embedded part).
 
@@
 
-The meta-logical rules of the framework are quite simple as they only
-involve dealing with object-level equations:
+Having defined what make a valid rule, we now present the meta-logic
+of the framework. This is quite simple as they only involve dealing
+with object-level equations:
 
@@
 \prooftree
-\Gamma \vdash A = B
+\Gamma \vdash A = B \quad\quad
 \Gamma \vdash C = D
@@
 
-The proof of adequancy ...
-
 The completeness of the resulting system, with respect to the intended
 underlying semantics, depends on the completeness of the set of given
-rules.
-
-\subsection{Derived Rules}
-
-\subsection{Simplification}
-
-\subsection{Adequacy}
+rules. Completeness is a challenging and still open problem for graph
+based models of quantum computation. Even without completeness, we
+obersve in \S\ref{sec:case-study} that interesting and useful quantum
+computations can be expressed.
+
+\section{Implementation}
+\label{ssec:implementation}
+
+% \subsection{Simplification}
+
+% One of the main applications of
 
@@
 \section{A Case Study in Quantum Computation}
-
-The model of qantum computation in which we are interested is finite
-dimensional hilbert spaces.
+\label{sec:case-study}
+%The model of qantum computation in which we are interested is finite
+%dimensional hilbert spaces.
 
@@
 
-Completeness is a challenging and still open problem for graph based
-models of quantum computation. However, even without completeness
-interesting quantum computations can be expressed.
+\subsection{}
+\label{ssec:node-expressions}
 

### 2008-03-05 papers/reasoning-graphically/paper.tex
more work.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1402 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
   Recent graph-based formalisms of quantum computation provide an
-  abstract and symbolic way to represent and simulate the
+  abstract and symbolic way to represent and simulate
   computations. However, manual manipulation of such graphs is slow
-  and error prone. We present a formalism that supports mechanised
-  reasoning about such graphs. This involves a compositional account
-  of graph rewriting that preserves the underlying categorical
-  semantics. We present a generic system with a fixed logical kernel
-  for representing compact closed categories. This provides a
-  declarative account which supports derived rules. We illustrate the
-  framework by instantiating it for the graphical language of quantum
-  computation.
+  and error prone. We present a formalism, based on compact closed
+  categories, that supports mechanised reasoning about such
+  graphs. This gives a compositional account of graph rewriting that
+  preserves the underlying categorical semantics. Using this
+  representation, we describe a generic system with a fixed logical
+  kernel that supports reasoning about models of compact closed
+  category. A salient feature of the system is that it provides a
+  declarative account of derived results. We illustrate the framework
+  by instantiating it for a graphical language of quantum computation
+  and show how this can be used to perform symbolic computation.
 
@@
     interactive theorem proving, graphical calculi; concerning:
-    symbolic computation, automated reasoning, formal mathematics,
-    logic}
+    symbolic computation, automated reasoning, formal mathematics}
 \end{abstract}
@@
 The major problem with these graphical representations is the lack of
-machinery for automating the manipulation of graphs.  Existing
-approaches to graph transformation have a different underlying
-semantics corresponding to graphs in general rather than those which
-describe compact closed categories. The result is that they provide an
-unsound concept of rewriting for the kind of graphs we are interested
-in.
-
-This paper reviews a formalism for graphical representations of
+machinery for automating their manipulation. Existing approaches to
+graph transformation have a different underlying semantics
+corresponding to graphs in general rather than those which describe
+compact closed categories. The result is that they provide an unsound
+concept of rewriting for the kind of graphs we are interested in.
+
+The main contribution of this paper is a graph-vased formalism that is
+suitable for representing and evaluating quantum computations. We
+start by reviewing a formalism for graphical representations of
 compact closed categories. We also revisit a model of quantum
@@
 calculus in two significant ways driven by the need to express rules
-that could not be accounted for in the initial formalism. We provide a
-semantics for these in terms of the initial representation. By
-suitably combining these extensions to graphs, we provide a
-representation that captures an interesting and useful set of graph
-patterns. We use this as the foundation for a simple logical framework
-for manipulating compact closed categories as graphs. This has a
-suitable rewriting mechanism where the axioms of the underlying
-object-formalism are expressed as equations between graphs. We then
-present a short case study that illustrates the framework by
-instantiating it for the introduced model of quantum computation.
-This shows how the framework can be used to symbolically perform
-simplifications to quantum programs as well as simulate quantum
+that could not be accounted for in the initial formalism. By combining
+these extensions, we provide a representation that captures an
+interesting and useful set of graph patterns. Notably, it can express
+the Spider Theorem, which is normally written using informal elipses
+notation. We provide a semantics for our graph-based formalism in
+terms of the initial representation of compact closed categories.
+
+Using our graph-based formalism as the representational foundation, we
+develop a simple logical framework for manipulating models of compact
+closed categories. This has a suitable rewriting mechanism where the
+axioms of the underlying object-formalism are expressed as equations
+between graphs. We then present a short case study that illustrates
+the framework by instantiating it for the introduced model of quantum
+computation.  This shows how the framework can be used to symbolically
+perform simplifications of quantum programs as well as simulate
 computations.
@@
 the morphism: an edge connected to a node gets mapped to a new edge
-that must connected in the same way to the mapped node.
+that must be connected, in the same way, to the mapped node.
 
@@
 \begin{description}
-\item[Objects] the self dual object $Q$;
-\item[Arrows] Two families of arrows:
+\item[Objects:] the self dual object $Q$;
+\item[Arrows:] two families of arrows $X$ and $Z$:
   \begin{align*}
@@
 number of repetitions of a subgraph need to be matched. To support
-this we introduce an operation, !-boxing (pronounced band-boxing), on
+this we introduce an operation, !-boxing (pronounced bang-boxing), on
 graph representations. Given a graph representation, this introduces a
@@
 nodes outside the !-box. The !-boxes can also be instantiated with
-zero copies. This erases all edges to and from the !-box.
+zero copies. This erases all edges to and from the !-box.  The Spider
+Theorem in Figure~\ref{fig:informal-spider-law} for instance is
+represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.
 
@@
 
-\subsection{!-Matching.}
+\subsubsection{Matching.}
 
@@
 
-\subsection{!-Box Semantics.}
+\subsubsection{Semantics.}
 
@@
 
-\subsection{Matching}
-
-The purpose of graph patterns is to formally express, in a declarative
-manner, the intuitive idea of graph matching which commonly used in
-quantum computation. To do this, we combine the two extensions to the
-concrete graph representation, namely variable nodes and !-boxes, to
-provides a matching relation between graph patterns and concrete
-graphs as well as directly between graph patterns. The former relation
-gives a semantics for our system and ensures soundness, while the
-latter supports deirved equations and thus provides the theoretical
-basis for our formulation of rewriting.
-
-\subsubsection{Concrete Matches.}
+% \subsubsection{Matching.}
+
+% The purpose of graph patterns is to formally express, in a declarative
+% manner, the intuitive idea of graph matching which commonly used in
+% quantum computation. To do this, we combine the two extensions to the
+% concrete graph representation, namely variable nodes and !-boxes, to
+% provides a matching relation between graph patterns and concrete
+% graphs as well as directly between graph patterns. The former relation
+% gives a semantics for our system and ensures soundness, while the
+% latter supports deirved equations and thus provides the theoretical
+% basis for our formulation of rewriting.
+
+\subsubsection{Semantics as Concrete Matches.}
 
 The semantics for a graph pattern $G$ is a set of concrete graphs
-denoted by $\minterp{G}$. We say that members of this set are
-\emph{matched} by $G$. Formally, the concrete set of matches are
-defined as:
+denoted by $\minterp{G}$ and define it as:
 
@@
 
+\noindent this simply considers every interpretation of the !-boxes to
+give variable-node graphs for which we then appeal to their own
+semantics.
+
 \subsubsection{Graph Pattern Matches.}
 
-The specification for one pattern matching another one is simply that
-it is more general. Formally, a graph pattern, $G$, matches another
-graph pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This is
-the specification for a matching algorithm which we implement of
-in~\S\ref{sec:rewriting}. We can use the definition of open embedding
-from variable-node graphs as an intermediate mechanism and show that a
-suitable definition of graph pattern matching is simply the !-Box
-matching followed by variable-node matching:
-
-\begin{definition}
-  $G \leq H \Leftrightarrow \forall H' \in \binterp{H}.\,\exists G' \in \binterp{G}.\, G' \leq_v H'$
-\end{definition}
-
-Because variable-node matching preserves the semantics, it is trivial to
-prove that this definition meets the specification.
+The specification for one pattern, $G$, to match another one, $H$ is
+that it is more general with respect to the interpretation:
+$\minterp{G} \supseteq \minterp{H}$. However, graph patterns can
+correspond to a countably infinite number of concrete graphs. Thus
+matching between graph pattern cannot be implemented by simply
+unfolding them to their interpretated concrete graphs and checking the
+membership relation.
+
+*** add matching algo outline ***
+
+%We can use the definition of open embedding from variable-node graphs
+%as an intermediate mechanism and show that a suitable definition of
+%graph pattern matching is simply the !-Box matching followed by
+%variable-node matching:
+%
+%\begin{definition}
+%  $G \leq H \Leftrightarrow \forall H' \in \binterp{H}.\,\exists G' \in \binterp{G}.\, G' \leq_v H'$
+%\end{definition}
+%Because variable-node matching preserves the semantics, it is trivial to
+%prove that this definition meets the specification.
 
@@
 compact closed categories. Following the terminology of logical
-frameworks such as Isabelle~\cite{isabelle}, we call the specification
-provided by the underlying model an \emph{object-level} graph
-formalism. In particular, an object-level formalisation defines a set
-of rules which are treated as the axioms for the system. It also
-defines the data at the nodes and edges as well as corresponding
-data-matching behaviour. For it's part, the meta-level provides
-machinery to manipulate graphs and derive new rules. In this section
-we describe the meta-level framework and prove its adequacy for
-rewriting.
+frameworks such as Isabelle~\cite{isabelle}, we call the
+specification, provided by the underlying model, an
+\emph{object-level} graph formalism. An object-level formalisation
+defines a set of rules which are treated as the axioms for the
+system. It also defines the data at the nodes and edges as well as
+corresponding data-matching behaviour. For it's part, the meta-level
+provides generic machinery to manipulate graphs and derive new
+rules. In this section we describe the meta-level framework and prove
+its adequacy for rewriting.
 
@@
 
-\subsubsection{Adequacy.} An equation on graph patterns corresponds to
+\subsection{Lifting}
+
+*** lifting concrete laws to graph-pattern laws ***
+
+\subsection{Adequacy}
+
+An equation on graph patterns corresponds to
 a possibly infinite set of equations betweens concrete graphs. We
@@
 \begin{center}
-\prooftree
-A = B \in \Gamma
-\justifies
-\Gamma \vdash A = B
-\using\mbox{trivial}
-\endprooftree
-\quad\quad
+%% not needed, can be derived by refl + subst
+%\prooftree
+%A = B \in \Gamma
+%\justifies
+%\Gamma \vdash A = B
+%\using\mbox{trivial}
+%\endprooftree
+%\quad\quad
 \prooftree
@@
 \Gamma \vdash A = A
-\using\mbox{reflexivity}
+\using\mbox{refl}
 \endprooftree 
-\end{center}
-
-\begin{center}
+\quad\quad
 \prooftree
@@
 \Gamma \vdash B = A
-\using\mbox{symmetry}
+\using\mbox{sym}
 \endprooftree
@@
 \Gamma \vdash (C = D[A/B])
-\using\mbox{substitution}
+\using\mbox{subst}
 \endprooftree
@@
 
-This system allows derived rules which allow a combination of steps to
-be abbreviated by representing them as a single new rule.
+A new graph can be introduced using the reflexivity law. This can be
+the left hand side of a conjecture, or a graph that needs to be
+computed. By applying the \emph{subst} rule, intermediate results are
+derived which can themselves be used to rewrite other rules and
+conjetcures. In this way, the system allows derived rules to provide
+an abbreviation for a combination of steps.
 
@@
 
-\section{Implementation}
-\label{ssec:implementation}
+%\subsection{Implementation}
+%\label{ssec:implementation}
 
@@
 \subsection{Node Expressions}
-\label{ssec:node-expressions}
-
-\subsection{}
 \label{ssec:node-expressions}

### 2008-03-06 papers/reasoning-graphically/paper.tex
work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1403 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \begin{description}
-\item[Objects:] the self dual object $Q$;
-\item[Arrows:] two families of arrows $X$ and $Z$:
+\item[Objects:] are written $Q$ and are self dual;
+\item[Arrows:] there are families of arrows $X$ and $Z$:
   \begin{align*}
@@
 $$\begin{array}{ccc}
-\inlinegraphic{8em}{spider_lhs_patt} & \;=\; & \inlinegraphic{6em}{spider_rhs_patt}
+\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
 \end{array}$$
-\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using graph patterns.}
+\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using
+  graph patterns. The !-boxes are named $i$ and $j$. The variable
+  nodes are white and named $a$ and $b$. The internal node's data,
+  which is an angle, is written inside the node.}
 \end{figure}
@@
 
-*** add matching algo outline ***
+Fortunately, it is quite easy to provide a decidable matching. This is
+because the size of the unfolding that needs to be considered can be
+bounded. The key observation is that a graph $G_1$ will never match a
+graph with less non-variable nodes. Thus unfolding of $G_1$ can be
+bounded by the number of non-variable nodes in $G_2$. While this gives
+a generate and test style algorithm, it is not efficient. The
+intuition for an efficient algorithm is to search through one graph
+incrementally increasing the matched part and avoids symmetries in
+matching. The development of an efficient algotihm is ongoing work.
+
+%non-variable nodes than another graph $G_2$
+%
+%can algorithm as
+%
+%*** add matching algo outline ***
 
@@
 
-\subsection{Lifting}
-
-*** lifting concrete laws to graph-pattern laws ***
-
-\subsection{Adequacy}
-
-An equation on graph patterns corresponds to
-a possibly infinite set of equations betweens concrete graphs. We
-might expect that such equations express too much: they may allow
-rewrites which are not intended. We prove \emph{adequacy}, which
-states that this is not the case. In particular, given an equation
-between concrete graphs, $G = H$, the equations between the lifting of
-these to variable node graphs by replacing boundary nodes with
-variable nodes is a conservative extension. The proof follows directly
-from the obervation that every instance of the lifted variable node
-equation corresponds to compsition of $G$ and $H$ with some other
-graph (the non-embedded part).
+\subsection{Lifting Axioms and Adequacy}
+
+The axioms of an object system typically come from the semantics of
+the unerlying system. For instance, the equations given in
+Figure~\ref{fig:graph-quant-eqns} provide a system of reasoning about
+quantum computation. In some cases, such equations can be provdied by
+the underltying semantics. When such rules are expressed as graph
+patterns, we replace the concrete representations boundary nodes with
+variable nodes. This operation is called \emph{lifting}. An equation
+on graph patterns corresponds to a possibly infinite set of equations
+betweens concrete graphs. Thus we might worry that the lifted
+equations express too much: they may allow rewrites which are not
+intended.
+
+We call the property that the lifted representation is a conservative
+extension of the initial theory \emph{adequacy}. For models of compact
+closed categories, the proof of adequacy is quite simple: given an
+equation between concrete graphs, $G = H$, we observe that every
+instance of the lifted equation corresponds to the original equations
+composed with some graph. The graph is given by the unmapped subgraph
+using the open embedding from the lifted equations onto their
+instance. Thus if $G = H$ is true, then so is every instance of it's
+lifting, and thus the lifting produces an adequate representation.
 
@@
 \prooftree
+%\mathit{valid}(A)
 \justifies
@@
 
-\noindent where $\Gamma$ is the set of object-level axioms. For the
-reflexivity rule, $A$ must be a well-formed pattern graph.
-
-A new graph can be introduced using the reflexivity law. This can be
-the left hand side of a conjecture, or a graph that needs to be
-computed. By applying the \emph{subst} rule, intermediate results are
-derived which can themselves be used to rewrite other rules and
-conjetcures. In this way, the system allows derived rules to provide
-an abbreviation for a combination of steps.
+\noindent where $\Gamma$ is the set of object-level axioms and for the
+reflexivity rule, $A$ must be a well-formed pattern graph. We assume
+that axioms in $\Gamma$ meet the validity conditions described
+earlier. These rules all preserve the validity conditions on equations
+and thus the system as a whole ensures only valid rules are derived.
+
+A new graph can be introduced using the reflexivity law (\emph{refl}).
+By applying the \emph{subst} rule, intermediate results are derived
+which can themselves be used to rewrite other rules and conjetcures.
+In this way, the system allows derived rules to provide an
+abbreviation for a combination of steps. Sets of rules can be applied
+automatically to simulate computation in an the object domain. In
+\S\ref{sec:case-study} we illustrate this for quantum computation.
 
@@
 
-% One of the main applications of
-
 % An object-level graph system is sound if their is some pair of graphs
@@
 \label{sec:case-study}
+
+The model of quantum computation introduced in
+\S\ref{sec:quotients-rewriting} provides an object formalism for
+logical system. The equations from Figure~\ref{fig:graph-quant-eqns}
+are lifted to graph patterns and the spider rule is encoded as shown
+in Figure~\ref{fig:spider-thm-patt}. Although a complete graph-based
+account of the underlying quantum model is ongoing work, these rules
+provide an effective and useful initial set.
+
+As well as these equations, recall that the object level formalism
+must define the data held at the nodes and edges as well as provide a
+suitable macthing algorithm for these. Our model of quantum
+computation provides no data for the edges. The nodes, on the other
+hand, are either a $H$ with no additional data (a Hadamard gate), or
+have an \emph{angle} and are either the $X$ or $Z$ transforms. For
+their part, angles are expressed as rational numbers which correspond
+to the coefficient of $\pi$ in the corresponding matrix.
+
+The result is that we give the $X$ and $Z$ nodes an \emph{angle
+  expression}. When a node is within a !-box, this is a variable which
+gets instantiated to a new variable in each of the unfoldings of the
+!-box. When a node is not within a !-box, angle-expressions are a
+mapping from a set of angle-variables to the corresponding rational
+cooefficient. When an angle-expression contains a variable within a
+!-box, this is interpreted as a sum of the variables that result from
+it's unfolding. This rather simple expression language has a simple
+normal form by lexographically ordering the angle-variable. Matching
+then results in variables being instantiated and all affected nodes
+normalising the result of the instantiation. The substitution rule
+also has to make sure that a angle-variables in the rule being applied
+are distinct from those in the expression being rewritten.
+
 %The model of qantum computation in which we are interested is finite
@@
 
-\subsection{Node Expressions}
-\label{ssec:node-expressions}
+%\subsection{Node Expressions}
+%\label{ssec:cs:node-expressions}
+
+\subsection{Computing the Quantum Fourier Transform}
+\label{ssec:cs:quantum-transform}
+
+The quantum fourier transform is among the most important quantum
+algorithms, forming an essential part of Shor's
+algorithm~\cite{Shor:PolyTimeFact:1997}.
+
+\begin{figure}[t]
+$$\begin{array}{cccc}
+  & \inlinegraphic{2cm}{qft2} & = & \inlinegraphic{2cm}{qft3} \\ 
+ & & & \\
+= & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10}
+\end{array}$$
+\label{fig:quantum-transform}\caption{An example computation of the
+  Quantum Fourier Transform, performed symbollically by rewriting.}
+\end{figure}
 

### 2008-03-07 papers/reasoning-graphically/paper.tex
work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1405 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\begin{definition}
-  \emph{copy}: copies a !-box, $b$ in a graph to produce a new graph
+\begin{description}
+  \item[copy]: copies a !-box, $b$ in a graph to produce a new graph
   with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
@@
   is a new edge from $m$ to the new copy of $n$ in $b'$.
-\end{definition}
-
-\begin{definition}
-  \emph{drop} simply removes the !-box, but leaves its contents in the
+
+\item[drop]: simply removes the !-box, but leaves its contents in the
   graph.
-\end{definition}
-
-\begin{definition}
-  \emph{kill} removes from the graph all nodes in the !-box as well as
+
+\item[kill]: removes from the graph all nodes in the !-box as well as
   any incident edges.
-\end{definition}
-
-\begin{definition}
-  \emph{merge} combines two !-boxes, $B_1$ and $B_2$ into a single
+
+\item[merge]: combines two !-boxes, $B_1$ and $B_2$ into a single
   larger !-box $B_1 \cup B_2$.
-\end{definition}
+\end{description}
 
@@
 
-As well as these equations, recall that the object level formalism
-must define the data held at the nodes and edges as well as provide a
-suitable macthing algorithm for these. Our model of quantum
-computation provides no data for the edges. The nodes, on the other
-hand, are either a $H$ with no additional data (a Hadamard gate), or
-have an \emph{angle} and are either the $X$ or $Z$ transforms. For
-their part, angles are expressed as rational numbers which correspond
-to the coefficient of $\pi$ in the corresponding matrix.
+Our model of quantum computation requires no data for the edges. The
+nodes, on the other hand, are either a $H$ with no additional data (a
+Hadamard gate), or have an \emph{angle} and are either the $X$ or $Z$
+transforms. For their part, angles are expressed as rational numbers
+which correspond to the coefficient of $\pi$ in the corresponding
+matrix.
 
@@
 normal form by lexographically ordering the angle-variable. Matching
-then results in variables being instantiated and all affected nodes
-normalising the result of the instantiation. The substitution rule
-also has to make sure that a angle-variables in the rule being applied
-are distinct from those in the expression being rewritten.
+then results in variables being instantiated and the values of all
+affected nodes are then normalised. An additional implementation
+detail must also be observed for the substitution rule: it must ensure
+that angle-variables in the rule being applied are distinct from those
+in the expression being rewritten.
 
@@
 
-\subsection{Computing the Quantum Fourier Transform}
+\subsubsection{Computing the Quantum Fourier Transform.}
 \label{ssec:cs:quantum-transform}
@@
 algorithms, forming an essential part of Shor's
-algorithm~\cite{Shor:PolyTimeFact:1997}.
+algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
+polynomial time algorithm for factoring. In our graph patten calculus
+this ciruit becomes the top-left graph in
+Figure~\ref{fig:quantum-transform}. This figure shows how computation
+can be symbolically performed by rewriting with the lifted equations
+from Figure~\ref{fig:graph-quant-eqns} with the graph pattern version
+of the Spider Theorem.
 
 \begin{figure}[t]
-$$\begin{array}{cccc}
+\begin{tabular}{cccc}
+\hline
+ & & & \\
   & \inlinegraphic{2cm}{qft2} & = & \inlinegraphic{2cm}{qft3} \\ 
  & & & \\
-= & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10}
-\end{array}$$
+\hline
+ & & & \\
+= & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10} \\
+ & & & \\
+\hline
+\end{tabular}
 \label{fig:quantum-transform}\caption{An example computation of the
-  Quantum Fourier Transform, performed symbollically by rewriting.}
+  Quantum Fourier Transform with inputs 1 and 0, performed
+  symbollically by rewriting. }
 \end{figure}

### 2008-03-07 papers/reasoning-graphically/paper.tex
added related work and citations.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1406 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \end{array}$$
-\label{fig:informal-spider-law}\caption{An informal equation on graphs
+
+\label{fig:spider}\caption{An informal equation on graphs
   that expresses the Spider Theorem. }
@@
 this can be depicted graphically as the equation in
-Figure~\ref{fig:informal-spider-law}. Conversely, a spider may be
+Figure~\ref{fig:spider}. Conversely, a spider may be
 arbitrarily divided into subspiders, provided the total in- and
@@
 zero copies. This erases all edges to and from the !-box.  The Spider
-Theorem in Figure~\ref{fig:informal-spider-law} for instance is
+Theorem in Figure~\ref{fig:spider} for instance is
 represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.
@@
 
-Initially one might be tempted to think that the usual notion of
-subgraph might surfice for expressing matching, however it allows
-additional edges on nodes where, because of the tensorial semantics of
-our graphs, we do not. Our work provides a restriction of the usual
-subgraph definition that is sound for Compact Closed Categories.
-
-Link Graphs and their extention to BiGraphs also encapsualte a very
-different kind of semantics where they allow edges to go to multiple
-nodes. An interesting area of further work would be to consider our
-graph patterns for these, and other, graph based formalisms.
-
-Graph Grammars...
-
-Other graph rewriting...
-
-Matrix Based Graph Tansformation...
-
-Other Graph Transformation...
-
-Interaction nets and interaction combinators (Lafont and others)
+There are several foundational approaches to graph transformation,
+including algebraic appoaches~\cite{corradini97algebraic}, node-label
+controlled~\cite{Graphgrammars83}, matrix
+based~\cite{DBLP:conf/gg/VelascoL06}, and programmed graph
+replacement~\cite{progress97}. These provide general ways of
+interpretating graph transformations which can then be implemented to
+provide machinery for a specific application.  However, systems based
+on these theories do not provide machinery for the semantics of
+compact closed categories. Thus their notion of matching and
+replacement do not guarentee well-typed results. Furthermore, they do
+not provide machinery to for rewriting of graphs with elipses
+notation.
+
+The distinctive feature of our form of graph rewriting is that the
+graphs capture the structural properties of compact closed categories.
+In this respect, our graphs caputure the structural properties of
+compact closed categories and rewriting is compositional: it preserves
+the type of the rewritten subgraph. However, the various components of
+our system can be seen as instantiations for in a general graph
+rewriting system: our matching provides the embedding information and
+the node-data matching of the object formalism provides application
+conditions and the attribute transfer function.
+
+% Graph Grammars...
+% Other graph rewriting... 
+% Matrix Based Graph Tansformation...
+% Other Graph Transformation...
+% Interaction nets and interaction combinators (Lafont and others)
 

### 2008-03-07 papers/reasoning-graphically/paper.tex
added conclusion.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1407 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 There are several foundational approaches to graph transformation,
-including algebraic appoaches~\cite{corradini97algebraic}, node-label
+including algebraic approaches~\cite{corradini97algebraic}, node-label
 controlled~\cite{Graphgrammars83}, matrix
@@
 replacement~\cite{progress97}. These provide general ways of
-interpretating graph transformations which can then be implemented to
+interpretation graph transformations which can then be implemented to
 provide machinery for a specific application.  However, systems based
@@
 compact closed categories. Thus their notion of matching and
-replacement do not guarentee well-typed results. Furthermore, they do
-not provide machinery to for rewriting of graphs with elipses
+replacement do not guarantee well-typed results. Furthermore, they do
+not provide machinery to for rewriting of graphs with ellipses
 notation.
@@
 graphs capture the structural properties of compact closed categories.
-In this respect, our graphs caputure the structural properties of
+In this respect, our graphs capture the structural properties of
 compact closed categories and rewriting is compositional: it preserves
@@
 
-\section{Conclusions}
+\section{Conclusions and Further Work}
 \label{sec:conclusions}
 
-Ideas for further work
-\begin{itemize}
-\item simplification ordering
-\item confluence arguments for graphs
-\item formalise algorithms in a thm prover
-\item richer graph structures - quantification over edges
-\item richer expression language in vertices
-\item completeness of rewrite rules with respect to wpHilb
-\end{itemize}
+We have introduced a representation for graphs which can formally
+characterise the ellipses notation used informally to represent certain
+infinite families of graph rewrites, such as the Spider Theorem. This
+representation provides the foundation for a simple meta-logic for
+reasoning about models of compact closed categories. We illustrated
+this by providing an account of quantum computation and showing how
+computation can be performed. Having developed the basic
+representational machinery and shown matching to be decidable, we are
+left with several exciting avenues for further research. The most
+immediate is to provide a full implementation - a partial one is
+already
+available~\footnote{\url{http://dream.inf.ed.ac.uk/projects/quantomatic}}.
+Other areas of further work include considering confluence results for
+sets of rewrite rules, even richer representations for graph-patterns
+and completeness of the set of the rewrite rules for quantum
+computation.
+
+% simplification ordering
+% \item confluence arguments for graphs
+% \item formalise algorithms in a thm prover
+% \item richer graph structures - quantification over edges
+% \item richer expression language in vertices
+% \item completeness of rewrite rules with respect to wpHilb
+% \end{itemize}
 

### 2008-03-07 papers/reasoning-graphically/paper.tex
close to final edit - spell check 1
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1408 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\newcommand{\vinterp}[1]{\lfloor\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rfloor}
-\newcommand{\binterp}[1]{\lceil\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rceil}
+%\newcommand{\vinterp}[1]{\lfloor\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rfloor}
+%\newcommand{\binterp}[1]{\lceil\hspace{-0.27em}|\, #1\, |\hspace{-0.27em}\rceil}
+\newcommand{\vinterp}[1]{\llbracket #1 \rrbracket_v}
+\newcommand{\binterp}[1]{\llbracket #1 \rrbracket_!}
 \newcommand{\minterp}[1]{\llbracket #1 \rrbracket}
@@
   Recent graph-based formalisms of quantum computation provide an
-  abstract and symbolic way to represent and simulate
-  computations. However, manual manipulation of such graphs is slow
-  and error prone. We present a formalism, based on compact closed
-  categories, that supports mechanised reasoning about such
-  graphs. This gives a compositional account of graph rewriting that
-  preserves the underlying categorical semantics. Using this
-  representation, we describe a generic system with a fixed logical
-  kernel that supports reasoning about models of compact closed
-  category. A salient feature of the system is that it provides a
-  declarative account of derived results. We illustrate the framework
-  by instantiating it for a graphical language of quantum computation
-  and show how this can be used to perform symbolic computation.
+  abstract and symbolic way to represent and simulate computations.
+  However, manual manipulation of such graphs is slow and error prone.
+  We present a formalism, based on compact closed categories, that
+  supports mechanised reasoning about such graphs. This gives a
+  compositional account of graph rewriting that preserves the
+  underlying categorical semantics. Using this representation, we
+  describe a generic system with a fixed logical kernel that supports
+  reasoning about models of compact closed category. A salient feature
+  of the system is that it provides a formal and declarative account
+  of derived results that can include `ellipses'-style notation. We
+  illustrate the framework by instantiating it for a graphical
+  language of quantum computation and show how this can be used to
+  perform symbolic computation.
 
@@
 itself~\cite{AbrCoe:CatSemQuant:2004,Selinger:dagger:2005,Coecke2005Kindergarten-Qu,Coecke2006POVMs-and-Naima,Coecke2006Quantum-Measure}.
-These techniques have a number of advantages over the convential
+These techniques have a number of advantages over the conventional
 matrix-based approach to quantum mechanics:
@@
 \item The visual representation abstracts over the values in the
-  matricies. This removes detail that requires a lot of work for a
+  matrices. This removes detail that requires a lot of work for a
   human to interpret.
@@
 \item Many properties have a natural graphical representation. For
-  example, non entanglement is can be infered from disjoint subgraphs.
+  example, non entanglement is can be inferred from disjoint subgraphs.
 
@@
   vector spaces. In particular, it provides a representation for
-  compact closed cartegories.
+  compact closed categories.
 
@@
 
-The main contribution of this paper is a graph-vased formalism that is
+The main contribution of this paper is a graph-based formalism that is
 suitable for representing and evaluating quantum computations. We
@@
 interesting and useful set of graph patterns. Notably, it can express
-the Spider Theorem, which is normally written using informal elipses
-notation. We provide a semantics for our graph-based formalism in
-terms of the initial representation of compact closed categories.
+the Spider Theorem which is normally written using informal ellipses
+notation (see Figure~\ref{fig:spider}). We provide a semantics for our
+graph-based formalism in terms of the initial representation of
+compact closed categories.
 
@@
 $G$ determined by $V'$ should be preserved exactly by $f$; it should
-contain $G$ without any extra incident egdes into the verticies $V'$
+contain $G$ without any extra incident edges into the vertices $V'$
 whereas other vertices may be identified and may contain additional
-verticies.
+vertices.
 
@@
 $r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
-$r_E$ to its preimage satifies equations \eqref{eq:graph-hom1} and
+$r_E$ to its preimage satisfies equations \eqref{eq:graph-hom1} and
 \eqref{eq:graph-hom2} above.
@@
 \begin{gather*}
-  d_A : I \to A^* \otimes A\\
-  e_A : A \otimes A^* \to A
+  d_A : I \to A^* \otimes A \quad\quad\quad e_A : A \otimes A^* \to A
 \end{gather*}
@@
 \end{proposition}
-For each ordered signed set $A$, the identity map for $A$ has
-$\dom \id{A} = A^*$ and $\cod \id{A} = A$; its underlying
-graph has $E = A$ and $V = A^* + A$ with $t(a) = a$ and $s(a) =
-a^*$.  Given a pair of concrete graphs $f:A\to B$ and $g:B\to C$ their
-composition $g\circ f:A\to C$ is constructed by merging the two
-graphs, erasing the vertices of $\cod f$ and $\dom g$, and identifying
-the edges previously incident at the deleted vertices.  (Due to the
-opposite polarity of the domain and codomain the edges have compatible
-direction.)  The tensor product on objects $A,B$ is simply $A+B$;
-given $f:A\to B$, $g: C\to D$, the graph of $f \otimes g$ is the
-disjoint union of the graphs of $f$ and $g$.  The unit for the tensor
-is the empty set.  The morphisms $d_A : I \to A^* \otimes A$,
-$e_: A \otimes A^* \to I$ have the same underlying graph, but $\dom d =
-\emptyset$, $\cod d = A^*+A$, $\dom e = A+A^*$ and $\cod e = \emptyset$.
+For each ordered signed set $A$, the identity map for $A$ has $\dom
+\id{A} = A^*$ and $\cod \id{A} = A$; its underlying graph has $E = A$
+and $V = A^* + A$ with $t(a) = a$ and $s(a) = a^*$.  Given a pair of
+concrete graphs $f:A\to B$ and $g:B\to C$ their composition $g\circ
+f:A\to C$ is constructed by merging the two graphs, erasing the
+vertices of $\cod f$ and $\dom g$ (called the \emph{boundary
+  vertices}), and identifying the edges previously incident at the
+deleted vertices.  (Due to the opposite polarity of the domain and
+codomain the edges have compatible direction.)  The tensor product on
+objects $A,B$ is simply $A+B$; given $f:A\to B$, $g: C\to D$, the
+graph of $f \otimes g$ is the disjoint union of the graphs of $f$ and
+$g$.  The unit for the tensor is the empty set.  The morphisms $d_A :
+I \to A^* \otimes A$, $e_: A \otimes A^* \to I$ have the same
+underlying graph, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
+= A+A^*$ and $\cod e = \emptyset$.
 
@@
 
-As a final remark before moving on, note that the external structre of vertex
-in a concrete graph is esentially the same as that of a complete
-graph;  hence one can consistently view subgraphs as vertices, and
-abstract over the their internal structure.
+As a final remark before moving on, note that the external structure
+of a vertex in a concrete graph is essentially the same as that of a
+complete graph; hence one can consistently view subgraphs as vertices,
+and abstract over the their internal structure.
 
@@
     equation we have a second equation obtained by flipping the
-    diagram upside down.  In additon, we have a ``colour duality'':
+    diagram upside down.  In addition, we have a ``colour duality'':
     each equation shown here gives rise to second, which is obtained
@@
 \inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
+\end{array}\qquad
+\begin{array}{ccc}
+\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
 \end{array}$$
-
-\label{fig:spider}\caption{An informal equation on graphs
-  that expresses the Spider Theorem. }
+\label{fig:spider}\caption{ [Left] An informal equation on graphs that
+  expresses the Spider Theorem.  [Right] The Spider Theorem expressed
+  formally using graph patterns.  The !-boxes are named $i$ and $j$.
+  The variable nodes are white and named $a$ and $b$. The non-variable
+  node data (the angle) is written inside the node.}
+
 \end{figure}
@@
 Figure~\ref{fig:spider}. Conversely, a spider may be
-arbitrarily divided into subspiders, provided the total in- and
+arbitrarily divided into sub-spiders, provided the total in- and
 out-degree is preserved, along with and the sum of the $\alpha $s.
@@
 \begin{center}
-\inlinegraphic{2.5em}{compact1} \quad\quad
-\inlinegraphic{2.5em}{compact3} \quad\quad
-\inlinegraphic{2.5em}{compact5}
+\inlinegraphic{2.5em}{compact1} $=$
+\inlinegraphic{1.3em}{compact2} $\qquad\quad$
+\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
+$\qquad\quad$ \inlinegraphic{2.5em}{compact5} $=$
+\inlinegraphic{2.5em}{comonoid-unit4}
 \end{center}
+
+%\begin{center}
+%\inlinegraphic{2.5em}{compact1} \quad\quad
+%\inlinegraphic{2.5em}{compact3} \quad\quad
+%\inlinegraphic{2.5em}{compact5}
+%\end{center}
 
@@
 an equation involving one of the above to easily lead to a derivation
-corresponding to either of the other two. For instance, proving any of
-the following allows a trivial derivation of the others:
-
-\begin{center}
-\inlinegraphic{2.5em}{compact1} $=$
-\inlinegraphic{1.3em}{compact2} $\Leftrightarrow$
-\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
-$\Leftrightarrow$ \inlinegraphic{2.5em}{compact5} $=$
-\inlinegraphic{2.5em}{comonoid-unit4}
-\end{center}
-
-\noindent This motivates a characterisation that has only one
-representative for such a set of equations. More presicely, we
-formalise a representation that abstracts over the boundary nodes
-membership in the domain or co-domain. This gives rise to a
-\emph{variable-node graphs}, in which boundary nodes have been
+corresponding to either of the other two: proving any of the above
+allows a trivial derivation of the others.
+
+To address this, we formalise a representation that abstracts over the
+boundary nodes membership in the domain or co-domain. This gives rise
+to a \emph{variable-node graphs}, in which boundary nodes have been
 generalised to \emph{variable nodes}. The intuition of variable nodes
 is that they can replaced by concrete nodes in some graph in a process
-analgous to composition.
+analogous to composition.
 
@@
 matching, which captures the intuitive idea of a graph with variable
-nodes occuring within another graph.
+nodes occurring within another graph:
 
@@
 A graph with variable nodes, $G$, can be given a formal semantics by
-being interpretated as a set of concrete graphs, denoted by
+being interpreted as a set of concrete graphs, denoted by
 $\vinterp{G}$. The interpretation is simply the set of concrete graphs
-which the variable node graph matches. We refer to members of this set
-as instances. Matching is also well-defined between two variable-node
-graphs, in which case we denote that $G$ macthes $H$ by $G \leq_v H$.
+which the variable node graph matches. The above definition of
+matching also applies between two variable-node graphs, in which case
+we denote that $G$ matches $H$ by $G \leq_v H$.
 
@@
 $G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. The
-proof from left to right is a consequence of the trivial lemma that a
-compositon of open embeddings is an open embedding. This makes the
-embedding of $G$ into $H$ composed with $H$ into $\vinterp{H}$ an open
-embedding of $G$ into $\vinterp{H}$. From right to left, we compose
-the open embedding $\vinterp{G}$, restricted to it's subset
-$\vinterp{H}$, with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
+proof from left to right is a consequence of the lemma that a
+composition of open embeddings is an open embedding. From left to
+right, the embedding of $G \leq_v H$ composed with $\vinterp{H}$ is
+thus an open embedding of $G$ into $\vinterp{H}$. Hence $\vinterp{G}
+\supseteq \vinterp{H}$. From right to left, we compose the open
+embedding $\vinterp{G}$, restricted to it's subset $\vinterp{H}$, with
+the inverse of $\vinterp{H}$ to get $G \leq_v H$.
 \end{theorem}
@@
 \end{theorem}
+
+These theorems ensure that macthing is what we expect it to be and
+also give us the usual properties such as associativity.
 
@@
 
-A crucial feature of rules like the Spider Theorem is that an arbirary
+A crucial feature of rules like the Spider Theorem is that an arbitrary
 number of repetitions of a subgraph need to be matched. To support
@@
 in a !-box. Intuitively, the resulting !-boxed graph can be thought of
-as representing a set of graphs with an arbirary number of copies of
+as representing a set of graphs with an arbitrary number of copies of
 the !-boxed nodes, where every copy connects, in the same way, to the
 nodes outside the !-box. The !-boxes can also be instantiated with
-zero copies. This erases all edges to and from the !-box.  The Spider
-Theorem in Figure~\ref{fig:spider} for instance is
-represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.
+zero copies. This erases all edges to and from the !-box.
+
+% The Spider
+%Theorem in Figure~\ref{fig:spider} for instance is
+%represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.
 
@@
 
-More formaly, a !-box graph is a graph paired with a set !-boxes. Each
+More formally, a !-box graph is a graph paired with a set !-boxes. Each
 of these !-boxes is the set of nodes inside it and each !-box must be
@@
   slightly more complex, although also more expressive, semantics for
-  overlaping node sets in the !-boxes. While such expressivity is
+  overlapping node sets in the !-boxes. While such expressivity is
   interesting from a representational point of view, it is not needed
@@
 
-\subsubsection{Matching.}
-
-To formalise the intuitive notion that a !-box represents arbitrary
-number of copies of the subgraph made from it's nodes, we introduce
-\emph{!-box matching}. This is a binary relation on graphs, written
-infix as $G \leq_! H$ when $G$ matches $H$. The set matching !-box
-graphs is defined as the set closed by the following operations on the
-!-boxes in a graph:
+\textbf{Matching.} To formalise the intuitive notion that a !-box
+represents arbitrary number of copies of the subgraph made from it's
+nodes, we introduce \emph{!-box matching}. This is a binary relation
+on graphs, written infix as $G \leq_! H$ when $G$ matches $H$. The set
+matching !-box graphs is defined as the set closed by the following
+operations on the !-boxes in a graph:
 
@@
 
-\subsubsection{Semantics.}
-
-We give a formal semantics to a !-box graphs by defining the !-box
-graph in terms of a set of graphs in the underlying representation. We
-denote the interpretation of a !-box graph $G$ by $\binterp{G}$ and
-say that members of $\binterp{G}$ are instances of $G$.
+\textbf{Semantics.} We give a formal semantics to a !-box graphs by
+defining the !-box graph in terms of a set of graphs in the underlying
+representation. We denote the interpretation of a !-box graph $G$ by
+$\binterp{G}$ and say that members of $\binterp{G}$ are instances of
+$G$.
 
@@
 \Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
-simple consequence from the defintion of $\binterp{G}$ being a subset
+simple consequence from the definition of $\binterp{G}$ being a subset
 of the graphs that match $G$.
@@
 quantum computation. In particular, graphical rules such the Spider
-Theorem, from Figure~\ref{fig:informal-spider-law}, are frequently
-needed, but not expressible. In this section we combine the {\em
-  !-boxes} and {\em variable-node} extensions to graphs. We call this
-representation \emph{graph patterns}. This forms a representation that
-allows us to express, in a finite way, the infinite family of concrete
-graphs equations. In particular, the Spider Theroem can now be
-represented by the equation shown in Figure~\ref{fig:spider-thm-patt}.
-We also extend the notion of matching for graph pattens.
+Theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
+not expressible. In this section we combine the {\em !-boxes} and {\em
+  variable-node} extensions to graphs. We call this representation
+\emph{graph patterns}. This forms a representation that allows us to
+express, in a finite way, the infinite family of concrete graphs
+equations. In particular, the Spider Theorem can now be represented as
+shown in Figure~\ref{fig:spider}-right. We also extend the notion of
+matching for graph pattens.
 %as well as present a corresponding notion of plugging. 
@@
 quantum computation.
-
-\begin{figure}[t]
-$$\begin{array}{ccc}
-\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
-\end{array}$$
-\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using
-  graph patterns. The !-boxes are named $i$ and $j$. The variable
-  nodes are white and named $a$ and $b$. The internal node's data,
-  which is an angle, is written inside the node.}
-\end{figure}
 
@@
 
-\subsubsection{Semantics as Concrete Matches.}
-
-The semantics for a graph pattern $G$ is a set of concrete graphs
-denoted by $\minterp{G}$ and define it as:
+\textbf{Semantics as Concrete Matches.} The semantics for a graph
+pattern $G$ is a set of concrete graphs denoted by $\minterp{G}$ and
+define it as:
 
@@
 
-\subsubsection{Graph Pattern Matches.}
-
-The specification for one pattern, $G$, to match another one, $H$ is
-that it is more general with respect to the interpretation:
-$\minterp{G} \supseteq \minterp{H}$. However, graph patterns can
-correspond to a countably infinite number of concrete graphs. Thus
-matching between graph pattern cannot be implemented by simply
-unfolding them to their interpretated concrete graphs and checking the
-membership relation.
-
-Fortunately, it is quite easy to provide a decidable matching. This is
-because the size of the unfolding that needs to be considered can be
-bounded. The key observation is that a graph $G_1$ will never match a
-graph with less non-variable nodes. Thus unfolding of $G_1$ can be
-bounded by the number of non-variable nodes in $G_2$. While this gives
-a generate and test style algorithm, it is not efficient. The
-intuition for an efficient algorithm is to search through one graph
-incrementally increasing the matched part and avoids symmetries in
-matching. The development of an efficient algotihm is ongoing work.
+\textbf{Graph Pattern Matches.} The specification for one pattern,
+$G$, to match another one, $H$ is that it is more general with respect
+to the interpretation: $\minterp{G} \supseteq \minterp{H}$. However,
+graph patterns can correspond to a countably infinite number of
+concrete graphs. Thus matching between graph pattern cannot be
+implemented by simply unfolding all interpretions as concrete graphs
+and checking the membership relation.
+
+Fortunately, it is quite easy to provide decidable matching: the size
+of the unfolding that needs to be considered can be bounded. The key
+observation is that a graph $G_1$ will never match a graph with less
+non-variable nodes. Thus unfolding of $G_1$ can be bounded by the
+number of non-variable nodes in $G_2$. While this gives a generate and
+test style algorithm, it is not efficient. The intuition for an
+efficient algorithm is to search through one graph incrementally
+increasing the matched part and avoids symmetries in matching.  
+%The development of an efficient algorithm is ongoing work.
 
@@
 
-In this section we describe how the garph pattern formalism can
+In this section we describe how the graph pattern formalism can
 provide a \emph{meta-level} framework for reasoning about models of
 compact closed categories. Following the terminology of logical
-frameworks such as Isabelle~\cite{isabelle}, we call the
-specification, provided by the underlying model, an
-\emph{object-level} graph formalism. An object-level formalisation
-defines a set of rules which are treated as the axioms for the
-system. It also defines the data at the nodes and edges as well as
-corresponding data-matching behaviour. For it's part, the meta-level
-provides generic machinery to manipulate graphs and derive new
-rules. In this section we describe the meta-level framework and prove
-its adequacy for rewriting.
+frameworks such as Isabelle~\cite{isabelle}, we call the specification
+provided by the underlying model an \emph{object-level} graph
+formalism. An object-level formalisation defines a set of rules which
+are treated as the axioms for the system. It also defines the data at
+the nodes and edges as well as corresponding data-matching behaviour.
+For it's part, the meta-level provides generic machinery to manipulate
+graphs and derive new rules. We now describe the meta-level framework,
+noting the conditions for a rule to be valid, and prove the systems
+adequacy for rewriting.
 
@@
 
-The intuitive idea of substituon with a rule is to replace a subgraph
-that matches the left hand side with the right hand side. However, not
-all pairs of rules are make sense with respect to the underlying
-semantics. For an equation to be well defined with respect to the
-compact closed structure it must not be possible to changes the type
-of a concrete graph by rewriting. Mapping this restriction back to
-pairs of graph patterns results in the following conditions on rules:
+The intuitive idea of substitution with a rule is to replace a
+subgraph that matches the left hand side with the right hand side.
+However, not all pairs of rules are make sense with respect to the
+underlying semantics. For an equation to be well defined with respect
+to the compact closed structure it must not be possible to changes the
+type (the boundary nodes in the domain and co-domain) of a concrete
+graph by rewriting.  Mapping this restriction back to pairs of graph
+patterns results in the following conditions on rules:
 
@@
 \item There has to be a isomorphism between variable nodes in the left
-  and right hand subgraphs. More concretely, there must be a unique
-  naming of the variable nodes in the left hand side such that each
-  name occurs in a unique variable node on the right hand side. Given
-  a matching on the left hand side, the target subgraph is replaced
-  with the right hand side while keeping the same instantiations for
-  the isomorphic variable nodes.
-
-\item Rules must also define a partial 1-1 mapping between !-boxes on
-  the left and right hand sides. The intuition for this mapping is
-  that the unfolding used when matching a !-box on the left, is
-  applied to the mapped !-box on the right before replacement.
+  and right hand subgraphs. Given a matching on the left hand side,
+  the target subgraph is replaced with the right hand side while
+  keeping the same instantiations for the isomorphic variable nodes.
+
+\item Rules must also define a partial bijective mapping between
+  !-boxes on the left and right hand sides. The intuition for this
+  mapping is that the unfolding used when matching a !-box on the
+  left, is applied to the mapped !-box on the right before
+  replacement.
 
 \item The interplay between !-boxes and variable nodes means that when
-  a variable node appears within a !-box on one side of a rule it must
+  a variable node appears within a !-box on one side of a rule, it must
   also appear under a mapped !-box on the other side.
@@
 Notationally, and implementationally, we annotate !-boxes and variable
-nodes in a graphs with unique names. For example, see
-Figure~\ref{fig:spider-thm-patt} which illustrates the Spider Theorem
-as a rule between graph patterns. In this figure, the partial 1-1
-mapping for !-boxes is represented by !-boxes having the same name on
-the left and right of a rule.  Similarly, the isomorphism between
-variable nodes is captured by the set of variable node names on the
-left and right hand side being equal.
+nodes in a graphs with unique names. For example,
+Figure~\ref{fig:spider-thm-patt} shows the Spider Theorem. In this
+figure, the mapping between !-boxes is represented by !-boxes having
+the same name on the left and right of a rule.  Similarly, the
+isomorphism between variable nodes is captured by the set of variable
+node names on the left and right hand side being equal.
 
@@
 
-The axioms of an object system typically come from the semantics of
-the unerlying system. For instance, the equations given in
-Figure~\ref{fig:graph-quant-eqns} provide a system of reasoning about
-quantum computation. In some cases, such equations can be provdied by
-the underltying semantics. When such rules are expressed as graph
-patterns, we replace the concrete representations boundary nodes with
-variable nodes. This operation is called \emph{lifting}. An equation
-on graph patterns corresponds to a possibly infinite set of equations
-betweens concrete graphs. Thus we might worry that the lifted
-equations express too much: they may allow rewrites which are not
-intended.
+The axioms of an object formalism come from the semantics of the
+underlying system. For instance, the equations given in
+Figure~\ref{fig:graph-quant-eqns} can be proved by matrix
+calucluations oi the underlying model. When such rules are expressed
+as graph patterns, we replace the concrete representation's boundary
+nodes with variable nodes. This operation is called \emph{lifting}. An
+equation on graph patterns corresponds to a infinite (when their are
+variable nodes) set of equations between concrete graphs. Thus we
+might worry that the lifted equations express too much: they may allow
+rewrites which are not true.
 
@@
 composed with some graph. The graph is given by the unmapped subgraph
-using the open embedding from the lifted equations onto their
+using the open embedding from the lifted equations onto the considered
 instance. Thus if $G = H$ is true, then so is every instance of it's
-lifting, and thus the lifting produces an adequate representation.
+lifting, and thus lifting produces an adequate representation.
 
@@
 Having defined what make a valid rule, we now present the meta-logic
-of the framework. This is quite simple as they only involve dealing
+of the framework. This is quite simple as it only involves dealing
 with object-level equations:
@@
 
-\noindent where $\Gamma$ is the set of object-level axioms and for the
-reflexivity rule, $A$ must be a well-formed pattern graph. We assume
+\noindent $\Gamma$ is the set of object-level axioms.
+
+We assume
 that axioms in $\Gamma$ meet the validity conditions described
@@
 
-A new graph can be introduced using the reflexivity law (\emph{refl}).
-By applying the \emph{subst} rule, intermediate results are derived
-which can themselves be used to rewrite other rules and conjetcures.
-In this way, the system allows derived rules to provide an
-abbreviation for a combination of steps. Sets of rules can be applied
-automatically to simulate computation in an the object domain. In
-\S\ref{sec:case-study} we illustrate this for quantum computation.
-
-\subsection{Soundness and Completeness}
-
-The soundness of a system that uses our meta-level rests on two
-arguments:
-
-\begin{itemize}
-\item Representation of the underying model as a compact closed
-  category with a defined set of equations.
-
-\item The soundness and adequacy of the axioms when expressed within
-  the compact closed category. Adquency ensures that the encoding as a
-  graph pattern represents only the intended concrete rules. The
-  soundness of the rules is obviously needed and requires proving the
-  rule in terms of the underlying semantics.
-\end{itemize}
-
-The completeness of the resulting system, with respect to the intended
-underlying semantics, depends on the completeness of the set of given
-rules. Completeness is a challenging and still open problem for graph
-based models of quantum computation. Even without completeness, we
-obersve in \S\ref{sec:case-study} that interesting and useful quantum
-computations can be expressed.
+The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
+pattern graph. This rule allows a new graph can be introduced. By then
+applying the \emph{subst} rule, intermediate results are derived which
+can themselves be used to rewrite other rules and conjectures.  In
+this way, the system allows derived rules to provide an abbreviation
+for a combination of steps.
+
+A sets of rules can be applied automatically to simplify a graph or
+simulate computation in an the object domain. For this to terminate, a
+suitable left-to-right ordering on rules can be observed such as the
+size of the subgraph must decrease. In \S\ref{sec:case-study} we
+illustrate simulating a quantum computation.
+
+% \subsection{Soundness and Completeness}
+
+% The soundness of a system that uses our meta-level rests on two
+% arguments:
+
+% \begin{itemize}
+% \item Representation of the underlying model as a compact closed
+%   category with a defined set of equations.
+
+% \item The soundness and adequacy of the axioms when expressed within
+%   the compact closed category. Adequacy ensures that the encoding as a
+%   graph pattern represents only the intended concrete rules. The
+%   soundness of the rules is obviously needed and requires proving the
+%   rule in terms of the underlying semantics.
+% \end{itemize}
+
+% The completeness of the resulting system, with respect to the intended
+% underlying semantics, depends on the completeness of the set of given
+% rules. Completeness is a challenging and still open problem for graph
+% based models of quantum computation. Even without completeness, we
+% observe in \S\ref{sec:case-study} that interesting and useful quantum
+% computations can be expressed.
 
@@
 The model of quantum computation introduced in
-\S\ref{sec:quotients-rewriting} provides an object formalism for
-logical system. The equations from Figure~\ref{fig:graph-quant-eqns}
-are lifted to graph patterns and the spider rule is encoded as shown
-in Figure~\ref{fig:spider-thm-patt}. Although a complete graph-based
-account of the underlying quantum model is ongoing work, these rules
-provide an effective and useful initial set.
+\S\ref{sec:quotients-rewriting} provides an object formalism for our
+meta-level framework. The equations from
+Figure~\ref{fig:graph-quant-eqns} are lifted to graph patterns and the
+spider rule is encoded as shown in Figure~\ref{fig:spider-thm-patt}.
+
+%Although a complete graph-based account of the underlying quantum
+%model is ongoing work, these rules provide an effective and useful
+%initial set.
 
 Our model of quantum computation requires no data for the edges. The
-nodes, on the other hand, are either a $H$ with no additional data (a
+nodes, on the other hand, are either $H$, with no additional data (a
 Hadamard gate), or have an \emph{angle} and are either the $X$ or $Z$
 transforms. For their part, angles are expressed as rational numbers
-which correspond to the coefficient of $\pi$ in the corresponding
-matrix.
-
-The result is that we give the $X$ and $Z$ nodes an \emph{angle
-  expression}. When a node is within a !-box, this is a variable which
-gets instantiated to a new variable in each of the unfoldings of the
-!-box. When a node is not within a !-box, angle-expressions are a
-mapping from a set of angle-variables to the corresponding rational
-cooefficient. When an angle-expression contains a variable within a
-!-box, this is interpreted as a sum of the variables that result from
-it's unfolding. This rather simple expression language has a simple
-normal form by lexographically ordering the angle-variable. Matching
-then results in variables being instantiated and the values of all
-affected nodes are then normalised. An additional implementation
-detail must also be observed for the substitution rule: it must ensure
-that angle-variables in the rule being applied are distinct from those
-in the expression being rewritten.
+which correspond to the coefficient of $\pi$ in the underlying matrix.
+
+To allow composition of rules to compute the resulting angles we give
+the $X$ and $Z$ nodes an \emph{angle expression}. When a node is
+within a !-box, the expression is a single \emph{angle-variable} which
+gets instantiated to a new angle-variable in each of the unfoldings of
+the !-box.  When a node is not within a !-box, the angle-expression is
+a mapping from a set of angle-variables to the corresponding rational
+coefficient. When an angle-expression contains a angle-variable within
+a !-box, this is interpreted as a sum of the variables that result
+from its unfolding.  This rather simple expression language has a
+normal form by ordering the angle-variable by name.  Matching then
+results in angle-variables being instantiated and the expressions in
+all affected nodes are then re-normalised. An additional
+implementation detail must also be observed for the substitution rule:
+it must ensure that angle-variables in the rule being applied are
+distinct from those in the expression being rewritten.
 
@@
 
-\subsubsection{Computing the Quantum Fourier Transform.}
-\label{ssec:cs:quantum-transform}
-
-The quantum fourier transform is among the most important quantum
-algorithms, forming an essential part of Shor's
-algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
-polynomial time algorithm for factoring. In our graph patten calculus
-this ciruit becomes the top-left graph in
+\textbf{Computing the Quantum Fourier Transform.} The quantum Fourier
+transform is among the most important quantum algorithms, forming an
+essential part of Shor's algorithm~\cite{Shor:PolyTimeFact:1997},
+famous for providing a polynomial time algorithm for factoring. In our
+graph patten calculus this circuit becomes the top-left graph in
 Figure~\ref{fig:quantum-transform}. This figure shows how computation
 can be symbolically performed by rewriting with the lifted equations
-from Figure~\ref{fig:graph-quant-eqns} with the graph pattern version
+from Figure~\ref{fig:graph-quant-eqns} and the graph pattern version
 of the Spider Theorem.
@@
 \begin{tabular}{cccc}
-\hline
- & & & \\
   & \inlinegraphic{2cm}{qft2} & = & \inlinegraphic{2cm}{qft3} \\ 
  & & & \\
-\hline
  & & & \\
 = & \inlinegraphic{2cm}{qft6} & = & \inlinegraphic{2cm}{qft10} \\
- & & & \\
-\hline
 \end{tabular}
@@
   Quantum Fourier Transform with inputs 1 and 0, performed
-  symbollically by rewriting. }
+  symbolically by rewriting. }
 \end{figure}
@@
 replacement~\cite{progress97}. These provide general ways of
-interpretation graph transformations which can then be implemented to
+understanding graph transformations which can then be implemented to
 provide machinery for a specific application.  However, systems based
@@
 replacement do not guarantee well-typed results. Furthermore, they do
-not provide machinery to for rewriting of graphs with ellipses
-notation.
+not provide machinery for rewriting of graphs with ellipses notation
+which is needed to represent the Spider Theorem.
 
 The distinctive feature of our form of graph rewriting is that the
-graphs capture the structural properties of compact closed categories.
-In this respect, our graphs capture the structural properties of
-compact closed categories and rewriting is compositional: it preserves
-the type of the rewritten subgraph. However, the various components of
-our system can be seen as instantiations for in a general graph
-rewriting system: our matching provides the embedding information and
-the node-data matching of the object formalism provides application
+graphs capture the structural properties of compact closed categories
+and rewriting is compositional: it preserves the type of the rewritten
+subgraph. However, our system can also be seen as an instantiation of
+a general graph rewriting system: matching provides the embedding
+information and the object-level node matching defines the application
 conditions and the attribute transfer function.
@@
 We have introduced a representation for graphs which can formally
-characterise the ellipses notation used informally to represent certain
-infinite families of graph rewrites, such as the Spider Theorem. This
-representation provides the foundation for a simple meta-logic for
-reasoning about models of compact closed categories. We illustrated
-this by providing an account of quantum computation and showing how
-computation can be performed. Having developed the basic
+characterise the ellipses notation used informally to represent
+certain infinite families of graph rewrites, such as the Spider
+Theorem.  This representation provides the foundation for a simple
+meta-logic for reasoning about models of compact closed categories. We
+illustrated this by providing an account of quantum computation and
+showing how computation can be performed. Having developed the basic
 representational machinery and shown matching to be decidable, we are
 left with several exciting avenues for further research.  The most
-immediate is to provide a full implementation - a partial one is
-already
+immediate direction we are pursuing is to provide a full
+implementation - only a partial one is currently
 available~\footnote{\url{http://dream.inf.ed.ac.uk/projects/quantomatic}}.
 Other areas of further work include considering confluence results for
-sets of rewrite rules, even richer representations for graph-patterns
-and completeness of the set of the rewrite rules for quantum
-computation.
+sets of rewrite rules, increasing the expressiveness of the
+representation for graph-patterns, and finding a complete set of
+rewrite rules for the considered model of quantum computation.
 

### 2008-03-07 papers/reasoning-graphically/paper.tex
bibtex cleaning and squeesing to 15 pages.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1409 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\textbf{Matching.} To formalise the intuitive notion that a !-box
+%\textbf{Matching.} 
+To formalise the intuitive notion that a !-box
 represents arbitrary number of copies of the subgraph made from it's
@@
 
-\textbf{Semantics.} We give a formal semantics to a !-box graphs by
+%\textbf{Semantics.} 
+We give a formal semantics to a !-box graphs by
 defining the !-box graph in terms of a set of graphs in the underlying
@@
 
-\textbf{Semantics as Concrete Matches.} The semantics for a graph
+%\textbf{Semantics as Concrete Matches.} 
+The semantics for a graph
 pattern $G$ is a set of concrete graphs denoted by $\minterp{G}$ and
@@
 
-\textbf{Graph Pattern Matches.} The specification for one pattern,
+%\textbf{Graph Pattern Matches.} 
+The specification for one pattern,
 $G$, to match another one, $H$ is that it is more general with respect
@@
 
-\textbf{Computing the Quantum Fourier Transform.} The quantum Fourier
-transform is among the most important quantum algorithms, forming an
-essential part of Shor's algorithm~\cite{Shor:PolyTimeFact:1997},
-famous for providing a polynomial time algorithm for factoring. In our
-graph patten calculus this circuit becomes the top-left graph in
+The quantum Fourier transform is among the most important quantum
+algorithms, forming an essential part of Shor's
+algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
+polynomial time algorithm for factoring. In our graph patten calculus
+this circuit becomes the top-left graph in
 Figure~\ref{fig:quantum-transform}. This figure shows how computation

### 2008-03-07 papers/reasoning-graphically/paper.tex
title changed, dropped thanks until accepted.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1410 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 % first the title is needed
-\title{Reasoning Graphically about Quantum Computation}
+\title{Reasoning with Graphs for Quantum Computation}
 
@@
 %
-\author{Lucas Dixon\inst{1} \and Ross Duncan\inst{2}~\thanks{EPSRC Grants...}%
-}
+\author{Lucas Dixon\inst{1} \and Ross Duncan\inst{2}}
+%~\thanks{EPSRC Grants...}%
 %
@@
     interactive theorem proving, graphical calculi; concerning:
-    symbolic computation, automated reasoning, formal mathematics}
+    symbolic computation, automated reasoning, formal mathematics.}
 \end{abstract}

### 2008-03-07 papers/reasoning-graphically/paper.tex
minor spell check
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1411 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-These theorems ensure that macthing is what we expect it to be and
+These theorems ensure that matching is what we expect it to be and
 also give us the usual properties such as associativity.
@@
 concrete graphs. Thus matching between graph pattern cannot be
-implemented by simply unfolding all interpretions as concrete graphs
+implemented by simply unfolding all interpretations as concrete graphs
 and checking the membership relation.
@@
 Figure~\ref{fig:graph-quant-eqns} can be proved by matrix
-calucluations oi the underlying model. When such rules are expressed
+calculations oi the underlying model. When such rules are expressed
 as graph patterns, we replace the concrete representation's boundary

### 2008-03-07 papers/reasoning-graphically/paper.tex
bijective -> injective,
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1412 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
added when non-zero remark for node data.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1414 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
   The variable nodes are white and named $a$ and $b$. The non-variable
-  node data (the angle) is written inside the node.}
+  node data (the angle) is written inside the node when non-zero.}
 
@@
 
-\item Rules must also define a partial bijective mapping between
+\item Rules must also define a partial injective mapping between
   !-boxes on the left and right hand sides. The intuition for this

### 2008-03-07 papers/reasoning-graphically/paper.tex
edited up to !-boxes
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1415 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \begin{abstract}
-  Recent graph-based formalisms of quantum computation provide an
+  Graph-based formalisms of quantum computation provide an
   abstract and symbolic way to represent and simulate computations.
@@
 \item Many properties have a natural graphical representation. For
-  example, non entanglement is can be inferred from disjoint subgraphs.
-
-\item The graphical calculus generalises to domains other than just
+  example, separability of quantum states can be inferred from disjoint subgraphs.
+
+\item The of algebra of graphs generalises to domains other than
   vector spaces. In particular, it provides a representation for
@@
 The major problem with these graphical representations is the lack of
-machinery for automating their manipulation. Existing approaches to
-graph transformation have a different underlying semantics
-corresponding to graphs in general rather than those which describe
-compact closed categories. The result is that they provide an unsound
-concept of rewriting for the kind of graphs we are interested in.
+machinery for automating their manipulation. Unlike existing approaches to
+graph transformation, the graphs described here provide a 
+representation of compact closed categories; hence we require a new
+approach to rewriting, which is sound with respect to the underlying
+semantics.
 
@@
 \end{diagram}
-which we call \emph{source} and \emph{target}.  The definitions
-$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ express the
+which we call \emph{source} and \emph{target}.  Let
+$\text{in}(v) := t^{-1}(v)$ and $\text{out}(v) := s^{-1}(v)$ denote the
 \emph{incoming} and \emph{outgoing} edges at a vertex $v$.  The
@@
 that $f_E(e') = e$. Strictness ensures that there are no additional
-edges in $H$.
+edges connected to vertices in the image of $V'$.
 
@@
 \noindent The intuition behind this definition is that the subgraph of
-$G$ determined by $V'$ should be preserved exactly by $f$; it should
-contain $G$ without any extra incident edges into the vertices $V'$
+$G$ determined by $V'$ should be preserved exactly by $f$; 
 whereas other vertices may be identified and may contain additional
@@
 \rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
-partial graph morphism if whenever $r_E(e)$ is defined then
+\emph{partial graph morphism} if whenever $r_E(e)$ is defined then
 $r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
@@
 \begin{gather*}
-  d_A : I \to A^* \otimes A \quad\quad\quad e_A : A \otimes A^* \to A
+  d_A : I \to A^* \otimes A \quad\quad\quad e_A : A \otimes A^* \to I
 \end{gather*}
@@
 $\cod \Gamma$ and $\dom \Gamma$ are \emph{ordered signed sets}.  Given any
-ordered signed set $S$ we write $S^*$ for the same ordered set, but
+ordered signed set $S$ we write $S^*$ for the same ordered set with
 the opposite signing.   Given two such sets we can define their disjoint
@@
 \end{proposition}
-For each ordered signed set $A$, the identity map for $A$ has $\dom
+For each ordered signed set $A$, the identity map $\id{A}$ has $\dom
 \id{A} = A^*$ and $\cod \id{A} = A$; its underlying graph has $E = A$
@@
 I \to A^* \otimes A$, $e_: A \otimes A^* \to I$ have the same
-underlying graph, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
+underlying graph as $\id{A}$, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
 = A+A^*$ and $\cod e = \emptyset$.
@@
 \begin{description}
-\item[Objects:] are written $Q$ and are self dual;
+\item[Objects:] a single object, written $Q$.
 \item[Arrows:] there are families of arrows $X$ and $Z$:
@@
 free compact closed category is then given by all graphs formed by
-composing and tensoring these basic graphs.  Note  that since $Q$ is
-self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.
+composing and tensoring these basic graphs.
 
 To model the behaviour of quantum systems certain additional equations
-must be satisfied.  These are discussed in detail in
+must be satisfied.  At the  level of obejects, we ask that  $Q$ is
+self dual, i.e. that is $Q = Q^*$.  Hence we use \emph{undirected}
+graphs.  The equations between arrows are discussed in detail in
 \cite{Coecke:2008jo};  we present them in graphical form in
@@
 
-\begin{theorem}
-  The \emph{Spider Theorem}: let $G$ be a connected graph generated
+\begin{theorem}[Spider Theorem]
+  Let $G$ be a connected graph generated
   from $\delta_Z$, $\epsilon_Z$, $\alpha_Z$ and their adjoints; then
@@
 
-\noindent However, composition with semi-circles ($d$ and $e$ from
+\noindent However, composition with semi-circles ($d_Q$ and $e_Q$ from
 Definition~\ref{compactcat-def}), on the co-domain or domain, allows
@@
 \begin{definition}
-  \emph{Matching} is an open embedding
-  (Definition~\ref{open-embedding-def}) which is strict for the
-  non-variable nodes.
+  A variable-node (resp., concrete) graph $G$ \emph{matches} another graph
+  $H$ if there exists an open embedding $G\to H$ which is strict on
+  the non-variable (resp., non-boundary) nodes.  This open embedding
+  is called a \emph{matching}.
 \end{definition}
+
+If $G$ and $H$ are variable-node graphs such that  $G$ matches $H$
+then we write $G \leq_v H$.
 
@@
 $\vinterp{G}$. The interpretation is simply the set of concrete graphs
-which the variable node graph matches. The above definition of
-matching also applies between two variable-node graphs, in which case
-we denote that $G$ matches $H$ by $G \leq_v H$.
-
-\begin{theorem}
+which the variable node graph matches.
+
+\begin{proposition}
 \label{match-interp-thm}
-$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$. The
-proof from left to right is a consequence of the lemma that a
+$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$.
+\end{proposition}
+\begin{proof}
+  The proof from left to right is a consequence of the fact that a
   composition of open embeddings is an open embedding. From left to
@@
   \supseteq \vinterp{H}$. From right to left, we compose the open
-embedding $\vinterp{G}$, restricted to it's subset $\vinterp{H}$, with
-the inverse of $\vinterp{H}$ to get $G \leq_v H$.
-\end{theorem}
-
-\begin{theorem}
-\label{reflexive-match-thm}
-\emph{Matching is reflexive}: $G \leq_v G$ is a trivial instance of
-the same theorem for open-embeddings, where the embedding is the
-identity map.
-\end{theorem}
-
-These theorems ensure that matching is what we expect it to be and
-also give us the usual properties such as associativity.
+  embedding $\vinterp{G}$, restricted to it's subset $\vinterp{H}$,
+  with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
+\end{proof}
+
+It's easy to verify that $\leq_v$ is a preorder: i.e. that
+matching is transitive and reflexive.
 
@@
 
-A crucial feature of rules like the Spider Theorem is that an arbitrary
+If we wish formalise reasoning via spiders then an arbitrary
 number of repetitions of a subgraph need to be matched. To support

### 2008-03-07 papers/reasoning-graphically/paper.tex
typos
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1416 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-The representation of Compact Closed Categories as Graphs, discussed
+The representation of Compact Closed Categories as graphs, discussed
 in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
-quantum computation. In particular, graphical rules such the Spider
-Theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
+quantum computation. In particular, graphical rules such the spider
+theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
 not expressible. In this section we combine the {\em !-boxes} and {\em
@@
 of the unfolding that needs to be considered can be bounded. The key
-observation is that a graph $G_1$ will never match a graph with less
+observation is that a graph $G_1$ will never match a graph with fewer
 non-variable nodes. Thus unfolding of $G_1$ can be bounded by the

### 2008-04-15 papers/reasoning-graphically/paper.tex
typo
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1417 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
latest diagrams package fixes w.r.t. referee's comments. draft 1
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1447 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-\item The of algebra of graphs generalises to domains other than
+\item The algebra of graphs generalises to domains other than
   vector spaces. In particular, it provides a representation for
@@
 
-The major problem with these graphical representations is the lack of
+A major problem with these graphical representations is the lack of
 machinery for automating their manipulation. Unlike existing approaches to
@@
 
-\subsection{Graphs}
-\label{sec:graphs}
+\subsubsection*{Graphs}
+% \label{sec:graphs}
 
@@
 
-Let $f: G \to H$ be graph morphism and let $V' \subseteq V_G$. We say
+Let $f: G \to H$ be a graph morphism and let $V' \subseteq V_G$. We say
 that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
@@
 
-\subsection{Compact Closed Categories}
-\label{sec:comp-clos-categ}
+\subsubsection*{Compact Closed Categories}
+%\label{sec:comp-clos-categ}
 
@@
 \end{gather*}
-such that
+where $I$ is the tensor identity of the compact closed category, such
+that
 \begin{align}
@@
 which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
-$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
+$\coname{f} = e_B \circ (f \otimes \id{B^*})$.  Hence there are natural
 isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
@@
 \cite{Duncan:thesis:2006}.  Pictorial representations are in
-Fig.~\ref{fig:comcl-graphs}. We make the convention that domain of an
+Fig.~\ref{fig:comcl-graphs}. We make the convention that the domain of an
 arrow is at the top of the picture, and its codomain is at the bottom.
@@
 $g$.  The unit for the tensor is the empty set.  The morphisms $d_A :
-I \to A^* \otimes A$, $e_: A \otimes A^* \to I$ have the same
+I \to A^* \otimes A$, $e_A : A \otimes A^* \to I$ have the same
 underlying graph as $\id{A}$, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
@@
   \]
-  Call a concrete graph $\Gamma$ \emph{$T,F$-labellable} if there exists an 
-  $T,F$-labelling for it; if $\theta$ is a labelling for $\Gamma$, then
- the pair $(\Gamma,\theta)$ is an \emph{$T,F$-labelled graph}.
+  We say a concrete graph $\Gamma$ is \emph{$T,F$-labellable} if there exists an 
+  $T,F$-labelling for it; and if $\theta$ is a labelling for $\Gamma$, then
+ the pair $(\Gamma,\theta)$ is called a \emph{$T,F$-labelled graph}.
 \end{definition}
@@
 The $T,F$-labelled graphs form a compact closed category in the same
-way as the the concrete graphs, subject to the further restriction
+way as the concrete graphs, subject to the further restriction
 that arrows are composable only when their labellings agree.
@@
 \end{gather*}
-The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
-for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
-free compact closed category is then given by all graphs formed by
-composing and tensoring these basic graphs.
+The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
+Hadamard gate. (The adjoints for $H$, $\alpha_X$ and
+$\alpha_Z$ will be defined equationally.)  The free compact closed
+category is then given by all graphs formed by composing and tensoring
+these basic graphs.
 
 To model the behaviour of quantum systems certain additional equations
-must be satisfied.  At the  level of obejects, we ask that  $Q$ is
+must be satisfied.  At the  level of objects, we ask that  $Q$ is
 self dual, i.e. that is $Q = Q^*$.  Hence we use \emph{undirected}
@@
     by exchanging the two colours. The colour duality is derivable
-    from the equations for involving $H$.}
+    from the equations involving $H$.}
   \label{fig:graph-quant-eqns}
@@
 this can be depicted graphically as the equation in
-Figure~\ref{fig:spider}. Conversely, a spider may be
+Figure~\ref{fig:spider}-left. Conversely, a spider may be
 arbitrarily divided into sub-spiders, provided the total in- and
-out-degree is preserved, along with and the sum of the $\alpha $s.
+out-degree is preserved, along with the sum of the $\alpha$s.
 Further, one can derive, from the spider theorem, $n$-fold versions of
@@
 To formalise the semantics of variable-node graphs, we define
-matching, which captures the intuitive idea of a graph with variable
-nodes occurring within another graph:
+\emph{matching}, which captures the intuitive idea of a graph with
+variable nodes occurring within another graph:
 
 \begin{definition}
-  A variable-node (resp., concrete) graph $G$ \emph{matches} another graph
-  $H$ if there exists an open embedding $G\to H$ which is strict on
-  the non-variable (resp., non-boundary) nodes.  This open embedding
-  is called a \emph{matching}.
+  A variable-node (resp., concrete) graph $G$ \emph{matches} another
+  graph $H$ if there exists an open embedding $G\to H$ which is strict
+  on the non-variable (resp., non-boundary) nodes.  This open
+  embedding is called a \emph{matching}. The notation $G \leq_v H$ is
+  used for $G$ matches $H$.
 \end{definition}
-
-If $G$ and $H$ are variable-node graphs such that  $G$ matches $H$
-then we write $G \leq_v H$.
 
@@
 \begin{proof}
-  The proof from left to right is a consequence of the fact that a
-  composition of open embeddings is an open embedding. From left to
-  right, the embedding of $G \leq_v H$ composed with $\vinterp{H}$ is
-  thus an open embedding of $G$ into $\vinterp{H}$. Hence $\vinterp{G}
-  \supseteq \vinterp{H}$. From right to left, we compose the open
-  embedding $\vinterp{G}$, restricted to it's subset $\vinterp{H}$,
-  with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
+  The proof of the implication from left to right is a consequence of
+  the fact that a composition of open embeddings is an open embedding.
+  In particular, the embedding of $G \leq_v H$ composed with
+  $\vinterp{H}$ is thus an open embedding of $G$ into $\vinterp{H}$.
+  Hence $\vinterp{G} \supseteq \vinterp{H}$. From right to left, we
+  compose the open embedding $\vinterp{G}$, restricted to its subset
+  $\vinterp{H}$, with the inverse of $\vinterp{H}$ to get $G \leq_v
+  H$.
 \end{proof}
 
-It's easy to verify that $\leq_v$ is a preorder: i.e. that
+It is easy to verify that $\leq_v$ is a preorder: i.e. that
 matching is transitive and reflexive.
@@
 % \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
-% variable node is instantiated to some other node, it's type will have
+% variable node is instantiated to some other node, its type will have
 % at least the currently attached edges.
@@
 %Theorem in Figure~\ref{fig:spider} for instance is
-%represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.
+%represented with !-boxes in Figure~\ref{fig:spider}.
 
@@
 To formalise the intuitive notion that a !-box
-represents arbitrary number of copies of the subgraph made from it's
+represents arbitrary number of copies of the subgraph made from its
 nodes, we introduce \emph{!-box matching}. This is a binary relation
@@
 %\textbf{Semantics.} 
-We give a formal semantics to a !-box graphs by
-defining the !-box graph in terms of a set of graphs in the underlying
-representation. We denote the interpretation of a !-box graph $G$ by
-$\binterp{G}$ and say that members of $\binterp{G}$ are instances of
-$G$.
+We give a formal semantics to !-box graphs by defining them in terms
+of a set of graphs in the underlying representation. We denote the
+interpretation of a !-box graph $G$ by $\binterp{G}$ and say that
+members of $\binterp{G}$ are instances of $G$.
 
@@
 !-box with the natural number that defines how many copies are made of
-it. Thus the $\binterp{G}$ is isomorphic to the set of $k$-tuples of
+it. Thus $\binterp{G}$ is isomorphic to the set of $k$-tuples of
 natural numbers, where $k$ is the number of !-boxes.
@@
 graph patterns can correspond to a countably infinite number of
-concrete graphs. Thus matching between graph pattern cannot be
+concrete graphs. Thus matching between graph patterns cannot be
 implemented by simply unfolding all interpretations as concrete graphs
@@
 efficient algorithm is to search through one graph incrementally
-increasing the matched part and avoids symmetries in matching.  
+increasing the matched part.  
 %The development of an efficient algorithm is ongoing work.
@@
 the nodes and edges as well as corresponding data-matching behaviour.
-For it's part, the meta-level provides generic machinery to manipulate
+For its part, the meta-level provides generic machinery to manipulate
 graphs and derive new rules. We now describe the meta-level framework,
@@
 Axioms defined by an object-level model, as well as derived rules in
-our framework, are pairs of pattern graph. The pair represents the
+our framework, are pairs of graph patterns. The pair represents the
 left and right hand sides of an equation. Rules are declarative in
@@
 subgraph that matches the left hand side with the right hand side.
-However, not all pairs of rules are make sense with respect to the
+However, not all pairs of rules make sense with respect to the
 underlying semantics. For an equation to be well defined with respect
-to the compact closed structure it must not be possible to changes the
+to the compact closed structure it must not be possible to change the
 type (the boundary nodes in the domain and co-domain) of a concrete
@@
 nodes in a graphs with unique names. For example,
-Figure~\ref{fig:spider-thm-patt} shows the Spider Theorem. In this
+Figure~\ref{fig:spider} shows the Spider Theorem. In this
 figure, the mapping between !-boxes is represented by !-boxes having
@@
 Figure~\ref{fig:graph-quant-eqns} can be proved by matrix
-calculations oi the underlying model. When such rules are expressed
+calculations in the underlying model. When such rules are expressed
 as graph patterns, we replace the concrete representation's boundary
 nodes with variable nodes. This operation is called \emph{lifting}. An
-equation on graph patterns corresponds to a infinite (when their are
+equation on graph patterns corresponds to an infinite (when their are
 variable nodes) set of equations between concrete graphs. Thus we
@@
 using the open embedding from the lifted equations onto the considered
-instance. Thus if $G = H$ is true, then so is every instance of it's
+instance. Thus if $G = H$ is true, then so is every instance of its
 lifting, and thus lifting produces an adequate representation.
@@
 
-Having defined what make a valid rule, we now present the meta-logic
+Having defined what makes a valid rule, we now present the meta-logic
 of the framework. This is quite simple as it only involves dealing
@@
 \justifies
-\Gamma \vdash (C = D[A/B])
+\Gamma \vdash C = (D[A/B])
 \using\mbox{subst}
@@
 
-We assume
-that axioms in $\Gamma$ meet the validity conditions described
-earlier. These rules all preserve the validity conditions on equations
-and thus the system as a whole ensures only valid rules are derived.
+We assume that axioms in $\Gamma$ meet the validity conditions
+described earlier. These rules all preserve the validity conditions on
+equations and thus the system as a whole ensures only valid rules are
+derived.
 
 The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
-pattern graph. This rule allows a new graph can be introduced. By then
+pattern graph. This rule allows a new graph to be introduced. By then
 applying the \emph{subst} rule, intermediate results are derived which
@@
 Figure~\ref{fig:graph-quant-eqns} are lifted to graph patterns and the
-spider rule is encoded as shown in Figure~\ref{fig:spider-thm-patt}.
+spider rule is encoded as shown in Figure~\ref{fig:spider}.
 
@@
 algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
-polynomial time algorithm for factoring. In our graph patten calculus
+polynomial time algorithm for factoring. In our graph pattern calculus
 this circuit becomes the top-left graph in

### 2008-05-01 papers/reasoning-graphically/paper.tex
all entries now in bibfile; updated a few small things/fixes.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1448 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
*** empty log message ***
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1465 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
small changed to clarify wording of open embedding strict for...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1467 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 % first the title is needed
-\title{Reasoning with Graphs for Quantum Computation}
+\title{Extending Graphical Representations for Compact Closed Categories with
+Applications to Symbolic Quantum Computation}
 
@@
 \label{open-embedding-def}
-We call a graph morphism $f$ an \emph{open embedding} for $V'$ if:
+We call a graph morphism $f$ an \emph{open embedding which is stirct for} $V'$ if:
 \begin{enumerate}
@@
   exactly one graph morphism  $\tau:  f \to (g\circ f)$ such that
-  $\tau$ is an open embedding for the non-boundary nodes
+  $\tau$ is an open embedding which is strict for the non-boundary nodes
   of $f$. Indeed the intuition behind an open embedding, is that any
@@
 increasing the matched part.
+
 %The development of an efficient algorithm is ongoing work.
@@
 \bibliographystyle{plain}
-\bibliography{all,bibfile}
+\bibliography{bibfile}
 

### 2008-05-02 papers/reasoning-graphically/paper.tex
added some explaination about the connections to QC more generally. Need to add the references to the bibtex!
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1468 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
+A substantial strand of work in quantum computation is
+the development of high-level models of quantum processes based on
+compact closed  categories.  In this formalism, initiated in
+\cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
+logic gates, or the measurement of a qubit---correspond to arrows
+in a category, while the different quantum data types, usually just
+arrays of qubits, are the objects.  The graphical presentation
+described in the the preceding section thus provides a very expressive
+notation for representing quantum processes.
+
 While compact closed categories provide a suitable setting for
-reasoning about quantum computation \cite{AbrCoe:CatSemQuant:2004},
+reasoning about quantum computation,
 freely generated structure will not suffice:  we need additional
@@
 these basic graphs.
+
+Space does not permit a full exposition of this system, however we
+can outline the basic intuitions behind these generators.  Each edge
+in the picture represents a qubit, although several edges may
+represent the same physical qubit at different times.  An edge may
+even represent a ``virtual'' qubit which stands for a correlation
+between different parts of the  system.  The maps
+$\delta_Z$ and $\epsilon_Z$ represent quantum operations which
+respectively copy and delete the eigenstates
+of the  Pauli $Z$ operator.\footnote{Uniform copying operations are
+  forbidden by the no-cloning theorem \cite{no-cloning}, but such
+  operations are possible if we demand only the eigenstates of some
+  self-adjoint operator to be copied. Other states will not not
+  copied.  The same remarks hold true for erasing.}  
+Notice that $\delta_Z$ has one edge
+in its domain representing the qubit to copied, and two edges in its
+codomain for the two copies it produces.  Similarly, $\epsilon_Z$ has
+one qubit as input and no outputs.
+The adjoints $\delta^\dag_Z$ and  $\epsilon^\dag_Z$ correspond to an
+operation known as \emph{fusion}, and to the operation of preparing a
+fresh qubit in a certain state. The $\alpha_Z$ corresponds to phase
+shift of angle $\alpha$ in the $Z$ direction.  The family of maps
+indexed by $X$ are defined in exactly the same way, but relative to
+the Pauli $X$ rather than the $Z$.  All quantum operations may be
+defined by combining these simple operations---which are essentially
+classical---on two complementary observables.
+
+\begin{remark}
+  We emphasise that this is a mathematical notation is for
+  representing quantum processes, unlike the so-called graph states
+  used in measurement based quantum computation \cite{Briegel etc}
+  which is based on certain quantum states.  Their graphical notation
+  is dual to ours, with the vertices representing qubits and the edges
+  representing entanglement between the individual quibits.  Within
+  the formalism of this paper, a state is simply a process with no
+  inputs; that  is, a concrete graph with empty domain.  An important
+  application is to show that states or computations implmented
+  differently are equivalent:  this includes graph states, and
+  measurement based computations more generally. 
+\end{remark}
 

### 2008-05-02 papers/reasoning-graphically/paper.tex
working on images...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1470 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 %\textbf{Matching.} 
-To formalise the intuitive notion that a !-box
-represents arbitrary number of copies of the subgraph made from its
-nodes, we introduce \emph{!-box matching}. This is a binary relation
-on graphs, written infix as $G \leq_! H$ when $G$ matches $H$. The set
-matching !-box graphs is defined as the set closed by the following
-operations on the !-boxes in a graph:
+To formalise the intuitive notion that a !-box represents arbitrary
+number of copies of the subgraph made from its nodes, we introduce
+\emph{!-box matching}. This is a binary relation on graphs, written
+infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
+matching $G$ is defined as the set of graphs closed by the following
+operations on $G$:
 
@@
 
+\begin{figure}[t]
+$$\begin{array}{ccccccc}
+\inlinegraphic{3.8em}{bbox-ex1} & 
+\frac{\leq}{copy} & \inlinegraphic{3.8em}{bbox-ex2}  & 
+\frac{\leq}{merge} & \inlinegraphic{3.8em}{bbox-ex3}  & 
+\frac{\leq}{drop} & \inlinegraphic{3.8em}{bbox-ex4}
+\end{array}$$
+\label{fig:bang-box-example}\caption{An illustration of !-box graph
+  matching using the !-box operations. This involves first copying
+  !-box $i_0$ twice, then merging $i_0$ and $i_1$ and finally dropping
+  $i_3$.}
+\end{figure}
+
 %\textbf{Semantics.} 
+An example of matching with these operations is illustrated in
+Figure~\ref{fig:bang-box-example}.
+
 We give a formal semantics to !-box graphs by defining them in terms

### 2008-05-02 papers/reasoning-graphically/paper.tex
slight tweak of pic and added sentence about !-box needing to be not k-tuples.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1473 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \item[merge]: combines two !-boxes, $B_1$ and $B_2$ into a single
-  larger !-box $B_1 \cup B_2$.
+  larger !-box $B_1 \cup B_2$. To ensure that copying after merging
+  commutes with copying before, merging is restricted to !-boxes which
+  do not have an edge between their nodes.
 \end{description}
@@
 $$\begin{array}{ccccccc}
-\inlinegraphic{3.8em}{bbox-ex1} & 
-\frac{\leq}{copy} & \inlinegraphic{3.8em}{bbox-ex2}  & 
-\frac{\leq}{merge} & \inlinegraphic{3.8em}{bbox-ex3}  & 
-\frac{\leq}{drop} & \inlinegraphic{3.8em}{bbox-ex4}
+\inlinegraphic{4em}{bbox-ex1} &  
+\stackrel{\leq}{\stackrel{\ }{\small{copy}}} 
+& \inlinegraphic{4em}{bbox-ex2}  & 
+\stackrel{\leq}{\stackrel{\ }{\small{merge}}} 
+& \inlinegraphic{4em}{bbox-ex3}  & 
+\stackrel{\leq}{\stackrel{\ }{\small{drop}}} 
+& \inlinegraphic{4em}{bbox-ex4}
 \end{array}$$ 
@@
 
-Observe that an instance of a !-box graph is defined by pairing each
-!-box with the natural number that defines how many copies are made of
-it. Thus $\binterp{G}$ is isomorphic to the set of $k$-tuples of
-natural numbers, where $k$ is the number of !-boxes.
+Observe that every concrete instance of a !-box graph can be defined
+by pairing each !-box with the natural number that defines how many
+copies are made of it. Thus $\binterp{G}$ is isomorphic to the set of
+$k$-tuples of natural numbers, where $k$ is the number of !-boxes.
 
@@
 \end{theorem}
+
+The need for the !-box matching operation, rather than using a direct
+$k$-tuple interpretation, is to allow matching between !-box graphs,
+and thus to provide a mechanism for derived rules.
 

### 2008-05-02 papers/reasoning-graphically/paper.tex
fixed a few typos.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1474 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-A substantial strand of work in quantum computation is
-the development of high-level models of quantum processes based on
-compact closed  categories.  In this formalism, initiated in
+A substantial strand of work in quantum computation has involved the
+development of high-level models of quantum processes based on compact
+closed categories.  In these formalisms, initiated in
 \cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
-logic gates, or the measurement of a qubit---correspond to arrows
-in a category, while the different quantum data types, usually just
-arrays of qubits, are the objects.  The graphical presentation
-described in the the preceding section thus provides a very expressive
-notation for representing quantum processes.
+logic gates, or the measurement of a qubit---correspond to arrows in a
+category, while the different quantum data types, usually just arrays
+of qubits, are the objects.  The graphical representation described in
+the preceding section thus provides a very expressive notation quantum
+processes.
 
@@
 \to A$.  Each arrow is represented as a small graph; its adjoint is
-the same graph written upside down.  We use colours to denote two families:
+the same graph written upside down by reflection in the x-axis.  We
+use colours to denote two families:
 \begin{gather*}
@@
 The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
-Hadamard gate. (The adjoints for $H$, $\alpha_X$ and
-$\alpha_Z$ will be defined equationally.)  The free compact closed
+Hadamard gate. The adjoints for $H$, $\alpha_X$ and
+$\alpha_Z$ will be defined equationally.  The free compact closed
 category is then given by all graphs formed by composing and tensoring
@@
 
-Space does not permit a full exposition of this system, however we
-can outline the basic intuitions behind these generators.  Each edge
-in the picture represents a qubit, although several edges may
-represent the same physical qubit at different times.  An edge may
-even represent a ``virtual'' qubit which stands for a correlation
-between different parts of the  system.  The maps
-$\delta_Z$ and $\epsilon_Z$ represent quantum operations which
-respectively copy and delete the eigenstates
-of the  Pauli $Z$ operator.\footnote{Uniform copying operations are
-  forbidden by the no-cloning theorem \cite{no-cloning}, but such
-  operations are possible if we demand only the eigenstates of some
-  self-adjoint operator to be copied. Other states will not not
-  copied.  The same remarks hold true for erasing.}  
-Notice that $\delta_Z$ has one edge
-in its domain representing the qubit to copied, and two edges in its
-codomain for the two copies it produces.  Similarly, $\epsilon_Z$ has
-one qubit as input and no outputs.
-The adjoints $\delta^\dag_Z$ and  $\epsilon^\dag_Z$ correspond to an
-operation known as \emph{fusion}, and to the operation of preparing a
-fresh qubit in a certain state. The $\alpha_Z$ corresponds to phase
-shift of angle $\alpha$ in the $Z$ direction.  The family of maps
-indexed by $X$ are defined in exactly the same way, but relative to
-the Pauli $X$ rather than the $Z$.  All quantum operations may be
-defined by combining these simple operations---which are essentially
+In terms of quantum processes, each edge in a graph represents a
+qubit, although several edges may represent the same physical qubit at
+different times.  An edge may even represent a ``virtual'' qubit which
+stands for a correlation between different parts of the system.  The
+maps $\delta_Z$ and $\epsilon_Z$ represent quantum operations which
+respectively copy and delete the eigenstates of the Pauli $Z$
+operator.\footnote{Uniform copying operations are forbidden by the
+  no-cloning theorem \cite{no-cloning}, but such operations are
+  possible if we demand only the eigenstates of some self-adjoint
+  operator to be copied.  Other states will not not copied.  The same
+  remarks hold true for erasing.}  Notice that $\delta_Z$ has one edge
+in its domain representing the qubit to be copied, and two edges in
+its codomain for the two copies it produces.  Similarly, $\epsilon_Z$
+has one qubit as input and no outputs.  The adjoints $\delta^\dag_Z$
+and $\epsilon^\dag_Z$ correspond to an operation known as
+\emph{fusion}, and to the operation of preparing a fresh qubit in a
+certain state.  The $\alpha_Z$ corresponds to phase shift of angle
+$\alpha$ in the $Z$ direction.  The family of maps indexed by $X$ are
+defined in exactly the same way, but relative to the Pauli $X$
+operator rather than the $Z$.  All quantum operations may be defined
+by combining these simple operations---which are essentially
 classical---on two complementary observables.
@@
   representing quantum processes, unlike the so-called graph states
-  used in measurement based quantum computation \cite{Briegel etc}
-  which is based on certain quantum states.  Their graphical notation
-  is dual to ours, with the vertices representing qubits and the edges
-  representing entanglement between the individual quibits.  Within
-  the formalism of this paper, a state is simply a process with no
-  inputs; that  is, a concrete graph with empty domain.  An important
-  application is to show that states or computations implmented
-  differently are equivalent:  this includes graph states, and
-  measurement based computations more generally. 
+  used in measurement based quantum computation \cite{Briegel etc}.
+  Their graphical notation is dual to ours, with the vertices
+  representing qubits and the edges representing entanglement between
+  the individual quibits.  Within the formalism of this paper, a state
+  is simply a process with no inputs; that is, a concrete graph with
+  empty domain.  An important application is to show that states or
+  computations implemented differently are equivalent: this includes
+  graph states, and measurement based computations more generally.
 \end{remark}
@@
 
-Consider the equations shown in Figure \ref{fig:graph-quant-eqns} which
-involve only one colour:  these allow the remarkable \emph{spider theorem},
-first noted in \cite{Coecke2006Quantum-Measure}, to be proved:
+Consider the equations from Figure \ref{fig:graph-quant-eqns} which
+involve only one colour: these allow the remarkable \emph{spider
+  theorem}, first noted in \cite{Coecke2006Quantum-Measure}, to be
+proved:
 
@@
 
-Hence any connected subgraph involving nodes of only one colour,
-wherever it occurs in a graph, may be collapsed to a single vertex
-labelled by a single value $\alpha$, giving a ``spider''. Informally,
-this can be depicted graphically as the equation in
-Figure~\ref{fig:spider}-left. Conversely, a spider may be
+Hence any connected subgraph involving nodes of only one colour may be
+collapsed to a single vertex, with a single value $\alpha$, giving a
+``spider''. Informally, this can be depicted graphically as the
+equation in Figure~\ref{fig:spider}-left. Conversely, a spider may be
 arbitrarily divided into sub-spiders, provided the total in- and
 out-degree is preserved, along with the sum of the $\alpha$s.
-Further, one can derive, from the spider theorem, $n$-fold versions of
+Furthermore, one can derive, from the spider theorem, $n$-fold versions of
 many of the other equations.

### 2008-05-03 papers/reasoning-graphically/paper.tex
updated images, deleted unused images, small edits to paper.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1475 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 Augmented by some additional structure, graphs form a \emph{compact closed
-category}.  In section \ref{sec:graph-repr-comp} we will describe this
+category}.  In section~\ref{sec:graph-repr-comp} we will describe this
 structure, but first we review the basic properties of compact closed
@@
 \label{compactcat-def}
-A strict symmetric monoidal  category
-\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
-closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
-object $A^*$, and morphisms
+A strict symmetric monoidal category~\cite{AspLon:CatTypStruct:1991}
+is called compact closed~\cite{KelLap:comcl:1980} when each object $A$
+has a chosen dual object $A^*$, and morphisms
 \begin{gather*}
@@
       &\qquad\qquad&
-      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
+      \inlinegraphic{3.3em}{Heq-alpha} &=\;&
       \inlinegraphic{1.5em}{redalpha}
@@
 \inlinegraphic{1.3em}{compact2} $\qquad\quad$
-\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
-$\qquad\quad$ \inlinegraphic{2.5em}{compact5} $=$
-\inlinegraphic{2.5em}{comonoid-unit4}
+\inlinegraphic{2.5em}{compact3} $=$ 
+\inlinegraphic{1.3em}{compact4} $\qquad\quad$ 
+\inlinegraphic{2.5em}{compact5} $=$
+\inlinegraphic{2.5em}{compact6}
 \end{center}
-
-%\begin{center}
-%\inlinegraphic{2.5em}{compact1} \quad\quad
-%\inlinegraphic{2.5em}{compact3} \quad\quad
-%\inlinegraphic{2.5em}{compact5}
-%\end{center}
 
@@
 formalism. An object-level formalisation defines a set of rules which
-are treated as the axioms for the system. It also defines the data at
-the nodes and edges as well as corresponding data-matching behaviour.
-For its part, the meta-level provides generic machinery to manipulate
-graphs and derive new rules. We now describe the meta-level framework,
-noting the conditions for a rule to be valid, and prove the systems
-adequacy for rewriting.
+are treated as the axioms for the system; for instance, the equation
+from Figure~\ref{}. It also defines the data at the nodes and edges as
+well as corresponding data-matching behaviour.  For its part, the
+meta-level provides generic machinery to manipulate graphs and derive
+new rules.  We now describe the meta-level framework, noting the
+conditions for a rule to be valid, and prove the systems adequacy for
+rewriting. The resulting system forms the basis for an interactive
+proof assistant that supports reasoning about compact closed
+categories.
 
@@
 \S\ref{sec:quotients-rewriting} provides an object formalism for our
-meta-level framework. The equations from
-Figure~\ref{fig:graph-quant-eqns} are lifted to graph patterns and the
-spider rule is encoded as shown in Figure~\ref{fig:spider}.
+meta-level framework. In particular, the object level axioms come from
+the lifting the equations in Figure~\ref{fig:graph-quant-eqns} and the
+from the encoding of the spider law as shown in
+Figure~\ref{fig:spider}.
 
@@
 Our model of quantum computation requires no data for the edges. The
-nodes, on the other hand, are either $H$, with no additional data (a
-Hadamard gate), or have an \emph{angle} and are either the $X$ or $Z$
+nodes on the other hand are either $H$ (a Hadamard gate), with no
+additional data, or have an \emph{angle} and are either the $X$ or $Z$
 transforms. For their part, angles are expressed as rational numbers

### 2008-05-03 papers/reasoning-graphically/paper.tex
tiny thing on keywords fixed.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1479 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
   \keywords{graph rewriting, quantum computing, categorical logic,
-    interactive theorem proving, graphical calculi; concerning:
-    symbolic computation, automated reasoning, formal mathematics.}
+    interactive theorem proving, graphical calculi.}
+% concerning:
+%    symbolic computation, automated reasoning, formalised mathematics.
 \end{abstract}

### 2008-05-04 papers/reasoning-graphically/paper.tex
final edits by me
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1481 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 
-Let $f: G \to H$ be a graph morphism and let $V' \subseteq V_G$. We say
-that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
-\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
-that $f_E(e') = e$. Strictness ensures that there are no additional
+%%% TODO removed this because we don't need partial maps
+%%% unless we are talking about plugging.
+%%%
+% Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
+% \rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
+% \emph{partial graph morphism} if whenever $r_E(e)$ is defined then
+% $r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
+% $r_E$ to its preimage satisfies equations \eqref{eq:graph-hom1} and
+% \eqref{eq:graph-hom2} above.
+
+\begin{definition}
+  Let $f: G \to H$ be a graph morphism and let $V' \subseteq V_G$. We
+  say that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if
+  $s_H(e) \in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in
+  E_G$ such that $f_E(e') = e$.
+\end{definition}
+\noindent Strictness ensures that there are no additional
 edges connected to vertices in the image of $V'$.
@@
 vertices.
-
-Given two graphs, $G$ and $H$, a pair of partial maps, $r_V: V_G
-\rightharpoonup V_H$ and $r_E: E_G \rightharpoonup E_H$, is called a
-\emph{partial graph morphism} if whenever $r_E(e)$ is defined then
-$r_V(s_G(e))$ and $r_V(t_G(e))$ are defined, and the restriction of
-$r_E$ to its preimage satisfies equations \eqref{eq:graph-hom1} and
-\eqref{eq:graph-hom2} above.
 
@@
 of qubits, are the objects.  The graphical representation described in
-the preceding section thus provides a very expressive notation quantum
+the preceding section thus provides a very expressive notation for quantum
 processes.
@@
 operator.\footnote{Uniform copying operations are forbidden by the
-  no-cloning theorem \cite{no-cloning}, but such operations are
+  no-cloning theorem \cite{Wootters1982A-single-quantu}, but such operations are
   possible if we demand only the eigenstates of some self-adjoint
   operator to be copied.  Other states will not not copied.  The same
-  remarks hold true for erasing.}  Notice that $\delta_Z$ has one edge
+  remarks hold true for erasing \cite{Pati2000Impossibility-o}.}  Notice that $\delta_Z$ has one edge
 in its domain representing the qubit to be copied, and two edges in
@@
 
-\begin{remark}
-  We emphasise that this is a mathematical notation is for
-  representing quantum processes, unlike the so-called graph states
-  used in measurement based quantum computation \cite{Briegel etc}.
-  Their graphical notation is dual to ours, with the vertices
-  representing qubits and the edges representing entanglement between
-  the individual quibits.  Within the formalism of this paper, a state
-  is simply a process with no inputs; that is, a concrete graph with
-  empty domain.  An important application is to show that states or
-  computations implemented differently are equivalent: this includes
-  graph states, and measurement based computations more generally.
-\end{remark}
+We emphasise that this is a notation is for representing quantum
+processes, rather than quantum states.  In this setting a state is
+simply a process with no inputs; that is, a concrete graph with
+empty domain.
+  Since our formalism is based on the
+underlying mathematical structure rather than any particular model of
+quantum computation, it is capable of representing  quantum circuits,
+and measurement-based quantum computations, among other models.
+Indeed, an important application is to show that states or 
+computations implemented differently are equivalent.
 
@@
 
-More formally, a !-box graph is a graph paired with a set !-boxes. Each
-of these !-boxes is the set of nodes inside it and each !-box must be
-disjoint from the others.\footnote{The disjointness condition avoids a
-  slightly more complex, although also more expressive, semantics for
-  overlapping node sets in the !-boxes. While such expressivity is
-  interesting from a representational point of view, it is not needed
-  to express the currently known rules for quantum computation.}  We
-now define the concept of !-box matching which we then use to provide
-a formal semantics for the !-boxed graphs.
-
-%\textbf{Matching.} 
+More formally, a \emph{!-box graph} is a pair $(G, {\cal B})$ where $G$ is a
+graph and $\cal B$ is a set of subsets of $V$ such that  whenever
+$b_1, b_2 \in {\cal B}$ then $b_1 \cap b_2 =\emptyset$,\footnote{One
+  could consider more expressive notions of nested, or overlapping,
+  node sets in the !-boxes. While such expressivity is 
+  interesting,  it is not required for the system we formalise here.}  and
+subject to the further restriction that no edge may connect vertices
+in distinct !-boxes, i.e. if $v_1 \in b_1$, $v_2 \in b_2$ and $s(e) =
+t(e)$ then $b_1 = b_2$.  
 To formalise the intuitive notion that a !-box represents arbitrary
 number of copies of the subgraph made from its nodes, we introduce
-\emph{!-box matching}. This is a binary relation on graphs, written
-infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
-matching $G$ is defined as the set of graphs closed by the following
-operations on $G$:
-
+\emph{!-box matching}.   This binary relation, written infix as
+$\leq_!$, is defined such that 
+$(G,{\cal B}) \leq_! (H, {\cal C})$ whenever $(H, {\cal C})$ can be
+obtained from $(G,{\cal B})$ be a sequence the following operations:
+
+%
+% TODO there ought to be formal definitions of these but not enough
+% space in this version
 \begin{description}
@@
 of a set of graphs in the underlying representation. We denote the
-interpretation of a !-box graph $G$ by $\binterp{G}$ and say that
-members of $\binterp{G}$ are instances of $G$.
+interpretation of a !-box graph $(G,{\cal B}$ by $\binterp{(G,{\cal
+    B})}$ and say that members of $\binterp{G}$ are instances of $G$.
 
 \begin{definition}
-  \emph{Interpretation of !-box Graphs}: $\binterp{G}$ is the subset
-  of graphs matched by the !-box graph that have no !-boxes.
+  The \emph{interpretation} $\binterp{(G,{\cal B})}$ is a
+  set of concrete graphs defined 
+  \[
+  \binterp{(G,{\cal B})} = \{ H \;\;|\;\; (G,{\cal B}) \leq_! (H,\emptyset) \}
+  \]
+  i.e. the those graphs matched by the !-box graph that have no
+  !-boxes.
 \end{definition}
@@
 
-\begin{theorem}
+\begin{proposition}
 \label{thm:bang-box-respect}
@@
 of the graphs that match $G$.
-\end{theorem}
+\end{proposition}
 
@@
 For the purposes of this paper, the underlying graph representation of
-!-boxed graphs is variable-node graphs. This gives rise to graph
-patterns which we now discuss in more detail. 
+!-boxed graphs is variable-node graphs. This gives rise to \emph{graph
+patterns} which we now discuss in more detail. 
 % The set of variable-node
@@
 nodes on the other hand are either $H$ (a Hadamard gate), with no
-additional data, or have an \emph{angle} and are either the $X$ or $Z$
-transforms. For their part, angles are expressed as rational numbers
+additional data, or have an \emph{angle} and a \emph{colour}, expressing that
+they are defined in the $X$ or $Z$ basis. For their part, angles are
+expressed as rational numbers 
 which correspond to the coefficient of $\pi$ in the underlying matrix.
@@
 conditions and the attribute transfer function.
+
+We note that our graphical notation has little connection
+to \emph{graph states} as used in various approaches to
+measurement-based quantum computation \cite{Raussendorf-2001}.  In
+that approach the graph structure is used to provide a description of
+the entanglement in a state:  it does not provide a complete
+description of a computation.
 

### 2008-05-04 papers/reasoning-graphically/paper.tex
fixed one last thing
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1482 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
final small edits.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1483 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@
 \label{open-embedding-def}
-We call a graph morphism $f$ an \emph{open embedding which is stirct for} $V'$ if:
+We call a graph morphism $f$ an \emph{open embedding which is strict for} $V'$ if:
 \begin{enumerate}
@@
 whereas other vertices may be identified and may contain additional
-vertices.
+incident edges.
 
@@
   linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
-Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
+Furthermore,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
 \[
@@
 the same graph written upside down by reflection in the x-axis.  We
-use colours to denote two families:
+use colours (light green and a darker red) to denote the two families:
 \begin{gather*}
@@
 
-We emphasise that this is a notation is for representing quantum
+We emphasise that this is a notation for representing quantum
 processes, rather than quantum states.  In this setting a state is
@@
 and measurement-based quantum computations, among other models.
-Indeed, an important application is to show that states or 
+Indeed, an important application of this work is to show that states or 
 computations implemented differently are equivalent.
@@
 
+\noindent Hence any connected subgraph involving nodes of only one colour may be
+collapsed to a single vertex, with a single value $\alpha$, giving a
+``spider''. Informally, this can be depicted graphically as the
+equation in Figure~\ref{fig:spider}-left. Conversely, a spider may be
+arbitrarily divided into sub-spiders, provided the total in- and
+out-degree is preserved, along with the sum of the $\alpha$s.
+Furthermore, one can derive, from the Spider Theorem, $n$-fold versions of
+many of the other equations.
+
 \begin{figure}[t]
@@
   node data (the angle) is written inside the node when non-zero.}
-
 \end{figure}
-
-Hence any connected subgraph involving nodes of only one colour may be
-collapsed to a single vertex, with a single value $\alpha$, giving a
-``spider''. Informally, this can be depicted graphically as the
-equation in Figure~\ref{fig:spider}-left. Conversely, a spider may be
-arbitrarily divided into sub-spiders, provided the total in- and
-out-degree is preserved, along with the sum of the $\alpha$s.
-Furthermore, one can derive, from the spider theorem, $n$-fold versions of
-many of the other equations.
 
@@
 where nodes may have arbitrarily many edges, and there are infinitely
-many concrete rewrite rules. The desire to retain these intuitive
-reasoning methods motivates the extension from concrete graphs to
-\emph{graph patterns}. These will comprise the main subject of this
-paper.
+many concrete rewrite rules. The desire to retain intuitive reasoning
+methods for these infinite families of rewrites motivates the
+extension from concrete graphs to \emph{graph patterns}, the main
+subject of this paper.
 
@@
 an equation involving one of the above to easily lead to a derivation
-corresponding to either of the other two: proving any of the above
-allows a trivial derivation of the others.
+corresponding to either of the other two: given any one of the allows
+a trivial derivation of the others.
 
@@
 
-It is easy to verify that $\leq_v$ is a preorder: i.e. that
-matching is transitive and reflexive.
+%It is easy to verify that $\leq_v$ is a preorder: i.e. that
+%matching is transitive and reflexive.
 
@@
 
-More formally, a \emph{!-box graph} is a pair $(G, {\cal B})$ where $G$ is a
-graph and $\cal B$ is a set of subsets of $V$ such that  whenever
-$b_1, b_2 \in {\cal B}$ then $b_1 \cap b_2 =\emptyset$,\footnote{One
+More formally, a \emph{!-box graph} is a pair $(G, {\cal B})$ where
+$G$ is a graph and $\cal B$ is a set of disjoint subsets of $V$ i.e.
+$b_1, b_2 \in {\cal B}$ then $b_1 \cap b_2 =\emptyset$.\footnote{One
   could consider more expressive notions of nested, or overlapping,
-  node sets in the !-boxes. While such expressivity is 
-  interesting,  it is not required for the system we formalise here.}  and
-subject to the further restriction that no edge may connect vertices
-in distinct !-boxes, i.e. if $v_1 \in b_1$, $v_2 \in b_2$ and $s(e) =
-t(e)$ then $b_1 = b_2$.  
+  node sets in the !-boxes. While such expressivity is interesting, it
+  is not required for the system we formalise here.}
+%% -- we only need the unconnected !-box restriction for the merge operation. 
+%  and
+%subject to the further restriction that no edge may connect vertices
+%in distinct !-boxes, i.e. if $v_1 \in b_1$, $v_2 \in b_2$ and $s(e) =
+%t(e)$ then $b_1 = b_2$.  
 To formalise the intuitive notion that a !-box represents arbitrary
@@
 $(G,{\cal B}) \leq_! (H, {\cal C})$ whenever $(H, {\cal C})$ can be
-obtained from $(G,{\cal B})$ be a sequence the following operations:
+obtained from $(G,{\cal B})$ by a sequence the following operations:
 
@@
 \inlinegraphic{4em}{bbox-ex1} &  
-\stackrel{\leq}{\stackrel{\ }{\small{copy}}} 
+\stackrel{\leq}{\stackrel{\ }{\mbox{\small{copy}}}} 
 & \inlinegraphic{4em}{bbox-ex2}  & 
-\stackrel{\leq}{\stackrel{\ }{\small{merge}}} 
+\stackrel{\leq}{\stackrel{\ }{\mbox{\small{merge}}}} 
 & \inlinegraphic{4em}{bbox-ex3}  & 
-\stackrel{\leq}{\stackrel{\ }{\small{drop}}} 
+\stackrel{\leq}{\stackrel{\ }{\mbox{\small{drop}}}} 
 & \inlinegraphic{4em}{bbox-ex4}
@@
 We give a formal semantics to !-box graphs by defining them in terms
-of a set of graphs in the underlying representation. We denote the
-interpretation of a !-box graph $(G,{\cal B}$ by $\binterp{(G,{\cal
-    B})}$ and say that members of $\binterp{G}$ are instances of $G$.
+of a set of graphs in the underlying representation. In particular, we
+denote the interpretation of a !-box graph $(G,{\cal B})$ by
+$\binterp{(G,{\cal B})}$ and say that its members are instances.
 
@@
   The \emph{interpretation} $\binterp{(G,{\cal B})}$ is a
-  set of concrete graphs defined 
+  set of concrete graphs defined by
   \[
@@
 $k$-tuples of natural numbers, where $k$ is the number of !-boxes.
+The need for the !-box matching operation, rather than using a direct
+$k$-tuple interpretation, is to allow matching between !-box graphs,
+and thus to provide a mechanism for derived rules.
 
@@
 
-The need for the !-box matching operation, rather than using a direct
-$k$-tuple interpretation, is to allow matching between !-box graphs,
-and thus to provide a mechanism for derived rules.
-
 For the purposes of this paper, the underlying graph representation of
@@
 in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
-quantum computation. In particular, graphical rules such the spider
-theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
+quantum computation. In particular, graphical rules such the Spider
+Theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
 not expressible. In this section we combine the {\em !-boxes} and {\em
@@
 \emph{graph patterns}. This forms a representation that allows us to
-express, in a finite way, the infinite family of concrete graphs
-equations. In particular, the Spider Theorem can now be represented as
-shown in Figure~\ref{fig:spider}-right. We also extend the notion of
-matching for graph pattens.
+express, in a finite way, certain infinite families of equations
+between concrete graphs. In particular, the Spider Theorem can now be
+represented as shown in Figure~\ref{fig:spider}-right. We also extend
+the notion of matching for graph pattens.
 %as well as present a corresponding notion of plugging. 
 This provides the foundations for the rewriting machinery
-in~\S\ref{sec:rewriting} which can then be used to reasoning about
+in~\S\ref{sec:rewriting} which can then be used to reason about
 quantum computation.
@@
 formalism. An object-level formalisation defines a set of rules which
-are treated as the axioms for the system; for instance, the equation
-from Figure~\ref{}. It also defines the data at the nodes and edges as
-well as corresponding data-matching behaviour.  For its part, the
-meta-level provides generic machinery to manipulate graphs and derive
-new rules.  We now describe the meta-level framework, noting the
-conditions for a rule to be valid, and prove the systems adequacy for
-rewriting. The resulting system forms the basis for an interactive
-proof assistant that supports reasoning about compact closed
-categories.
+are treated as the axioms for the system; for instance, the equations
+from Figure~\ref{fig:graph-quant-eqns}. It also defines the data at
+the nodes and edges as well as corresponding data-matching behaviour.
+For its part, the meta-level provides generic machinery to manipulate
+graphs and derive new rules.  We now describe the meta-level
+framework, noting the conditions for a rule to be valid, and prove the
+systems adequacy for rewriting. The resulting system forms the basis
+for an interactive proof assistant that supports reasoning about
+compact closed categories.
 
@@
 
-Axioms defined by an object-level model, as well as derived rules in
-our framework, are pairs of graph patterns. The pair represents the
-left and right hand sides of an equation. Rules are declarative in
+In our framework, the axioms defined by an object-level model, as well
+as derived rules, are pairs of graph patterns. Such a pair represents
+the left and right hand sides of an equation. Rules are declarative in
 that they denote a set of concrete equational rules.
@@
 \item There has to be a isomorphism between variable nodes in the left
-  and right hand subgraphs. Given a matching on the left hand side,
-  the target subgraph is replaced with the right hand side while
-  keeping the same instantiations for the isomorphic variable nodes.
+  and right hand subgraphs. Given a matching against the left hand
+  side of a rule, the target subgraph is replaced with the right hand
+  side while keeping the same instantiations for the isomorphic
+  variable nodes of the right hand side.
 
@@
 Notationally, and implementationally, we annotate !-boxes and variable
-nodes in a graphs with unique names. For example,
-Figure~\ref{fig:spider} shows the Spider Theorem. In this
-figure, the mapping between !-boxes is represented by !-boxes having
-the same name on the left and right of a rule.  Similarly, the
-isomorphism between variable nodes is captured by the set of variable
-node names on the left and right hand side being equal.
+nodes in a graph with unique names. For example,
+Figure~\ref{fig:spider} shows the Spider Theorem. In this figure, the
+mapping between !-boxes is represented by !-boxes having the same name
+on the left and right of a rule.  Similarly, the isomorphism between
+variable nodes is captured by the set of variable node names on the
+left and right hand side being equal.
 
@@
 
-\noindent $\Gamma$ is the set of object-level axioms.
-
-We assume that axioms in $\Gamma$ meet the validity conditions
+\noindent where $\Gamma$ is the set of object-level axioms.
+
+We assume that the axioms in $\Gamma$ meet the validity conditions
 described earlier. These rules all preserve the validity conditions on
 equations and thus the system as a whole ensures only valid rules are
-derived.
-
-The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
-pattern graph. This rule allows a new graph to be introduced. By then
-applying the \emph{subst} rule, intermediate results are derived which
-can themselves be used to rewrite other rules and conjectures.  In
-this way, the system allows derived rules to provide an abbreviation
-for a combination of steps.
+derived. For the reflexivity rule (\emph{refl}), we assumes that $A$
+is a well-formed pattern graph. This rule allows a new graph to be
+introduced. By then applying the \emph{subst} rule, intermediate
+results are derived which can themselves be used to rewrite other
+rules and conjectures.  In this way, the system allows derived rules
+to provide an abbreviation for a combination of steps.
 
 A sets of rules can be applied automatically to simplify a graph or
-simulate computation in an the object domain. For this to terminate, a
-suitable left-to-right ordering on rules can be observed such as the
-size of the subgraph must decrease. In \S\ref{sec:case-study} we
-illustrate simulating a quantum computation.
+simulate computation in the object domain. For such rewriting to
+terminate, a suitable left-to-right ordering on rules needs to be
+observed, such as a decrease in the size of the subgraph. In
+\S\ref{sec:case-study} we illustrate simulating a quantum computation.
 
@@
 meta-level framework. In particular, the object level axioms come from
-the lifting the equations in Figure~\ref{fig:graph-quant-eqns} and the
-from the encoding of the spider law as shown in
+lifting the equations in Figure~\ref{fig:graph-quant-eqns} and from
+the encoding of the Spider Theorem as shown in
 Figure~\ref{fig:spider}.
@@
 a mapping from a set of angle-variables to the corresponding rational
-coefficient. When an angle-expression contains a angle-variable within
+coefficient. When an angle-expression contains an angle-variable within
 a !-box, this is interpreted as a sum of the variables that result
@@
 results in angle-variables being instantiated and the expressions in
-all affected nodes are then re-normalised. An additional
+all affected nodes are then (re)normalised. An additional
 implementation detail must also be observed for the substitution rule:
@@
 algorithms, forming an essential part of Shor's
-algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
-polynomial time algorithm for factoring. In our graph pattern calculus
-this circuit becomes the top-left graph in
-Figure~\ref{fig:quantum-transform}. This figure shows how computation
-can be symbolically performed by rewriting with the lifted equations
-from Figure~\ref{fig:graph-quant-eqns} and the graph pattern version
-of the Spider Theorem.
+algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing
+polynomial factoring. In our graph pattern calculus this circuit
+becomes the top-left graph in Figure~\ref{fig:quantum-transform}. This
+figure shows how computation can be symbolically performed by
+rewriting with the lifted equations from
+Figure~\ref{fig:graph-quant-eqns} and the graph pattern version of the
+Spider Theorem.
 
@@
 implementation - only a partial one is currently
-available~\footnote{\url{http://dream.inf.ed.ac.uk/projects/quantomatic}}.
+available\footnote{\url{http://dream.inf.ed.ac.uk/projects/quantomatic}}.
 Other areas of further work include considering confluence results for

### 2008-06-11 gui/src/Edge.java
removed quanto dynamically created files from bin dir and small changes to other things
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1617 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -1,220 +1,0 @@
-import java.util.ArrayList;
-import java.awt.Point;
-
-class Edge extends PLib {
-	Vertex source, dest;
-	String id;
-	ArrayList<Point> controlPoints;
-
-	public Edge(String name, Vertex source, Vertex dest) {
-		this.source = source;
-		this.dest = dest;
-		source.addEdge(this);
-		dest.addEdge(this);
-		this.id = name;
-		controlPoints = new ArrayList<Point>();
-	}
-	
-	public void display() {
-		display(false);
-	}
-	
-	public void addControlPoint(int x, int y) {
-		//Point pt = new Point(x, y);
-		//if (controlPoints.size()==0 || !pt.equals(controlPoints.get(controlPoints.size()-1)))
-		controlPoints.add(new Point(x,y));
-	}
-	
-	public void clearControlPoints() {
-		controlPoints.clear();
-	}
-	
-	public void bspline() {
-		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
-		
-		int size = controlPoints.size();
-		float[] cx = new float[size];
-		float[] cy = new float[size];
-		
-		for (int i=0;i<size;++i) {
-			cx[i] = controlPoints.get(i).x;
-			cy[i] = controlPoints.get(i).y;
-		}
-		
-		p.noFill();
-		p.beginShape();
-		p.vertex(cx[0],cy[0]);
-		float x, y, c1x, c1y, c2x, c2y, tight=0.8f;
-		for (int i=0;i<size-1;++i) {
-			x = 0.5f*(cx[i]+cx[i+1]);
-			y = 0.5f*(cy[i]+cy[i+1]);
-			
-			if (abs(cx[i]-cx[i+1])<5.0f && abs(cy[i]-cy[i+1])<5.0f) {
-				p.vertex(x, y);
-			} else {
-				c1x = x + tight*(cx[i]-x);
-				c1y = y + tight*(cy[i]-y);
-				c2x = x + tight*(cx[i+1]-x);
-				c2y = y + tight*(cy[i+1]-y);
-				/*p.stroke(0,0,255);
-				p.ellipse(c1x,c1y, 2, 2);
-				p.ellipse(c2x,c2y, 2, 2);
[…]

### 2008-06-11 gui/src/Graph.java

@@ -1,82 +1,0 @@
-import java.util.*;
-
-public class Graph extends PLib {
-	public Map<String,Vertex> vertices;
-	public ArrayList<Vertex> vertexList;
-	public Map<String,Edge> edges;
-	public List<Edge> edgeList;
-	private java.util.Random rando = new java.util.Random();
-	public Vertex newestVertex;
-
-	public Graph() {
-		vertices = new HashMap<String,Vertex>();
-		vertexList = new ArrayList<Vertex>();
-		edges = new HashMap<String,Edge>();
-		edgeList = Collections.synchronizedList(new ArrayList<Edge>());
-		newestVertex = null;
-	}
-
-	
-	// DEPRECATED
-	// terrible fresh key generator -- not to be used
-	private String freshKey(Map<String,?> m) {
-		String fresh = String.valueOf(rando.nextInt());
-		while (m.containsKey(fresh)) {
-			fresh = String.valueOf(rando.nextInt());
-		}
-		return fresh;
-	}
-
-	// DEPRECATED -- don't call this
-	// only the back-end is allowed to generate names
-	public Vertex newVertex() {
-		System.out.println("Warning: deprecated method newVertex called");
-		Vertex n = new Vertex(freshKey(vertices), 50, 50);
-		addVertex(n);
-		return n;
-	}
-
-	public Vertex addVertex(Vertex n) {
-		vertices.put(n.id, n);
-		vertexList.add(n);
-		return n;
-	}
-
-	// DEPRECATED -- don't call this
-	// only the back-end is allowed to generate names
-	public Edge newEdge(Vertex source, Vertex dest) {
-		System.out.println("Warning: deprecated method newEdge called");
-		Edge e = new Edge(freshKey(edges), source, dest);
-		addEdge(e);
-		return e;
-	}
-
-	public Edge addEdge(Edge e) {
-		edges.put(e.id, e);
-		edgeList.add(e);
-		return e;
-	}
-
-	public void layoutGraph() {
-		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
-		
-		p.layout(this);
-	}
-
-	/* this copies coords and selection info from the old graph to this one. */
-	public Graph reconcileVertices(Graph old) {
-		Vertex w = null;
-		for (Vertex v : vertexList) {
-			w = old.vertices.get(v.id);
-			if (w != null) {
-				v.x = w.x;
-				v.y = w.y;
[…]

### 2008-06-11 gui/src/PLib.java

@@ -1,2794 +1,0 @@
-/** 
- * Library gives all the static utility methods from PApplet - 
- * you can extend this to have the same env as you would normally 
- * expect from processing. 
- * 
- * @author ldixon
- *
- */
-
-import java.awt.FileDialog;
-import java.awt.Frame;
-import java.awt.Toolkit;
-import java.io.BufferedInputStream;
-import java.io.BufferedOutputStream;
-import java.io.BufferedReader;
-import java.io.ByteArrayOutputStream;
-import java.io.File;
-import java.io.FileInputStream;
-import java.io.FileOutputStream;
-import java.io.IOException;
-import java.io.InputStream;
-import java.io.InputStreamReader;
-import java.io.OutputStream;
-import java.io.OutputStreamWriter;
-import java.io.PrintWriter;
-import java.io.UnsupportedEncodingException;
-import java.lang.reflect.Array;
-import java.text.NumberFormat;
-import java.util.Arrays;
-import java.util.Calendar;
-import java.util.StringTokenizer;
-import java.util.regex.Matcher;
-import java.util.regex.Pattern;
-import java.util.zip.GZIPInputStream;
-import java.util.zip.GZIPOutputStream;
-import processing.core.PConstants;
-import processing.core.PGraphics;
-
-
-public class PLib implements PConstants {
-	
-	  /**
-	   * Full name of the Java version (i.e. 1.5.0_11).
-	   * Prior to 0125, this was only the first three digits.
-	   */
-	  public static final String javaVersionName =
-	    System.getProperty("java.version");
-
-	  /**
-	   * Version of Java that's in use, whether 1.1 or 1.3 or whatever,
-	   * stored as a float.
-	   * <P>
-	   * Note that because this is stored as a float, the values may
-	   * not be <EM>exactly</EM> 1.3 or 1.4. Instead, make sure you're
-	   * comparing against 1.3f or 1.4f, which will have the same amount
-	   * of error (i.e. 1.40000001). This could just be a double, but
-	   * since Processing only uses floats, it's safer for this to be a float
-	   * because there's no good way to specify a double with the preproc.
-	   */
-	  public static final float javaVersion =
-	    new Float(javaVersionName.substring(0, 3)).floatValue();
-
-	  /**
-	   * Current platform in use.
-	   * <P>
-	   * Equivalent to System.getProperty("os.name"), just used internally.
-	   */
-	  static public String platformName =
-	    System.getProperty("os.name");
-
-	  /**
[…]

### 2008-06-11 gui/src/QuantoApplet.java

@@ -1,368 +1,0 @@
-import java.io.BufferedReader;
-import java.io.IOException;
-import java.io.InputStreamReader;
-import java.io.OutputStreamWriter;
-import java.util.StringTokenizer;
-import javax.swing.JFileChooser;
-import javax.swing.UIManager;
-import processing.core.*;
-
-
-public class QuantoApplet extends PApplet {
-
-	private static final long serialVersionUID = 1L;
-	public static final int WIDTH = 800;
-	public static final int HEIGHT = 600;
-
-	PFont helvetica;
-	PFont times;
-
-	Vertex selectedVertex = null;
-	int selectedIndex = -1;
-	char tool;
-	JFileChooser fileChooser;
-	Graph graph;
-	boolean paused;
-	boolean doSplines=true;
-	int rectX=-1, rectY=-1;
-	boolean shift=false;
-
-	QuantoBack backend;
-	XMLReader xml;
-	static QuantoApplet p; // the top level applet 
-
-	public void setup() {
-		p = this;
-		paused = false;
-		size(WIDTH, HEIGHT, JAVA2D);
-		smooth();
-		frameRate(30);
-		try {
-			UIManager.setLookAndFeel(UIManager.getSystemLookAndFeelClassName());
-		} catch (Exception e) {
-			e.printStackTrace();
-		}
-		fileChooser = new JFileChooser();
-
-		helvetica = loadFont("HelveticaNeue-14.vlw");
-		times = loadFont("Times-Italic-14.vlw");
-
-		graph = new Graph();
-		tool = 's';
-
-		// just some testing code here
-		/*Vertex H = new Vertex("testH", 100, 100);
-		H.setColor("H");
-		graph.addVertex(H);
-
-		Vertex bnd = new Vertex("boundary", 100, 200);
-		bnd.setColor("boundary");
-		graph.addVertex(bnd);
-
-		Vertex red = new Vertex("testR", 200, 100);
-		red.setColor("red");
-		red.setAngle("\u03B1 + \u03B2");
-		graph.addVertex(red);
-
-		Vertex green = new Vertex("testG", 200, 200);
-		green.setColor("green");
-		green.setAngle("x");
-		graph.addVertex(green);*/
-
-		backend = new QuantoBack();
-		xml = new XMLReader();
-		
-	}
-	
-	public void pause() {
-		if (!paused) {
-			noLoop();
-			paused = true;
-		} 
-	}
-	
-	public void play() {
[…]

### 2008-06-11 gui/src/QuantoBack.java

@@ -1,79 +1,0 @@
-import java.io.*;
-//import java.util.Map;
-
-public class QuantoBack {
-
-	//final static String ml_command = "isabelle";
-	//final static String local_quanto_heap = "/.quantomatic/quanto";
-
-	Process backEnd;
-	BufferedReader from_backEnd;
-	BufferedReader from_backEndError;
-	BufferedWriter to_backEnd;
-
-	public QuantoBack() {
-		try {
-			//String homedir = System.getProperty("user.home");
-			//String heap = homedir + local_quanto_heap;
-			ProcessBuilder pb = new ProcessBuilder("quantomatic");
-			
-			//Map<String,String> env = pb.environment();
-			//env.put("PATH", env.get("PATH") + ":/usr/local/bin");
-			backEnd = pb.start();
-			from_backEnd = new BufferedReader(new InputStreamReader(backEnd
-					.getInputStream()));
-			from_backEndError = new BufferedReader(new InputStreamReader(backEnd
-					.getErrorStream()));
-			to_backEnd = new BufferedWriter(new OutputStreamWriter(backEnd
-					.getOutputStream()));
-			
-			System.out.println("Initialising QuantoML...");
-			System.out.println(receive());
-			
-			send("H\n"); // ask for back end status
-			/*String ln = from_backEndError.readLine();
-			while (ln != null) {
-				System.out.println("ERROR: "+ln);
-				ln = from_backEndError.readLine();
-			}*/
-			
-			// Make sure we eat up any garbage output before the status.
-			String rcv = receive();
-			while (!rcv.contains("Hello from QUANTOMATIC")) {
-				System.out.println(rcv);
-				rcv = receive();
-			}
-			
-			System.out.println("status:");
-			System.out.println(rcv); //  print it out
-		} catch (IOException e) {
-			e.printStackTrace();
-		}
-	}
-
-	public void send(String command) {
-		try {
-			to_backEnd.write(command);
-			to_backEnd.newLine();
-			to_backEnd.flush();
-		} catch (IOException e) {
-			e.printStackTrace();
-		}
-	}
-
-	public String receive() {
-		StringBuffer message = new StringBuffer();
-		try {
-			String ln = from_backEnd.readLine();
-			while (!ln.equals("stop")) {
-				message.append(ln);
-				message.append('\n');
-				ln = from_backEnd.readLine();
-			}
-		} catch (IOException e) {
-			e.printStackTrace();
-		}
-		return message.toString();
-
-	}
-}

### 2008-06-11 gui/src/QuantoFrame.java

@@ -1,26 +1,0 @@
-import java.awt.BorderLayout;
-import java.awt.Frame;
-
-
-public class QuantoFrame extends Frame {
-	private static final long serialVersionUID = 1L;
-
-	public QuantoFrame() {
-		super("Quanto");
-
-        setLayout(new BorderLayout());
-        setSize(QuantoApplet.WIDTH,QuantoApplet.HEIGHT);
-        QuantoApplet embed = new QuantoApplet();
-        add(embed, BorderLayout.CENTER);
-
-        // important to call this whenever embedding a PApplet.
-        // It ensures that the animation thread is started and
-        // that other internal variables are properly set.
-        embed.init();
-	}
-	
-	public static void main(String[] args) {
-		new QuantoFrame().setVisible(true);
-	}
-
-}

### 2008-06-11 gui/src/Vertex.java

@@ -1,163 +1,0 @@
-import java.lang.Math;
-import java.util.ArrayList;
-
-class Vertex extends PLib  {
-	public float destX, destY, x, y;
-	public float speed;
-	public boolean selected;
-	private int lastTick;
-	public String id;
-	public String col;
-	public String angleexpr;
-	private ArrayList<Edge> edges;
-	QuantoApplet p; // instance of PApplet which has processing tools
-
-	public Vertex(String id, int x, int y) {
-		this.id = id;
-		this.x = x;
-		this.y = y;
-		destX = x;
-		destY = y;
-		speed = 0.0005f;
-		col = "red";
-		lastTick = -1;
-		QuantoApplet.p.play();
-		edges = new ArrayList<Edge>();
-	}
-	
-	public void addEdge(Edge e) {
-		edges.add(e);
-	}
-	
-	public void clearEdgeControlPoints() {
-		for (Edge e : edges) e.clearControlPoints();
-	}
-
-	public void setDest(int x, int y) {
-		destX = x;
-		destY = y;
-		lastTick = -1;
-		QuantoApplet.p.play();
-	}
-
-	public void setColor(String col) {
-		this.col = col;
-	}
-
-	public void setAngle(String expr) {
-		this.angleexpr = expr;
-	}
-
-	public boolean tick() {
-		if (lastTick == -1) lastTick = millis();
-		
-		int thisTick = millis();
-		float rate = (float)(thisTick - lastTick) * speed;
-		if (rate>1) rate = 1;
-		float dx = destX - x;
-		float dy = destY - y;
-		x += dx * rate;
-		y += dy * rate;
-
-		if (floor(dx)==0 && floor(dy)==0) {
-			lastTick = -1;
-			return false;
-		} else {
-			return true;
-		}
-	}
-
-	private void displayRed() {
-		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
-
-		p.stroke(255, 0, 0);
-		p.fill(255, 100, 100);
-		p.ellipse(x, y, 15, 15);
-		if (angleexpr != null) {
-			p.textFont(p.times);
-			p.fill(100, 0, 0);
-			p.text(angleexpr, x + 21, y + 6, 30, 10);
-		}
-	}
-
-	private void displayGreen() {
[…]

### 2008-06-11 gui/src/XMLReader.java

@@ -1,144 +1,0 @@
-import processing.xml.XMLElement;
-
-/*
- * Class to interpret XML messages received from the back end
- */
-public class XMLReader {
-
-	private Graph g = null;
-
-	class XMLReaderException extends RuntimeException {
-		private static final long serialVersionUID = 1L;
-
-		XMLReaderException(String m) {
-			super(m);
-		}
-	}
-
-	public Graph parseGraph(String s) {
-		return parseGraph(mkXMLFromString(s));
-	}
-
-	public Graph parseGraph(XMLElement xml) {
-		assertElement(xml, "graph");
-
-		g = new Graph();
-
-		// assume that the only interesting thing in a graph
-		// are its nodes and edges -- probably need to modify
-		// for !-boxes
-		XMLElement[] vertices = xml.getChildren("vertex");
-		for (int i = 0; i < vertices.length; i++) {
-			g.addVertex(parseVertex(vertices[i]));
-		}
-		XMLElement[] edges = xml.getChildren("edge");
-		for (int i = 0; i < edges.length; i++) {
-			g.addEdge(parseEdge(edges[i]));
-		}
-		return g;
-	}
-
-	public Vertex parseVertex(String s) {
-		return parseVertex(mkXMLFromString(s));
-	}
-
-	public Vertex parseVertex(XMLElement xml) {
-		assertElement(xml, "vertex");
-
-		String name = getChildContents(xml, "name", "ERROR", true);
-		Vertex v = new Vertex(name, 0, 0);
-
-		String bnd = getChildContents(xml, "boundary", "false");
-		if (bnd.equals("true")) {
-			v.setColor("boundary");
-			// no need to set any other properties
-		} else if (bnd.equals("false")) {
-			String col = getChildContents(xml, "colour", "ERROR", true);
-			if (col.equals("H")) {
-				v.setColor(col);
-			} else if (col.equals("red") || col.equals("green")) {
-				v.setColor(col);
-				// non-H vertices have angle exprs
-				XMLElement angle_expr = xml.getChild("angleexpr");
-				if (angle_expr != null)
-					v.setAngle(parseAngleExpr(angle_expr));
-				else
-					v.setAngle("");
-			} else { // colour was not valid
-				badContents("colour", col);
-			}
-		} else { // boundary is neither true nor false!
-			badContents("boundary", bnd);
-		}
-		return v;
-	}
-
-	public Edge parseEdge(String s) {
[…]

### 2008-06-11 gui/src/Edge.java
re-import of GUI code
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1624 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -1,0 +1,220 @@
+import java.util.ArrayList;
+import java.awt.Point;
+
+class Edge extends PLib {
+	Vertex source, dest;
+	String id;
+	ArrayList<Point> controlPoints;
+
+	public Edge(String name, Vertex source, Vertex dest) {
+		this.source = source;
+		this.dest = dest;
+		source.addEdge(this);
+		dest.addEdge(this);
+		this.id = name;
+		controlPoints = new ArrayList<Point>();
+	}
+	
+	public void display() {
+		display(false);
+	}
+	
+	public void addControlPoint(int x, int y) {
+		//Point pt = new Point(x, y);
+		//if (controlPoints.size()==0 || !pt.equals(controlPoints.get(controlPoints.size()-1)))
+		controlPoints.add(new Point(x,y));
+	}
+	
+	public void clearControlPoints() {
+		controlPoints.clear();
+	}
+	
+	public void bspline() {
+		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
+		
+		int size = controlPoints.size();
+		float[] cx = new float[size];
+		float[] cy = new float[size];
+		
+		for (int i=0;i<size;++i) {
+			cx[i] = controlPoints.get(i).x;
+			cy[i] = controlPoints.get(i).y;
+		}
+		
+		p.noFill();
+		p.beginShape();
+		p.vertex(cx[0],cy[0]);
+		float x, y, c1x, c1y, c2x, c2y, tight=0.8f;
+		for (int i=0;i<size-1;++i) {
+			x = 0.5f*(cx[i]+cx[i+1]);
+			y = 0.5f*(cy[i]+cy[i+1]);
+			
+			if (abs(cx[i]-cx[i+1])<5.0f && abs(cy[i]-cy[i+1])<5.0f) {
+				p.vertex(x, y);
+			} else {
+				c1x = x + tight*(cx[i]-x);
+				c1y = y + tight*(cy[i]-y);
+				c2x = x + tight*(cx[i+1]-x);
+				c2y = y + tight*(cy[i+1]-y);
+				/*p.stroke(0,0,255);
+				p.ellipse(c1x,c1y, 2, 2);
+				p.ellipse(c2x,c2y, 2, 2);
[…]

### 2008-06-11 gui/src/Graph.java

@@ -1,0 +1,82 @@
+import java.util.*;
+
+public class Graph extends PLib {
+	public Map<String,Vertex> vertices;
+	public ArrayList<Vertex> vertexList;
+	public Map<String,Edge> edges;
+	public List<Edge> edgeList;
+	private java.util.Random rando = new java.util.Random();
+	public Vertex newestVertex;
+
+	public Graph() {
+		vertices = new HashMap<String,Vertex>();
+		vertexList = new ArrayList<Vertex>();
+		edges = new HashMap<String,Edge>();
+		edgeList = Collections.synchronizedList(new ArrayList<Edge>());
+		newestVertex = null;
+	}
+
+	
+	// DEPRECATED
+	// terrible fresh key generator -- not to be used
+	private String freshKey(Map<String,?> m) {
+		String fresh = String.valueOf(rando.nextInt());
+		while (m.containsKey(fresh)) {
+			fresh = String.valueOf(rando.nextInt());
+		}
+		return fresh;
+	}
+
+	// DEPRECATED -- don't call this
+	// only the back-end is allowed to generate names
+	public Vertex newVertex() {
+		System.out.println("Warning: deprecated method newVertex called");
+		Vertex n = new Vertex(freshKey(vertices), 50, 50);
+		addVertex(n);
+		return n;
+	}
+
+	public Vertex addVertex(Vertex n) {
+		vertices.put(n.id, n);
+		vertexList.add(n);
+		return n;
+	}
+
+	// DEPRECATED -- don't call this
+	// only the back-end is allowed to generate names
+	public Edge newEdge(Vertex source, Vertex dest) {
+		System.out.println("Warning: deprecated method newEdge called");
+		Edge e = new Edge(freshKey(edges), source, dest);
+		addEdge(e);
+		return e;
+	}
+
+	public Edge addEdge(Edge e) {
+		edges.put(e.id, e);
+		edgeList.add(e);
+		return e;
+	}
+
+	public void layoutGraph() {
+		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
+		
+		p.layout(this);
+	}
+
+	/* this copies coords and selection info from the old graph to this one. */
+	public Graph reconcileVertices(Graph old) {
+		Vertex w = null;
+		for (Vertex v : vertexList) {
+			w = old.vertices.get(v.id);
+			if (w != null) {
+				v.x = w.x;
+				v.y = w.y;
[…]

### 2008-06-11 gui/src/QuantoApplet.java

@@ -1,0 +1,368 @@
+import java.io.BufferedReader;
+import java.io.IOException;
+import java.io.InputStreamReader;
+import java.io.OutputStreamWriter;
+import java.util.StringTokenizer;
+import javax.swing.JFileChooser;
+import javax.swing.UIManager;
+import processing.core.*;
+
+
+public class QuantoApplet extends PApplet {
+
+	private static final long serialVersionUID = 1L;
+	public static final int WIDTH = 800;
+	public static final int HEIGHT = 600;
+
+	PFont helvetica;
+	PFont times;
+
+	Vertex selectedVertex = null;
+	int selectedIndex = -1;
+	char tool;
+	JFileChooser fileChooser;
+	Graph graph;
+	boolean paused;
+	boolean doSplines=true;
+	int rectX=-1, rectY=-1;
+	boolean shift=false;
+
+	QuantoBack backend;
+	XMLReader xml;
+	static QuantoApplet p; // the top level applet 
+
+	public void setup() {
+		p = this;
+		paused = false;
+		size(WIDTH, HEIGHT, JAVA2D);
+		smooth();
+		frameRate(30);
+		try {
+			UIManager.setLookAndFeel(UIManager.getSystemLookAndFeelClassName());
+		} catch (Exception e) {
+			e.printStackTrace();
+		}
+		fileChooser = new JFileChooser();
+
+		helvetica = loadFont("HelveticaNeue-14.vlw");
+		times = loadFont("Times-Italic-14.vlw");
+
+		graph = new Graph();
+		tool = 's';
+
+		// just some testing code here
+		/*Vertex H = new Vertex("testH", 100, 100);
+		H.setColor("H");
+		graph.addVertex(H);
+
+		Vertex bnd = new Vertex("boundary", 100, 200);
+		bnd.setColor("boundary");
+		graph.addVertex(bnd);
+
+		Vertex red = new Vertex("testR", 200, 100);
+		red.setColor("red");
+		red.setAngle("\u03B1 + \u03B2");
+		graph.addVertex(red);
+
+		Vertex green = new Vertex("testG", 200, 200);
+		green.setColor("green");
+		green.setAngle("x");
+		graph.addVertex(green);*/
+
+		backend = new QuantoBack();
+		xml = new XMLReader();
+		
+	}
+	
+	public void pause() {
+		if (!paused) {
+			noLoop();
+			paused = true;
+		} 
+	}
+	
+	public void play() {
[…]

### 2008-06-11 gui/src/QuantoBack.java

@@ -1,0 +1,79 @@
+import java.io.*;
+//import java.util.Map;
+
+public class QuantoBack {
+
+	//final static String ml_command = "isabelle";
+	//final static String local_quanto_heap = "/.quantomatic/quanto";
+
+	Process backEnd;
+	BufferedReader from_backEnd;
+	BufferedReader from_backEndError;
+	BufferedWriter to_backEnd;
+
+	public QuantoBack() {
+		try {
+			//String homedir = System.getProperty("user.home");
+			//String heap = homedir + local_quanto_heap;
+			ProcessBuilder pb = new ProcessBuilder("quantomatic");
+			
+			//Map<String,String> env = pb.environment();
+			//env.put("PATH", env.get("PATH") + ":/usr/local/bin");
+			backEnd = pb.start();
+			from_backEnd = new BufferedReader(new InputStreamReader(backEnd
+					.getInputStream()));
+			from_backEndError = new BufferedReader(new InputStreamReader(backEnd
+					.getErrorStream()));
+			to_backEnd = new BufferedWriter(new OutputStreamWriter(backEnd
+					.getOutputStream()));
+			
+			System.out.println("Initialising QuantoML...");
+			System.out.println(receive());
+			
+			send("H\n"); // ask for back end status
+			/*String ln = from_backEndError.readLine();
+			while (ln != null) {
+				System.out.println("ERROR: "+ln);
+				ln = from_backEndError.readLine();
+			}*/
+			
+			// Make sure we eat up any garbage output before the status.
+			String rcv = receive();
+			while (!rcv.contains("Hello from QUANTOMATIC")) {
+				System.out.println(rcv);
+				rcv = receive();
+			}
+			
+			System.out.println("status:");
+			System.out.println(rcv); //  print it out
+		} catch (IOException e) {
+			e.printStackTrace();
+		}
+	}
+
+	public void send(String command) {
+		try {
+			to_backEnd.write(command);
+			to_backEnd.newLine();
+			to_backEnd.flush();
+		} catch (IOException e) {
+			e.printStackTrace();
+		}
+	}
+
+	public String receive() {
+		StringBuffer message = new StringBuffer();
+		try {
+			String ln = from_backEnd.readLine();
+			while (!ln.equals("stop")) {
+				message.append(ln);
+				message.append('\n');
+				ln = from_backEnd.readLine();
+			}
+		} catch (IOException e) {
+			e.printStackTrace();
+		}
+		return message.toString();
+
+	}
+}

### 2008-06-11 gui/src/Vertex.java

@@ -1,0 +1,163 @@
+import java.lang.Math;
+import java.util.ArrayList;
+
+class Vertex extends PLib  {
+	public float destX, destY, x, y;
+	public float speed;
+	public boolean selected;
+	private int lastTick;
+	public String id;
+	public String col;
+	public String angleexpr;
+	private ArrayList<Edge> edges;
+	QuantoApplet p; // instance of PApplet which has processing tools
+
+	public Vertex(String id, int x, int y) {
+		this.id = id;
+		this.x = x;
+		this.y = y;
+		destX = x;
+		destY = y;
+		speed = 0.0005f;
+		col = "red";
+		lastTick = -1;
+		QuantoApplet.p.play();
+		edges = new ArrayList<Edge>();
+	}
+	
+	public void addEdge(Edge e) {
+		edges.add(e);
+	}
+	
+	public void clearEdgeControlPoints() {
+		for (Edge e : edges) e.clearControlPoints();
+	}
+
+	public void setDest(int x, int y) {
+		destX = x;
+		destY = y;
+		lastTick = -1;
+		QuantoApplet.p.play();
+	}
+
+	public void setColor(String col) {
+		this.col = col;
+	}
+
+	public void setAngle(String expr) {
+		this.angleexpr = expr;
+	}
+
+	public boolean tick() {
+		if (lastTick == -1) lastTick = millis();
+		
+		int thisTick = millis();
+		float rate = (float)(thisTick - lastTick) * speed;
+		if (rate>1) rate = 1;
+		float dx = destX - x;
+		float dy = destY - y;
+		x += dx * rate;
+		y += dy * rate;
+
+		if (floor(dx)==0 && floor(dy)==0) {
+			lastTick = -1;
+			return false;
+		} else {
+			return true;
+		}
+	}
+
+	private void displayRed() {
+		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
+
+		p.stroke(255, 0, 0);
+		p.fill(255, 100, 100);
+		p.ellipse(x, y, 15, 15);
+		if (angleexpr != null) {
+			p.textFont(p.times);
+			p.fill(100, 0, 0);
+			p.text(angleexpr, x + 21, y + 6, 30, 10);
+		}
+	}
+
+	private void displayGreen() {
[…]

### 2008-06-11 gui/src/XMLReader.java

@@ -1,0 +1,144 @@
+import processing.xml.XMLElement;
+
+/*
+ * Class to interpret XML messages received from the back end
+ */
+public class XMLReader {
+
+	private Graph g = null;
+
+	class XMLReaderException extends RuntimeException {
+		private static final long serialVersionUID = 1L;
+
+		XMLReaderException(String m) {
+			super(m);
+		}
+	}
+
+	public Graph parseGraph(String s) {
+		return parseGraph(mkXMLFromString(s));
+	}
+
+	public Graph parseGraph(XMLElement xml) {
+		assertElement(xml, "graph");
+
+		g = new Graph();
+
+		// assume that the only interesting thing in a graph
+		// are its nodes and edges -- probably need to modify
+		// for !-boxes
+		XMLElement[] vertices = xml.getChildren("vertex");
+		for (int i = 0; i < vertices.length; i++) {
+			g.addVertex(parseVertex(vertices[i]));
+		}
+		XMLElement[] edges = xml.getChildren("edge");
+		for (int i = 0; i < edges.length; i++) {
+			g.addEdge(parseEdge(edges[i]));
+		}
+		return g;
+	}
+
+	public Vertex parseVertex(String s) {
+		return parseVertex(mkXMLFromString(s));
+	}
+
+	public Vertex parseVertex(XMLElement xml) {
+		assertElement(xml, "vertex");
+
+		String name = getChildContents(xml, "name", "ERROR", true);
+		Vertex v = new Vertex(name, 0, 0);
+
+		String bnd = getChildContents(xml, "boundary", "false");
+		if (bnd.equals("true")) {
+			v.setColor("boundary");
+			// no need to set any other properties
+		} else if (bnd.equals("false")) {
+			String col = getChildContents(xml, "colour", "ERROR", true);
+			if (col.equals("H")) {
+				v.setColor(col);
+			} else if (col.equals("red") || col.equals("green")) {
+				v.setColor(col);
+				// non-H vertices have angle exprs
+				XMLElement angle_expr = xml.getChild("angleexpr");
+				if (angle_expr != null)
+					v.setAngle(parseAngleExpr(angle_expr));
+				else
+					v.setAngle("");
+			} else { // colour was not valid
+				badContents("colour", col);
+			}
+		} else { // boundary is neither true nor false!
+			badContents("boundary", bnd);
+		}
+		return v;
+	}
+
+	public Edge parseEdge(String s) {
[…]

### 2008-06-11 gui/src/QuantoApplet.java
Added support for saving PDF files of the window; just press f
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1625 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -6,6 +6,7 @@ import java.util.StringTokenizer;
 import javax.swing.JFileChooser;
 import javax.swing.UIManager;
 import processing.core.*;
+import processing.pdf.*;
 
 
 public class QuantoApplet extends PApplet {
@@ -20,13 +21,22 @@ public class QuantoApplet extends PApplet {
 	Vertex selectedVertex = null;
 	int selectedIndex = -1;
 	char tool;
-	JFileChooser fileChooser;
 	Graph graph;
 	boolean paused;
 	boolean doSplines=true;
 	int rectX=-1, rectY=-1;
 	boolean shift=false;
 
+	boolean saveNextFrame = false;
+	int nextPDFFile = 1;
+	String pdfDirName = "";
+	JFileChooser fileChooser;
+
+	String getNextPDFFileName() { 
+		return pdfDirName + "/quanto-frame-" + (nextPDFFile++) + ".pdf";		
+	}
+	
+	
 	QuantoBack backend;
 	XMLReader xml;
 	static QuantoApplet p; // the top level applet 
@@ -196,6 +206,16 @@ public class QuantoApplet extends PApplet {
 		case 'p':
 			doSplines = !doSplines;
 			break;
+		case 'f':  // we're going to dump the screen to PDF
+			fileChooser.setFileSelectionMode(JFileChooser.DIRECTORIES_ONLY);
+			fileChooser.setDialogTitle("Choose the dump directory");
+			fileChooser.setMultiSelectionEnabled(false);
+			int retVal = fileChooser.showOpenDialog(this);
+			if(retVal == JFileChooser.APPROVE_OPTION){
+				saveNextFrame = true;
+				pdfDirName = fileChooser.getSelectedFile().getAbsolutePath();
+			}
+			break;
 		case CODED:
 			if (keyCode == SHIFT) shift = true;
 			break;
@@ -243,6 +263,11 @@ public class QuantoApplet extends PApplet {
 			break;
 		}
 
+		//	if we are going to save this frame start recording after the 
+		// interface fluff has been drawn
+		if ( saveNextFrame ) {
+			p.beginRecord(p.PDF, getNextPDFFileName());
+		}
 		
 		boolean moved = false;
 		for (Vertex v : graph.vertexList) {
@@ -265,8 +290,15 @@ public class QuantoApplet extends PApplet {
 			stroke(100,100,255);
[…]

### 2008-06-11 gui/src/QuantoApplet.java
added support for QT movies.  Press v to start and stop recording.  Also changed the key for PDF to c
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1626 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -7,6 +7,7 @@ import javax.swing.JFileChooser;
 import javax.swing.UIManager;
 import processing.core.*;
 import processing.pdf.*;
+import processing.video.*;
 
 
 public class QuantoApplet extends PApplet {
@@ -27,16 +28,18 @@ public class QuantoApplet extends PApplet {
 	int rectX=-1, rectY=-1;
 	boolean shift=false;
 
-	boolean saveNextFrame = false;
+	String outDirName = "";
+	JFileChooser fileChooser;	
+	
+	String nextPDFFileName = "";
 	int nextPDFFile = 1;
-	String pdfDirName = "";
-	JFileChooser fileChooser;
+	boolean saveNextFrame = false;
+	
+	String nextQTFileName = "";
+	int nextQTFile = 1;
+	boolean recordingVideo = false;
+	MovieMaker mm;  // to be initialised when recording starts
 
-	String getNextPDFFileName() { 
-		return pdfDirName + "/quanto-frame-" + (nextPDFFile++) + ".pdf";		
-	}
-	
-	
 	QuantoBack backend;
 	XMLReader xml;
 	static QuantoApplet p; // the top level applet 
@@ -206,15 +209,12 @@ public class QuantoApplet extends PApplet {
 		case 'p':
 			doSplines = !doSplines;
 			break;
-		case 'f':  // we're going to dump the screen to PDF
-			fileChooser.setFileSelectionMode(JFileChooser.DIRECTORIES_ONLY);
-			fileChooser.setDialogTitle("Choose the dump directory");
-			fileChooser.setMultiSelectionEnabled(false);
-			int retVal = fileChooser.showOpenDialog(this);
-			if(retVal == JFileChooser.APPROVE_OPTION){
-				saveNextFrame = true;
-				pdfDirName = fileChooser.getSelectedFile().getAbsolutePath();
-			}
+		case 'c':  // "capture" the screen to PDF
+			startSaveNextFrame();
+			break;		
+		case 'v': // v is for video
+			if (recordingVideo) stopRecordVideo();
+			else startRecordVideo();
 			break;
 		case CODED:
 			if (keyCode == SHIFT) shift = true;
@@ -224,6 +224,7 @@ public class QuantoApplet extends PApplet {
 		case 'e': /* these are tools that require mouse input too */
 			tool = key;
 			break;
+
 		}
 		
 		play();
[…]

### 2008-06-11 gui/src/Edge.java
reasonable curves
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1627 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -44,25 +44,13 @@ class Edge extends PLib {
 		p.noFill();
 		p.beginShape();
 		p.vertex(cx[0],cy[0]);
-		float x, y, c1x, c1y, c2x, c2y, tight=0.8f;
+		int x, y, c1x, c1y, c2x, c2y;
+		float tight=1.0f;
 		for (int i=0;i<size-1;++i) {
-			x = 0.5f*(cx[i]+cx[i+1]);
-			y = 0.5f*(cy[i]+cy[i+1]);
+			x = floor(0.5f*(cx[i]+cx[i+1]));
+			y = floor(0.5f*(cy[i]+cy[i+1]));
 			
-			if (abs(cx[i]-cx[i+1])<5.0f && abs(cy[i]-cy[i+1])<5.0f) {
-				p.vertex(x, y);
-			} else {
-				c1x = x + tight*(cx[i]-x);
-				c1y = y + tight*(cy[i]-y);
-				c2x = x + tight*(cx[i+1]-x);
-				c2y = y + tight*(cy[i+1]-y);
-				/*p.stroke(0,0,255);
-				p.ellipse(c1x,c1y, 2, 2);
-				p.ellipse(c2x,c2y, 2, 2);
-				p.stroke(255,0,0);
-				p.ellipse(0.5f*(cx[i]+cx[i+1]), 0.5f*(cy[i]+cy[i+1]), 3, 3);*/
-				p.bezierVertex(c1x,c1y,c2x,c2y,x,y);
-			}
+			p.curveVertex(x, y);
 		}
 		p.vertex(cx[size-1], cy[size-1]);
 		p.stroke(0);

### 2008-06-12 gui/src/Edge.java
back to b-splines, also, added some nice tabbing
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1628 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -33,24 +33,28 @@ class Edge extends PLib {
 		QuantoApplet p = QuantoApplet.p; // instance of PApplet which has all processing tools
 		
 		int size = controlPoints.size();
-		float[] cx = new float[size];
-		float[] cy = new float[size];
+		float[] cx = new float[size+2];
+		float[] cy = new float[size+2];
 		
 		for (int i=0;i<size;++i) {
 			cx[i] = controlPoints.get(i).x;
 			cy[i] = controlPoints.get(i).y;
 		}
+		cx[size] = controlPoints.get(size-1).x;
+		cy[size] = controlPoints.get(size-1).y;
+		cx[size+1] = cx[size];
+		cy[size+1] = cy[size];
 		
-		p.noFill();
+		/*p.noFill();
 		p.beginShape();
 		p.vertex(cx[0],cy[0]);
-		int x, y, c1x, c1y, c2x, c2y;
-		float tight=1.0f;
+		int x, y; //, c1x, c1y, c2x, c2y;
+		//float tight=1.0f;
 		for (int i=0;i<size-1;++i) {
 			x = floor(0.5f*(cx[i]+cx[i+1]));
 			y = floor(0.5f*(cy[i]+cy[i+1]));
-			
-			p.curveVertex(x, y);
+			p.vertex(x,y);
+			//p.curveVertex(x, y);
 		}
 		p.vertex(cx[size-1], cy[size-1]);
 		p.stroke(0);
@@ -94,11 +98,10 @@ class Edge extends PLib {
 		
 		
 		// CUBIC B-SPLINE
-		/*
 		float tx3, tx2, tx1, tx0, ty3, ty2, ty1, ty0;
 		float x, y, lastX=cx[0], lastY=cy[0];
 		for (int i=1; i<cx.length-2; ++i) {
-		  p.ellipse(cx[i], cy[i], 3, 3);
+		  //p.ellipse(cx[i], cy[i], 3, 3);
 		  tx3 = 0.1666666667f * (-cx[i-1] + 3f * cx[i] - 3f * cx[i+1] + cx[i+2]);
 		  tx2 = 0.1666666667f * (3f * cx[i-1] - 6f * cx[i] + 3f * cx[i+1]);
 		  tx1 = 0.1666666667f * (-3f * cx[i-1] + 3f * cx[i+1]);
@@ -120,7 +123,7 @@ class Edge extends PLib {
 		    lastY = y;
[…]

### 2008-06-12 gui/src/Graph.java

@@ -2,7 +2,7 @@ import java.util.*;
 
 public class Graph extends PLib {
 	public Map<String,Vertex> vertices;
-	public ArrayList<Vertex> vertexList;
+	public List<Vertex> vertexList;
 	public Map<String,Edge> edges;
 	public List<Edge> edgeList;
 	private java.util.Random rando = new java.util.Random();
@@ -10,7 +10,7 @@ public class Graph extends PLib {
 
 	public Graph() {
 		vertices = new HashMap<String,Vertex>();
-		vertexList = new ArrayList<Vertex>();
+		vertexList = Collections.synchronizedList(new ArrayList<Vertex>());
 		edges = new HashMap<String,Edge>();
 		edgeList = Collections.synchronizedList(new ArrayList<Edge>());
 		newestVertex = null;
@@ -77,6 +77,10 @@ public class Graph extends PLib {
 			}
 		}
 		return this;
+	}
+	
+	public void sortVertices() {
+		Collections.sort(vertexList);
 	}
 
 }

### 2008-06-12 gui/src/QuantoApplet.java

@@ -48,6 +48,7 @@ public class QuantoApplet extends PApplet {
 		p = this;
 		paused = false;
 		size(WIDTH, HEIGHT, JAVA2D);
+		
 		smooth();
 		frameRate(30);
 		try {
@@ -219,6 +220,26 @@ public class QuantoApplet extends PApplet {
 		case CODED:
 			if (keyCode == SHIFT) shift = true;
 			break;
+		case TAB:
+			synchronized(graph.vertexList) {
+				if (graph.vertexList.size()>0) {
+					graph.sortVertices();
+					Vertex sel = null;
+					boolean pickNext = false;
+					for (Vertex v : graph.vertexList) {
+						if (v.selected) {
+							pickNext = true;
+						} else if (pickNext) {
+							sel = v;
+							pickNext = false;
+						}
+						v.selected = false;
+					}
+					if (sel==null) sel = graph.vertexList.get(0);
+					sel.selected = true;
+				}
+			}
+			break;
 		case 's':
 		case 'm':
 		case 'e': /* these are tools that require mouse input too */
@@ -272,10 +293,13 @@ public class QuantoApplet extends PApplet {
 		}
 		
 		boolean moved = false;
-		for (Vertex v : graph.vertexList) {
-			moved = moved || v.tick();
-			v.tick();
-			v.display();
+		
+		synchronized(graph.vertexList) {
+			for (Vertex v : graph.vertexList) {
+				moved = moved || v.tick();
+				v.tick();
+				v.display();
+			}
 		}
 		
 		synchronized(graph.edgeList) {
@@ -364,8 +388,8 @@ public class QuantoApplet extends PApplet {
 						n = new Vertex(name, 50, 50);
 						graph.addVertex(n);
 					}
-					x = (int) (Float.parseFloat(tk.nextToken()) * 50.0) + 20;
-					y = (int) (Float.parseFloat(tk.nextToken()) * 50.0) + 20;
+					x = (int) (Float.parseFloat(tk.nextToken()) * 50.0) + 40;
+					y = (int) (Float.parseFloat(tk.nextToken()) * 50.0) + 40;
 
 					tk.nextToken();
 					tk.nextToken();
@@ -385,8 +409,8 @@ public class QuantoApplet extends PApplet {
 						int controlCount = Integer.parseInt(tk.nextToken());
 						
 						for (int i=0;i<controlCount;++i) {
[…]

### 2008-06-12 gui/src/Vertex.java

@@ -1,7 +1,7 @@
 import java.lang.Math;
 import java.util.ArrayList;
 
-class Vertex extends PLib  {
+class Vertex extends PLib implements Comparable<Vertex>  {
 	public float destX, destY, x, y;
 	public float speed;
 	public boolean selected;
@@ -159,5 +159,12 @@ class Vertex extends PLib  {
 			return true;
 		else
 			return false;
+	}
+	
+	// a lexicographic order on y, x (with a minor fudge factor)
+	public int compareTo(Vertex v) {
+		if (v.x==x && v.y==y) return 0;
+		else if (abs(v.y-y)<4) return (v.x<x) ? 1 : -1;
+		else return (v.y<y) ? -1 : 1;
 	}
 }

### 2008-06-13 gui/src/Graph.java
added snap-to-grid and a nifty menu bar
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1630 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -7,6 +7,8 @@ public class Graph extends PLib {
 	public List<Edge> edgeList;
 	private java.util.Random rando = new java.util.Random();
 	public Vertex newestVertex;
+	public static final int GRID_X = 25;
+	public static final int GRID_Y = 25;
 
 	public Graph() {
 		vertices = new HashMap<String,Vertex>();
@@ -63,7 +65,7 @@ public class Graph extends PLib {
 		p.layout(this);
 	}
 
-	/* this copies coords and selection info from the old graph to this one. */
+	/* this copies coordinates and selection info from the old graph to this one. */
 	public Graph reconcileVertices(Graph old) {
 		Vertex w = null;
 		for (Vertex v : vertexList) {
@@ -77,6 +79,21 @@ public class Graph extends PLib {
 			}
 		}
 		return this;
+	}
+	
+	public void enableSnap() {
+		synchronized(vertexList) {
+			for (Vertex v : vertexList) {
+				v.snap=true;
+				v.setDest(v.x, v.y); // force the vertex to move to the snapped position
+			}
+		}
+	}
+	
+	public void disableSnap() {
+		synchronized(vertexList) {
+			for (Vertex v : vertexList) v.snap=false;
+		}
 	}
 	
 	public void sortVertices() {

### 2008-06-13 gui/src/QuantoApplet.java

@@ -27,6 +27,7 @@ public class QuantoApplet extends PApplet {
 	boolean doSplines=true;
 	int rectX=-1, rectY=-1;
 	boolean shift=false;
+	boolean snapToGrid=false;
 
 	String outDirName = "";
 	JFileChooser fileChooser;	
@@ -114,20 +115,24 @@ public class QuantoApplet extends PApplet {
 	}
 
 	public void mousePressed() {
-		Vertex n;
 		switch (tool) {
 		case 's':
 			rectX = mouseX;
 			rectY = mouseY;
 			selectedVertex = null;
-			// IMPROVE: use tree of locations for sub-object matching:
-			// get log-time search for finding object from coordinates instead
-			// of linear time.
-			for (int i = 0; i < graph.vertexList.size(); ++i) {
-				n = (Vertex) graph.vertexList.get(i);
-				if (!shift) n.selected = false;
-				if (n.at(mouseX, mouseY)) {
-					n.selected = !n.selected;
+			/*
+			 * we wish to only toggle the selection of one node, so toggle
+			 * actionPerformed when that is done.
+			 */
+			boolean actionPerformed = false;
+			synchronized(graph.vertexList) {
+				for (Vertex v : graph.vertexList) {
+					if (!actionPerformed && v.at(mouseX, mouseY)) {
+						actionPerformed = true;
+						v.selected = !v.selected;
+					} else if (!shift) {
+						v.selected = false;
+					}
 				}
 			}
 			break;
@@ -240,6 +245,11 @@ public class QuantoApplet extends PApplet {
 				}
 			}
 			break;
+		case 'G':
+			snapToGrid = !snapToGrid;
+			if (snapToGrid) graph.enableSnap();
+			else graph.disableSnap();
+			break;
 		case 's':
 		case 'm':
 		case 'e': /* these are tools that require mouse input too */
@@ -258,12 +268,13 @@ public class QuantoApplet extends PApplet {
 	void modifyGraph(String cmd) {
 		backend.send(cmd + "\n");
 		println(backend.receive());
-		// here send D to backend and dump the graph
+		// here send D to back-end and dump the graph
 		// then rebuild it via the XML parser.
 		backend.send("D\n");
[…]

### 2008-06-13 gui/src/QuantoFrame.java

@@ -1,8 +1,9 @@
-import java.awt.BorderLayout;
-import java.awt.Frame;
+import java.awt.*;
+import java.awt.event.*;
+import processing.core.PConstants;
 
 
-public class QuantoFrame extends Frame {
+public class QuantoFrame extends Frame implements PConstants {
 	private static final long serialVersionUID = 1L;
 
 	public QuantoFrame() {
@@ -10,13 +11,75 @@ public class QuantoFrame extends Frame {
 
         setLayout(new BorderLayout());
         setSize(QuantoApplet.WIDTH,QuantoApplet.HEIGHT);
-        QuantoApplet embed = new QuantoApplet();
+        
+        final QuantoApplet embed = new QuantoApplet();
         add(embed, BorderLayout.CENTER);
 
         // important to call this whenever embedding a PApplet.
         // It ensures that the animation thread is started and
         // that other internal variables are properly set.
         embed.init();
+        
+        MenuBar mb = new MenuBar();
+        Menu tools = new Menu("Tools");
+        Menu actions = new Menu("Actions");
+        mb.add(tools);
+        mb.add(actions);
+        
+        final String[] toolNames = {"Select (s)", "Move (m)", "Edge (e)"};
+        final char[] toolKeys = {'s', 'm', 'e'};
+        
+        final String[] actionNames = {
+        		"New graph (n)",
+        		"Undo (u)",
+        		"Layout (l)",
+        		"add Red (r)",
+        		"add Green (g)",
+        		"add Hadamard (h)",
+        		"add Boundary (b)",
+        		"Delete vertex (d)",
+        		"toggle sPlines (p)",
+        		"Capture to pdf (c)",
+        		"start/stop Video (v)",
+        		"toggle snap to Grid (G)",
+        		"quit (q)"
+        	};
+        final char[] actionKeys = {
+        		'n', 'u', 'l', 'r', 'g', 'h', 'b',
+        		'd', 'p', 'c', 'v', 'G', 'q'
+        	};
+        
+        MenuItem mi;
+        
+        for (int i=0;i<toolNames.length;++i) {
+        	// we need to ensure this char stays allocated...
+        	final char c = toolKeys[i];
+        	mi = new MenuItem(toolNames[i]);
+        	mi.addActionListener(new ActionListener() {
+				public void actionPerformed(ActionEvent e) {
+					embed.key = c;
+					embed.keyPressed();
+				}
+        		
+        	});
+        	tools.add(mi);
+        }
[…]

### 2008-06-13 gui/src/Vertex.java

@@ -10,6 +10,7 @@ class Vertex extends PLib implements Comparable<Vertex>  {
 	public String col;
 	public String angleexpr;
 	private ArrayList<Edge> edges;
+	public boolean snap;
 	QuantoApplet p; // instance of PApplet which has processing tools
 
 	public Vertex(String id, int x, int y) {
@@ -23,6 +24,7 @@ class Vertex extends PLib implements Comparable<Vertex>  {
 		lastTick = -1;
 		QuantoApplet.p.play();
 		edges = new ArrayList<Edge>();
+		snap = false;
 	}
 	
 	public void addEdge(Edge e) {
@@ -32,10 +34,19 @@ class Vertex extends PLib implements Comparable<Vertex>  {
 	public void clearEdgeControlPoints() {
 		for (Edge e : edges) e.clearControlPoints();
 	}
-
+	
 	public void setDest(int x, int y) {
-		destX = x;
-		destY = y;
+		setDest((float)x,(float)y);
+	}
+	
+	public void setDest(float x, float y) {
+		if (snap) {
+			destX = (Graph.GRID_X * round(x/(float)Graph.GRID_X));
+			destY = (Graph.GRID_Y * round(y/(float)Graph.GRID_Y));
+		} else {
+			destX = x;
+			destY = y;
+		}
 		lastTick = -1;
 		QuantoApplet.p.play();
 	}

### 2008-06-13 gui/src/Graph.java
added latex and dot to clipboard
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1631 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -97,7 +97,81 @@ public class Graph extends PLib {
 	}
 	
 	public void sortVertices() {
-		Collections.sort(vertexList);
+		synchronized(vertexList) {
+			Collections.sort(vertexList);
+		}
+	}
+	
+	String toDot() {
+		StringBuffer g = new StringBuffer();
+		g.append("digraph { nodesep=0.65; ranksep=0.65;\n");
+
+		for (Vertex v : vertexList) {
+			g.append(v.id);
+			g.append(" [color=\"");
+			g.append(v.col);
+			g.append("\",label=\"\",width=0.35,height=0.35,shape=circle];\n");
+		}
+
+		for (Edge e : edgeList) {
+			g.append(e.source.id);
+			g.append("->");
+			g.append(e.dest.id);
+			//g.append(" [arrowhead=none,headclip=false,tailclip=false];\n");
+			g.append(" [arrowhead=none];\n");
+		}
+
+		g.append("\n}\n");
+		return g.toString();
+	}
+	
+	String toLatex() {
+		if (vertexList.size()==0) return "\\begin{tikzpicture}\n\\end{tikzpicture}\n";
+		StringBuffer g = new StringBuffer();
+		g.append("\\begin{tikzpicture}\n");
+		sortVertices();
+		float xOrigin = vertexList.get(0).x;
+		float yOrigin = vertexList.get(0).y;
+		synchronized(vertexList) {
+			for (Vertex v : vertexList) {
+				g.append("    \\node ");
+				if (v.col.equals("red")) {
+					g.append("[rn] ");
+				} else if (v.col.equals("green")) {
+					g.append("[gn] ");
+				} else if (v.col.equals("boundary")) {
+					g.append("[dom] ");
+				}
+				g.append("(");
+				g.append(v.id);
+				g.append(") at (");
+				g.append(Float.toString((v.x-xOrigin) / (float)GRID_X).replaceFirst(".0$", ""));
+				g.append(", ");
+				g.append(Float.toString((yOrigin-v.y) / (float)GRID_Y).replaceFirst(".0$", ""));
+				g.append(") {};\n");
+			}
+		}
+		
+		g.append("\n    \\draw");
+		int counter = 0;
+		
+		synchronized (edgeList) {
+			for (Edge e : edgeList) {
+				if (counter==4) {
+					counter = 0;
+					g.append("\n        ");
+				}
+				g.append(" (");
+				g.append(e.source.id);
[…]

### 2008-06-13 gui/src/QuantoApplet.java

@@ -1,3 +1,6 @@
+import java.awt.Toolkit;
+import java.awt.datatransfer.Clipboard;
+import java.awt.datatransfer.StringSelection;
 import java.io.BufferedReader;
 import java.io.IOException;
 import java.io.InputStreamReader;
@@ -179,6 +182,8 @@ public class QuantoApplet extends PApplet {
 			return;
 		}
 		
+		Clipboard cb;
+		StringSelection data;
 		switch (key) {
 		case 'l':
 			layout(graph);
@@ -250,12 +255,21 @@ public class QuantoApplet extends PApplet {
 			if (snapToGrid) graph.enableSnap();
 			else graph.disableSnap();
 			break;
+		case 'x':
+			cb = Toolkit.getDefaultToolkit().getSystemClipboard();
+		    data = new StringSelection(graph.toLatex());
+		    cb.setContents(data, data);
+			break;
+		case 't':
+			cb = Toolkit.getDefaultToolkit().getSystemClipboard();
+		    data = new StringSelection(graph.toDot());
+		    cb.setContents(data, data);
+			break;
 		case 's':
 		case 'm':
 		case 'e': /* these are tools that require mouse input too */
 			tool = key;
 			break;
-
 		}
 		
 		play();
@@ -348,33 +362,8 @@ public class QuantoApplet extends PApplet {
 	}
 
 
-	String makeDot(Graph graph) {
-		StringBuffer g = new StringBuffer();
-		g.append("digraph { nodesep=0.65; ranksep=0.65;\n");
-		
-
-		
-		for (Vertex v : graph.vertexList) {
-			g.append(v.id);
-			g.append(" [color=\"");
-			g.append(v.col);
-			g.append("\",label=\"\",width=0.35,height=0.35,shape=circle];\n");
-		}
-
-		for (Edge e : graph.edgeList) {
-			g.append(e.source.id);
-			g.append("->");
-			g.append(e.dest.id);
-			//g.append(" [arrowhead=none,headclip=false,tailclip=false];\n");
-			g.append(" [arrowhead=none];\n");
-		}
-
-		g.append("\n}\n");
-		return g.toString();
-	}
-
 	void layout(Graph graph) {
-		layout(makeDot(graph), graph);
+		layout(graph.toDot(), graph);
 	}
 
 	void layout(String viz, Graph graph) {

### 2008-06-13 gui/src/QuantoFrame.java

@@ -33,20 +33,23 @@ public class QuantoFrame extends Frame implements PConstants {
         		"New graph (n)",
         		"Undo (u)",
         		"Layout (l)",
-        		"add Red (r)",
-        		"add Green (g)",
-        		"add Hadamard (h)",
-        		"add Boundary (b)",
+        		"Add red (r)",
+        		"Add green (g)",
+        		"Add hadamard (h)",
+        		"Add boundary (b)",
         		"Delete vertex (d)",
-        		"toggle sPlines (p)",
+        		"Toggle splines (p)",
         		"Capture to pdf (c)",
-        		"start/stop Video (v)",
-        		"toggle snap to Grid (G)",
-        		"quit (q)"
+        		"Start/stop Video (v)",
+        		"Toggle snap to Grid (G)",
+        		"Latex to clipboard (x)",
+        		"Dot to clipboard (t)",
+        		"Quit (q)"
         	};
         final char[] actionKeys = {
-        		'n', 'u', 'l', 'r', 'g', 'h', 'b',
-        		'd', 'p', 'c', 'v', 'G', 'q'
+        		'n', 'u', 'l', 'r', 'g', 'h',
+        		'b', 'd', 'p', 'c', 'v', 'G',
+        		'x', 't', 'q'
         	};
         
         MenuItem mi;

### 2008-06-13 gui/src/Vertex.java

@@ -67,13 +67,15 @@ class Vertex extends PLib implements Comparable<Vertex>  {
 		if (rate>1) rate = 1;
 		float dx = destX - x;
 		float dy = destY - y;
-		x += dx * rate;
-		y += dy * rate;
 
 		if (floor(dx)==0 && floor(dy)==0) {
 			lastTick = -1;
+			x = destX;
+			y = destY;
 			return false;
 		} else {
+			x += dx * rate;
+			y += dy * rate;
 			return true;
 		}
 	}

### 2008-06-14 gui/src/Graph.java
indentation tweak.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1633 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@ -159,7 +159,7 @@ public class Graph extends PLib {
 			for (Edge e : edgeList) {
 				if (counter==4) {
 					counter = 0;
-					g.append("\n        ");
+					g.append("\n         ");
 				}
 				g.append(" (");
 				g.append(e.source.id);

### 2012-03-02 gui/src/quanto/gui/GraphVisualizationViewer.java
Respond to changes in the graph saved status
The "Save" action is no longer disabled when it shouldn't be.

@@ -74,6 +74,7 @@ public class GraphVisualizationViewer
 		graph.addChangeListener(new ChangeListener() {
 			public void stateChanged(ChangeEvent e) {
 				modifyLayout();
+                fireStateChanged();
 			}
 		});
 	}

### 2012-03-02 gui/src/quanto/gui/InteractiveGraphView.java

@@ -506,6 +506,16 @@ public class InteractiveGraphView
 		viewer.setBoundingBoxEnabled(false);
 		
 		buildActionMap();
+
+        g.addChangeListener(new ChangeListener() {
+            public void stateChanged(ChangeEvent e) {
+                if (saveEnabled && isAttached()) {
+                    getViewPort().setCommandEnabled(CommandManager.Command.Save,
+                        !getGraph().isSaved()
+                        );
+                }
+            }
+        });
 	}
 	
 	public boolean isVertexLocked(Vertex v) {
@@ -825,17 +835,15 @@ public class InteractiveGraphView
 			errorDialog(e.getMessage());
 		}
 	}
-	
-	public void cleanUp() {
+
+    public void removeOldLabels() {
 		((QVertexLabeler) viewer.getRenderContext().getVertexLabelRenderer()).cleanup();
 		((QBangBoxLabeler) viewer.getRenderContext().getBangBoxLabelRenderer()).cleanup();
-		if (saveEnabled && isAttached()) {
-			getViewPort().setCommandEnabled(CommandManager.Command.Save,
-				!getGraph().isSaved()
-				);
-		}
+    }
+	
+	public void cleanUp() {
+        removeOldLabels();
 	}
-
 	
 	public void cacheVertexPositions(){
 		verticesCache= new HashMap<String, Point2D>();
@@ -882,23 +890,22 @@ public class InteractiveGraphView
 				}
 				viewer.getGraphLayout().lock(v, true);
 			}			
-		cleanUp();
 		}
 		int count=0;
 		for(Vertex v: getGraph().getVertices())	{					
-			if(verticesCache.get(v.getCoreName())==null)
+			if(verticesCache.get(v.getCoreName())==null) {
 				if(rewriteRect!=null) {
 					viewer.shift(rewriteRect, v, new Point2D.Double(0, 20*count));
 					setVerticesPositionData();
 					count++;
 				}
-			cleanUp();
+            }
 		}
 		
 		forceLayout.startModify();
 		viewer.modifyLayout();
 		forceLayout.endModify();
-		cleanUp();	
+		removeOldLabels();	
 		viewer.update();
 		//locking and unlocking used internally to notify the layout which vertices have user data
 		unlockVertices(getGraph().getVertices());
@@ -1246,7 +1253,7 @@ public class InteractiveGraphView
 					Set<Vertex> picked = viewer.getPickedVertexState().getPicked();
 					if (!picked.isEmpty()) {
[…]

### 2012-03-02 gui/src/quanto/gui/GraphVisualizationViewer.java
Respond to changes in the graph saved status
The "Save" action is no longer disabled when it shouldn't be.
Conflicts:

	gui/src/quanto/gui/InteractiveGraphView.java

@@ -72,6 +72,7 @@ public class GraphVisualizationViewer
 		graph.addChangeListener(new ChangeListener() {
 			public void stateChanged(ChangeEvent e) {
 				modifyLayout();
+                fireStateChanged();
 			}
 		});
 	}

### 2012-03-02 gui/src/quanto/gui/InteractiveGraphView.java

@@ -372,6 +372,16 @@ public class InteractiveGraphView
 		viewer.setBoundingBoxEnabled(false);
 		
 		buildActionMap();
+
+        g.addChangeListener(new ChangeListener() {
+            public void stateChanged(ChangeEvent e) {
+                if (saveEnabled && isAttached()) {
+                    getViewPort().setCommandEnabled(CommandManager.Command.Save,
+                        !getGraph().isSaved()
+                        );
+                }
+            }
+        });
 	}
 	
 	public boolean isVertexLocked(Vertex v) {
@@ -691,16 +701,14 @@ public class InteractiveGraphView
 			errorDialog(e.getMessage());
 		}
 	}
+
+    public void removeOldLabels() {
+		((QVertexLabeler) viewer.getRenderContext().getVertexLabelRenderer()).cleanup();
+    }
 	
 	public void cleanUp() {
-		((QVertexLabeler) viewer.getRenderContext().getVertexLabelRenderer()).cleanup();
-		if (saveEnabled && isAttached()) {
-			getViewPort().setCommandEnabled(CommandManager.Command.Save,
-				!getGraph().isSaved()
-				);
-		}
+        removeOldLabels();
 	}
-
 	
 	public void cacheVertexPositions(){
 		verticesCache= new HashMap<String, Point2D>();
@@ -747,23 +755,22 @@ public class InteractiveGraphView
 				}
 				viewer.getGraphLayout().lock(v, true);
 			}			
-		cleanUp();
 		}
 		int count=0;
 		for(Vertex v: getGraph().getVertices())	{					
-			if(verticesCache.get(v.getCoreName())==null)
+			if(verticesCache.get(v.getCoreName())==null) {
 				if(rewriteRect!=null) {
 					viewer.shift(rewriteRect, v, new Point2D.Double(0, 20*count));
 					setVerticesPositionData();
 					count++;
 				}
-			cleanUp();
+            }
 		}
 		
 		forceLayout.startModify();
 		viewer.modifyLayout();
 		forceLayout.endModify();
-		cleanUp();	
+		removeOldLabels();	
 		viewer.update();
 		//locking and unlocking used internally to notify the layout which vertices have user data
 		unlockVertices(getGraph().getVertices());
@@ -1111,7 +1118,7 @@ public class InteractiveGraphView
 					Set<Vertex> picked = viewer.getPickedVertexState().getPicked();
 					if (!picked.isEmpty()) {
[…]

### 2012-03-04 gui/src/quanto/core/protocol/ProtocolManager.java
Fix Copy/Paste bug

@@ -708,7 +708,7 @@ public class ProtocolManager {
         getOkResponse();
     }
 
-    public void insertGraph(String target, String source) throws CoreException {
+    public void insertGraph(String source, String target) throws CoreException {
         if (backend == null) {
             throw new IllegalStateException("The core is not running");
         }

### 2012-03-04 gui/src/quanto/gui/RulesBar.java
Delete rules (all/tag/selected) from the gui

@@ -86,15 +86,19 @@ public class RulesBar extends JPanel {
 	private DefaultListModel rulesModel;
 	private JToggleButton enableButton;
 	private JToggleButton disableButton;
+	private JToggleButton deleteButton;
 	private JComboBox tagsCombo;
 	private JPopupMenu enableMenu = new JPopupMenu();
 	private JPopupMenu disableMenu = new JPopupMenu();
-
 	private JMenuItem enableAllJMenuItem  = new JMenuItem("All");
 	private JMenuItem disableAllJMenuItem  = new JMenuItem("All");
 	private JMenuItem enableSelectionJMenuItem  = new JMenuItem("Selection");
 	private JMenuItem disableSelectionJMenuItem  = new JMenuItem("Selection");
 
+	private JPopupMenu deleteMenu = new JPopupMenu();
+	private JMenuItem deleteSelectionJMenuItem = new JMenuItem("Selection");
+	private JMenuItem deleteAllJMenuItem = new JMenuItem("All");
+	
 	private boolean suppressTagComboCallback = false;
 
 	private JPopupMenu createRuleContextualMenu() {
@@ -198,7 +202,7 @@ public class RulesBar extends JPanel {
 					}
 				}
 				catch (CoreException ex) {
-                    System.err.println(ex.getMessage());
+				     logger.log(Level.WARNING, "Could not disable rules");
 				}
 			}
 		});
@@ -215,10 +219,46 @@ public class RulesBar extends JPanel {
 					ruleset.deactivateRules(ruleNames);
 				}
 				catch (CoreException ex) {
+				     logger.log(Level.WARNING, "Could not disable selected rules");
 				}
 			}
 		});
 		disableMenu.add(disableSelectionJMenuItem);
+		
+		
+	     deleteAllJMenuItem.addActionListener(new ActionListener() {
+
+	               public void actionPerformed(ActionEvent e) {
+	                    try {
+	                         if (tagsCombo.getSelectedIndex() != 0) {
+	                              ruleset.deleteRulesByTag((String) tagsCombo.getSelectedItem());
+	                         }
+	                         else {
+	                              for(String r: ruleset.getRules())
+	                                   ruleset.deleteRule(r);
+	                         }
+	                    }
+	                    catch (CoreException ex) {
+	                         logger.log(Level.WARNING, "Could not delete rules");
+	                    }
+	               }
+	          });
+	     deleteMenu.add(deleteAllJMenuItem);
[…]

### 2012-03-05 docs/cqm_practical/cqm_practical.tex
added latex template

\documentclass{article}

\input{preamble.tex}
\input{tikzstyles.tex}

\begin{document}

test

\end{document}

### 2012-03-05 docs/cqm_practical/preamble.tex
added latex template

\usepackage{tikzfig}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{figures/}}

\usepackage{algpseudocode}
\usepackage{algorithm}

\newtheorem{theorem}{Theorem}[section]
\newtheorem*{theorem*}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}\newtheorem{example}[theorem]{Example}
\theoremstyle{definition}\newtheorem{examples}[theorem]{Examples}
\theoremstyle{definition}\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}\newtheorem{definitions}[theorem]{Definitions}
\theoremstyle{definition}\newtheorem{remark}[theorem]{Remark}
\theoremstyle{definition}\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}\newtheorem{impremark}[theorem]{Important remark}
\theoremstyle{definition}\newtheorem{remarks}[theorem]{Remarks}
\theoremstyle{definition}\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{definition}\newtheorem{notation}[theorem]{Notation}

\def\bR{\begin{color}{red}} 
\def\bB{\begin{color}{blue}}
\def\bM{\begin{color}{magenta}}
\def\bC{\begin{color}{cyan}}
\def\bW{\begin{color}{white}}
\def\bBl{\begin{color}{black}} 
\def\bG{\begin{color}{green}}
\def\bY{\begin{color}{yellow}}
\def\e{\end{color}}

### 2012-03-05 docs/cqm_practical/tikzstyles.tex
added latex template

\tikzstyle{every picture}=[baseline=-0.25em]
\tikzstyle{dotpic}=[scale=0.6]

\tikzstyle{dot}=[inner sep=0.7mm,minimum width=0pt,minimum height=0pt,fill=black,draw=black,shape=circle]
\tikzstyle{white dot}=[dot,fill=white]
\tikzstyle{red dot}=[dot,fill=red]
\tikzstyle{green dot}=[dot,fill=green]

### 2012-03-05 docs/cqm_practical/cqm_practical.tex
added figures and gitignore to practical sheet

@@
 
+\title{CQM Practical: MBQC in Quantomatic}
+
 \begin{document}
 
-test
+\maketitle
+
+There are various models of quantum computation. You most likely already know about the \textit{circuit model}, where quantum computation is performed in three steps:
+
+\begin{enumerate}
+  \item Prepare some collection of qubits in an initial state.
+  \item Perform unitary \textit{quantum gates} on these qubits.
+  \item Measure the result.
+\end{enumerate}
+
+In this model, steps 1 and 3 are more or less fixed, whereas step 2 is where the interesting stuff happens.
+in this picture, the quantum programs is represented by a circuit diagram.
+however this is not the only paradigm for quantum computing.
+another choice is \textit{measurement based quantum computing}.
+in this model, computation consists of these steps:
+
+\begin{enumerate}
+  \item Prepare some qubits in the $\ket{+}$ state.
+  \item Entangle pairs of qubits using controlled-$Z$ operations.
+  \item Perform a series of measurements at some angles $\alpha_i$.
+\end{enumerate}
+
+The family of states that can be prepared using steps 1 and 2 are called \textit{graph states}.
+This is because they are often represented as graphs, using nodes to represent qubits and edges to represent controlled-$Z$ operations.
+Usually, the graph state is taken to be fixed, and the interesting part of an MBQC program is step 3.
+Perhaps surprisingly, MBQC is just as powerful as the circuit model.
+That is, it is \textit{universal for quantum computing}.
+In this practical, we will prove this result using Quantomatic.
+
+\section{The Z/X Calculus}
+
+The Z/X calculus extends the normal red/green calculus with three new generators. The first two are the phase gates $Z_\alpha$ and $X_\alpha$. The third is the Hadamard gate.
+\begin{center}
+    $Z_\alpha :=$ \tikzfig{green_phase}
+    \qquad\qquad
+    $X_\alpha :=$ \tikzfig{red_phase}
+    \qquad\qquad
+    $H := $ \tikzfig{hadamard}
+\end{center}
 

### 2012-03-05 docs/cqm_practical/cqm_practical.tex
pi-copy exercise done

@@
 
-\section{The Z/X Calculus}
+\section*{The Z/X Calculus}
 
@@
 
+The Z/X-calculus satisfies all of the axioms of a strongly complementary pair, plus several additional axioms. Firstly, the caps and cups of the two colours coincide:
+\ctikzfig{caps_cups}
+
+Maps of the form $Z_\alpha$ and are phases for the $Z$ observable, so they can be used to decorate green spiders. Similarly for $X$:
+\ctikzfig{phase_spider}
+
+The Hadamard gate is unitary and self-adjoint, and it interchanges the two colours:
+\begin{center}
+  \tikzfig{h_self_adjoint}
+  \qquad\qquad
+  \tikzfig{h_colour_change}
+\end{center}
+
+Furthermore, the green points associated with phases $0$ and $\pi$ are classical for red and vice-versa.
+\ctikzfig{classical_points}
+
+\section*{Translating MBQC Program into Z/X graphs}
+
+An MBQC program is an expression consisting of the following operations:
+
+\begin{enumerate}
+  \item $N_i$ prepare a new (ancilla) qubit in the $\ket{+}$ state at position $i$.
+  \item $E_{i,j}$ entangle the $i$-th qubit to the $j$-th qubit using a controlled-$Z$ gate.
+  \item $M_i^\alpha$ measure at angle $\alpha$.
+\end{enumerate}
+
+MBQC programs can be written as composition of these operations. Note that they are all linear maps, so they are read from right-to-left. For example:
+
+\[ M_2^0 M_4^0 E_{1,3} E_{2,3} E_{3,4} N_3 N_4 \]
+
+This program prepares ancillae at 3 and 4, entangles 3 to 4, 2 to 3, and 1 to 3, then measures qubits 2 and 4 at angles 0 and 0. These operations can be encoded in the Z/X-calculus as:
+\ctikzfig{encoding}
+
+Note that measurements are non-deterministic, so either of the two outcomes shown may happen when we measure a qubit.
+The main point of MBQC is we can recover \textit{deterministic} computation by first performing measurements, then performing \textit{corrections} later by applying unitaries or adjusting the angles at which we perform later measurements.
+
+\section*{Quantomatic Basics}
+
+We will implement a fragment of the Z/X calculus in Quantomatic as a \textit{graph rewrite system}.
+The difference in a rewrite system and a set of graphical axioms is that equations $G = H$ are replaced with \textit{reduction rules} $G \rewritesto H$.
+First, we'll see how to input graphs and rules into Quantomatic, and how to express collections of edges with ellipses using graph patterns.
+
+\medskip
+
+\noindent{\color{red} \textbf{Disclaimer:} Quantomatic is a very ``researchy'' piece of software.
+Some features are partially implemented, and some things may not work as advertised.
+If you encounter a bug, don't panic.
+There should be a work-around that suffices to complete the exercise.}
+
+\medskip
+
+Launch Quantomatic using:
+
+\begin{verbatim}
+  $ cd /path/to/quantomatic/
+  $ ant run.gui
+\end{verbatim}
+
+You will now be looking at a blank graph.
+Along the left is the theory editor panel.
+In this panel, click on the arrow to the right of ``Rules'' to open the toolbox.
+Here, you can add dots of either colour, Hadamard gates, and boundary vertices.
+Click each of these buttons several times to add various kinds of nodes to a graph.
+Set phases for red and green nodes by double-clicking on the $0$.
+Note that it will recognise latex for Greek letters (\verb!\alpha!, \verb!\beta!, etc.).
+
+Add edges by selected the Directed Edge tool from the toolbar. Click-and-drag from source to target. You may need to turn on arrow heads in the View menu.
+
+Boundary vertices provide named inputs and outputs for a graph, corresponding to dangling wires in the graphical calculus.
+\begin{center}
+  \tikzfig{red_dot_sample} \ \ $:=$\ \ 
+  \raisebox{-1.5cm}{\includegraphics[scale=0.6]{graph_w_bds.png}}
+\end{center}
+
+Graphs with variable inputs and outputs (like spiders) can be represented using \textit{pattern graphs}.
+These graphs indicate sub-graphs which can be repeated many times by using \textit{bang-boxes} (by analogy to the ``bang'' operation $!(-)$ from linear logic).
+To make an input or an output variable, select the boundary vertex and click ``Bang vertices'' in the toolbox.
+We can then represent a spider of any arity as follows:
+\begin{center}
+  \tikzfig{red_spider} \ \ $:=$\ \ 
+  \raisebox{-1.65cm}{\includegraphics[scale=0.6]{spider_quanto.png}}
+\end{center}
+
+Note that bang-boxes have names as well.
+These and the names of boundaries can be changed by double-clicking on them.
+This will become important when defining rules.
+To define a rule, click the left arrow next to ``Toolbox'' to open the rules panel, and click ``New Rule''.
+Give the rule a name, and it will open as a pair of (empty) graphs, side-by-side.
+Click on the left or the right panel to focus on the LHS or RHS and edit the graph as before.
+Click the ``Save'' icon to save the rule back to the ruleset.
+Note that this will only succeed if you have input a valid rule.
+A valid rule must have:
+
+\begin{enumerate}
+  \item the same number of boundary vertices and bang-boxes on both sides,
+  \item the same \textit{names} for boundary vertices and bang-boxes on both sides, and
+  \item a \textit{compatible boxing} of boundary vertices.
+That is, if a boundary vertex named ``Va'' is in the box named ``Ba'' on the LHS, it must also be in the box named ``Ba'' on the RHS and vice-versa.
+\end{enumerate}
+
+For our purposes, we will only be placing single boundary vertices.
+Once rules are implemented, activate them using the button in the rule panel.
+Using the dropdown box below the toolbar, select a graph.
+Rules can be applied to a graph in one of two ways.
+The first is to apply rules manually, using ``Graph $>$ Show Rewrites''.
+The second is to apply rules automatically, using ``Graph $>$ Normalise'' or ``Graph $>$ Fast Normalise''.
+
+\section*{Exercise 1: Implementing the Spider Laws}
+
+The spider theorem for green dots looks like this:
+\ctikzfig{spider_green}
+
+Rather than implement the spider theorem as a single reduction rule, we can implement it as:
+
+\begin{enumerate}
+  \item A rule that merges adjacent dots of the same colour.
+  \item A rule that removes self-loops.
+  \item A rule that turns spiders with 1 input and 1 output
+\end{enumerate}
+
+Note the the first and second rule need to handle any number of inputs and outputs on the spiders. Bang-boxes can be used for this.
+
+\medskip
+\noindent{\color{blue} Implement the spider laws for red and green dots as rules (6 in total) in Quantomatic.
+Use these rules to simplify a graph consisting of red and green spiders, as well as some phases to a normal form.}
+\medskip
+
+\section*{Exercise 2: $\pi$-copy}
+
+As a consequence of $\ket{z_\pi}$ being a classical point, the map $Z_\pi$ is copied through the red dot. In the process, any angle inside picks up a minus-1 phase shift.
+
+\ctikzfig{pi_copy}
+
+\medskip
+\noindent{\color{blue} Implement the $\pi$-copy rule in Quantomatic.}
+\medskip
+
+However, once we implement this rule, we notice a problem! Consider graphs that look like this:
+
+\ctikzfig{pi_copy_critical}
+
+We now have a choice to apply the spider-merge rule to the two green vertices or to apply the copy rule to the red and green vertex.
+If we apply the spider rule, the copy will no longer be applicable and vice-versa.
+Therefore we have two distinct, equivalent normal forms.
+In rewriting lingo, this is called a \textit{critical pair}.
+We can close a critical pair by introducing a \textit{completion}.
+
+\ctikzfig{pi_copy_completion}
+
+\medskip
+\noindent{\color{blue} Implement the completion above as a rule in Quantomatic. Show that the original $\pi$-copy rule is now redundant and remove it.}
+\medskip
+
 \end{document}

### 2012-03-05 docs/cqm_practical/preamble.tex
added figures and gitignore to practical sheet
pi-copy exercise done

@@
 
+% REWRITING OPERATORS
+\newcommand{\normalised}{\ensuremath{\!\downarrow\,}}
+\newcommand{\cmdrewritesto}{\tikz[baseline=-0.25em] { \draw [-open triangle 45, line width=0.2pt] (0,0) -- (0.5,0); }\,}
+\newcommand{\cmdrewriteequiv}{\tikz[baseline=-0.25em] { \draw [open triangle 45-open triangle 45, line width=0.2pt] (0,0) -- node [auto,yshift=-1.2mm] {$*$} (0.7,0); }\,}
+\newcommand{\cmdrewritetrans}{\tikz[baseline=-0.25em] { \draw [-open triangle 45, line width=0.2pt] (0,0) -- node [auto,pos=0.3,yshift=-1.2mm] {$*$} (0.5,0); }\,}
+
+\newcommand{\cmdleftrewritesto}{\tikz[baseline=-0.25em] { \draw [open triangle 45-, line width=0.2pt] (0,0) -- (0.5,0); }\,}
+\newcommand{\cmdleftrewritetrans}{\tikz[baseline=-0.25em] { \draw [open triangle 45-, line width=0.2pt] (0,0) -- node [auto,pos=0.7,yshift=-1.2mm] {$*$} (0.5,0); }\,}
+
+\DeclareMathOperator{\rewritesto}{\cmdrewritesto}
+\DeclareMathOperator{\rewriteequiv}{\cmdrewriteequiv}
+\DeclareMathOperator{\rewritetrans}{\cmdrewritetrans}
+\DeclareMathOperator{\leftrewritesto}{\cmdleftrewritesto}
+\DeclareMathOperator{\leftrewritetrans}{\cmdleftrewritetrans}
+
+% BRAS AND KETS
+\newcommand{\bra}[1]{\ensuremath{\left\langle #1 \right|}}
+\newcommand{\ket}[1]{\ensuremath{\left| #1 \right\rangle}}
+\newcommand{\braket}[2]{\ensuremath{\langle#1|#2\rangle}}
+\newcommand{\ketbra}[2]{\ensuremath{\ket{#1}\!\bra{#2}}}
+
 \def\bR{\begin{color}{red}} 

### 2012-03-06 docs/cqm_practical/cqm_practical.tex
finished practical

@@
 however this is not the only paradigm for quantum computing.
-another choice is \textit{measurement based quantum computing}.
+another choice is \textit{measurement based quantum computing} (MBQC).
 in this model, computation consists of these steps:
@@
 
-\section*{The Z/X Calculus}
+\section{The Z/X Calculus}\label{sec:zx}
 
@@
 
-\section*{Translating MBQC Program into Z/X graphs}
+\section{Translating MBQC Program into Z/X graphs}\label{sec:mbqc}
 
@@
 
-\section*{Quantomatic Basics}
+\section{Quantomatic Basics}\label{sec:quanto}
 
@@
 
+If we can close all critical pairs in a (terminating) rewrite system, it is possible to show that all normal forms are unique. That is, the system is \textit{confluent and terminating}.
+
+\section*{Exercise 3: The CNOT Gate Using MBQC}
+
+It is a well-known result that single-qubit unitaries and the CNOT gate are universal for quantum computing. In the Z/X calculus, we can write the CNOT gate as:
+\begin{equation}\label{eq:cnot}
+  \tikzfig{cnot}
+\end{equation}
+
+Recall the MBQC program from section~\ref{sec:mbqc}.
+
+\[ M_2^0 M_4^0 E_{1,3} E_{2,3} E_{3,4} N_3 N_4 \]
+
+\medskip
+\noindent{\color{blue} Use the translation from section~\ref{sec:mbqc} to translate this program into a graph in Quantomatic, for measurement outcomes $\{ M_2^0 \rightarrow 0, M_4^0 \rightarrow 0 \}$.
+Implement enough rules from section~\ref{sec:zx} to normalise the measurement pattern into (\ref{eq:cnot}).}
+\medskip
+
+Even if we do not get outcome $00$, we can still produce a CNOT gate by performing single-qubit corrections.
+
+\medskip
+\noindent{\color{blue} For each of the other three outcomes, find red and green phases to apply to qubits 1 and 3 to produce a CNOT gate.
+Show that all four cases reduce to (\ref{eq:cnot}) in Quantomatic.}
+\medskip
+
+Let $M_{i,A}^\alpha$ be a measurement labelled $A$ of the $i$-th qubit at angle $\alpha$.
+Let $Z_j^A$ be the identity for the outcome $0$ of the measurement labelled $A$ and a $Z_\pi$ gate for outcome $1$.
+Define $X_j^A$ similarly.
+These are called \textit{conditional corrections}.
+
+\medskip
+\noindent{\color{blue} Using conditional corrections, write an MBQC program that produces the CNOT gate, regardless of the measurement outcomes.}
+\medskip
+
+\section*{Bonus: Implementing Single Qubit Unitaries}
+
+Any single qubit unitary can be written using three phase gates, $U = Z_\alpha X_\beta Z_\gamma$.
+
+\medskip
+\noindent{\color{blue} Write an MBQC program that implements an arbitrary single qubit unitary.
+Show that for several combinations of outcomes that the program reduces to a composition $Z_\alpha X_\beta Z_\gamma$ in Quantomatic.}
+\medskip
+
+Note that measurement angles can be chosen based on previous measurement outcomes.
+This is known as \textit{feed-forward} in MBQC.
+The existence of these two programs shows that MBQC is as powerful as the circuit model.
+
 \end{document}

### 2012-03-06 gui/src/quanto/core/protocol/ProtocolReader.java
Better error messages

@@ -30,11 +30,11 @@ public class ProtocolReader {
         int gotCh = input.read();
         if (gotCh != ESC) {
             if (gotCh == -1)
-                throw new ProtocolException("Expected ESC, got EOF");
+                throw new ProtocolException("Expected ESC from core, got EOF");
             else if (Character.isISOControl(gotCh))
-                throw new ProtocolException("Expected ESC, got \\u" + Integer.toHexString(gotCh));
+                throw new ProtocolException("Expected ESC from core, got \\u" + String.format("%1$04x", gotCh));
             else
-                throw new ProtocolException("Expected ESC, got " + (char)gotCh);
+                throw new ProtocolException("Expected ESC from core, got " + (char)gotCh);
         }
     }
 

### 2012-03-06 gui/src/quanto/core/protocol/ProtocolReader.java
Close streams correctly

@@ -26,6 +26,10 @@ public class ProtocolReader {
                 new BufferedInputStream(input), "quanto.core.protocol.stream");
     }
 
+    public void close() throws IOException {
+        input.close();
+    }
+
     private void eatEsc() throws ProtocolException, IOException {
         int gotCh = input.read();
         if (gotCh != ESC) {

### 2012-03-06 gui/src/quanto/core/protocol/RequestWriter.java

@@ -29,6 +29,10 @@ public class RequestWriter
                 "quanto.core.protocol.stream");
     }
 
+    public void close() throws IOException {
+        output.close();
+    }
+
     private byte[] convertInt(int i)
     {
         return stringToAscii(Integer.toString(i));

### 2012-03-06 gui/src/quanto/core/protocol/ProtocolManager.java
Close streams correctly

@@ -24,8 +24,6 @@ public class ProtocolManager {
     private final static Logger logger = Logger.getLogger("quanto.core.protocol");
 
     public static String quantoCoreExecutable = "quanto-core";
-    private LoggingInputStream dbgInputStream;
-    private LoggingOutputStream dbgOutputStream;
     private RequestWriter writer;
     private ProtocolReader reader;
     private Process backend;
@@ -126,8 +124,8 @@ public class ProtocolManager {
         if (backend != null) {
             logger.log(Level.FINEST, "Shutting down the core process");
             try {
-                dbgInputStream.close();
-                dbgOutputStream.close();
+                reader.close();
+                writer.close();
                 new ProcessCleanupThread(backend).start();
             } catch (IOException ex) {
                 logger.log(Level.WARNING, "Failed to close communication channels to the core");

### 2012-03-06 gui/src/quanto/core/protocol/ProtocolReader.java
Discard excess data when there is a problem
This may include useful information, such as a polyml backtrace

@@ -4,6 +4,7 @@
  */
 package quanto.core.protocol;
 
+import java.util.logging.Logger;
 import java.io.BufferedInputStream;
 import java.io.IOException;
 import java.io.InputStream;
@@ -17,6 +18,7 @@ import static quanto.core.protocol.Utils.*;
  */
 public class ProtocolReader {
     private static final char ESC = '\u001b';
+    private final static Logger logger = Logger.getLogger("quanto.core.protocol");
 
     private LoggingInputStream input;
     private String version;
@@ -333,10 +335,32 @@ public class ProtocolReader {
             return resp;
         } catch (IOException ex) {
             input.writeLog(Level.SEVERE, "Received partial message");
+            consumeStream();
             throw ex;
         } catch (ProtocolException ex) {
             input.writeLog(Level.SEVERE, "Received invalid message");
+            consumeStream();
             throw ex;
+        }
+    }
+
+    private void consumeStream() {
+        try {
+            byte[] b = new byte[1024];
+            int count = 0;
+            int avail = input.available();
+            while (avail > 0) {
+                if (avail > b.length)
+                    avail = b.length;
+                count = input.read(b, 0, avail);
+                if (count != -1 && logger.isLoggable(Level.INFO)) {
+                    String strVal = new String(b, 0, count)
+                            .replace('\u001b', '\u00a4');
+                    logger.log(Level.INFO, "Discarding data: \"{0}\"", strVal);
+                }
+                avail = input.available();
+            }
+        } catch (IOException ex) {
         }
     }
 }

### 2012-03-06 docs/cqm_practical/cqm_practical.tex
fixed problem w practical sheet

@@
 \begin{verbatim}
-  $ cd /path/to/quantomatic/
-  $ ant run.gui
+  $ /usr/local/practicals/cqm/quantomatic
 \end{verbatim}
@@
   \item A rule that removes self-loops.
-  \item A rule that turns spiders with 1 input and 1 output
+  \item A rule that turns spiders with 1 input and 1 output into identities.
 \end{enumerate}
@@
 
-\[ M_2^0 M_4^0 E_{1,3} E_{2,3} E_{3,4} N_3 N_4 \]
+\[ M_2^0 M_3^0 E_{1,3} E_{2,3} E_{3,4} N_3 N_4 \]
 
 \medskip
-\noindent{\color{blue} Use the translation from section~\ref{sec:mbqc} to translate this program into a graph in Quantomatic, for measurement outcomes $\{ M_2^0 \rightarrow 0, M_4^0 \rightarrow 0 \}$.
+\noindent{\color{blue} Use the translation from section~\ref{sec:mbqc} to translate this program into a graph in Quantomatic, for measurement outcomes $\{ M_2^0 \rightarrow 0, M_3^0 \rightarrow 0 \}$.
 Implement enough rules from section~\ref{sec:zx} to normalise the measurement pattern into (\ref{eq:cnot}).}

### 2012-03-06 gui/src/quanto/gui/InteractiveGraphView.java
Ask before overwriting files

@@ -1183,15 +1183,6 @@ public class InteractiveGraphView
 
 					File outputFile =  QuantoApp.getInstance().saveFile(InteractiveGraphView.this);
 					if (outputFile != null) {
-						if (outputFile.exists()) {
-							int overwriteAnswer = JOptionPane.showConfirmDialog(
-								InteractiveGraphView.this,
-								"Are you sure you want to overwrite \"" + outputFile.getName() + "\"?",
-								"Overwrite file?",
-								JOptionPane.YES_NO_OPTION);
-							if (overwriteAnswer != JOptionPane.YES_OPTION)
-								return;
-						}
 						OutputStream file = new FileOutputStream(outputFile);
                                                 PdfGraphVisualizationServer server = new PdfGraphVisualizationServer(core.getActiveTheory(), getGraph());
 						server.renderToPdf(file);

### 2012-03-06 gui/src/quanto/gui/InteractiveGraphView.java
Ask before overwriting files

@@ -1318,15 +1318,6 @@ public class InteractiveGraphView
 
 					File outputFile =  QuantoApp.getInstance().saveFile(InteractiveGraphView.this);
 					if (outputFile != null) {
-						if (outputFile.exists()) {
-							int overwriteAnswer = JOptionPane.showConfirmDialog(
-								InteractiveGraphView.this,
-								"Are you sure you want to overwrite \"" + outputFile.getName() + "\"?",
-								"Overwrite file?",
-								JOptionPane.YES_NO_OPTION);
-							if (overwriteAnswer != JOptionPane.YES_OPTION)
-								return;
-						}
 						OutputStream file = new FileOutputStream(outputFile);
                                                 PdfGraphVisualizationServer server = new PdfGraphVisualizationServer(core.getActiveTheory(), getGraph());
 						server.renderToPdf(file);

### 2012-03-06 gui/src/quanto/gui/QuantoApp.java

@@ -476,6 +476,15 @@ public class QuantoApp {
         fileChooser[type].setDialogType(JFileChooser.SAVE_DIALOG);
         if (retVal == JFileChooser.APPROVE_OPTION) {
             File f = fileChooser[type].getSelectedFile();
+            if (f.exists()) {
+                int overwriteAnswer = JOptionPane.showConfirmDialog(
+                    parent,
+                    "Are you sure you want to overwrite \"" + f.getName() + "\"?",
+                    "Overwrite file?",
+                    JOptionPane.YES_NO_OPTION);
+                if (overwriteAnswer != JOptionPane.YES_OPTION)
+                    return null;
+            }
             if (f.getParent() != null) {
                 setPreference(QuantoApp.LAST_OPEN_DIRS[type], f.getParent());
             }

### 2012-03-07 gui/src/quanto/gui/RulesBar.java
No menus for ruleset buttons (c.f. issue #19)

@@ -34,12 +34,9 @@ import javax.swing.JOptionPane;
 import javax.swing.JPanel;
 import javax.swing.JPopupMenu;
 import javax.swing.JScrollPane;
-import javax.swing.JToggleButton;
 
 import javax.swing.event.ChangeEvent;
 import javax.swing.event.ChangeListener;
-import javax.swing.event.PopupMenuEvent;
-import javax.swing.event.PopupMenuListener;
 
 import quanto.core.CoreException;
 import quanto.core.Ruleset;
@@ -84,20 +81,12 @@ public class RulesBar extends JPanel {
 	};
 	private JList listView;
 	private DefaultListModel rulesModel;
-	private JToggleButton enableButton;
-	private JToggleButton disableButton;
-	private JToggleButton deleteButton;
+	private JButton enableButton;
+	private JButton disableButton;
+	private JButton deleteButton;
+	private JButton createRuleButton;
+	private JButton refreshButton;
 	private JComboBox tagsCombo;
-	private JPopupMenu enableMenu = new JPopupMenu();
-	private JPopupMenu disableMenu = new JPopupMenu();
-	private JMenuItem enableAllJMenuItem  = new JMenuItem("All");
-	private JMenuItem disableAllJMenuItem  = new JMenuItem("All");
-	private JMenuItem enableSelectionJMenuItem  = new JMenuItem("Selection");
-	private JMenuItem disableSelectionJMenuItem  = new JMenuItem("Selection");
-
-	private JPopupMenu deleteMenu = new JPopupMenu();
-	private JMenuItem deleteSelectionJMenuItem = new JMenuItem("Selection");
-	private JMenuItem deleteAllJMenuItem = new JMenuItem("All");
 	
 	private boolean suppressTagComboCallback = false;
 
@@ -155,111 +144,6 @@ public class RulesBar extends JPanel {
 	    }
 	    return popupMenu;
 	}
-	
-	private void createMenus() {
-		enableAllJMenuItem.addActionListener(new ActionListener() {
-
-			public void actionPerformed(ActionEvent e) {
-				try {
-					if (tagsCombo.getSelectedIndex() != 0) {
-						ruleset.activateRulesByTag((String) tagsCombo.getSelectedItem());
-					}
-					else {
-						ruleset.activateAllRules();
-					}
-				}
-				catch (CoreException ex) {
-				}
-			}
-		});
-		enableMenu.add(enableAllJMenuItem);
[…]

### 2012-03-07 gui/src/quanto/gui/RulesBar.java
No menus for ruleset buttons (c.f. issue #19)

@@ -84,20 +84,12 @@ public class RulesBar extends JPanel {
 	};
 	private JList listView;
 	private DefaultListModel rulesModel;
-	private JToggleButton enableButton;
-	private JToggleButton disableButton;
-	private JToggleButton deleteButton;
+	private JButton enableButton;
+	private JButton disableButton;
+	private JButton deleteButton;
+	private JButton createRuleButton;
+	private JButton refreshButton;
 	private JComboBox tagsCombo;
-	private JPopupMenu enableMenu = new JPopupMenu();
-	private JPopupMenu disableMenu = new JPopupMenu();
-	private JMenuItem enableAllJMenuItem  = new JMenuItem("All");
-	private JMenuItem disableAllJMenuItem  = new JMenuItem("All");
-	private JMenuItem enableSelectionJMenuItem  = new JMenuItem("Selection");
-	private JMenuItem disableSelectionJMenuItem  = new JMenuItem("Selection");
-
-	private JPopupMenu deleteMenu = new JPopupMenu();
-	private JMenuItem deleteSelectionJMenuItem = new JMenuItem("Selection");
-	private JMenuItem deleteAllJMenuItem = new JMenuItem("All");
 	
 	private boolean suppressTagComboCallback = false;
 
@@ -155,111 +147,6 @@ public class RulesBar extends JPanel {
 	    }
 	    return popupMenu;
 	}
-	
-	private void createMenus() {
-		enableAllJMenuItem.addActionListener(new ActionListener() {
-
-			public void actionPerformed(ActionEvent e) {
-				try {
-					if (tagsCombo.getSelectedIndex() != 0) {
-						ruleset.activateRulesByTag((String) tagsCombo.getSelectedItem());
-					}
-					else {
-						ruleset.activateAllRules();
-					}
-				}
-				catch (CoreException ex) {
-				}
-			}
-		});
-		enableMenu.add(enableAllJMenuItem);
-		enableSelectionJMenuItem.addActionListener(new ActionListener() {
-
-			public void actionPerformed(ActionEvent e) {
-				try {
-					Object[] descs = listView.getSelectedValues();
-					List<String> ruleNames = new LinkedList<String>();
-					for (Object d : descs) {
-						ruleNames.add(((RuleDescription) d).rulename);
-					}
-					ruleset.activateRules(ruleNames);
-				}
-				catch (CoreException ex) {
-				}
-			}
-		});
[…]

### 2012-03-07 gui/src/quanto/gui/RulesBar.java
No menus for ruleset buttons (c.f. issue #19)
Conflicts:

	gui/src/quanto/gui/RulesBar.java

@@ -34,12 +34,9 @@ import javax.swing.JOptionPane;
 import javax.swing.JPanel;
 import javax.swing.JPopupMenu;
 import javax.swing.JScrollPane;
-import javax.swing.JToggleButton;
 
 import javax.swing.event.ChangeEvent;
 import javax.swing.event.ChangeListener;
-import javax.swing.event.PopupMenuEvent;
-import javax.swing.event.PopupMenuListener;
 
 import quanto.core.CoreException;
 import quanto.core.Ruleset;
@@ -205,7 +202,6 @@ public class RulesBar extends JPanel {
                public void actionPerformed(ActionEvent e) {
                     try {
                          Object[] descs = listView.getSelectedValues();
-                         List<String> ruleNames = new LinkedList<String>();
                          for (Object d : descs) {
                               ruleset.deleteRule(((RuleDescription) d).rulename);
                          }

### 2012-03-07 gui/src/quanto/gui/InteractiveGraphView.java
- Append/Remove asterisk in graph name - When a graph is saved change its core name to the new name. (c.f. issue #21)

@@ -513,6 +513,7 @@ public class InteractiveGraphView
                     getViewPort().setCommandEnabled(CommandManager.Command.Save,
                         !getGraph().isSaved()
                         );
+                    firePropertyChange("saved", !getGraph().isSaved(), getGraph().isSaved());
                 }
             }
         });
@@ -1168,8 +1169,10 @@ public class InteractiveGraphView
 		if (f != null) {
 			try {
 				core.saveGraph(getGraph(), f);
+				core.renameGraph(getGraph(), f.getName());
 				getGraph().setFileName(f.getAbsolutePath());
 				getGraph().setSaved(true);
+				firePropertyChange("saved", !getGraph().isSaved(), getGraph().isSaved());
 				setTitle(f.getName());
 			}
 			catch (CoreException e) {
@@ -1186,6 +1189,7 @@ public class InteractiveGraphView
 			try {
 				core.saveGraph(getGraph(), new File(getGraph().getFileName()));
 				getGraph().setSaved(true);
+				firePropertyChange("saved", !getGraph().isSaved(), getGraph().isSaved());
 			}
 			catch (CoreException e) {
 				errorDialog(e.getMessage());

### 2012-03-07 gui/src/quanto/gui/InteractiveGraphView.java
- Append/Remove asterisk in graph name - When a graph is saved change its core name to the new name. (c.f. issue #21)

@@ -379,6 +379,7 @@ public class InteractiveGraphView
                     getViewPort().setCommandEnabled(CommandManager.Command.Save,
                         !getGraph().isSaved()
                         );
+                    firePropertyChange("saved", !getGraph().isSaved(), getGraph().isSaved());
                 }
             }
         });
@@ -1033,8 +1034,10 @@ public class InteractiveGraphView
 		if (f != null) {
 			try {
 				core.saveGraph(getGraph(), f);
+				core.renameGraph(getGraph(), f.getName());
 				getGraph().setFileName(f.getAbsolutePath());
 				getGraph().setSaved(true);
+				firePropertyChange("saved", !getGraph().isSaved(), getGraph().isSaved());
 				setTitle(f.getName());
 			}
 			catch (CoreException e) {
@@ -1051,6 +1054,7 @@ public class InteractiveGraphView
 			try {
 				core.saveGraph(getGraph(), new File(getGraph().getFileName()));
 				getGraph().setSaved(true);
+				firePropertyChange("saved", !getGraph().isSaved(), getGraph().isSaved());
 			}
 			catch (CoreException e) {
 				errorDialog(e.getMessage());

### 2012-03-07 gui/src/quanto/gui/RulesBar.java
-Dialog when deleting a rule -Delete rules from contextual menu (c.f. issue #38)

@@ -93,31 +93,38 @@ public class RulesBar extends JPanel {
 	private JPopupMenu createRuleContextualMenu() {
 		JPopupMenu popupMenu = new JPopupMenu();
 		JMenuItem menuItem = new JMenuItem("Edit rule");
-	    popupMenu.add(menuItem);
-	    menuItem.addActionListener(new ActionListener() {
+		popupMenu.add(menuItem);
+		menuItem.addActionListener(new ActionListener() {
 			public void actionPerformed(ActionEvent e) {
 				editRule(listView.getSelectedValue().toString());
 			}
-	    });
-	    
-	    JMenu subMenu = new JMenu("Add tag");
-	    
-	    try {
-	    	Collection<String> allTags = ruleset.getTags();
-	    	for (String tag: allTags) {
-	    		menuItem = new JMenuItem(tag);
-	    		menuItem.addActionListener(new ActionListener() {
-	    			public void actionPerformed(ActionEvent e) {
-	    				ruleset.tagRule(listView.getSelectedValue().toString(), e.getActionCommand());
-	    			}
-    			});
-    			subMenu.add(menuItem);
-    		}
-	    } catch (CoreException e) {
+		});
+		
+		menuItem = new JMenuItem("Delete rule");
+          popupMenu.add(menuItem);
+          menuItem.addActionListener(new ActionListener() {
+               public void actionPerformed(ActionEvent e) {
+                    deleteSelectedRules();
+               }
+          });
+          
+		JMenu subMenu = new JMenu("Add tag");
+		try {
+		     Collection<String> allTags = ruleset.getTags();
+		     for (String tag: allTags) {
+		          menuItem = new JMenuItem(tag);
+		          menuItem.addActionListener(new ActionListener() {
+		               public void actionPerformed(ActionEvent e) {
+		                    ruleset.tagRule(listView.getSelectedValue().toString(), e.getActionCommand());
+		               }
+		          });
+		          subMenu.add(menuItem);
+		     }
+		} catch (CoreException e) {
 	   		 logger.warning("Could not load tags from the core.");
-	   	}
-	    menuItem = new JMenuItem("New Tag...");
-	    menuItem.addActionListener(new ActionListener() {	
+		}
+		menuItem = new JMenuItem("New Tag...");
+		menuItem.addActionListener(new ActionListener() {	
 			public void actionPerformed(ActionEvent e) {
 				String tag = JOptionPane.showInputDialog("Tag:");
[…]

### 2012-03-07 gui/src/quanto/gui/RewriteViewer.java
Fix highlighting bug in rewrite viewer (c.f. issue #33)

@@ -84,7 +84,7 @@ public class RewriteViewer extends JFrame {
 				public void mouseEntered(MouseEvent e) {
 					CoreGraph match =
 						RewriteViewer.this.rewrites
-							.get(thisIndex).getLhs();
+							.get(thisIndex).getNewGraph();
 					RewriteViewer.this.vis.highlightSubgraph(match);
 				}
 				

### 2012-03-07 gui/src/quanto/gui/QuantoFrame.java
Make the splitpane divider a bit thicker (cf. issue #46)

@@ -232,6 +232,7 @@ public class QuantoFrame extends JFrame implements ViewPortHost {
         splitPane.setLeftComponent(sidebar);
         splitPane.setRightComponent(viewPort);
         splitPane.setDividerLocation(150);
+        splitPane.setDividerSize(15);
         splitPane.setOneTouchExpandable(true);
 
         getContentPane().add(splitPane, BorderLayout.CENTER);

### 2012-05-10 docs/matching_algo/matching_algo.tex
added matching algo doc

\documentclass{article}
\usepackage{fullpage}
\usepackage{graphicx}

\begin{document}

\section{Match State}

A match state consists of:

itemize

\section{Concrete Matching}

\begin{center}
    \includegraphics[width=16cm]{concrete_part.pdf}
\end{center}

\end{document}

### 2012-05-11 docs/matching_algo/matching_algo.tex
working on algo

@@
 
-A match state consists of:
+For a string graph $G$, let $N(G)$ be the set of node-vertices and $W(G)$ the wire-vertices.
 
-itemize
+Let $L$ be the pattern and $G$ the target. A match state consists of:
 
-\section{Concrete Matching}
+\begin{itemize}
+    \item $U_c \subseteq W(L)$ : a set of unmatched circles
+    \item $U_w \subseteq W(L)$ : a set of unmatched (non-circle) wire-vertices
+    \item $U_n \subseteq N(L)$ : a set of unmatched node-vertices
+    \item $P \subseteq N(L)$ : a set of partially-matched node-vertices
+    \item $P_s \subseteq P$ : a set of partially-matched node-vertices scheduled for re-checking
+    \item $m : V_L \rightarrow V_G$ : a partial injective function of the current match. When this is total and local iso, this match is complete.
+\end{itemize}
+
+\section{Inner Loop: Concrete Matching}
+
+The goal of the concrete matching step is to match all of the concrete vertices in the graph.
+If a node-vertex is not connected to a !-box, its complete neighbourhood will also be matched.
+If it is connected to one or more !-boxes, only some if its neighbourhood is matched.
+We refer to these as \textit{partially matched} node-vertices.
+A node-vertex $v$ such that the neighbourhood of $m(v)$ is totally covered is called \textit{completely matched}.
 
 \begin{center}
-    \includegraphics[width=16cm]{concrete_part.pdf}
+    \includegraphics[height=16cm]{concrete_part.pdf}
+\end{center}
+
+\section{Outer Loop: Pattern Expansion}
+
+\begin{center}
+    \includegraphics[height=16cm]{pattern_part.pdf}
 \end{center}

### 2012-05-11 docs/matching_algo/matching_algo.tex
Add pre- and post-conditions for concrete matching

@@
 \documentclass{article}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{fullpage}
@@
 
+\DeclareMathOperator{\dom}{dom}
+\DeclareMathOperator{\codom}{cod}
+
 \begin{document}
+
+\section{Normalisation}
+
+We normalise both the pattern and target graphs by ensuring that
+\begin{itemize}
+\item each wire connected to a node-vertex at both ends has exactly two wire-vertices on it
+\item boundary vertices are adjacent to either node-vertices or another boundary vertex
+\item circles (wires in a loop with no node-vertices) have exactly two wire-vertices on them
+\end{itemize}
 
@@
 
+\subsection{Preconditions}
+\begin{itemize}
+\item $L$ and $G$ are both normalised
+\item $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
+\item each vertex in $U_w$ is adjacent to something in $U_n\cup P$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P$)
+\item $P \subseteq \dom(m)$ ($P$ contains either partially-matched or completely-matched vertices)
+\item $U_c$ genuinely does contain circles
+\end{itemize}
+
+\subsection{Postconditions}
+These are true for the match state that results from each successful branch:
+\begin{itemize}
+\item $U_c = U_w = U_n = P = \varnothing$ (we cleaned up after ourselves)
+\item $P$ is exactly the set of vertices in $\dom(m)$ that are adjacent to a vertex not in $\dom(m)$ ($P$ contains exactly the partially-matched vertices)
+\item $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched everything we were asked to, and not lost anything in the process)
+\item $m$ induces a graph monomorphism from the full subgraph of $L$ described by $\dom(m)$ to the full subgraph of $G$ described by $\codom(m)$ ($m$ is actually a partial matching from $L$ to $G$)
+\end{itemize}
+
 \section{Outer Loop: Pattern Expansion}

### 2012-05-12 docs/matching_algo/matching_algo.tex
writing more on matching

@@
 
+Note that circles are given as the name of smallest wire-vertex in the circle, w.r.t. the name ordering.
+
 \section{Inner Loop: Concrete Matching}
@@
 
-\begin{center}
-    \includegraphics[height=16cm]{concrete_part.pdf}
-\end{center}
+The concrete matching routine is given a function ``doomed'', which returns true if a match of a concrete node-vertex is impossible.
+For !-box matching this returns true if the (concrete) neighbourhood of $v$ contains fewer in-edges (resp.
+out-edges) than that of $m(v)$ and there are no !-boxes connected to $v$ by an in-edge (resp.
+out-edge).
+It also takes a match state, subject to the following preconditions:
 
@@
 
+\begin{center}
+  \includegraphics[height=16cm]{concrete_part.pdf}
+\end{center}
+
 \subsection{Postconditions}
@@
 \begin{itemize}
-\item $U_c = U_w = U_n = P = \varnothing$ (we cleaned up after ourselves)
+  \item $U_c = U_w = U_n = P_s = \varnothing$ (every vertex marked for matching was handled)
+  \item $m$ is a partial matching from $L$ to $G$ that satisfies local iso condition for concrete vertices not in $P$,
   \item $P$ is exactly the set of vertices in $\dom(m)$ that are adjacent to a vertex not in $\dom(m)$ ($P$ contains exactly the partially-matched vertices)
   \item $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched everything we were asked to, and not lost anything in the process)
-\item $m$ induces a graph monomorphism from the full subgraph of $L$ described by $\dom(m)$ to the full subgraph of $G$ described by $\codom(m)$ ($m$ is actually a partial matching from $L$ to $G$)
 \end{itemize}
 
-\section{Outer Loop: Pattern Expansion}
+\section{Testing Wrapper: Concrete Matching Only}
+
+For testing purposes, it is straight-forward to wrap the above procedure to just do concrete matching.
+
+\begin{center}
+  \includegraphics[height=12cm]{concrete_wrapper.pdf}
+\end{center}
+
+\section{Full Wrapper: Lazy !-Box Expansion}
+
+Since we are only matching !-graphs on to concrete graphs, it suffices to restrict !-box operations to only two: KILL and EXPAND (COPY, then DROP the new box).
 

### 2012-05-14 docs/matching_algo/matching_algo.tex
Start showing correctness of matching inner loop

--- 2012-05-11 docs/matching_algo/matching_algo.tex
@@
 
+\DeclareMathOperator{\dom}{dom}
+
 \begin{document}
 
+\section{Normalisation}
+\label{sec:normalisation}
+
+We normalise both the pattern and target graphs by ensuring that
+\begin{itemize}
+\item each wire connected to a node-vertex at both ends has exactly two wire-vertices on it
+\item boundary vertices are adjacent to either node-vertices or another boundary vertex
+\item circles (wires in a loop with no node-vertices) have exactly two wire-vertices on them
+\end{itemize}
+
+\section{Bare Wires and Circles}
+\label{sec:wires-and-circles}
+
+Two forms of graph component require special attention: \emph{circles}, which are wires forming a closed loop with no node-vertices, and \emph{bare wires}, which are graph components equivalent to isolated points.
+A circle is normalised to two wire-vertices with two edges between them, one in each direction.
+A bare wire is normalised to an input connected directly to an output.
+Normalisation will ensure that any wire-vertex not in a circle or bare wire will be adjacent to exactly one node-vertex.
+
+Circles can only ever match against other circles, and all matchings for a given rule and target that are identical except on circles will produce isomorphic graphs after rewriting.
+Therefore, all that is relevant when matching circles is the number of circles in the pattern and the target.
+Without $!$-boxes, it is sufficient to check that there are at least as many circles in the target as in the pattern, and pick an arbitrary matching of circles; this does not require any branching.
+Circle matching is therefore always done first in any round of conrete matching.
+More circle matching may be necessary after expanding $!$-boxes in the pattern, as this may produce more circles.
+
+Bare wires are more complicated, as they can match any wire in the target, not just bare wires.
+This can create a huge amount of branching.
+We therefore leave wire matching until right at the end of the matching process, so that the ``hard'' part of the matching is not repeated unnecessarily.
+When matching a bare wire, we need to choose an unmatched edge in the target graph (one where either the edge's source or target are not matched, or where the edge's source and target are matched by an output and an input respectively) and split it into three edges joined with two wire-vertices.
+The bare wire is then matched against these two new wire-vertices and the edge between them.
+
 \section{Match State}
+\label{sec:match-state}
 
@@
 \begin{itemize}
-    \item $U_c \subseteq W(L)$ : a set of unmatched circles
+    \item $U_c \subseteq W(L)\times W(L)$ : a set of unmatched circles, as pairs of wire-vertices
     \item $U_w \subseteq W(L)$ : a set of unmatched (non-circle) wire-vertices
@@
     \item $P_s \subseteq P$ : a set of partially-matched node-vertices scheduled for re-checking
-    \item $m : V_L \rightarrow V_G$ : a partial injective function of the current match. When this is total and local iso, this match is complete.
+    \item $m : V_L \rightarrow V_G$ : a partial function describing the match so far
 \end{itemize}
 
+Throughout the entire procedure, we maintain the following global invariants:
+\begin{enumerate}
+  \renewcommand{\theenumi}{(\arabic{enumi})}
+  \renewcommand{\labelenumi}{\theenumi}
+  \item \label{enum:m-inv-inj} $m$ is injective and respects vertex types
+  \item \label{enum:m-inv-local-iso} For any node-vertex $v \in \dom(m)\setminus P$, all vertices adjacent to $m(v)$ in $G$ are in the image of $m$
+  \item \label{enum:m-inv-edge-match} For any $v,w \in \dom(m)$, if there is an edge from $v$ to $w$ in $L$, then there is an edge of the same type from $m(v)$ to $m(w)$ in $G$
+\end{enumerate}
+
+Matching is complete when $m$ is total and $P$ is empty.
+Providing we maintain the above invariants and ensure that $L$ and $G$ are simple graphs, $m$ will induce a unique graph monomorphism $\hat m:L \rightarrow G$, and this will be a local isomorphism.
+
+That $\hat m$ exists and is the unique morphism from $L$ to $G$ that restricts to $m$ can be seen by construction: invariant \ref{enum:m-inv-edge-match} ensures that a mapping consistent with $m$ can be found for each edge in $L$ and simplicity of $G$ ensures that there is only one such mapping.
+Invariant \ref{enum:m-inv-inj} ensures the vertex-function (ie: $m$) is injective, and so simplicity of $L$ ensures that the edge function is also injective, and invariants \ref{enum:m-inv-inj} and \ref{enum:m-inv-edge-match} ensure that both functions respect typing, and hence $\hat m$ is a monomorphism.
+Invariant \ref{enum:m-inv-local-iso} ensures that $\hat m$ will be a local isomorphism.
+
+Providing the match state satisfies the above invariants, $P$ is empty and $m$ is total everywhere except on bare wires, the bare wire matching procedure given in section \ref{sec:wires-and-circles} maintains the invariants and makes $m$ total on $V_L$.
+
 \section{Inner Loop: Concrete Matching}
+\label{sec:inner-loop}
 
@@
 
+The concrete matching routine is given a function ``doomed'', which returns true if a match of a concrete node-vertex is impossible.
+For !-box matching this returns true if the (concrete) neighbourhood of $v$ contains fewer in-edges (resp.
+out-edges) than that of $m(v)$ and there are no !-boxes connected to $v$ by an in-edge (resp.
+out-edge).
+It also takes a match state, subject to the following preconditions:
+
+\subsection{Preconditions}
+\begin{enumerate}
+  \renewcommand{\theenumi}{(\arabic{enumi})}
+  \renewcommand{\labelenumi}{\theenumi}
+  \item \label{enum:il-precond-norm} $L$ and $G$ are both normalised
+  \item \label{enum:il-precond-inv} The match state satisfies the global invariants
+  \item \label{enum:il-precond-unmatched} $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
+  \item \label{enum:il-precond-reachable} Each vertex in $U_w$ is adjacent to something in $U_n\cup P$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P$)
+  \item \label{enum:il-precond-P-matched} $P \subseteq \dom(m)$ ($P$ contains either partially-matched or completely-matched vertices)
+  \item \label{enum:il-precond-circles} $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+  \item \label{enum:il-precond-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+\end{enumerate}
+
+\subsection{Algorithm}
 \begin{center}
@@
 
-\section{Outer Loop: Pattern Expansion}
+\subsection{Postconditions}
+These are true for the match state that results from each successful branch:
+\begin{enumerate}
+  \renewcommand{\theenumi}{(\arabic{enumi})}
+  \renewcommand{\labelenumi}{\theenumi}
+  \item \label{enum:il-postcond-inv} The match state satisfies the global invariants
+  \item \label{enum:il-postcond-no-unmatched} $U_c = U_w = U_n = P_s = \varnothing$ (every vertex marked for matching was handled)
+  \item \label{enum:il-postcond-P} $P$ is exactly the set of vertices in $\dom(m)$ that are adjacent to a vertex not in $\dom(m)$ ($P$ contains exactly the partially-matched vertices)
+  \item \label{enum:il-postcond-all-matched} $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched everything we were asked to, and not lost anything in the process)
+  \item \label{enum:il-postcond-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+\end{enumerate}
+
+\subsection{Correctness}
+
+\subsubsection{Invariants}
+
+\begin{enumerate}
+  \renewcommand{\theenumi}{(\roman{enumi})}
+  \renewcommand{\labelenumi}{\theenumi}
+  \item \label{enum:il-inv-circles} $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+
+	This is guaranteed at the start by precondition \ref{enum:il-precond-circles}, and trivially maintained by the fact that nothing is ever added to either of these sets.
+  \item \label{enum:il-inv-unmatched} $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$
+
+	This is guaranteed at the start by precondition \ref{enum:il-precond-unmatched}.
+Nothing is ever added to $U_c$, $U_w$ or $U_n$.
+There are three places where vertices are added to $m$.
+The first is when matching circles, when we match the contents of $U_c$ and set $U_c$ to be empty.
+We cannot add anything from $U_w$ at this point, due to \ref{enum:il-inv-circles}, and we cannot add anything from $U_n$, since that contains only node-vertices.
+The second is when $P_s$ is empty and we take a vertex from $U_n$ and add its matching to $m$, which clearly maintains the invariant.
+In the final case all the wire-vertices added to $m$ are removed from $U_w$ at the same time and if a node-vertex is added to $m$ it has already been removed from $U_n$.
+In both the final two cases, $U_c$ is already empty.
+  \item \label{enum:il-inv-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+
+This is satisfied at the start of the algorithm by precondition \ref{enum:il-precond-wire-connected}.
+There are only two places we add wire-vertices: when matching circles and when matching wires connected to a partially matched node-vertex.
+In the former case, each wire-vertex we add to $m$ is adjacent to exactly one other wire-vertex, which we also add to $m$.
+In the latter case, we add the entire wire, including the end-point not already in $m$, to $m$.
+So both cases preserve the invariant.
+\end{enumerate}
+
+One consequence of these invariants is that $m$ is only ever extended by this algorithm: we never alter what any vertex already in $\dom(m)$ maps to.
+This can be seen by noting that when matching circles, we only ever add wire-vertices that are in $U_c$; when $P_s$ is empty we only ever add a single node-vertex, which is drawn from $U_n$, and when exploring wires connected to a vertex in $P$, normalisation ensures that if the wire terminates with a wire-vertex, it must be the only wire-vertex on the wire, which means that we only add a single vertex to $m$, and that vertex is drawn from $U_w$.
+Otherwise, it must terminated with a node-vertex and the wire must contain exactly two wire-vertices.
+The first is in $U_w$, and hence not in $\dom(m)$, and so the second (which it is adjacent to) cannot be in $\dom(m)$ by invariant \ref{enum:il-inv-wire-connected}.
+The only other vertex added to $m$ is the terminating node-vertex.
+If this is in $P$, it is only added if the matchings agree (which doesn't change anything).
+If not, the node-vertex cannot be in $\dom(m)$ by global invariant \ref{enum:m-inv-local-iso} and the fact that its image is adjacent to a wire-vertex not in the image of $m$ (since we only consider partial matchings whose image does not intersect with the image of $m$).
+
+\subsubsection{Global Invariants}
+
+That $m$ remains injective and continues to respect typing follows from the fact that it is only ever extended with valid partial matchings whose image is disjoint from the full subgraph defined by the image of $m$.
+
+There are three points where we extend $m$, and we will consider them independently for the other global invariants.
+
+One of the two simpler cases is where we match circles.
+The partial match that is added encompasses the entire graph component connected to the vertices we add to $m$, and so global invariant \ref{enum:m-inv-edge-match} is maintained.
+We only add wire-vertices, so global invariant \ref{enum:m-inv-local-iso} is trivially maintained.
+
+The other simple case is where $P_s$ is empty and we extend $m$ with a matching from an element of $U_n$.
+The vertex is added to $P$, and so global invariant \ref{enum:m-inv-local-iso} is trivially satisfied.
+Global invariant \ref{enum:m-inv-edge-match} is also maintained as none of the adjacent (wire-)vertices can be in $\dom(m)$ by local invariant \ref{enum:il-inv-wire-connected}.
+
+The final case we need to consider is where we add a wire connected to something in $P$.
+If we add a node-vertex that is not already in $P$ to $m$, it is added to $P$, and so global invariant \ref{enum:m-inv-local-iso} is trivially satisfied.
+For global invariant \ref{enum:m-inv-edge-match}, if the matched wire ended in a wire-vertex, only one vertex was added, and the only incident edge to that vertex has a corresponding edge in the target.
+Otherwise, a wire terminated with a node-vertex was added.
+All the incident edges of the two wire-vertices on the wire have corresponding edges in the target (by the matching), so all we have to consider are the incident edges of the node-vertex.
+If it was already in $\dom(m)$, we already know the invariant holds.
+Otherwise, none of the incident wire-vertices could have already been in $\dom(m)$, by invariant \ref{enum:il-inv-wire-connected}, and so global invariant \ref{enum:m-inv-edge-match} must be maintained.
+
+\subsubsection{Termination}
+
+For termination of the inner loop exploring the wires incident to a node-vertex $v$, we consider the variant $n = |U_w\cap N_L(v)|$, where $N_L(v)$ is the neighbourhood of $v$ in $L$.
+In each of these branches, $n$ decreases by either $1$ or $2$, and nothing is ever added to $U_w$ (and $N_L(v)$ is fixed).
+So this loop must terminate.
+
+The loop working through $P_s$ has two branches: when it is not empty and when it is not.
+Either may add more things to $P_s$.
+We will take as variant $m = 2|U_n| + |P_s|$.
+When this is $0$, the algorithm will terminate.
+Otherwise, if $P_s$ is empty, $U_n$ will have one vertex removed and added to $P_s$, decreasing $m$ by one.
+If $P_s$ is not empty, one vertex will be removed from $P_s$, decreasing $m$ by one.
+Every vertex that is added to $P_s$ in this loop is taken from $U_n$, further decreasing $m$.
+As a result, $m$ will decrease by at least one with every iteration of the loop.
+
+\subsubsection{Postconditions}
+
+Postcondition \ref{enum:il-postcond-inv} has already been shown.  Postcondition \ref{enum:il-postcond-wire-connected} is guaranteed by invariant \ref{enum:il-inv-wire-connected}
+
+\section{Testing Wrapper: Concrete Matching Only}
+\label{sec:test-wrapper}
+
+For testing purposes, it is straight-forward to wrap the above procedure to just do concrete matching.  When we populate the match state, we do it in the following manner:
+\begin{itemize}
+\item $U_c$ will contain all circles
+\item $U_w$ will contain all other wire vertices, except those forming bare wires (boundary vertices connected to another boundary vertex)
+\item $U_n$ will contain all node vertices
+\item $P$, $P_s$ and $m$ will be empty
+\end{itemize}
+
+This will ensure the preconditions for the concrete matching inner-loop will be satisfied (assuming normalisation).
+The only precondition that is not trivially satisfied is \ref{enum:il-precond-reachable}, but this follows from the fact that normalisation ensures that the only wire-vertices not adjacent to a node-vertex are those in circles and bare wires.
+
+\begin{center}
+  \includegraphics[height=12cm]{concrete_wrapper.pdf}
+\end{center}
+
+The postconditions of the concrete matching inner-loop, together with how we created the initial match state, ensure that a successful termination of this algorithm produces a valid matching, since $m$ will be total after matching bare wires and the global invariants from section \ref{sec:match-state} are maintained.
+
+\section{Full Wrapper: Lazy !-Box Expansion}
+\label{sec:full-wrapper}
+
+Since we are only matching pattern graphs on to concrete graphs, it suffices to restrict $!$-box operations to only two: KILL and EXPAND (COPY, then DROP the new box).
 

### 2012-05-15 docs/matching_algo/matching_algo.tex

--- 2012-05-12 docs/matching_algo/matching_algo.tex
@@
 \end{itemize}
+
+\section{Bare Wires and Circles}
+
+Two forms of graph component require special attention: \emph{circles}, which are wires forming a closed loop with no node-vertices, and \emph{bare wires}, which are graph components equivalent to isolated points.
+A circle is normalised to two wire-vertices with two edges between them, one in each direction.
+A bare wire is normalised to an input connected directly to an output.
+
+Circles can only ever match against other circles, and all matchings for a given rule and target that are identical except on circles will produce isomorphic graphs after rewriting.
+Therefore, all that is relevant when matching circles is the number of circles in the pattern and the target.
+Without $!$-boxes, it is sufficient to check that there are at least as many circles in the target as in the pattern, and pick an arbitrary matching of circles; this does not require any branching.
+Circle matching is therefore always done first in any round of conrete matching.
+More circle matching may be necessary after copying $!$-boxes in the pattern, as this may produce more circles.
+
+Bare wires are more complicated, as they can match any wire in the target, not just bare wires.
+This can create a huge amount of branching.
+We therefore leave wire matching until right at the end of the matching process, so that the ``hard'' part of the matching is not repeated unnecessarily.
+When matching a bare wire, we need to choose an unmatched edge in the target graph (one where either the edge source or target are not matched, or where the edge source and target are matched by an input and an output respectively) and split it into three edges joined with two wire-vertices.
+The bare wire is then matched against these two new wire-vertices and the edge between them.
 
@@
 \subsection{Preconditions}
-\begin{itemize}
+\begin{enumerate}
   \item $L$ and $G$ are both normalised
   \item $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
-  \item each vertex in $U_w$ is adjacent to something in $U_n\cup P$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P$)
+  \item Each vertex in $U_w$ is adjacent to something in $U_n\cup P$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P$)
   \item $P \subseteq \dom(m)$ ($P$ contains either partially-matched or completely-matched vertices)
   \item $U_c$ genuinely does contain circles
-\end{itemize}
+\end{enumerate}
 
+\subsection{Algorithm}
 \begin{center}
@@
 These are true for the match state that results from each successful branch:
-\begin{itemize}
+\begin{enumerate}
   \item $U_c = U_w = U_n = P_s = \varnothing$ (every vertex marked for matching was handled)
@@
   \item $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched everything we were asked to, and not lost anything in the process)
-\end{itemize}
+\end{enumerate}
 
@@
 
-For testing purposes, it is straight-forward to wrap the above procedure to just do concrete matching.
+For testing purposes, it is straight-forward to wrap the above procedure to just do concrete matching.  When we populate the match state, we do it in the following manner:
+\begin{itemize}
+\item $U_c$ will contain all circles
+\item $U_w$ will contain all other wire vertices, except those forming bare wires (boundary vertices connected to another boundary vertex)
+\item $U_n$ will contain all node vertices
+\item $P$, $P_s$ and $m$ will be empty
+\end{itemize}
+
+This will ensure the preconditions for the concrete matching inner-loop will be satisfied (assuming normalisation).
+The only precondition that is not trivially satisfied is (3), but this follows from the fact that normalisation ensures that the only wire-vertices not adjacent to a node-vertex are those in circles and bare wires.
 
@@
 \end{center}
+
+The postconditions of the concrete matching inner-loop, together with how we created the initial match state, ensure that, if $P$ is empty, $m$ is a valid matching (satisfying the local isomorphism condition) from all of the pattern graph, except its bare wires, to the target graph.
+Once bare wires are added, we have only valid matchings.
 

### 2012-05-15 docs/matching_algo/matching_algo.tex
Finish proving correctness of matching inner loop

--- 2012-05-14 docs/matching_algo/matching_algo.tex
@@
 \usepackage{amssymb}
+\usepackage{enumitem}
 \usepackage{fullpage}
@@
 \DeclareMathOperator{\dom}{dom}
+\DeclareMathOperator{\im}{im}
 
@@
 
-Throughout the entire procedure, we maintain the following global invariants:
+Throughout the entire procedure, in addition to the invariants implicit in the typing of the components of the match state, we maintain the following global invariants:
 \begin{enumerate}
@@
   \renewcommand{\labelenumi}{\theenumi}
-  \item \label{enum:m-inv-inj} $m$ is injective and respects vertex types
-  \item \label{enum:m-inv-local-iso} For any node-vertex $v \in \dom(m)\setminus P$, all vertices adjacent to $m(v)$ in $G$ are in the image of $m$
-  \item \label{enum:m-inv-edge-match} For any $v,w \in \dom(m)$, if there is an edge from $v$ to $w$ in $L$, then there is an edge of the same type from $m(v)$ to $m(w)$ in $G$
+  \item \label{enum:g-inv-inj} $m$ is injective and respects vertex types
+  \item \label{enum:g-inv-local-iso} $P \subseteq \dom(m)$ and for any node-vertex $v \in \dom(m)\setminus P$, all vertices adjacent to $m(v)$ in $G$ are in the image of $m$
+  \item \label{enum:g-inv-edge-match} For any $v,w \in \dom(m)$, if there is an edge from $v$ to $w$ in $L$, then there is an edge of the same type from $m(v)$ to $m(w)$ in $G$
 \end{enumerate}
@@
 
-That $\hat m$ exists and is the unique morphism from $L$ to $G$ that restricts to $m$ can be seen by construction: invariant \ref{enum:m-inv-edge-match} ensures that a mapping consistent with $m$ can be found for each edge in $L$ and simplicity of $G$ ensures that there is only one such mapping.
-Invariant \ref{enum:m-inv-inj} ensures the vertex-function (ie: $m$) is injective, and so simplicity of $L$ ensures that the edge function is also injective, and invariants \ref{enum:m-inv-inj} and \ref{enum:m-inv-edge-match} ensure that both functions respect typing, and hence $\hat m$ is a monomorphism.
-Invariant \ref{enum:m-inv-local-iso} ensures that $\hat m$ will be a local isomorphism.
+That $\hat m$ exists and is the unique morphism from $L$ to $G$ that restricts to $m$ can be seen by construction: invariant \ref{enum:g-inv-edge-match} ensures that a mapping consistent with $m$ can be found for each edge in $L$ and simplicity of $G$ ensures that there is only one such mapping.
+Invariant \ref{enum:g-inv-inj} ensures the vertex-function (ie: $m$) is injective, and so simplicity of $L$ ensures that the edge function is also injective, and invariants \ref{enum:g-inv-inj} and \ref{enum:g-inv-edge-match} ensure that both functions respect typing, and hence $\hat m$ is a monomorphism.
+Invariant \ref{enum:g-inv-local-iso} ensures that $\hat m$ will be a local isomorphism.
 
@@
 \subsection{Preconditions}
+\label{sec:il-precond}
+
+These preconditions are actually invariants of the algorithm.
+
 \begin{enumerate}
-  \renewcommand{\theenumi}{(\arabic{enumi})}
+  \renewcommand{\theenumi}{(\roman{enumi})}
   \renewcommand{\labelenumi}{\theenumi}
-  \item \label{enum:il-precond-norm} $L$ and $G$ are both normalised
-  \item \label{enum:il-precond-inv} The match state satisfies the global invariants
-  \item \label{enum:il-precond-unmatched} $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
-  \item \label{enum:il-precond-reachable} Each vertex in $U_w$ is adjacent to something in $U_n\cup P$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P$)
-  \item \label{enum:il-precond-P-matched} $P \subseteq \dom(m)$ ($P$ contains either partially-matched or completely-matched vertices)
-  \item \label{enum:il-precond-circles} $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
-  \item \label{enum:il-precond-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+  \item \label{enum:il-inv-norm} $L$ and $G$ are both normalised
+  \item \label{enum:il-inv-inv} The match state satisfies the global invariants
+  \item \label{enum:il-inv-unmatched} $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
+  \item \label{enum:il-inv-reachable} Each vertex in $U_w$ is adjacent to something in $U_n\cup P_s$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P_s$)
+  \item \label{enum:il-inv-Ps} If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_s$ (everything that is completely matched and in $P$ is also in $P_s$)
+  \item \label{enum:il-inv-wire-uw} If $v \in U_w$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_w$.
+  \item \label{enum:il-inv-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+  \item \label{enum:il-inv-circles} $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
 \end{enumerate}
@@
 \subsection{Algorithm}
+
+A note on the algorithm presentation: red wires (with labels starting ``foreach'') indicate a branching point.
+A branch is created for each possibility.
+If there are no possibilities, the current branch dies.
+So, for example, if we take a vertex from $U_n$ and there are no possible matchings for it in the target graph, we kill off the current branch.
+
+Note also that when the algorithm calls for a matching, it is implicitly assumed that the image of the matching should not intersect with the image of $m$.
+
 \begin{center}
@@
 \begin{enumerate}
-  \renewcommand{\theenumi}{(\arabic{enumi})}
+  \renewcommand{\theenumi}{(\Roman{enumi})}
   \renewcommand{\labelenumi}{\theenumi}
@@
   \item \label{enum:il-postcond-no-unmatched} $U_c = U_w = U_n = P_s = \varnothing$ (every vertex marked for matching was handled)
-  \item \label{enum:il-postcond-P} $P$ is exactly the set of vertices in $\dom(m)$ that are adjacent to a vertex not in $\dom(m)$ ($P$ contains exactly the partially-matched vertices)
-  \item \label{enum:il-postcond-all-matched} $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched everything we were asked to, and not lost anything in the process)
-  \item \label{enum:il-postcond-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+  \item \label{enum:il-postcond-all-matched} $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched exactly what we were asked to)
+  \item \label{enum:il-postcond-P} $P$ is exactly the set of vertices in $\dom(m)$ whose image is adjacent to a vertex not in $\im(m)$ ($P$ contains exactly the partially-matched vertices)
 \end{enumerate}
+
+Postconditions \ref{enum:il-postcond-inv} and \ref{enum:il-postcond-all-matched} state that the algorithm has completed the requested work.
+Postcondition \ref{enum:il-postcond-no-unmatched} just states that we left the match state ``clean''; it still satisfies all the preconditions and repeating the algorithm without altering the state further would do nothing.
 
@@
 
+First, we note that our data structure constraints are never broken: we never add anything to $U_c$, $U_w$ or $U_n$, so these cannot contain objects of the wrong type.
+When we add things to $P$, they are always either taken from $U_n$ or are the end of a wire that does not terminate in a vertex in $U_w$; normalisation and precondition \ref{enum:il-inv-wire-uw} guarantee that the vertex is a node-vertex in this case.
+Finally, when removing vertices from $P$ they are removed from $P_s$ simultaneously, and when adding vertices to $P_s$ they are added to $P$ at the same time.
+
 \subsubsection{Invariants}
 
-\begin{enumerate}
-  \renewcommand{\theenumi}{(\roman{enumi})}
-  \renewcommand{\labelenumi}{\theenumi}
-  \item \label{enum:il-inv-circles} $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+The preconditions (section \ref{sec:il-precond}) are maintained throughout the algorithm as invariants:
 
-	This is guaranteed at the start by precondition \ref{enum:il-precond-circles}, and trivially maintained by the fact that nothing is ever added to either of these sets.
-  \item \label{enum:il-inv-unmatched} $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$
+\begin{enumerate}[label=(\roman*),ref=(\roman*)]
+  \item $L$ and $G$ are both normalised
 
-	This is guaranteed at the start by precondition \ref{enum:il-precond-unmatched}.
+  This is trivially maintained as we alter neither graph.
+  \item The match state satisfies the global invariants
+
+  See section \ref{sec:il-global-inv}.
+  \item $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$
+
   Nothing is ever added to $U_c$, $U_w$ or $U_n$.
@@
 The first is when matching circles, when we match the contents of $U_c$ and set $U_c$ to be empty.
-We cannot add anything from $U_w$ at this point, due to \ref{enum:il-inv-circles}, and we cannot add anything from $U_n$, since that contains only node-vertices.
+We cannot add anything from $U_w$ at this point, due to invariant \ref{enum:il-inv-circles}, and we cannot add anything from $U_n$, since that contains only node-vertices.
 The second is when $P_s$ is empty and we take a vertex from $U_n$ and add its matching to $m$, which clearly maintains the invariant.
@@
 In both the final two cases, $U_c$ is already empty.
-  \item \label{enum:il-inv-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+  \item Each vertex in $U_w$ is adjacent to something in $U_n\cup P_s$
 
-This is satisfied at the start of the algorithm by precondition \ref{enum:il-precond-wire-connected}.
+  We never add anything to $U_w$, so we only have to consider the removal of vertices from $U_n$ and $P_s$.
+When we remove a vertex from $U_n$, we always immediately add it to $P_s$, and we only remove a vertex from $P_s$ when there are no adjacent vertices in $U_w$.
+  \item If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_s$ (everything that is completely matched and in $P$ is also in $P_s$)
+
+  When we add anything to $P$, we also add it to $P_s$, and when we remove anything from $P_s$, we remove it from $P$ if it is completely matched.
+  \item If $v \in U_w$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_w$.
+
+  When we remove a vertex from $U_w$, we remove the entire wire that it is on.
+  \item If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+
+  We never remove anything from $\dom(m)$ (invariant \ref{enum:il-inv-extend}).
 There are only two places we add wire-vertices: when matching circles and when matching wires connected to a partially matched node-vertex.
-In the former case, each wire-vertex we add to $m$ is adjacent to exactly one other wire-vertex, which we also add to $m$.
+In the former case, each wire-vertex we add to $m$ is adjacent to exactly one other vertex, which we also add to $m$.
 In the latter case, we add the entire wire, including the end-point not already in $m$, to $m$.
 So both cases preserve the invariant.
+  \item $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+
+  This is trivially maintained as we never add anything to either set.
 \end{enumerate}
 
-One consequence of these invariants is that $m$ is only ever extended by this algorithm: we never alter what any vertex already in $\dom(m)$ maps to.
-This can be seen by noting that when matching circles, we only ever add wire-vertices that are in $U_c$; when $P_s$ is empty we only ever add a single node-vertex, which is drawn from $U_n$, and when exploring wires connected to a vertex in $P$, normalisation ensures that if the wire terminates with a wire-vertex, it must be the only wire-vertex on the wire, which means that we only add a single vertex to $m$, and that vertex is drawn from $U_w$.
+To these, we add an additional invariant:
+\begin{enumerate}[resume*]
+  \item \label{enum:il-inv-extend} $m$ restricts to its initial state (it is only ever extended, never changed in any other way)
+
+  This is clearly true at the start of the algorithm.
+When matching circles, we only ever add vertices that are in $U_c$, which is disjoint from $\dom(m)$ by invariant \ref{enum:il-inv-unmatched}.
+When $P_s$ is empty we only ever add a single node-vertex, which is drawn from $U_n$ (which, again by \ref{enum:il-inv-unmatched}, is disjoint from $\dom(m)$).
+
+  When exploring wires connected to a vertex in $P$, normalisation ensures that if the wire terminates with a wire-vertex, it must be the only wire-vertex on the wire, which means that we only add a single vertex to $m$, and that vertex is drawn from $U_w$, which is disjoint from $\dom(m)$.
 Otherwise, it must terminated with a node-vertex and the wire must contain exactly two wire-vertices.
-The first is in $U_w$, and hence not in $\dom(m)$, and so the second (which it is adjacent to) cannot be in $\dom(m)$ by invariant \ref{enum:il-inv-wire-connected}.
+The first is in $U_w$, and so the second (which it is adjacent to) must also be in $U_w$ by invariant \ref{enum:il-inv-wire-uw}, and hence neither can be in $\dom(m)$ by invariant \ref{enum:il-inv-unmatched}.
 The only other vertex added to $m$ is the terminating node-vertex.
 If this is in $P$, it is only added if the matchings agree (which doesn't change anything).
-If not, the node-vertex cannot be in $\dom(m)$ by global invariant \ref{enum:m-inv-local-iso} and the fact that its image is adjacent to a wire-vertex not in the image of $m$ (since we only consider partial matchings whose image does not intersect with the image of $m$).
+If not, the node-vertex cannot be in $\dom(m)$ by global invariant \ref{enum:g-inv-local-iso} and the fact that its image is adjacent to a wire-vertex not in the image of $m$ (since we only consider partial matchings whose image does not intersect with the image of $m$).
+\end{enumerate}
 
 \subsubsection{Global Invariants}
+\label{sec:il-global-inv}
 
-That $m$ remains injective and continues to respect typing follows from the fact that it is only ever extended with valid partial matchings whose image is disjoint from the full subgraph defined by the image of $m$.
+\begin{enumerate}
+  \renewcommand{\theenumi}{(\arabic{enumi})}
+  \renewcommand{\labelenumi}{\theenumi}
+  \item $m$ is injective and respects vertex types
 
-There are three points where we extend $m$, and we will consider them independently for the other global invariants.
+  This follows from the fact that it is only ever extended with valid partial matchings whose image is disjoint from the full subgraph defined by the image of $m$.
+  \item $P \subseteq \dom(m)$ and for any node-vertex $v \in \dom(m)\setminus P$, all vertices adjacent to $m(v)$ in $G$ are in the image of $m$
 
-One of the two simpler cases is where we match circles.
-The partial match that is added encompasses the entire graph component connected to the vertices we add to $m$, and so global invariant \ref{enum:m-inv-edge-match} is maintained.
-We only add wire-vertices, so global invariant \ref{enum:m-inv-local-iso} is trivially maintained.
+  The first part is maintained as any vertex added to $m$ is added to $P$.
+There are three points where we extend $m$.
+When we match circles, the partial match that is added encompasses the entire graph component connected to the vertices we add to $m$.
+In both the other cases, the vertex is added to $P$, and so the invariant is trivially satisfied.
 
-The other simple case is where $P_s$ is empty and we extend $m$ with a matching from an element of $U_n$.
-The vertex is added to $P$, and so global invariant \ref{enum:m-inv-local-iso} is trivially satisfied.
-Global invariant \ref{enum:m-inv-edge-match} is also maintained as none of the adjacent (wire-)vertices can be in $\dom(m)$ by local invariant \ref{enum:il-inv-wire-connected}.
+  \item For any $v,w \in \dom(m)$, if there is an edge from $v$ to $w$ in $L$, then there is an edge of the same type from $m(v)$ to $m(w)$ in $G$
+
+  There are three points where we alter $m$, and we will consider them independently.
+When we match circles, we only add wire-vertices, so the invariant is trivially maintained.
+When $P_s$ is empty and we extend $m$ with a matching from an element of $U_n$, the invariant is also maintained as none of the adjacent (wire-)vertices can be in $\dom(m)$, by invariant \ref{enum:il-inv-wire-connected}.
 
 The final case we need to consider is where we add a wire connected to something in $P$.
-If we add a node-vertex that is not already in $P$ to $m$, it is added to $P$, and so global invariant \ref{enum:m-inv-local-iso} is trivially satisfied.
-For global invariant \ref{enum:m-inv-edge-match}, if the matched wire ended in a wire-vertex, only one vertex was added, and the only incident edge to that vertex has a corresponding edge in the target.
+If the matched wire ended in a wire-vertex, only one vertex was added, and the only incident edge to that vertex has a corresponding edge in the target.
 Otherwise, a wire terminated with a node-vertex was added.
@@
 If it was already in $\dom(m)$, we already know the invariant holds.
-Otherwise, none of the incident wire-vertices could have already been in $\dom(m)$, by invariant \ref{enum:il-inv-wire-connected}, and so global invariant \ref{enum:m-inv-edge-match} must be maintained.
+Otherwise, none of the incident wire-vertices could have already been in $\dom(m)$, by invariant \ref{enum:il-inv-wire-connected}, and so global invariant \ref{enum:g-inv-edge-match} must be maintained.
+\end{enumerate}
 
 \subsubsection{Termination}
+\label{sec:termination}
 
@@
 
-Postcondition \ref{enum:il-postcond-inv} has already been shown.  Postcondition \ref{enum:il-postcond-wire-connected} is guaranteed by invariant \ref{enum:il-inv-wire-connected}
+\begin{enumerate}
+  \renewcommand{\theenumi}{(\Roman{enumi})}
+  \renewcommand{\labelenumi}{\theenumi}
+  \item The match state satisfies the global invariants
+
+  See section \ref{sec:il-global-inv}.
+  \item $U_c = U_w = U_n = P_s = \varnothing$
+
+  $U_c$ is emptied by the circle-matching step and is never added to.
+$P_s$ and $U_n$ must be empty when we return the match due to the termination condition ($m = 0$ in section \ref{sec:termination}).
+Invariant \ref{enum:il-inv-reachable} guarantees that if $U_n$ and $P_s$ are empty, then so must $U_w$ be.
+  \item $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$
+
+  Every time we remove a vertex from $U_c$, $U_w$ or $U_n$, we add it to $m$.
+Conversely, every time we add a vertex to $m$, it must be drawn from one of these sets (given invariant \ref{enum:il-inv-wire-uw}).
+This, coupled with postcondition \ref{enum:il-postcond-no-unmatched} and the fact that we only ever extend $m$, gives us postcondition \ref{enum:il-postcond-all-matched}.
+
+  \item $P$ is exactly the set of vertices in $\dom(m)$ whose image is adjacent to a vertex not in $\im(m)$
+
+  Global invariant \ref{enum:g-inv-local-iso} gives us most of what we need.
+We just need that if $v$ is completely matched, it is not in $P$.
+But this is guaranteed by invariant \ref{enum:il-inv-Ps} and the fact that $P_s = \varnothing$.
+\end{enumerate}
 
@@
 This will ensure the preconditions for the concrete matching inner-loop will be satisfied (assuming normalisation).
-The only precondition that is not trivially satisfied is \ref{enum:il-precond-reachable}, but this follows from the fact that normalisation ensures that the only wire-vertices not adjacent to a node-vertex are those in circles and bare wires.
+The only precondition that is not trivially satisfied is \ref{enum:il-inv-reachable}, but this follows from the fact that normalisation ensures that the only wire-vertices not adjacent to a node-vertex are those in circles and bare wires.
 

### 2012-05-16 docs/matching_algo/matching_algo.tex
Remove latex key from too-recent version of enumitem
Didn't build on Fedora 15
Prove concrete part preconds satisfied for pattern matching algorithm

@@
 To these, we add an additional invariant:
-\begin{enumerate}[resume*]
+\begin{enumerate}[resume,label=(\roman*),ref=(\roman*)]
   \item \label{enum:il-inv-extend} $m$ restricts to its initial state (it is only ever extended, never changed in any other way)
@@
 
+This algorithm performs lazy $!$-box expansion to find all possible matchings of a pattern graph against a string graph.
+In order for the algorithm to be deterministic, we disallow pattern graphs with empty $!$-boxes or $!$-boxes containing only bare wires, as these may be expanded any number of times and still match any graph that contains a wire.
+
 Since we are only matching pattern graphs on to concrete graphs, it suffices to restrict $!$-box operations to only two: KILL and EXPAND (COPY, then DROP the new box).
@@
 
+\subsection{Termination}
+
+Each step terminates, and there is only one loop.  As a variant, we take
+\[ n = |V_G| + |V_L\setminus\dom(m)| - |\dom(m)| \]
+
+We calculate the variant at the point where we consider terminating the loop, immediately after the concrete matching step.
+If the previous iteration involved killing one or more $!$-boxes, $V_L\setminus\dom(m)$ will decrease (by the combined size of the $!$-boxes).
+If the previous iteration involved expanding a $!$-box, $|\dom(m)|$ will increase (by the size of the $!$-box, excluding any bare wires).
+$V_L\setminus\dom(m)$ is clearly bounded below by $0$, and $|\dom(m)|$ is bounded below by $|V_G|$, by global invariant \ref{enum:g-inv-inj}, so $n$ is bounded below by $0$ and decreases with every iteration.
+Hence the algorithm terminates.
+
+\subsection{Concrete Matching Preconditions}
+
+We need to show that the preconditions of the concrete matching step are satisfied whenever it is reached.
+
+\begin{enumerate}[label=(\roman*),ref=(\roman*)]
+  \item $L$ and $G$ are both normalised
+
+  We normalise the graphs at the start.
+We never alter $G$ after that.
+Because $!$-boxes are open subgraphs, we always remove or copy an entire wire at once, which ensures that $L$ remains normalised throughout.
+  \item The match state satisfies the global invariants
+
+  Only the concrete matching step alters $m$ or $P$ in any way, and this maintains the global invariants.
+  \item $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$
+
+  We only ever add unmatched vertices to these sets.
+  \item Each vertex in $U_w$ is adjacent to something in $U_n\cup P_s$
+
+  This follows from openness of $!$-boxes, which (together with normalisation) ensures that any concrete wire-vertex that is not on a circle or bare wire must be adjacent to a concrete node-vertex.
+The first time concrete matching happens, all such node-vertices are put in $U_n$.
+
+  Since all non-bare-wire, non-circle concrete vertices are added to $U_w$ each time, and so added to $\dom(m)$ by the concrete matching step, on subsequent iterations of concrete matching any wire-vertex $v$ in $U_w$ must have arisen from the expansion of a $!$-box (since this is the only way concrete vertices can be added).
+If $v$ is adjacent to a node-vertex that came from the $!$-box expansion, that node-vertex will be added to $U_n$.
+Otherwise, it must be adjacent to a node-vertex in $\dom(m)$, and hence in $P$ (by the postconditions of the concrete matching step).
+Then this node-vertex (being adjacent to the expanded $!$-box) will be added to $P_s$.
+  \item If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_s$ (everything that is completely matched and in $P$ is also in $P_s$)
+
+  The first time we run the concrete matching step, $P$ is empty.
+After each concrete matching step, the postconditions ensure that there is no completely matched vertex in $P$.
+As the rest of the algorithm does not add anything to $P$ or alter $m$, this is maintained throughout.
+  \item If $v \in U_w$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_w$.
+
+  This is guaranteed by openness of $!$-boxes: we only ever add entire wires to $U_w$.
+  \item If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
+
+  This is guaranteed by the postcondition \ref{enum:il-postcond-all-matched} in combination with precondition \ref{enum:il-inv-wire-uw} and the fact that $m$ is only altered by the concrete matching step.
+  \item $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+
+  This is implicit in the step where we populate $U_w$ and $U_c$.
+\end{enumerate}
+
 \end{document}

### 2012-05-16 docs/matching_algo/matching_algo.tex
Add some TODOs to the matching algorithm notes
Fill in the last correctness proof for matching algorithm
Just need to show completeness now.

@@
 \usepackage{amssymb}
+\usepackage{color}
 \usepackage{enumitem}
@@
 \DeclareMathOperator{\im}{im}
+
+\newcommand{\TODOitem}{%
+\typeout{WARNING!!! there is still a TODO left}
+\color{blue}\item
+}
+\newcommand{\TODO}[1]{%
+\typeout{WARNING!!! there is still a TODO left}
+\marginpar{\textbf{!TODO: }\emph{#1}}
+}
+\newcommand{\TODOinline}[1]{%
+\typeout{WARNING!!! there is still a TODO left}
+{\color{blue}\textbf{!TODO: }#1}
+}
 
@@
 
+\subsubsection{Completeness}
+
+\TODOinline{all possible ways of satisfying the postconditions are found}
 \section{Testing Wrapper: Concrete Matching Only}
@@
 
+\subsection{Correctness}
+
+At the end of the algorithm, everything that was ever in the concrete part of the graph except for bare wires has been added to $m$ by the concrete matching step (postcondition \ref{enum:il-postcond-all-matched}, together with the fact that we always add everything concrete and unmatched to $U_w$, $U_c$ and $U_n$, except bare wires).
+We have no more $!$-boxes, so the entire graph is concrete.
+$P$ is also empty.
+We matched bare wires at the last stage, so $m$ is in fact total on $L$.
+So, by the argument in section \ref{sec:match-state}, $m$ is a valid match from the final state of $L$ to $G$.
+Since we have only ever performed valid $!$-box operations on $L$, this is a valid instance of $L$, and hence we have found a valid matching from $L$ to $G$.
+
+\subsection{Completeness}
+
+\TODOinline{all possible matchings are found}
+
 \end{document}

### 2012-05-17 docs/matching_algo/matching_algo.tex
added nonstrict wrapper

@@
 \DeclareMathOperator{\im}{im}
+
+\DeclareMathOperator{\N}{N}
+\DeclareMathOperator{\PN}{PN}
+\DeclareMathOperator{\W}{W}
 
@@
 
+\section{Alternate Wrapper: Partial (Non-strict) Graph Matching}
+
+The advantage of the concrete/wrapper separation is it allows easy implementation of alternative pattern graph paradigms.
+The one described in this section is strictly weaker than $!$-box matching, yet in many cases could perform better, as the pattern expansion does not require an indeterminism.
+That is, it is uniquely determined by a single run of the concrete matching algorithm.
+
+A partial string graph $(L, \PN(L))$ is a string graph with a marked subset of node-vertices $\PN(L) \subseteq \N(L)$ that are allowed to be matched partially.
+A partial string graph rewrite rule is a normal string graph rewrite rule $(L,R)$ with a surjection $p: \PN(L) \twoheadrightarrow \PN(R)$.
+\textit{Reversible} PSG rules are those where $p$ is a bijection.
+
+A PSG match is a normal match, with the exception that local iso need only be satisfied for node-vertices in $\N(L) - \PN(L)$.
+Instantiation of a PSG is done by adding new edges with boundaries for every unmatched edge in the neighbourhood of $m(n)$ for all $n \in \PN(L)$.
+For rule instantiation, the new edge/boundary pairs are added both the the neighbourhood of $n \in \PN(L)$ and $p(n) \in \PN(R)$.
+
+Matching can be done as a wrapper for the concrete matching algorithm.
+
+\begin{center}
+  \includegraphics[height=12cm]{nonstrict.pdf}
+\end{center}
+
+The result is the same as DPO rewriting for more general spans $L \leftarrow I \rightarrow R$ where $I$ contains the shared boundary of $L$ and $R$ along with some node-vertices.
+For non-reversible rules, the map $I \rightarrow R$ does not need to be mono.
+
 \end{document}

### 2012-05-18 docs/matching_algo/matching_algo.tex
added componentwise matching wrapper
made subscripts capital letters. added description for component-wise

@@
 \begin{itemize}
-    \item $U_c \subseteq W(L)\times W(L)$ : a set of unmatched circles, as pairs of wire-vertices
-    \item $U_w \subseteq W(L)$ : a set of unmatched (non-circle) wire-vertices
-    \item $U_n \subseteq N(L)$ : a set of unmatched node-vertices
+    \item $U_C \subseteq W(L)\times W(L)$ : a set of unmatched circles, as pairs of wire-vertices
+    \item $U_W \subseteq W(L)$ : a set of unmatched (non-circle) wire-vertices
+    \item $U_N \subseteq N(L)$ : a set of unmatched node-vertices
     \item $P \subseteq N(L)$ : a set of partially-matched node-vertices
-    \item $P_s \subseteq P$ : a set of partially-matched node-vertices scheduled for re-checking
+    \item $P_S \subseteq P$ : a set of partially-matched node-vertices scheduled for re-checking
     \item $m : V_L \rightarrow V_G$ : a partial function describing the match so far
@@
   \item \label{enum:il-inv-inv} The match state satisfies the global invariants
-  \item \label{enum:il-inv-unmatched} $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
-  \item \label{enum:il-inv-reachable} Each vertex in $U_w$ is adjacent to something in $U_n\cup P_s$ (ensures we can always reach wire vertices in $U_w$ by starting from a node-vertex in $U_n$ or $P_s$)
-  \item \label{enum:il-inv-Ps} If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_s$ (everything that is completely matched and in $P$ is also in $P_s$)
-  \item \label{enum:il-inv-wire-uw} If $v \in U_w$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_w$.
+  \item \label{enum:il-inv-unmatched} $U_C\cap\dom(m) = U_W\cap\dom(m) = U_N\cap\dom(m) = \varnothing$ (ie: the vertices in the unmatched sets are not part of the existing matching)
+  \item \label{enum:il-inv-reachable} Each vertex in $U_W$ is adjacent to something in $U_N\cup P_S$ (ensures we can always reach wire vertices in $U_W$ by starting from a node-vertex in $U_N$ or $P_S$)
+  \item \label{enum:il-inv-Ps} If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_S$ (everything that is completely matched and in $P$ is also in $P_S$)
+  \item \label{enum:il-inv-wire-uw} If $v \in U_W$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_W$.
   \item \label{enum:il-inv-wire-connected} If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
-  \item \label{enum:il-inv-circles} $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+  \item \label{enum:il-inv-circles} $U_C$ contains only circles, and $U_W$ contains no wire-vertices that are in circles
 \end{enumerate}
@@
 If there are no possibilities, the current branch dies.
-So, for example, if we take a vertex from $U_n$ and there are no possible matchings for it in the target graph, we kill off the current branch.
+So, for example, if we take a vertex from $U_N$ and there are no possible matchings for it in the target graph, we kill off the current branch.
 
@@
   \item \label{enum:il-postcond-inv} The match state satisfies the global invariants
-  \item \label{enum:il-postcond-no-unmatched} $U_c = U_w = U_n = P_s = \varnothing$ (every vertex marked for matching was handled)
-  \item \label{enum:il-postcond-all-matched} $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$ (we have matched exactly what we were asked to)
+  \item \label{enum:il-postcond-no-unmatched} $U_C = U_W = U_N = P_S = \varnothing$ (every vertex marked for matching was handled)
+  \item \label{enum:il-postcond-all-matched} $\dom(m)$ is the union of the initial states of $U_C$, $U_W$, $U_N$ and $\dom(m)$ (we have matched exactly what we were asked to)
   \item \label{enum:il-postcond-P} $P$ is exactly the set of vertices in $\dom(m)$ whose image is adjacent to a vertex not in $\im(m)$ ($P$ contains exactly the partially-matched vertices)
@@
 
-First, we note that our data structure constraints are never broken: we never add anything to $U_c$, $U_w$ or $U_n$, so these cannot contain objects of the wrong type.
-When we add things to $P$, they are always either taken from $U_n$ or are the end of a wire that does not terminate in a vertex in $U_w$; normalisation and precondition \ref{enum:il-inv-wire-uw} guarantee that the vertex is a node-vertex in this case.
-Finally, when removing vertices from $P$ they are removed from $P_s$ simultaneously, and when adding vertices to $P_s$ they are added to $P$ at the same time.
+First, we note that our data structure constraints are never broken: we never add anything to $U_C$, $U_W$ or $U_N$, so these cannot contain objects of the wrong type.
+When we add things to $P$, they are always either taken from $U_N$ or are the end of a wire that does not terminate in a vertex in $U_W$; normalisation and precondition \ref{enum:il-inv-wire-uw} guarantee that the vertex is a node-vertex in this case.
+Finally, when removing vertices from $P$ they are removed from $P_S$ simultaneously, and when adding vertices to $P_S$ they are added to $P$ at the same time.
 
@@
   See section \ref{sec:il-global-inv}.
-  \item $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$
+  \item $U_C\cap\dom(m) = U_W\cap\dom(m) = U_N\cap\dom(m) = \varnothing$
 
-  Nothing is ever added to $U_c$, $U_w$ or $U_n$.
+  Nothing is ever added to $U_C$, $U_W$ or $U_N$.
 There are three places where vertices are added to $m$.
-The first is when matching circles, when we match the contents of $U_c$ and set $U_c$ to be empty.
-We cannot add anything from $U_w$ at this point, due to invariant \ref{enum:il-inv-circles}, and we cannot add anything from $U_n$, since that contains only node-vertices.
-The second is when $P_s$ is empty and we take a vertex from $U_n$ and add its matching to $m$, which clearly maintains the invariant.
-In the final case all the wire-vertices added to $m$ are removed from $U_w$ at the same time and if a node-vertex is added to $m$ it has already been removed from $U_n$.
-In both the final two cases, $U_c$ is already empty.
-  \item Each vertex in $U_w$ is adjacent to something in $U_n\cup P_s$
+The first is when matching circles, when we match the contents of $U_C$ and set $U_C$ to be empty.
+We cannot add anything from $U_W$ at this point, due to invariant \ref{enum:il-inv-circles}, and we cannot add anything from $U_N$, since that contains only node-vertices.
+The second is when $P_S$ is empty and we take a vertex from $U_N$ and add its matching to $m$, which clearly maintains the invariant.
+In the final case all the wire-vertices added to $m$ are removed from $U_W$ at the same time and if a node-vertex is added to $m$ it has already been removed from $U_N$.
+In both the final two cases, $U_C$ is already empty.
+  \item Each vertex in $U_W$ is adjacent to something in $U_N\cup P_S$
 
-  We never add anything to $U_w$, so we only have to consider the removal of vertices from $U_n$ and $P_s$.
-When we remove a vertex from $U_n$, we always immediately add it to $P_s$, and we only remove a vertex from $P_s$ when there are no adjacent vertices in $U_w$.
-  \item If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_s$ (everything that is completely matched and in $P$ is also in $P_s$)
+  We never add anything to $U_W$, so we only have to consider the removal of vertices from $U_N$ and $P_S$.
+When we remove a vertex from $U_N$, we always immediately add it to $P_S$, and we only remove a vertex from $P_S$ when there are no adjacent vertices in $U_W$.
+  \item If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_S$ (everything that is completely matched and in $P$ is also in $P_S$)
 
-  When we add anything to $P$, we also add it to $P_s$, and when we remove anything from $P_s$, we remove it from $P$ if it is completely matched.
-  \item If $v \in U_w$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_w$.
+  When we add anything to $P$, we also add it to $P_S$, and when we remove anything from $P_S$, we remove it from $P$ if it is completely matched.
+  \item If $v \in U_W$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_W$.
 
-  When we remove a vertex from $U_w$, we remove the entire wire that it is on.
+  When we remove a vertex from $U_W$, we remove the entire wire that it is on.
   \item If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
@@
 So both cases preserve the invariant.
-  \item $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+  \item $U_C$ contains only circles, and $U_W$ contains no wire-vertices that are in circles
 
@@
   This is clearly true at the start of the algorithm.
-When matching circles, we only ever add vertices that are in $U_c$, which is disjoint from $\dom(m)$ by invariant \ref{enum:il-inv-unmatched}.
-When $P_s$ is empty we only ever add a single node-vertex, which is drawn from $U_n$ (which, again by \ref{enum:il-inv-unmatched}, is disjoint from $\dom(m)$).
+When matching circles, we only ever add vertices that are in $U_C$, which is disjoint from $\dom(m)$ by invariant \ref{enum:il-inv-unmatched}.
+When $P_S$ is empty we only ever add a single node-vertex, which is drawn from $U_N$ (which, again by \ref{enum:il-inv-unmatched}, is disjoint from $\dom(m)$).
 
-  When exploring wires connected to a vertex in $P$, normalisation ensures that if the wire terminates with a wire-vertex, it must be the only wire-vertex on the wire, which means that we only add a single vertex to $m$, and that vertex is drawn from $U_w$, which is disjoint from $\dom(m)$.
+  When exploring wires connected to a vertex in $P$, normalisation ensures that if the wire terminates with a wire-vertex, it must be the only wire-vertex on the wire, which means that we only add a single vertex to $m$, and that vertex is drawn from $U_W$, which is disjoint from $\dom(m)$.
 Otherwise, it must terminated with a node-vertex and the wire must contain exactly two wire-vertices.
-The first is in $U_w$, and so the second (which it is adjacent to) must also be in $U_w$ by invariant \ref{enum:il-inv-wire-uw}, and hence neither can be in $\dom(m)$ by invariant \ref{enum:il-inv-unmatched}.
+The first is in $U_W$, and so the second (which it is adjacent to) must also be in $U_W$ by invariant \ref{enum:il-inv-wire-uw}, and hence neither can be in $\dom(m)$ by invariant \ref{enum:il-inv-unmatched}.
 The only other vertex added to $m$ is the terminating node-vertex.
@@
 When we match circles, we only add wire-vertices, so the invariant is trivially maintained.
-When $P_s$ is empty and we extend $m$ with a matching from an element of $U_n$, the invariant is also maintained as none of the adjacent (wire-)vertices can be in $\dom(m)$, by invariant \ref{enum:il-inv-wire-connected}.
+When $P_S$ is empty and we extend $m$ with a matching from an element of $U_N$, the invariant is also maintained as none of the adjacent (wire-)vertices can be in $\dom(m)$, by invariant \ref{enum:il-inv-wire-connected}.
 
@@
 
-For termination of the inner loop exploring the wires incident to a node-vertex $v$, we consider the variant $n = |U_w\cap N_L(v)|$, where $N_L(v)$ is the neighbourhood of $v$ in $L$.
-In each of these branches, $n$ decreases by either $1$ or $2$, and nothing is ever added to $U_w$ (and $N_L(v)$ is fixed).
+For termination of the inner loop exploring the wires incident to a node-vertex $v$, we consider the variant $n = |U_W\cap N_L(v)|$, where $N_L(v)$ is the neighbourhood of $v$ in $L$.
+In each of these branches, $n$ decreases by either $1$ or $2$, and nothing is ever added to $U_W$ (and $N_L(v)$ is fixed).
 So this loop must terminate.
 
-The loop working through $P_s$ has two branches: when it is not empty and when it is not.
-Either may add more things to $P_s$.
-We will take as variant $m = 2|U_n| + |P_s|$.
+The loop working through $P_S$ has two branches: when it is not empty and when it is not.
+Either may add more things to $P_S$.
+We will take as variant $m = 2|U_N| + |P_S|$.
 When this is $0$, the algorithm will terminate.
-Otherwise, if $P_s$ is empty, $U_n$ will have one vertex removed and added to $P_s$, decreasing $m$ by one.
-If $P_s$ is not empty, one vertex will be removed from $P_s$, decreasing $m$ by one.
-Every vertex that is added to $P_s$ in this loop is taken from $U_n$, further decreasing $m$.
+Otherwise, if $P_S$ is empty, $U_N$ will have one vertex removed and added to $P_S$, decreasing $m$ by one.
+If $P_S$ is not empty, one vertex will be removed from $P_S$, decreasing $m$ by one.
+Every vertex that is added to $P_S$ in this loop is taken from $U_N$, further decreasing $m$.
 As a result, $m$ will decrease by at least one with every iteration of the loop.
@@
   See section \ref{sec:il-global-inv}.
-  \item $U_c = U_w = U_n = P_s = \varnothing$
+  \item $U_C = U_W = U_N = P_S = \varnothing$
 
-  $U_c$ is emptied by the circle-matching step and is never added to.
-$P_s$ and $U_n$ must be empty when we return the match due to the termination condition ($m = 0$ in section \ref{sec:termination}).
-Invariant \ref{enum:il-inv-reachable} guarantees that if $U_n$ and $P_s$ are empty, then so must $U_w$ be.
-  \item $\dom(m)$ is the union of the initial states of $U_c$, $U_w$, $U_n$ and $\dom(m)$
+  $U_C$ is emptied by the circle-matching step and is never added to.
+$P_S$ and $U_N$ must be empty when we return the match due to the termination condition ($m = 0$ in section \ref{sec:termination}).
+Invariant \ref{enum:il-inv-reachable} guarantees that if $U_N$ and $P_S$ are empty, then so must $U_W$ be.
+  \item $\dom(m)$ is the union of the initial states of $U_C$, $U_W$, $U_N$ and $\dom(m)$
 
-  Every time we remove a vertex from $U_c$, $U_w$ or $U_n$, we add it to $m$.
+  Every time we remove a vertex from $U_C$, $U_W$ or $U_N$, we add it to $m$.
 Conversely, every time we add a vertex to $m$, it must be drawn from one of these sets (given invariant \ref{enum:il-inv-wire-uw}).
@@
 We just need that if $v$ is completely matched, it is not in $P$.
-But this is guaranteed by invariant \ref{enum:il-inv-Ps} and the fact that $P_s = \varnothing$.
+But this is guaranteed by invariant \ref{enum:il-inv-Ps} and the fact that $P_S = \varnothing$.
 \end{enumerate}
@@
 \begin{itemize}
-\item $U_c$ will contain all circles
-\item $U_w$ will contain all other wire vertices, except those forming bare wires (boundary vertices connected to another boundary vertex)
-\item $U_n$ will contain all node vertices
-\item $P$, $P_s$ and $m$ will be empty
+\item $U_C$ will contain all circles
+\item $U_W$ will contain all other wire vertices, except those forming bare wires (boundary vertices connected to another boundary vertex)
+\item $U_N$ will contain all node vertices
+\item $P$, $P_S$ and $m$ will be empty
 \end{itemize}
@@
   Only the concrete matching step alters $m$ or $P$ in any way, and this maintains the global invariants.
-  \item $U_c\cap\dom(m) = U_w\cap\dom(m) = U_n\cap\dom(m) = \varnothing$
+  \item $U_C\cap\dom(m) = U_W\cap\dom(m) = U_N\cap\dom(m) = \varnothing$
 
   We only ever add unmatched vertices to these sets.
-  \item Each vertex in $U_w$ is adjacent to something in $U_n\cup P_s$
+  \item Each vertex in $U_W$ is adjacent to something in $U_N\cup P_S$
 
   This follows from openness of $!$-boxes, which (together with normalisation) ensures that any concrete wire-vertex that is not on a circle or bare wire must be adjacent to a concrete node-vertex.
-The first time concrete matching happens, all such node-vertices are put in $U_n$.
+The first time concrete matching happens, all such node-vertices are put in $U_N$.
 
-  Since all non-bare-wire, non-circle concrete vertices are added to $U_w$ each time, and so added to $\dom(m)$ by the concrete matching step, on subsequent iterations of concrete matching any wire-vertex $v$ in $U_w$ must have arisen from the expansion of a $!$-box (since this is the only way concrete vertices can be added).
-If $v$ is adjacent to a node-vertex that came from the $!$-box expansion, that node-vertex will be added to $U_n$.
+  Since all non-bare-wire, non-circle concrete vertices are added to $U_W$ each time, and so added to $\dom(m)$ by the concrete matching step, on subsequent iterations of concrete matching any wire-vertex $v$ in $U_W$ must have arisen from the expansion of a $!$-box (since this is the only way concrete vertices can be added).
+If $v$ is adjacent to a node-vertex that came from the $!$-box expansion, that node-vertex will be added to $U_N$.
 Otherwise, it must be adjacent to a node-vertex in $\dom(m)$, and hence in $P$ (by the postconditions of the concrete matching step).
-Then this node-vertex (being adjacent to the expanded $!$-box) will be added to $P_s$.
-  \item If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_s$ (everything that is completely matched and in $P$ is also in $P_s$)
+Then this node-vertex (being adjacent to the expanded $!$-box) will be added to $P_S$.
+  \item If $v \in P$ and $N_G(m(v)) \subseteq \im(m)$, $v \in P_S$ (everything that is completely matched and in $P$ is also in $P_S$)
 
@@
 As the rest of the algorithm does not add anything to $P$ or alter $m$, this is maintained throughout.
-  \item If $v \in U_w$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_w$.
+  \item If $v \in U_W$ and $w$ is a wire-vertex adjacent to $v$ in $L$, $w \in U_W$.
 
-  This is guaranteed by openness of $!$-boxes: we only ever add entire wires to $U_w$.
+  This is guaranteed by openness of $!$-boxes: we only ever add entire wires to $U_W$.
   \item If $v \in \dom(m)$ is a wire-vertex, then any vertices it is adjacent to are also in $\dom(m)$
@@
   This is guaranteed by the postcondition \ref{enum:il-postcond-all-matched} in combination with precondition \ref{enum:il-inv-wire-uw} and the fact that $m$ is only altered by the concrete matching step.
-  \item $U_c$ contains only circles, and $U_w$ contains no wire-vertices that are in circles
+  \item $U_C$ contains only circles, and $U_W$ contains no wire-vertices that are in circles
 
-  This is implicit in the step where we populate $U_w$ and $U_c$.
+  This is implicit in the step where we populate $U_W$ and $U_C$.
 \end{enumerate}
@@
 
-At the end of the algorithm, everything that was ever in the concrete part of the graph except for bare wires has been added to $m$ by the concrete matching step (postcondition \ref{enum:il-postcond-all-matched}, together with the fact that we always add everything concrete and unmatched to $U_w$, $U_c$ and $U_n$, except bare wires).
+At the end of the algorithm, everything that was ever in the concrete part of the graph except for bare wires has been added to $m$ by the concrete matching step (postcondition \ref{enum:il-postcond-all-matched}, together with the fact that we always add everything concrete and unmatched to $U_W$, $U_C$ and $U_N$, except bare wires).
 We have no more $!$-boxes, so the entire graph is concrete.
@@
 \TODOinline{all possible matchings are found}
+
+\section{Improved Pattern Wrapper: Component-wise !-box Expansion}
+
+The previous algorithm greedily matches all concrete vertices in each iteration.
+However, in a situation where different concrete components of a graph are connected only by a path containing !-boxes, this could produce many spurious match candidates that only get ruled out when the connecting !-box is killed or expanded.
+The following algorithm removed this problem by doing concrete matching component-wise, only moving to the next component once all !-boxes connected to the currently matched concrete components have been expanded.
+
+\begin{center}
+  \includegraphics[height=16cm]{component_pattern_part.pdf}
+\end{center}
 

### 2012-05-19 gui/src/quanto/core/protocol/ProtocolManager.java
- Rules Priorities are used to decide which rule should be applied - Bug Fixes on metrics - Metrics can be called from the core and the gui - Removes the java implementation of metrics and strategies.

@@ -1721,4 +1721,101 @@ public class ProtocolManager {
 
         return getXmlResponse();
     }
-}
+
+    public String[] listMetricsNames() throws CoreException {
+
+        if (backend == null) {
+            throw new IllegalStateException("The core is not running");
+        }
+
+        try {
+            writer.addHeader("MNL", generateRequestId());
+            writer.closeMessage();
+        } catch (IOException ex) {
+            throw writeFailure(ex);
+        }
+
+        return getNameListResponse();
+    }
+        public String[] listMetricsDescs() throws CoreException {
+
+        if (backend == null) {
+            throw new IllegalStateException("The core is not running");
+        }
+
+        try {
+            writer.addHeader("MDL", generateRequestId());
+            writer.closeMessage();
+        } catch (IOException ex) {
+            throw writeFailure(ex);
+        }
+
+        return getNameListResponse();
+    }
+
+     public String getCurrentMetricNameRequest() throws CoreException {
+
+        if (backend == null) {
+            throw new IllegalStateException("The core is not running");
+        }
+
+        try {
+            writer.addHeader("MCMN", generateRequestId());
+            writer.closeMessage();
+        } catch (IOException ex) {
+            throw writeFailure(ex);
+        }
+
+        return getNameResponse();
+    }
+    public void setMetricRequest(String metricName) throws CoreException {
+
+        if (backend == null) {
+            throw new IllegalStateException("The core is not running");
+        }
+
+        try {
+            writer.addHeader("MSM", generateRequestId());
+            writer.addStringArg(metricName);
+            writer.closeMessage();
+        } catch (IOException ex) {
+            throw writeFailure(ex);
+        }
+
+        getOkResponse();
+    }
+     public void computeMetricRequest(String graphName) throws CoreException {
+
+        if (backend == null) {
+            throw new IllegalStateException("The core is not running");
+        }
+
+        try {
+            writer.addHeader("MCM", generateRequestId());
+            writer.addStringArg(graphName);
+            writer.closeMessage();
+        } catch (IOException ex) {
+            throw writeFailure(ex);
+        }
+
+        getOkResponse();
+    }
+
+    public int ordMetricRequest(int[] metric1, int[] metric2) throws CoreException {
+
+        if (backend == null) {
+            throw new IllegalStateException("The core is not running");
+        }
+
+        try {
+            writer.addHeader("MOM", generateRequestId());
+            writer.addIntListArg(metric1);
[…]

### 2012-05-19 gui/src/quanto/core/strategies/DFSStrategy.java

@@ -1,52 +1,0 @@
-package quanto.core.strategies;
-
-import java.util.List;
-
-import quanto.core.Core;
-import quanto.core.data.AttachedRewrite;
-import quanto.core.data.CoreGraph;
-import quanto.core.strategies.QuantoMetrics.comparisonType;
-
-/*
- * Strategy based on metrics on graphs: Look n moves ahead and apply the rewrites that
- * minimize the metric after n rewrites
- */
-
-public class DFSStrategy extends GenericStrategy implements QuantoStrategy {
-	
-	String strategyName = "DFSStrategy";
-	private QuantoMetrics metrics = null;
-	private int depth = 0;
-	
-	public DFSStrategy(Core core, QuantoMetrics metrics, int depth) {
-		super(core);
-		this.metrics = metrics;
-		/* TODO for now we look look only one move ahead */
-		this.depth = 1; //depth
-	}
-	
-	@Override
-	public int getNext(List<AttachedRewrite<CoreGraph>> rws, CoreGraph graph) {
-		/*
-		 * TODO: Since for now we just look one move ahead simply loop through all the
-		 * rewrites and get the one that minimizes the score of the graph
-		 */
-		/* We have a list not a collection: we can rely on the order the elements */
-		int min = -1;
-		int count = 0;
-		CoreGraph bestGraph = graph;
-		for (AttachedRewrite<CoreGraph> rw : rws) {
-			switch (this.metrics.compareGraphs(bestGraph, rw.getNewGraph())) {
-				/* Unfortunately our comparison type forces to do that... */
-				case GREATER:
-					min = count;
-					bestGraph = rw.getNewGraph();
-				default: //Do nothing
-			}
-			count++;
-		}
-		if (min < 0)
-			return 0;
-		return min;
-	}
-}

### 2012-05-19 gui/src/quanto/core/strategies/GenericStrategy.java

@@ -1,28 +1,0 @@
-package quanto.core.strategies;
-
-import java.util.List;
-
-import quanto.core.Core;
-import quanto.core.data.AttachedRewrite;
-import quanto.core.data.CoreGraph;
-
-//Generic strategy: always apply the first rewrite
-
-public class GenericStrategy implements QuantoStrategy {
-
-	public String strategyName = "GenericStrategy";
-	private Core core = null;
-	
-	public GenericStrategy(Core core) {
-		this.core = core;
-	}
-	
-	public int getNext(List<AttachedRewrite<CoreGraph>> rws, CoreGraph graph) {
-		//Always apply the 1st rewrite: equivalent to the lexical order
-		return 0;
-	}
-
-	public String getName() {
-		return this.strategyName;
-	}
-}

### 2012-05-19 gui/src/quanto/core/strategies/QuantoMetrics.java

@@ -1,29 +1,0 @@
-package quanto.core.strategies;
-
-import quanto.core.data.CoreGraph;
-
-/*
- * Metrics take a graph an return a score which should represent
- * a given feature.
- * NOTE: Good metrics should take into account vertex types etc...
- * and therefore are theory specific. The definitions of such metrics
- * should be implemented in the core along with the GraphicalTheory: A metric can
- * have access to the core.
- */
-
-public interface QuantoMetrics {
-	
-	String metricsName = null;
-	
-	public enum comparisonType {
-		GREATER, 
-		SMALLER, 
-		EQUAL, 
-		NONE /*Could be a partial order...*/
-	};
-	
-	public String getName();
-	public int getScore(CoreGraph graph);
-	
-	public comparisonType compareGraphs(CoreGraph g1, CoreGraph g2);
-}

### 2012-05-19 gui/src/quanto/core/strategies/QuantoStrategy.java

@@ -1,18 +1,0 @@
-package quanto.core.strategies;
-
-import java.util.List;
-
-import quanto.core.data.AttachedRewrite;
-import quanto.core.data.CoreGraph;
-
-/*
- * Gui strategies decide in which order the rules should be applied.
- * They have access to the core but should *not* apply the rules. They return
- */
-
-public interface QuantoStrategy {
-	final String strategyName = null;
-	
-	int getNext(List<AttachedRewrite<CoreGraph>> rws, CoreGraph graph);
-	String getName();
-}

### 2012-05-19 gui/src/quanto/core/strategies/RandomStrategy.java

@@ -1,28 +1,0 @@
-package quanto.core.strategies;
-
-import java.util.List;
-import java.util.Random;
-
-import quanto.core.Core;
-import quanto.core.data.AttachedRewrite;
-import quanto.core.data.CoreGraph;
-
-/* 
- * Example of a strategy: return a random int
- */
-
-public class RandomStrategy extends GenericStrategy implements QuantoStrategy {
-
-	String strategyName = "RandomStrategy";
-	private Random r = null;
-	
-	public RandomStrategy(Core core) {
-		super(core);
-		this.r = new Random();
-	}
-	
-	@Override
-	public int getNext(List<AttachedRewrite<CoreGraph>> rws, CoreGraph graph) {
-		return r.nextInt(rws.size());
-	}
-}

### 2012-05-19 gui/src/quanto/core/strategies/RulesPriorityStrategy.java

@@ -1,46 +1,0 @@
-package quanto.core.strategies;
-
-import quanto.core.Core;
-import quanto.core.CoreException;
-import quanto.core.data.AttachedRewrite;
-import quanto.core.data.CoreGraph;
-
-import java.util.List;
-
-public class RulesPriorityStrategy implements QuantoStrategy{
-    String strategyName = "SimplePriorityStrategy";
-    Core core = null;
-    public RulesPriorityStrategy(Core core) {
-        this.core = core;
-    }
-
-    public int getNext(List<AttachedRewrite<CoreGraph>> rws, CoreGraph graph) {
-        //Get the priorities of all the applicable rules and apply the last one in case of equality
-        int max = -100;
-        int count = 0;
-        int toApply = 0;
-        for (AttachedRewrite<CoreGraph> rw : rws) {
-            int priority = getRulePriority(rw.getRuleName());
-            if (priority >= max) {
-                max = priority;
-                toApply = count;
-            }
-            count++;
-        }
-        return toApply;
-    }
-
-    public String getName() {
-        return this.strategyName;
-    }
-
-    private int getRulePriority(String ruleName) {
-        try {
-            String prioStr = this.core.getTalker().ruleUserData(ruleName, "quanto-gui:priority");
-            return  Integer.parseInt(prioStr);
-        } catch (CoreException e) {
-            //We could not get the priority for that rule... set it to 5
-            return 5;
-        }
-    }
-}

### 2012-05-19 gui/src/quanto/core/strategies/SimpleMetrics.java

@@ -1,40 +1,0 @@
-package quanto.core.strategies;
-
-import quanto.core.data.CoreGraph;
-
-/*
- * Simple Metrics based on the number of vertices and the number of edges.
- * If two graphs have the same number of vertices, then look at the number of edges.
- */
-
-public class SimpleMetrics implements QuantoMetrics {
-
-	String metricsName = "SimpleMetrics";
-	
-	public SimpleMetrics() {}
-	
-	public int getScore(CoreGraph graph) {
-		/* Simply return the number of vertices */
-		return graph.getVertexCount();
-	}
-
-	public comparisonType compareGraphs(CoreGraph g1, CoreGraph g2) {
-		if (g1.getVertexCount() > g2.getVertexCount())
-			return comparisonType.GREATER;
-		else if (g1.getVertexCount() < g2.getVertexCount())
-			return comparisonType.SMALLER;
-		else { /* total order here */
-			if (g1.getEdgeCount() > g2.getEdgeCount())
-				return comparisonType.GREATER;
-			else if (g1.getEdgeCount() < g2.getEdgeCount())
-				return comparisonType.SMALLER;
-			else
-				return comparisonType.EQUAL; /* Let the strategy make a choice */
-		}
-	}
-
-	public String getName() {
-		return this.metricsName;
-	}
-	
-}

### 2012-05-19 gui/src/quanto/gui/InteractiveGraphView.java

@@ -94,8 +94,6 @@ public class InteractiveGraphView
 	private QuantoForceLayout forceLayout;
 	private QuantoDotLayout initLayout;
 
-	private QuantoStrategy strategy = null;
-	
 	public boolean viewHasParent() {
 		return this.getParent() != null;
 	}
@@ -436,8 +434,6 @@ public class InteractiveGraphView
 		add(new ViewZoomScrollPane(viewer), BorderLayout.CENTER);
 
 		this.core = core;
-		//FIXME WE whould make the choice of the strategy a global preference
-		strategy = new RulesPriorityStrategy(getCore());
 		Relaxer r = viewer.getModel().getRelaxer();
 		if (r != null) {
 			r.setSleepTime(10);
@@ -1132,6 +1128,16 @@ public class InteractiveGraphView
 			}
 		}
 
+        private int getRulePriority(String ruleName) {
+            try {
+                String prioStr = getCore().getTalker().ruleUserData(ruleName, "quanto-gui:priority");
+                return  Integer.parseInt(prioStr);
+            } catch (CoreException e) {
+                //We could not get the priority for that rule... set it to 5
+                return 5;
+          }
+        }
+
 		@Override
 		public void run() {
 			try {
@@ -1144,7 +1150,17 @@ public class InteractiveGraphView
 				int rw = 0;
 				while (rws.size() > 0
 					&& !Thread.interrupted()) {
-					rw = strategy.getNext(rws, getGraph());
+                    int max = -100;
+                    int c = 0;
+                    int toApply = 0;
+                    for (AttachedRewrite<CoreGraph> r : rws) {
+                        int priority = getRulePriority(r.getRuleName());
+                        if (priority >= max) {
+                            max = priority;
+                            rw = c;
+                        }
+                        c++;
+                    }
 					invokeHighlightSubgraphAndWait(rws.get(rw).getNewGraph());
 					sleep(1500);
 					invokeApplyRewriteAndWait(rw);	

### 2012-05-21 gui/src/quanto/gui/InteractiveGraphView.java
Fix build (remove outdated import)

@@ -64,7 +64,6 @@ import javax.swing.event.EventListenerList;
 import quanto.core.data.AttachedRewrite;
 import quanto.core.protocol.CopyOfUserDataSerializer;
 import quanto.core.protocol.Point2DUserDataSerialiazer;
-import quanto.core.strategies.*;
 import quanto.core.Core;
 import quanto.gui.graphhelpers.ConstrainedMutableAffineTransformer;
 import quanto.gui.graphhelpers.Labeler;

### 2012-05-21 docs/matching_algo/matching_algo.tex
tweaked for improved circle handling
More matching algo docs

@@
 
+\subsection{Termination}
+\label{sec:termination}
+
+For termination of the inner loop exploring the wires incident to a node-vertex $v$, we consider the variant $n = |U_W\cap N_L(v)|$, where $N_L(v)$ is the neighbourhood of $v$ in $L$.
+In each of these branches, $n$ decreases by either $1$ or $2$, and nothing is ever added to $U_W$ (and $N_L(v)$ is fixed).
+So this loop must terminate.
+
+The loop working through $P_S$ has two branches: when it is not empty and when it is not.
+Either may add more things to $P_S$.
+We will take as variant $m = 2|U_N| + |P_S|$.
+When this is $0$, the algorithm will terminate.
+Otherwise, if $P_S$ is empty, $U_N$ will have one vertex removed and added to $P_S$, decreasing $m$ by one.
+If $P_S$ is not empty, one vertex will be removed from $P_S$, decreasing $m$ by one.
+Every vertex that is added to $P_S$ in this loop is taken from $U_N$, further decreasing $m$.
+As a result, $m$ will decrease by at least one with every iteration of the loop.
+
 \subsection{Correctness}
@@
 
-\subsubsection{Termination}
-\label{sec:termination}
-
-For termination of the inner loop exploring the wires incident to a node-vertex $v$, we consider the variant $n = |U_W\cap N_L(v)|$, where $N_L(v)$ is the neighbourhood of $v$ in $L$.
-In each of these branches, $n$ decreases by either $1$ or $2$, and nothing is ever added to $U_W$ (and $N_L(v)$ is fixed).
-So this loop must terminate.
-
-The loop working through $P_S$ has two branches: when it is not empty and when it is not.
-Either may add more things to $P_S$.
-We will take as variant $m = 2|U_N| + |P_S|$.
-When this is $0$, the algorithm will terminate.
-Otherwise, if $P_S$ is empty, $U_N$ will have one vertex removed and added to $P_S$, decreasing $m$ by one.
-If $P_S$ is not empty, one vertex will be removed from $P_S$, decreasing $m$ by one.
-Every vertex that is added to $P_S$ in this loop is taken from $U_N$, further decreasing $m$.
-As a result, $m$ will decrease by at least one with every iteration of the loop.
-
 \subsubsection{Postconditions}
@@
 
-\subsubsection{Completeness}
+\subsection{Completeness}
+\label{sec:il-complete}
 
-\TODOinline{all possible ways of satisfying the postconditions are found}
+\TODO{need some restrictions on doomed(v)}
+Let $f$ be a matching of the full subgraph of $L$ given by the vertices $U_C \cup U_W \cup U_N \cup \dom(m)$ into $G$ that restricts to $m$ on the vertices in $\dom(m)$.
+Then we need to show that, after completion of the algorithm, one of the match states produced will agree with $f$ on vertices, except on circles.
+Note that, by the argument in section \ref{sec:match-state}, this will then uniquely extend to $f$.
+
+Since $f$ is a matching with $U_C$ in its domain (and $U_C\cap\dom(m) = \varnothing$ and $f$ restricts to $m$), there must be at least $|U_C|$ unmatched circles in $G$, and so the step where we match circles must succeed.
+
+Invariant: the vertex map part of $f$ restricts to $m$.
+$m$ is extended in two places (after circle-matching).
+Each of these follows a branching point.
+We need to show that in both of these places, there is a branch that will maintain this invariant.
+We also need to show that none of the ``die'' branches are taken.
+
+When adding something from $U_N$, it is clear that $f(v)$ is a valid matching for $v$, and so this is a branch that will maintain the invariant.
+
+\TODOinline{doomed(v)}
+
+When adding something from $U_W$, the wire starting $f(w)$ must be a valid matching for the wire starting with $w$, as the entire wire must be in the domain of $f$: if $w$ is an input or output, this is trivial.
+Otherwise, there must be a wire-vertex $w'$ adjacent to $w$, and a node-vertex $n$ adjacent to $w'$.
+Then $w'$ is in $U_W$, and hence in $\dom(f)$, by precondition \ref{enum:il-inv-wire-uw}, and then $n$ must be in either $P_S$ or $U_N$, by precondition \ref{enum:il-inv-reachable}, and so in $\dom(m)$ or $U_N$, and hence $n$ is in $\dom(f)$.
+Once we have matched $w$ against $f(w)$, the only possible matching for $w'$ is $f(w')$, and similarly $n$ must match against $f(n)$.
+If $n$ is already in $\dom(m)$, we know it maps to $f(m)$, so the matchings must agree.
+So this branch maintains the invariant (and does not die).
+
+Thus, regardless of how vertices are chosen from $P_S$, $U_N$ or $U_W$, there must be a trace that returns a match where $m$ agrees with $f$ on vertices.
+
 \section{Testing Wrapper: Concrete Matching Only}
@@
 
+\subsection{Correctness}
+
 The postconditions of the concrete matching inner-loop, together with how we created the initial match state, ensure that a successful termination of this algorithm produces a valid matching, since $m$ will be total after matching bare wires and the global invariants from section \ref{sec:match-state} are maintained.
+
+\subsection{Completeness}
+
+Let $f:L \rightarrow G'$ be a matching of $L$ (after normalisation) onto a graph equivalent to $G$ (up to wire expansion/contraction).
+
+following the argument in section \ref{sec:il-complete}, as if $f:L \rightarrow G'$ is a matching to a graph equivalent to $G'$, then after the concrete step (which is given an empty $m$), there must be a branch where $m$ agrees with the restriction of $f$ to all of $L$ except bare wires.
+$P$ must then be empty by postcondition \ref{enum:il-postcond-P}.
 
@@
 However, in a situation where different concrete components of a graph are connected only by a path containing !-boxes, this could produce many spurious match candidates that only get ruled out when the connecting !-box is killed or expanded.
-The following algorithm removed this problem by doing concrete matching component-wise, only moving to the next component once all !-boxes connected to the currently matched concrete components have been expanded.
+The following algorithm removes this problem by doing concrete matching component-wise, only moving to the next component once all !-boxes connected to the currently matched concrete components have been expanded.
+
+Here, concrete component means a connected subgraph containing no vertices in !-boxes, excluding bare wires and circles.
 

### 2012-07-30 docs/matching_algo/matching_algo.tex

@@
 
-\section{Inner Loop: Concrete Matching}
+\section{Inner Loop: (Partial) Concrete Matching}
 \label{sec:inner-loop}
@@
 
-\section{Testing Wrapper: Concrete Matching Only}
+\section{String Graph to String Graph Matching}
 \label{sec:test-wrapper}
 
-For testing purposes, it is straight-forward to wrap the above procedure to just do concrete matching.  When we populate the match state, we do it in the following manner:
+For testing purposes, and as a general procedure for matching string-graphs against other string-graphs, it is straight-forward to wrap the above procedure to just do concrete matching.
+When we populate the match state, we do it in the following manner:
 \begin{itemize}
@@
 
-\section{Full Wrapper: Lazy !-Box Expansion}
+\section{Greedy $!$-Graph to String Graph Matching}
 \label{sec:full-wrapper}
 
-This algorithm performs lazy $!$-box expansion to find all possible matchings of a pattern graph against a string graph.
-In order for the algorithm to be deterministic, we disallow pattern graphs with empty $!$-boxes or $!$-boxes containing only bare wires, as these may be expanded any number of times and still match any graph that contains a wire.
+This algorithm lazily expands or kills $!$-boxes to find all possible matchings of a pattern graph against a string graph, greedily matching everything not in a $!$-box at each stage.
+The algorithm is complete and terminating only for those pattern graphs where no sequence of $!$-operations can result in a $!$-box that is empty or contains only bare wires (the implementation in Quantomatic actually allows such pattern graphs, but will fail to find all possible matchings for such graphs).
 
@@
 
+The algorithm terminates, on the assumption that no sequence of $!$-box operations will ever produce a \textit{wild} $!$-box (ie: a $!$-box that is either empty or contains only bare wires), by the following argument.
+
 Each step terminates, and there is only one loop.  As a variant, we take
@@
 We calculate the variant at the point where we consider terminating the loop, immediately after the concrete matching step.
-If the previous iteration involved killing one or more $!$-boxes, $V_L\setminus\dom(m)$ will decrease (by the combined size of the $!$-boxes).
+If the previous iteration involved killing one or more $!$-boxes, $V_L\setminus\dom(m)$ will decrease (by the combined size of the $!$-boxes, less bare wires --- this amount is positive due to the above assumption).
 If the previous iteration involved expanding a $!$-box, $|\dom(m)|$ will increase (by the size of the $!$-box, excluding any bare wires).
@@
 Hence the algorithm terminates.
+
+Note that Quantomatic's implementation ensures termination (at the expense of completeness) for all pattern graphs by killing any wild $!$-boxes as soon as they are produced.
 
@@
 
-\section{Improved Pattern Wrapper: Component-wise !-box Expansion}
+Note that Quantomatic's implementation is incomplete when the pattern graph is \textit{wild} --- ie, when there is a sequence of $!$-box operations on the pattern graph that can produce a wild $!$-box.
+
+\section{Component-wise $!$-Graph to String Graph Matching}
 
@@
 
-\section{Alternate Wrapper: Partial (Non-strict) Graph Matching}
+\section{Partial (Non-strict) String Graph to String Graph Matching}
 

### 2012-10-15 gui/src/quanto/core/Theory.java
Some fixes to the JSON stuff
Adding vertices to graphs crashes the core.  Will chase down later.

@@ -99,7 +99,9 @@ public class Theory {
 	 */
 	public VertexType getVertexType(JsonNode data) {
 		if (vertexTypePath != null) {
-			JsonNode typeNode = data.get(vertexTypePath);
+			JsonNode typeNode = data;
+			if (vertexTypePath.length() > 0)
+				typeNode = data.get(vertexTypePath);
 			if (typeNode == null || !typeNode.isTextual()) {
 				throw new IllegalArgumentException("Data did not have a type selector at '" + vertexTypePath + "'");
 			}
@@ -345,49 +347,53 @@ public class Theory {
 		boolean resDirIsTheoryParent = resourceDirectory.equals(theoryFile.getParentFile());
 
 		JsonGenerator jg = jf.createGenerator(theoryFile, JsonEncoding.UTF8);
-		jg.writeStartObject();
-		jg.writeObjectField("name", name);
-		jg.writeObjectField("coreName", coreName);
-		if (vertexTypePath != null) {
-			jg.writeObjectField("vertexTypePath", vertexTypePath);
-		}
-		jg.writeFieldName("vertexTypes");
-		jg.writeStartObject();
-		for (VertexType vt : vertexTypes.values()) {
-			jg.writeFieldName(vt.getTypeName());
+		try {
 			jg.writeStartObject();
-			GraphElementDataType dt = vt.getDataType();
-			if (dt != null) {
-				jg.writeObjectField("labelPath", dt.getDataPath());
-				jg.writeObjectField("labelDataType", dt.getTypeName());
+			jg.writeObjectField("name", name);
+			jg.writeObjectField("coreName", coreName);
+			if (vertexTypePath != null) {
+				jg.writeObjectField("vertexTypePath", vertexTypePath);
 			}
-			if (vt.getMnemonic() != null) {
-				jg.writeObjectField("mnemonic", vt.getMnemonic().toString());
-			}
-			jg.writeFieldName("visualization");
+			jg.writeFieldName("vertexTypes");
 			jg.writeStartObject();
-			SvgVertexVisualizationData svgVisData = (SvgVertexVisualizationData) vt.getVisualizationData();
-			SVGDocument svgDoc = svgVisData.getSvgDocument();
-			String svgFileName = encodeString(vt.getTypeName()) + ".svg";
-			File svgFile = new File(resourceDirectory, svgFileName);
-			writeXMLResource(svgDoc.getDocument(), svgFile);
-			if (resDirIsTheoryParent) {
[…]

### 2012-10-15 gui/src/quanto/core/data/SVGDocument.java

@@ -85,6 +85,12 @@ public class SVGDocument {
 		testParse();
 	}
 
+	public Icon createIcon() {
+		Dimension2D size = getSize();
+		return createIcon((int)Math.ceil(size.getWidth()),
+				          (int)Math.ceil(size.getHeight()));
+	}
+
 	public Icon createIcon(int width, int height) {
 		try {
 			BufferedImage img = new BufferedImage(width, height, BufferedImage.TYPE_INT_ARGB);
@@ -113,6 +119,30 @@ public class SVGDocument {
 				return new Dimension2D.Double(viewBox.getWidth(), viewBox.getHeight());
 			}
 			return SVGUtils.getSize(svgRoot);
+		} catch (InvalidXMLException ex) {
+			throw new IllegalStateException(ex);
+		}
+	}
+	
+	public Rectangle2D getBounds() {
+		try {
+			Element svgRoot = document.getDocumentElement();
+			Attr viewBoxAttr = svgRoot.getAttributeNode("viewBox");
+			if (viewBoxAttr != null) {
+				String tokens[] = viewBoxAttr.getValue().trim().split("[ \t\n]+");
+				return new Rectangle2D.Double(
+						Double.parseDouble(tokens[0]),
+						Double.parseDouble(tokens[1]),
+						Double.parseDouble(tokens[2]),
+						Double.parseDouble(tokens[3]));
+			} else {
+				Dimension2D dim = SVGUtils.getSize(svgRoot);
+				return new Rectangle2D.Double(
+						0,
+						0,
+						dim.getWidth(),
+						dim.getHeight());
+			}
 		} catch (InvalidXMLException ex) {
 			throw new IllegalStateException(ex);
 		}
@@ -207,16 +237,8 @@ public class SVGDocument {
 			return null;
 		}
 
-		Element svgRoot = document.getDocumentElement();
-		Attr viewBoxAttr = svgRoot.getAttributeNode("viewBox");
 		AffineTransform transform = new AffineTransform();
 		applyAllTransforms(e, transform);
-		if (viewBoxAttr != null) {
-			String tokens[] = viewBoxAttr.getValue().trim().split("[ \t\n]+");
-			transform.concatenate(AffineTransform.getTranslateInstance(
-					-Double.parseDouble(tokens[0]),
-					-Double.parseDouble(tokens[1])));
-		}
 		return transform.createTransformedShape(shape);
 	}
 	

### 2012-10-15 gui/src/quanto/core/data/SvgVertexVisualizationData.java

@@ -27,10 +27,9 @@ public class SvgVertexVisualizationData implements VertexVisualizationData {
 		this.labelColor = labelColor;
 		shape = doc.getElementShape("boundary");
 		if (shape == null) {
-			Dimension2D dim = doc.getSize();
-			shape = new Rectangle2D.Double(0, 0, dim.getWidth(), dim.getHeight());
+			shape = doc.getBounds();
 		}
-		cachedIcon = doc.createIcon(24, 24);
+		cachedIcon = doc.createIcon();
 	}
 
 	public SVGDocument getSvgDocument() {

### 2012-10-15 gui/src/quanto/core/data/Vertex.java

@@ -101,13 +101,16 @@ public class Vertex extends GraphElement {
 	}
 	
 	void updateFromJson(Theory theory, boolean isWireVertex, JsonNode node) throws ParseException {
-		if (!node.isObject())
-			throw new ParseException("Expected object");
-
 		if (isWireVertex) {
 			vertexType = null;
 			data = null;
+
+			if (!node.isObject())
+				return;
 		} else {
+			if (!node.isObject())
+				throw new ParseException("Expected object");
+
 			JsonNode dataNode = node.get("data");
 			vertexType = theory.getVertexType(dataNode);
 			GraphElementDataType dataType = vertexType.getDataType();

### 2012-10-16 gui/lindenb-src/org/lindenb/svg/SVGRenderer.java
Improve theory loading efficiency
The XML parser was previously trying to load the SVG DTD from the internet, which took a lot of time for something we don't care about.

@@ -154,13 +154,11 @@ public class SVGRenderer
 			else if(key.equals("transform"))
 				{
 				AffineTransform tr=svgToaffineTransform(value);
-				//tr.concatenate(this.transform);
-				//this.transform=tr;
 				this.transform.concatenate(tr);
 				}
 			else if(key.equals("font-size"))
 				{
-				this.fontSize= Float.parseFloat(value);
+				this.fontSize= (float)castUnit(value);
 				}
 			else if(key.equals("font-family"))
 				{

### 2012-10-16 gui/lindenb-src/org/lindenb/svg/SVGUtils.java

@@ -4,7 +4,6 @@ import java.awt.Shape;
 import java.awt.geom.AffineTransform;
 import java.awt.geom.GeneralPath;
 import java.awt.geom.PathIterator;
-import java.awt.geom.Rectangle2D;
 
 import java.io.PrintWriter;
 import java.io.StringReader;

### 2012-10-16 gui/src/quanto/core/Theory.java

@@ -3,9 +3,6 @@ package quanto.core;
 import com.fasterxml.jackson.core.JsonEncoding;
 import com.fasterxml.jackson.core.JsonFactory;
 import com.fasterxml.jackson.core.JsonGenerator;
-import com.fasterxml.jackson.core.JsonParser;
-import com.fasterxml.jackson.core.JsonParser.NumberType;
-import com.fasterxml.jackson.core.JsonToken;
 import com.fasterxml.jackson.databind.JsonNode;
 import com.fasterxml.jackson.databind.ObjectMapper;
 import java.awt.Color;
@@ -20,7 +17,6 @@ import java.util.Collections;
 import java.util.HashMap;
 import java.util.Iterator;
 import java.util.LinkedList;
-import java.util.List;
 import java.util.Map;
 import javax.xml.transform.Transformer;
 import javax.xml.transform.TransformerException;
@@ -231,7 +227,7 @@ public class Theory {
 		SVGDocument svgdoc;
 		try {
 			URL svgURL = new URL(contextUrl, svgNode.asText());
-			svgdoc = new SVGDocument(svgURL.toString());
+			svgdoc = new SVGDocument(svgURL);
 		} catch (MalformedURLException e) {
 			throw new ParseException("Malformed URL for SVG file");
 		} catch (InvalidXMLException e) {

### 2012-10-16 gui/src/quanto/core/data/SVGDocument.java

@@ -14,7 +14,8 @@ import java.awt.geom.Rectangle2D;
 import java.awt.image.BufferedImage;
 import java.io.File;
 import java.io.IOException;
-import java.net.URI;
+import java.io.InputStream;
+import java.net.URL;
 import javax.swing.Icon;
 import javax.swing.ImageIcon;
 import javax.xml.parsers.DocumentBuilder;
@@ -59,7 +60,9 @@ public class SVGDocument {
 			domFactory.setIgnoringComments(true);
 			domFactory.setNamespaceAware(true);
 			domFactory.setValidating(false);
-			domFactory.setNamespaceAware(true);
+			// we turn off external DTD loading, since that slows things right down
+			domFactory.setFeature("http://apache.org/xml/features/nonvalidating/load-dtd-grammar", false);
+			domFactory.setFeature("http://apache.org/xml/features/nonvalidating/load-external-dtd", false);
 			return domFactory.newDocumentBuilder();
 		} catch (ParserConfigurationException ex) {
 			throw new Error(ex);
@@ -73,9 +76,14 @@ public class SVGDocument {
 		testParse();
 	}
 
-	public SVGDocument(URI uri) throws SAXException, IOException, InvalidXMLException {
+	public SVGDocument(URL url) throws SAXException, IOException, InvalidXMLException {
 		DocumentBuilder domBuilder = createDocumentBuilder();
-		document = domBuilder.parse(uri.toString());
+		InputStream is = url.openStream();
+		try {
+			document = domBuilder.parse(is, url.toExternalForm());
+		} finally {
+			is.close();
+		}
 		testParse();
 	}
 

### 2012-10-16 gui/src/quanto/core/data/SvgVertexVisualizationData.java

@@ -8,7 +8,6 @@ import java.awt.Color;
 import java.awt.Shape;
 import java.awt.geom.Rectangle2D;
 import javax.swing.Icon;
-import org.lindenb.awt.Dimension2D;
 
 /**
  *

### 2012-10-16 gui/src/quanto/core/Theory.java
More info when handling exceptions

@@ -229,13 +229,13 @@ public class Theory {
 			URL svgURL = new URL(contextUrl, svgNode.asText());
 			svgdoc = new SVGDocument(svgURL);
 		} catch (MalformedURLException e) {
-			throw new ParseException("Malformed URL for SVG file");
+			throw new ParseException("Malformed URL for SVG file", e);
 		} catch (InvalidXMLException e) {
-			throw new ParseException("Malformed SVG file");
+			throw new ParseException("Malformed SVG file", e);
 		} catch (SAXException e) {
-			throw new ParseException("Malformed SVG file");
+			throw new ParseException("Malformed SVG file", e);
 		} catch (IOException e) {
-			throw new ParseException("Could not open SVG file \"" + svgNode.asText() + "\"");
+			throw new ParseException("Could not open SVG file \"" + svgNode.asText() + "\"", e);
 		}
 		JsonNode labelNode = visNode.get("label");
 		Color labelFill = null;

### 2012-10-16 gui/src/quanto/core/Theory.java
Vertex data parsing fixes

@@ -96,9 +96,12 @@ public class Theory {
 	public VertexType getVertexType(JsonNode data) {
 		if (vertexTypePath != null) {
 			JsonNode typeNode = data;
-			if (vertexTypePath.length() > 0)
-				typeNode = data.get(vertexTypePath);
-			if (typeNode == null || !typeNode.isTextual()) {
+			if (vertexTypePath.length() > 0) {
+				for (String fieldName : vertexTypePath.split("\\.")) {
+					typeNode = typeNode.path(fieldName);
+				}
+			}
+			if (!typeNode.isTextual()) {
 				throw new IllegalArgumentException("Data did not have a type selector at '" + vertexTypePath + "'");
 			}
 			String typeName = typeNode.asText();

### 2012-10-16 gui/src/quanto/core/data/GraphElementDataType.java

@@ -5,6 +5,7 @@
 package quanto.core.data;
 
 import com.fasterxml.jackson.databind.JsonNode;
+import com.fasterxml.jackson.databind.node.MissingNode;
 import quanto.core.ParseException;
 
 /**
@@ -16,6 +17,17 @@ public abstract class GraphElementDataType {
 
 	public GraphElementDataType(String dataPath) {
 		this.dataPath = dataPath;
+	}
+	
+	protected JsonNode findData(JsonNode node) {
+		if (dataPath == null)
+			return MissingNode.getInstance();
+		if (dataPath.length() == 0)
+			return node;
+		for (String fieldName : dataPath.split("\\.")) {
+			node = node.path(fieldName);
+		}
+		return node;
 	}
 
 	public String getDataPath() {
@@ -37,7 +49,7 @@ public abstract class GraphElementDataType {
 
 		@Override
 		public GraphElementData parseData(JsonNode node) throws ParseException {
-			JsonNode strNode = node.findPath(dataPath);
+			JsonNode strNode = findData(node);
 			if (!strNode.isTextual())
 				throw new ParseException("Expected string value");
 			return new GraphElementData(strNode.asText());
@@ -56,7 +68,7 @@ public abstract class GraphElementDataType {
 
 		@Override
 		public GraphElementData parseData(JsonNode node) throws ParseException {
-			JsonNode strNode = node.findPath(dataPath);
+			JsonNode strNode = findData(node);
 			if (!strNode.isTextual())
 				throw new ParseException("Expected string value");
 			return new GraphElementMathsData(strNode.asText());

### 2012-10-16 gui/jung-src/edu/uci/ics/jung/visualization/renderers/Renderer.java
Allow vertex label position decisions to be based on the vertex

@@ -57,15 +57,15 @@ public interface Renderer<V,E> {
 		void labelVertex(RenderContext<V,E> rc, Layout<V,E> layout, V v, String label);
 		Position getPosition();
 		void setPosition(Position position);
-		void setPositioner(Positioner positioner);
-		Positioner getPositioner();
+		void setPositioner(Positioner<V> positioner);
+		Positioner<V> getPositioner();
 		class NOOP implements VertexLabel {
 			public void labelVertex(RenderContext rc, Layout layout, Object v, String label) {}
 			public Position getPosition() { return Position.CNTR; }
 			public void setPosition(Position position) {}
 			public Positioner getPositioner() {
 				return new Positioner() {
-					public Position getPosition(float x, float y, Dimension d) {
+					public Position getPosition(Object v, float x, float y, Dimension d) {
 						return Position.CNTR;
 					}};
 			}
@@ -73,8 +73,8 @@ public interface Renderer<V,E> {
 			}
 		}
 		enum Position { N, NE, E, SE, S, SW, W, NW, CNTR, AUTO }
-	    interface Positioner {
-	    	Position getPosition(float x, float y, Dimension d);
+	    interface Positioner<V> {
+	    	Position getPosition(V vertex, float x, float y, Dimension d);
 	    }
 
 	}

### 2012-10-16 gui/jung-src/edu/uci/ics/jung/visualization/renderers/VertexLabelAsShapeRenderer.java

@@ -104,9 +104,9 @@ public class VertexLabelAsShapeRenderer<V,E>
 		return Renderer.VertexLabel.Position.CNTR;
 	}
 
-	public Renderer.VertexLabel.Positioner getPositioner() {
-		return new Positioner() {
-			public Renderer.VertexLabel.Position getPosition(float x, float y, Dimension d) {
+	public Renderer.VertexLabel.Positioner<V> getPositioner() {
+		return new Positioner<V>() {
+			public Renderer.VertexLabel.Position getPosition(V v, float x, float y, Dimension d) {
 				return Renderer.VertexLabel.Position.CNTR;
 			}};
 	}
@@ -115,7 +115,7 @@ public class VertexLabelAsShapeRenderer<V,E>
 		// noop
 	}
 
-	public void setPositioner(Renderer.VertexLabel.Positioner positioner) {
+	public void setPositioner(Renderer.VertexLabel.Positioner<V> positioner) {
 		//noop
 	}
 }

### 2012-10-18 docs/programming-quanto/manual.tex

\documentclass{article}
\usepackage[usenames]{color}
\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{latexsym}
%\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}
\usepackage{allmtt}
\usepackage{xspace}
%\usepackage{pdfsetup}
\usepackage[a4paper,hscale=0.67,vscale=0.76]{geometry}
\usepackage{index}
\usepackage{framed}
\usepackage{boxedminipage}

\usepackage{tikzfig}
\usepackage{tikzstyles}
\usepackage{todo}

\newenvironment{graybox}
  {\def\FrameCommand{\fboxsep=0pt\colorbox{gray!20}}\MakeFramed{\smallskip\FrameRestore}\begin{quote}}
  {\end{quote}\smallskip\endMakeFramed}
\newenvironment{code}
  {\begin{graybox}\begin{allmtt}}
  {\end{allmtt}\end{graybox}}

\title{Programming Quantomatic}
\author{Lucas Dixon}

\begin{document}
\maketitle

\section{Introduction}

This is an introduction to programming Quantomatic. It is intended for
those wanting to extend the system or adapt it for some other use. For
a description of the mathematics that the system is based on,
see~\cite{2010:DDK:DCM}.

\section{Architecture}

Quantomatic consists of two programs that communicate with the each
other:

\begin{description}
\item[Quanto-Core:] is an ML program that is responsible for
  manipulating the graphs, rewrite rules and theories. It performs the
  rewriting and has a protocol for interacting with it. Quanto core
  makes use of a collection of libraries for ML called \emph{isaplib}.
\item[Quanto-Gui:] is a Java program that provide the user-interface
  for quantomatic. It starts, and then communicates with the
  Quanto-Core program.
\end{description}

Each of the programs is built on a number of libraries and contains a
number of comonents.

\subsection{Quanto-Core}

\begin{figure}[t]
\centering{\scalebox{0.9}{\includegraphics{arch0-core-overview}}}
\caption{The architecture of the main components of the core-part of
  quantomatic. Arrows indicate the target component using the
  source. The dotted line is the communication protocol between the
  Core component, written in ML, and the GUI one, written in Java.}
\label{fig:arch-overview1}
\end{figure}

An overview of the libraries and components, from the perspective of
the core program, is shown in Figure~\ref{fig:arch-overview1}. There
are 8 main components to the core program, all built using the common
functions provided by isaplib. These are:

\begin{description}
\item[Expression] defines a linear arithmetic expressions and
  operations on them. These expressions are used as the data stored in
  certain kinds of vertices.
\item[Graph] defines representations of graph. This provides
  structures and signatures for graphs, open graphs and bang-box
  graphs. There is a basic signature for each kinds of graph that is
  essentially only the most elementry operations needed to manipulate
  the data. The is also a signature that extends this basic
  information to provide a range of algorithms and higher-level
  functinoality.
\item[Matching] provides data structure and algorithms for matching
  graphs.
\item[Rules] defines what a graphical rewrite rule is and provides
  the functionality to applying rules to graph, using the matching
  algorithms. 
\item[Theories] are collections of rules with some meta-information,
  such as which ones should be used for automatic rewriting.
\item[I/O] provides the tools for loading and saving graphs, theories,
  expresions, etc. 
\item[Controller] defines a protocol for interacting with
  Quantomatic. This is the component that interacts with Java, and
  through which the quantomatic-core is controlled.
\item[Hilb] defines a representation of finite dimensional hilbert
  spaces using matricies. It also provides a way to interpret a
  red-green quantum graph as a such a matrix. This is used to support
  interaction with other tools, such as Mathematica.
\end{description}

An overview of the essential components for graphs, matching and
rewriting, along with the relevant components of isaplib is shown in
Figure~\ref{fig:arch-overview-in-core}.

\begin{figure}[p]
\centering{\scalebox{0.9}{\includegraphics{arch1-names-graphs-rewriting}}}
\caption{The architecture for the construction of Graphs, Matching and
  Rewriting in Quantomatic. White-headed arrows are inheritance
  relations, and dark-headed arrows are dependency relations: the
  target uses the source.}
\label{fig:arch-overview-in-core}
\end{figure}

\subsection{Quanto-GUI}

\section{Installing, Building and Running}

\section{ML/Core}

\subsection{The SML programming language}

There is a good collection of books on programming in Standard ML including:

\begin{itemize}
\item Paulson's ``ML for the working programmer''~\cite{Paulson:ML},
  and 
\item Harper's in progress ``Programming in Standard ML'' which has a
  draft available online
  \url{http://www.cs.cmu.edu/~rwh/smlbook/online.pdf}
\end{itemize}

\subsection{PolyML}

PolyML (\url{http://www.polyml.org/}) is an efficient implementation
of Standard ML which supports efficient multi-threaded programming.

\subsection{Editing code: jEdit and Emacs}

\begin{itemize}
\item An ML mode for emacs
  (\url{http://www.smlnj.org/doc/Emacs/sml-mode.html}), and 
\item An in-progress PolyML plugin for jEdit
  (\url{http://dream.inf.ed.ac.uk/projects/polyml/})
\end{itemize}

\subsection{Standard Basis Library}

The Standard ML Basis Library (\url{http://www.standardml.org/Basis/})
contains basic tool for interacting with the operating system.

\subsection{IsaLib Library}

{\tt isaplib/basic}: this directory contains additional libraries
beyond those of the Standard ML basis library. It builds on the
library provided by the Isabelle theorem prover. In particular, it
defines a useful collection of top-level functions and infix
operators. The most important are the following:

\begin{code}
fun I x = x

fun K x = fn \_ => x

fun x |> f = f x

fun map f [] = []
  | map f (x::xs) = (f x)::(map f xs)

fun fold f [] y = y
  | fold f (x::xs) y = fold f xs (f x y)
\end{code}

In particular, it is worth noting that the {\tt fold} function can be
understood as a more diciplined way to write a for-loop. It iterates
over all elements in a list updating the value {\tt y}, by applying
{\tt f x} to it.

\subsection{Names, name-sets, name-tables}

Rich data structures can be built up using finite sets and
mappings. For example, finite sets of names are used to identiy the
edges in a graph. A graph can then be defined as a mapping from
edge-names to the source and target vertex name. Data can be
associated to a vertex by defining a mapping from the vertex name to
the data associated with a vertex.

The key idea of name is that given a set of names, a new, fresh name
can be constructed. This allows finite sets and maps to be extended
programatically.

%% Lets use string-based names for vertices and edges. The code for this
%% is:

%% \begin{code}
%% type graph = StrNames.NTab.T 
%% \end{code}

%%  A implementation of finite
%% sets of objects

The signature that shows most of the functions for working with finite
sets of names can be found in: \begin{center}{\tt
  isaplib/names/basic\_nameset.ML}\end{center}

Map from a finite set of names to some other type are called
name-tables and the signature showing most of the functions for
working with them can be found in: \begin{center}{\tt
  isaplib/names/basic\_nametab.ML}\end{center}

\subsection{Renaming}

\subsection{Maps, Isomorphisms and Binary Relations}

\subsection{Graphs}

\subsection{Matching}

\subsection{I/O}

\subsection{Controller}

\bibliographystyle{plain}
\bibliography{bibfile}

\end{document}

### 2012-10-18 gui/src/quanto/gui/InteractiveGraphView.java
Only resolve the first match when normalising
Before, we forced the core the find every match of every rule, even
though we only ever used the first one.

This tends to be much faster when there are many matches to be found.

@@ -860,21 +860,14 @@ public class InteractiveGraphView
 
 		private boolean highlight = false;
 
-		private void attachNextRewrite() {
+		private boolean attachNextRewrite() {
 			try {
-				core.attachOneRewrite(
+				return core.attachOneRewrite(
 						getGraph(),
 						getGraph().getVertices());
 			} catch (CoreException e) {
 				coreErrorDialog("Could not attach the next rewrite", e);
-			}
-		}
-
-		private void attachAllRewrites() {
-			try {
-				core.attachRewrites(getGraph(), getGraph().getVertices());
-			} catch (CoreException e) {
-				coreErrorDialog("Could not attach the next rewrite", e);
+				return false;
 			}
 		}
 
@@ -940,18 +933,14 @@ public class InteractiveGraphView
 				// FIXME: communicating with the core: is this
 				//        really threadsafe?  Probably not.
 
-				attachAllRewrites();
-				List<AttachedRewrite<CoreGraph>> rws = getRewrites();
 				int count = 0;
-				int rw = 0;
-				while (rws.size() > 0
-						&& !Thread.interrupted()) {
-					invokeHighlightSubgraphAndWait(rws.get(rw).getNewGraph());
+				while (!Thread.interrupted() && attachNextRewrite()) {
+					List<AttachedRewrite<CoreGraph>> rws = getRewrites();
+					CoreGraph newGraph = rws.get(0).getNewGraph();
+					invokeHighlightSubgraphAndWait(newGraph);
 					sleep(1500);
-					invokeApplyRewriteAndWait(rw);
+					invokeApplyRewriteAndWait(0);
 					++count;
-					attachAllRewrites();
-					rws = getRewrites();
 				}
 
 				fireJobFinished();

### 2012-10-20 console/quanto.rb
added ruby script that will serve as console. IO not working right yet

@@ -1,0 +1,63 @@
+#!/usr/bin/env ruby
+
+require "readline"
+require "open3"
+
+$seq = 0
+$controller = "red_green"
+
+def send_request(qin, modl, fun, json)
+  puts "\e<#{$seq}\e,#{$controller}\e,#{modl}\e,#{fun}\e,#{json}\e>"
+  sz = qin.write "\e<#{$seq}\e,#{$controller}\e,#{modl}\e,#{fun}\e,#{json}\e>"
+  puts "wrote #{sz} chars"
+  qin.flush
+  $seq+=1
+end
+
+def read_until_esc(qout, code)
+  str = ''
+  loop do
+    puts 'about to readchar'
+    c = qout.readchar.chr
+    puts "[#{c}]"
+    if c == "\e"
+      c = qout.readchar.chr
+      puts "[#{c}]"
+      
+      if c == "\e"
+        str << c # escaped ESC
+      elsif c == code
+        return str
+      else
+        raise Exception("Expected escape code: [#{code}], got: [#{c}]")
+      end
+    else
+      str << c
+    end
+  end
+end
+
+def get_response(qout)
+  read_until_esc(qout, '<')
+  rid = read_until_esc(qout, ',')
+  status = read_until_esc(qout, ',')
+  json_out = read_until_esc(qout, '>')
+  
+  return [rid, status, json_out]
+end
+
+Open3.popen3("../core/bin/quanto-core") do |qin, qout, qerr, wait_thr|
+  qin.sync = true
+  qout.sync = true
+  qerr.sync = true
+  send_request(qin, '!!', 'version', 'null')
+  puts 'sent request'
+  puts get_response(qout)
+  puts 'done'
+end
+
+# loop do
+#   s = Readline.readline('quanto> ')
+#   break if s == nil
+#   puts s
+# end

### 2013-05-16 scala/src/main/scala/quanto/core/CoreProcess.scala
parallel core

@@ -44,20 +44,20 @@ class CoreProcess(parallel: Boolean = false) {
   
   def killCore(waitForExit: Boolean = true) {
     if (backend != null) {
-        logger.log(Level.FINEST, "Shutting down the core process");
+        logger.log(Level.FINEST, "Shutting down the core process")
         val killThread = new Thread() {
           override def run() {
             if (waitForExit) {
               try {
-                logger.log(Level.FINER, "Waiting for 5 seconds for the core to exit");
-                Thread.sleep(5000);
+                logger.log(Level.FINER, "Waiting for 5 seconds for the core to exit")
+                Thread.sleep(5000)
               } catch {
                 case e: InterruptedException =>
-                  logger.log(Level.FINER, "Thread interrupted");
+                  logger.log(Level.FINER, "Thread interrupted")
               }
             }
-            logger.log(Level.FINER, "Forcibly terminating the core process");
-            backend.destroy();
+            logger.log(Level.FINER, "Forcibly terminating the core process")
+            backend.destroy()
           }
         }
         

### 2013-05-16 scala/src/main/scala/quanto/core/CoreState.scala

@@ -1,17 +1,21 @@
 package quanto.core
 
 import akka.actor._
-import quanto.util.json.Json
+import quanto.util.json._
 
-abstract class CoreMessage
-case object StartCore extends CoreMessage
-case object CoreInitialized extends CoreMessage
-case class CoreResponse(requestId: Int, resp: Json) extends CoreMessage
-case class UnhandledRequest(requestId: Int, reason: String) extends CoreMessage
-abstract class CoreRequest extends CoreMessage {
+case object StartCore
+case object CoreInitialized
+case class CoreResponse(requestId: Int, resp: Json)
+case class UnhandledRequest(requestId: Int, reason: String)
+abstract class CoreRequest {
   def requestId: Int
   def json : Json
   def decode(resp: Json): Any
+}
+case class SimpleRequest(jsonString: String) extends CoreRequest {
+  val json = Json.parse(jsonString)
+  def requestId = (json / "request_id").intValue
+  def decode(resp: Json): Json = resp
 }
 
 class CoreState(executable: String) extends Actor with ActorLogging {
@@ -20,43 +24,71 @@ class CoreState(executable: String) extends Actor with ActorLogging {
   var writer: ActorRef = _
   val listeners = collection.mutable.Map[Int, (ActorRef,CoreRequest)]()
 
-  val coreDown: Receive = {
-    case StartCore =>
-      coreProcess.startCore(executable)
-      reader = context.actorOf(Props[CoreReader], name = "core_reader")
-      writer = context.actorOf(Props[CoreWriter], name = "core_writer")
-      context.become(coreWaitForInit)
-    case req : CoreRequest =>
-      sender ! UnhandledRequest(req.requestId, "Core down")
-    case x => log.warning("Unexpected message: " + x + " in state coreDown.")
-  }
+  coreProcess.startCore(executable)
+  reader = context.actorOf(Props { new CoreReader(coreProcess) }, name = "core_reader")
+  writer = context.actorOf(Props { new CoreWriter(coreProcess) }, name = "core_writer")
 
-  val coreWaitForInit: Receive = {
-    case CoreInitialized => context.become(coreRunning)
-    case req : CoreRequest =>
-      sender ! UnhandledRequest(req.requestId, "Core initializing")
-    case x => log.warning("Unexpected message: " + x + " in state coreWaitForInit.")
-  }
+  log.info("fired up")
 
-  val coreRunning: Receive = {
+//  val coreDown: Receive = {
+//    case StartCore =>
[…]

### 2013-05-16 scala/src/main/scala/quanto/gui/JsonConsole.scala

@@ -1,0 +1,107 @@
+package quanto.gui
+
+import swing._
+import event.{TableUpdated, Key, KeyPressed}
+import java.awt.{Font => AWTFont, Color}
+import javax.swing.border.LineBorder
+import apple.laf.JRSUIConstants.AlignmentVertical
+
+sealed abstract class CoreOutputItem
+case class Waiting(rid: Int) extends CoreOutputItem {
+  override def toString = "waiting for request " + rid + "..."
+}
+
+case class ErrorMessage(rid: Int, output: String) extends CoreOutputItem {
+  override def toString = output
+}
+case class Output(rid: Int, output: String) extends CoreOutputItem {
+  override def toString = output
+}
+
+object JsonConsole extends SimpleSwingApplication {
+
+//  object OutputModel {
+//
+//    case class Cell(row: Int, column: Int) {
+//      override def toString = "foo"
+//    }
+//    val cells = Array.ofDim[Cell](3,1)
+//    for (i <- 0 until 3) cells(i)(0) = Cell(i,0)
+//  }
+
+  val JsonOutput = new ListView[CoreOutputItem] {
+    renderer = new ListView.Renderer[CoreOutputItem] {
+      val label = new Label
+      label.horizontalAlignment = Alignment.Left
+      val panel = new BorderPanel {
+        add(label, BorderPanel.Position.Center)
+        border = new LineBorder(Color.WHITE, 1)
+      }
+
+      def componentFor(list: ListView[_], isSelected: Boolean,
+                       focused: Boolean, a: CoreOutputItem, index: Int): Component =
+      {
+
+        a match {
+          case Waiting(_) =>
+            panel.background = new Color(0.9f,0.9f,1.0f)
+            label.border = new LineBorder(new Color(0.2f,0.2f,0.4f))
+          case ErrorMessage(_,_) =>
+            panel.background = new Color(1.0f,0.5f,0.5f)
+            label.border = new LineBorder(new Color(0.4f,0.2f,0.2f))
+          case Output(_,_) =>
+            panel.background = new Color(0.9f,1.0f,0.9f)
+            label.border = new LineBorder(new Color(0.2f,0.6f,0.2f))
+        }
+
+        label.text = "<html><body style='width:100%;padding:1px 4px;text-align:left;font-family:monospace'><pre>" + a.toString
+
+        panel
+      }
+    }
+  }
+
[…]

### 2013-05-16 scala/src/test/scala/quanto/core/test/ParCoreSpec.scala

@@ -1,0 +1,29 @@
+package quanto.core.test
+import org.scalatest._
+import akka.actor._
+import akka.pattern.ask
+import quanto.core._
+import scala.concurrent._
+import ExecutionContext.Implicits.global
+
+class ParCoreSpec extends FlatSpec {
+  val sys = ActorSystem("Quanto-Test")
+  val core = sys.actorOf(Props { new CoreState("../core/bin/quanto-core") }, "core_state")
+
+  def testReq(rid: Int) = new SimpleRequest(
+    s"""
+      |{
+      |  "request_id": ${rid},
+      |  "controller": "!!",
+      |  "module": "system",
+      |  "function": "version"
+      |}
+    """.stripMargin)
+
+  "CoreState actor" should "accept some requests" in {
+    core ! testReq(0)
+    core ! testReq(1)
+  }
+
+  //sys.shutdown()
+}
